### 2012-08-31 Dail_Mad_ext.tex
first commit

\documentclass{article}

\usepackage[a4paper, hmargin={2.4cm,2.4cm}, vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{parskip}

\bibliographystyle{ecology}

\begin{document}

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or management actions. State-space
models, are widely used for these purposes, and they allow for
estimating parameters of classical growth models, such as exponential,
logistic, and Gompertz, while accounting for demographic stochasticity
and observation error. Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models for spatially-replicated
time-series data recently proposed by \citet{dail_madsen:2011}. We
expand this class of models to accommodate classical growth
models, zero-inflation, and random effects such as observer-specific
detection probabilities. Furthermore, we describe classical and
Bayesian methods of parameter estimation. These new developments will
allow researchers to apply these methods to address important
questions regarding the factors affecting spatial and temporal
variation in abundance. We demonstrate these developments by analyzing
data from the North American Breeding Bird Survey. Code to fit these
models using the R package unmarked and JAGS is also provided.
}

\textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
Gompertz model, immigration, open population point count models,
random observer effects, range, Ricker model, zero-inflated

\newpage

Theoretical ecology requires robust models for testing hypotheses
regarding density dependence and spatial variation in population
dynamics. Such models are also required in applied ecological research
for predicting species' responses to environmental change or
management actions. For example, many researchers are currently
investigating the influence of weather on population dynamics so that
predictions can be made about the effects of future climate
change. Several complicating factors must be confronted when
addressing these questions. First, deterministic models of population
dynamics are virtually always inadequate due to ``process variation'',
the inherent stochasticity in demographic parameters and environmental
conditions. Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of ``observation error'', such as imperfect
detection. Failure to account for  and heterogeneity in observation
error can bias the estimators of abundance and related parameters.

State-space models have become the most widespread approach for
studying population dynamics while accounting for process variation
and observation error (de Valpine and Hastings 2002, Buckland et
al. 2004, Dennis et al. 2006). Conventional state-space models are
simply time-series models in which the true state of the system is
observed imperfectly. Originally developed for xxx (cite Kalman filter
papers), the ability to model both the ecological state process and
the observation process has made them relevant in numerous ecological
applications [need to clean this up]. State-space models have been
defined differently by different authors, but we follow Buckland et
al. (2004) using the definition : ``xxx'' . Typically, the ecological
process is described by a Markovian model that includes a
deterministic description of population change as well as two random
sources of variation for ``process variation''. The model for the
observation process is typically phenomenological in the sense that
observation error is not defined explicitly. As an example, let  be
the abundance of a species during year  , and   be the observed data,
which differs from  due to observation error ( ). A simple state space
model that includes exponential growth and process variation ( ) could
be written as
	 	(1)
Here   is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions:   and   (de
Valpine and Hastings 2002, Dennis et al. 2006). Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace   with   where A is the area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus . Generally, zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.

The problems mentioned thus far are minor in comparison to the more
serious issue that the parameters of the model are often not
identifiable. Specifically, the Markovian nature of the model implies
the following likelihood
\[
LIKELIHOOD
\]
The first term is the probability density for the first observation in
the time series,  , yet it is intuitively not possible to estimate the
parameter(s) of this distribution using only a single
observation. This problem has resulted in several papers with
interesting names like ``multi-modal likelihoods and ridges etc...''
The workaround here is to assume that the population is at equilibrium
so that that the x can be replaced with the equilibrium
distribution. However, assuming equilibrium defeats the objective of
many studies of population dynamics, namely determining if and why a
population is at equilibrium.

A third problem with these models is that they do not admit
spatial variation. This is not a fair criticism because this
lies beyond the scope of traditional state-space models, but
it does restrict the utility of their since it is typically
impossible assess the impacts of factors such as habitat
fragmentation or climate change on a population without
considering spatial variation. Indeed many populations are
regulated by spatial processes such as source-sink dynamics.

Dail and Madsen (2011) developed a model (henceforth the DM model)
that resolves each of these problems with traditional state-space
models. The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error using
only spatially and temporally replicated count data.  Their model is
an extension of the closed-population N-mixture model (Royle
2004)(2004), which was designed specifically to address the problem of
modeling spatial variation in abundance when detection probability is
less than unity. The DM extension relaxes the assumption of population
closure and includes explicit parameters describing population change
as a first-order Markovian process. [more details]

[applications / potential etc... ] The DM model is relevant to a huge
number of problems confronting basic and applied ecological
research. Answering questions about climate change etc... require
spatially and temporally extensive datasets such as result from the
North American Breeding Bird Survey (BBS) (Robbins et
al. 1986). However, most of the datasets of the appropriate scale are
collected by volunteers and using protocols that are not amenable to
traditional approaches of modeling abundance and detection probability
based on standard capture-recapture methods. Instead they have, at
best, simple count data for which, when it can be interpreted as
counts of unique individuals, the DM model can be applied.. Yadi ya

Although the DM model was designed explicitly for these purposes, the
original formulation of the model makes strict distributional
assumptions and does not acknowledge many of the sources of variation
inherent to existing ecological time-series data.  We propose
variations on these models that are likely to be more realistic and
useful in many cases.  We demonstrate the usefulness of these new
models with simulated and real BBS data.

\bibliography{Dail_Mad_ext}

\end{document}

### 2012-08-31 Dail_Mad_ext.tex
Word to tex

@@
 \usepackage{fancyvrb}
-\usepackage{parskip}
 
@@
 
+\section{Methods}
+
+\subsection{Model Development}
+
+As originally formulated, the DM model includes submodels for four
+conditionally related processes: the initial abundance state, apparent
+survival, recruitment, and detection probability. Dail and Madsen
+proposed two options for initial abundance (Poisson and negative
+binomial):
+(2)
+(3)
+where $N_{i1}$  is the abundance at site $i$  during year 1 and $\Lambda$ is
+the expected abundance. The negative binomial model includes the
+overdispersion parameter $\alpha$.  The survival and recruitment terms allow
+abundance to change over time:
+(4)
+where $\omega$ is the survival probability and $\gamma$ is the arrival rate (which
+can depend on Nit-1). Dail and Madsen propose three models for $\gamma$: the
+constant model, where $\gamma$ does not depend on Nit-1, and which simulates
+a ``propagule rain'' of recruitment; the autoregressive model , which
+simulates geometric or density independent growth; and the ``no-trend''
+model, which keeps abundance fixed over time.
+The number seen is a function of detection probability:
+(5)
+Initial abundance, apparent survival, recruitment, and detection can
+all be modeled as functions of covariates, allowing fairly
+sophisticated inference about population processes and dynamics from
+point count data.
+Range model of initial abundance
+Dail and Madsen (2011) suggested two distributions for modeling
+initial abundance: Poisson and negative binomial.  We have extended
+their model to include another distribution, the zero-inflated
+Poisson.  This could be useful when, for example, one is modeling the
+abundance of several bird species with the same set of BBS routespoint
+count surveys for all species, but this set includes routes sites
+outside the range of some of the bird species.  The distribution of
+initial abundances can be represented as:
+(6)
+where $\psi$ represents the proportion of extra zeros.
+These models allow three sources of zero counts by observers: a
+species was at a route but not detected; the route was within the
+species' range but there were no birds at that site in that year; and
+the route was outside the species' range.  Furthermore, detection,
+abundance, and zero-inflation can be modeled separately as functions
+of different (or the same) covariates.  For example, detection of
+species x at site y in year z might depend on wind speed, abundance on
+forest type and weather, and zero-inflation upon elevation and
+climate.  This approach combines elements of occupancy modeling
+(MacKenzie et al. 2006) and abundance modeling, and could prove useful
+for distinguishing species' fundamental and realized niches
+(Hutchinson 1957).   The zero inflation factor can also affect
+recruitment or population dynamics (see below).
+We also considered a zero-inflated negative binomial model of initial
+abundance, but this model did not perform well in preliminary tests.
+Population growth models
+Preliminary DM model runs for several species tended to lead to
+estimates of survival that were unrealistically high and recruitment
+that were unrealistically low, or the reverse (compared to independent
+demographic analyses).  The DM models are able to partition changes in
+abundance to survival and recruitment in part by making strong
+distributional assumptions (Dail and Madsen 2011).  When those
+assumptions are heavily violated the models may proportion population
+growth incorrectly into survival and recruitment, even if they
+estimates population growth accurately.
+Although partitioning population growth into survival and recruitment
+is informative, it is not needed for all applications.  Furthermore, a
+simpler model would have several merits: faster running time, fewer
+total model combinations, and possibly more realistic estimates.
+Therefore, we developed a version of the DM model that estimates and
+models population growth directly.  In this model, equation 4 is
+simplified to:
+(7)
+where r represents the instantaneous population growth rate.  This,
+like the autoregressive version of the model, is a variant on a simple
+density-independent exponential model of population growth.
+Density-dependent versions of the model are also possible.  For
+example:
+(8)
+where K is the stable equilibrium of the population and r is the
+instantaneous population growth rate at low population densities, and
+both parameters are constrained to be positive.  This model is based
+on Ricker's (1954) discrete time version of the logistic population
+growth model.  We also implemented Gompertz's (1825) density-dependent
+model:
+(9)
+where the log of Nit-1 is only taken when Nit-1 is greater than 0
+(when Nit-1 is 0, the full expression simplifies to 0 anyway).  Here
+the interpretations of r and K are similar to in the Ricker model, but
+K is constrained to be greater than one.
+Because a single Poisson distribution controls the
+distribution of Nit in each of these models, the discrete
+convolution used by Dail and Madsen (2011) is not required,
+speeding up processing time.
+Immigration and colonization models
+The autoregressive, population growth, Ricker, and Gompertz
+versions of the DM models all share a common feature (or bug):
+once the population at a site reaches 0, it must remain at 0.
+This is because all contributions to population growth are
+local in these models.  We generalized each these models that
+include both internal and external (immigration) contributions
+to population growth.  The population growth + immigration
+model is:
+(10)
+or
+(11)
+where $\iota$ represents the immigration rate (equations 10 and 11 are
+equivalent).  This model is close to the constant DM model (equation
+4), with exp(r) instead of $\omega$ and $\iota$ instead of $\gamma$, except that the first
+process is Poisson distributed instead of binomial.  Similarly, the
+Ricker + immigration model can be represented as:
+(12)
+the Gompertz + immigration model as:
+(13)
+and the autoregressive + immigration model as:
+(14)
+These immigration models are based on the assumption that numbers
+of immigrants are not dependent on local density.  Other versions
+of these models with different assumptions are possible. The
+Ricker + immigration and Gompertz + immigration models can be
+justified by the further assumption that immigrants arrive as
+adults immediately before the point-count survey (so that
+density-dependent processes do not have a chance to act on the
+numbers of immigrants; Otto and Day 2007 p. 161).  Deterministic
+versions of these models do reach stable equilibriums, but the
+equilibriums are not at K and cannot be solved for analytically
+(Otto and Day 2007).  We have implemented all preceding models in
+a maximum likelihood framework by extending the unmarked package
+(Fiske and Chandler 2011) in R (R Development Core Team 2011).
+Random eEffects of oObservers
+Differences in observers' ability to see, hear, or identify
+birds has long been recognized as a potential source of error
+in avian point count surveys such as the BBS (Robbins et
+al. 1986, Sauer et al. 1994, Diefenbach et al. 2003, Alldredge
+et al. 2007, Campbell and Francis 2011).  Estimating a
+separate detection probability for each observer can be
+difficult and reduces one's ability to estimate the quantities
+of interest.  This problem is compounded by the fact that
+observers differ greatly in the number of surveys they have
+run (so that many observers' separate detection probabilities
+could not be accurately estimated).
+Current BBS trend estimators deal with this problem by
+treating observer identity as a random (as opposed to a fixed)
+effect (Link and Sauer 2002, Sauer and Link 2011).  This
+allows observer-specific differences in detection probability,
+but assumes that observers are selected at random from a pool
+of potential observers.  Models that contain both random and
+fixed effects are referred to as mixed models.  Often in mixed
+models (as in this case), the random effect is modeled not
+because it of interest in itself, but to avoid bias in the
+estimates of the fixed effects.
+To include random observer effects in DM models and these extensions,
+equation 5 can be modified to:
+(15)
+(16)
+where nijt is the number of bird recorded at site i by observer j in
+year t, pj is observer-specific detection probability, $\mu_p$ is the mean
+detection probability (on the logit scale), and $\sigma_p$ is the standard
+deviation of the random observer effects (also on the logit scale).
+We have implemented random observer effects in the Bayesian framework
+using program JAGS (version 3.2.0; Plummer 2003) with the R (R
+Development Core Team 2011) package rjags (Plummer 2011) as an
+interface.
+Dynamic range models
+The zero-inflated Poisson distribution can be applied to not only
+initial abundance but also to recruitment and population growth terms.
+For example, the recruitment term of the constant DM model (equation
+4) can be modified as follows:
+(17)
+When both initial abundance and dynamics are modeled as zero-inflated,
+$\psi$ can be modeled as time-varying or not (in either case, $\psi$ can also be
+modeled as varying over space).  In the latter case, it makes sense
+for the model to "remember" which sites are inside the range of the
+species between the initial abundance stage and each time step of the
+dynamics stage (see code sample?).  Adding zero-inflation to the
+dynamics of the autoregressive, trend, Ricker, or Gompertz models
+(without immigration) should have no effect when initial abundance is
+also zero-inflated unless $\psi$ is time-varying (because these dynamic
+models cannot recover from zero abundance).  We have implemented
+zero-inflated dynamics in the Bayesian framework using program JAGS
+(version 3.2.0; Plummer 2003)
+
+\section{Applications}
+
+\subsection{Simulation Study}
+
+We simulated data for 100 sites over 40 years.  All
+simulations assumed initial abundance was Poisson distributed
+and no covariates affected initial abundance, dynamics, or
+detection probability.  Our first series of simulations
+assumed dynamics were exponential.  We ran 500 simulations for
+each combination of low, medium, and high $\Lambda$, $r$, and
+$p$ (Table 1; 27 total combinations).
+
+Our second series of simulations changed dynamics to the
+Ricker model.  We used an initial abundance of 10, a maximum
+growth rate of 0.005, and a detection probability of 0.25, and
+simulated low, medium, and high values of equilibrium
+abundance, each with 500 simulations (Table 1).  Our third
+series of simulations was based on the Ricker + immigration
+dynamics model; here we fixed all parameters the same as the
+Ricker model (with K = 10) and simulated low, medium, and high
+values of immigration rate with 500 simulation each (Table
+1).
+
+We estimated the parameters for each simulation using the same
+initial abundance (Poisson) and dynamics models as were
+simulated, implemented in the unmarked library in R.  When run
+in a maximum likelihood framework, these models require a
+maximum abundance to integrate over (Royle 2004, Dail and
+Madsen 2011); we used 150.  We report bias of estimates, root
+mean squared error, and coverage (percentage of 95% confidence
+% intervals for parameters that
+% overlap the true values).
+Ideas:
+I.	Test importance of random effects modeling, and other differences between JAGS and unmarked
+II.	Model with low p values, or a range of p values
+III.	Ability to distinguish correct initial abundance and dynamics models
+IV.	When can you detect DD if it exists (power)?
+V.	Building on the last two: what if your model set doesn't
+include the correct DD model?  When will the wrong DD model be favored
+over the wrong DID model?
+
+\subsection{Analysis of Breeding Bird Survey Data}
+
+We applied these models to North American Breeding Bird Survey (BBS)
+data from 1966-2010 for two species in the bordering US states
+Maryland and Virginia.  For our focal species, we selected ovenbirds
+(Seiurus aurocapilla), an abundant and widespread forest breeding
+migrant with a stable or increasing trend in the region (CITE) and
+golden-winged warblers (Vermivora chrysoptera), which only breeds in
+the western parts of these states and has been declining in the region
+(CITE).  MORE INFORMATION ABOUT THE SPECIES?
+
+GENERAL ABOUT BBS DATA.  We only used data marked as acceptable for
+use in the annual BBS analysis.  We summed the total birds of each
+species seen on each route and year and used the routes (rather than
+the individual stops) as our sites.
+Strong winds can interfere with point count observers' ability to hear
+birds (Simons et al. 2007)(cite); we tested the effects of wind speed
+on detection probability.  BBS volunteers record wind conditions at
+the beginning and end of each route on the Beaufort Scale (start and
+end wind; Robbins et al. 1986)(start and end wind; 0-9; cite).  When
+start or end wind was not recorded we imputed those values with the
+start or end mean.  We took the mean of start and end wind for each
+route and put this average wind scale value into four categories: 0 $\leq$
+wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind $\geq$ 3 (maximum of 3.5).
+Following Link and Sauer (2002), we also included the first time an
+observer ran a route as a predictor variable for detection
+probability.
+
+HOW WE OBTAINED AND MANIPULATED PRISM DATA
+We ran a series of maximum-likelihood based models for each species,
+and then a series of Bayesian models.  We started by testing three
+models of initial abundance (Poisson, negative binomial, and
+zero-inflated Poisson) with population growth dynamics (equation 7)
+and no covariates.  We selected the minimum Akaike's Information
+Criterion (AIC) model from that set to test three additional models
+for p: wind, first, and wind + first.   We selected the minimum AIC
+model from that set to test eight additional models of dynamics:
+constant, autoregressive, Ricker, Gompertz, autoregressive +
+immigration, population growth + immigration, Ricker + immigration,
+and Gompertz + immigration.  We then tested the effect of average
+minimum temperature for June and July on all dynamics parameters from
+the minimum AIC dynamics model.  When run in a maximum likelihood
+framework, these models require a maximum abundance to integrate over
+(Royle 2004, Dail and Madsen 2011); we used 600 for ovenbirds and 350
+for GWWA.
+
+We ran the top ranked models from the maximum likelihood analyses in a
+Bayesian framework.  We added random observer effects and, where
+appropriate, zero-inflated dynamics.  WILL ADD MORE HEREWe used
+non-informative priors.  We tested for lack of convergence using X
+Markov chains for each model (Gelman and Rubin 1992).  For each chain
+we sampled the MCMC for Y iterations, after Z tuning samples.  Latent
+inclusion factors for temperature effects?
+
+Results
+
+The negative binomial distribution was strongly supported for ovenbird
+initial abundance over the Poisson and zero-inflated Poisson (Table
+1A, models A.1 - A.3).  Even compared to the Poisson, there was no
+evidence to support a zero-inflation factor for this species.  The
+best supported model for p was additive effects of wind speed and
+first run (Table 1A, model A.4).  First run and increasing wind speeds
+both decreased p.  All dynamics models with immigration were better
+supported than models without immigration (Table 1A, models A.8 -
+A.16); the best supported of these was the Ricker + Immigration.  The
+best supported model of those without immigration was the Gompertz,
+and the autoregressive and the constant model had the least support.
+There was little support for an effect of minimum June and July
+temperature on dynamics parameters of the Ricker + Immigration model
+(Table 1A, models A.17 - A.18).
+
+Estimates for the top ranked model for OVEN
+(NB[$\Lambda (.) \alpha (.)$]Ricker+Imm.[$r(.)K(.) \iota (.)$]p(wind+1st)) were similar when
+run in the Bayesian framework, except that the estimate of r more than
+halved (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for Bayesian model
+estimates we present mean and SD]).  When random observer effects were
+added, estimates for $\Lambda$ and K increased dramatically (from 31.6 $\pm$ 4.9
+to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1, respectively), and
+the estimate for r was intermediate (0.018 $\pm$ 0.006).  The estimate of
+the intercept for p (on the logit scale) dropped from -1.5 $\pm$ 0.1 to
+-2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  [Richard, I ran
+the temperature model in the Bayesian framework (w/ and w/o random
+effect) too.  Should we include those results?  As earlier, very
+little evidence for an effect of temperature on any of the parameters,
+although I couldn't get the latent inclusion parameter to work for r.]
+Gelman and Rubin diagnostics and visual examination of the ? plots
+provided some evidence for a lack of convergence in estimates of the
+intercept for p for all Bayesian models ran.
+
+There was slightly more support for the ZIP distribution for initial
+abundance of GWWA than for the negative binomial; both were strongly
+supported over the Poisson (Table 1B, models B.1 - B.3).  The best
+supported model for p was an effect of first run (Table 1B, model
+B.4).  The best supported dynamics model was population growth (Table
+1B, model B.8); the estimate of r from this model was -0.058 $\pm$ 0.017
+(SE).  There was no support for immigration added to the population
+growth or autoregressive models (Table 1B, models B.10 and B.13).
+However, the Gompertz + immigration and Ricker + immigration models
+were supported over the corresponding models without immigration
+(Table 1B, models B.11, B.12, B.14, and B.15), despite low estimates
+of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$ 0.002,
+respectively).  There was considerable support for an effect of
+average minimum June and July temperature on r in the population
+growth model (Table 1B, model B.17), with a negative slope.
+
+Because of the closeness of rankings for the ZIP and negative binomial
+distributions for GWWA, we also ran the subsequent models with the
+negative binomial initial abundance.  Rankings of subsequent models
+were similar, but with a p covariate (first) the negative binomial
+distribution outranked the ZIP.
+
+STILL WORKING ON JAGS GWWA RESULTS
+Estimated detection probabilities were generally low.  For OVEN, the
+estimated probability of detecting a bird with no wind and not the
+first time an observer had run a route varied between 0.046 $\pm$ 0.002
+(model A.16) and 0.192 $\pm$ 0.015 (model A.11), among models that
+accounted for both.  For GWWA, the estimated probability of detecting
+a bird by a non-first time observer varied between 0.025 $\pm$ 0.008
+(model B.18) and 0.141 $\pm$ 0.027 (model B.16), among models that
+accounted for first but not wind.
+
+For both species, constant dynamics models estimated very high
+survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05 for OVEN and 0.935 $\pm$ 0.011 for
+GWWA).   The autoregressive and autoregressive + immigration models
+estimated very low survival probabilities for OVEN ($\omega$ = 0.058 $\pm$ 0.046
+and 0.026 $\pm$ 0.060, respectively), but not for GWWA ($\omega$ = 0.715 $\pm$ 0.288
+for both models).
+
+\section{Discussion}
+
+	We made four important developments of the open population
+        N-mixture model proposed by Dail and Madsen that make the
+        model more applicable to long-term data such as is collected
+        by the North American Breeding Bird Survey. First, we have
+        reconciled the objectives of traditional state-space models
+        with open population N-mixture models by illustrating how
+        classical models of population dynamics can be embedded within
+        the framework. Second, we have demonstrated methods of
+        accounting for zero-inflation in the time-series. Third, we
+        have illustrated how additional random effects such as
+        observer-specific detection probabilities can be
+        accommodated. Fourth, we presented a Bayesian analysis of the
+        model, which makes it much easier to estimate parameters and
+        to fine tune the model for specific purposes.
+
+A unique aspect of the Dail and Madsen model as originally devised was
+that it allowed for the estimation of demographic parameters under
+strict distributional assumptions and with the assumption of
+geographic closure. Clearly, estimating demographic parameters from
+count data is a loftyn ambitious goal, and the required assumption
+will not be valid in many cases. Nonetheless, their approach is
+important because it allows for the combination of count data with
+other data types such as capture-recapture data , which offers a
+means of making inferences about demographic processes at broad
+spatial scales. Models combining count data and demographic data are
+often referred to as integrated population models, and This topic has
+received much attention lately under t
+
+In the absence of direct information about demographic parameters, and
+when the original assumptions of the model do not hold, we have
+offered extensions of the model to estimate derived parameters such as
+population growth rate. This has been the traditional emphasis of
+state-space models, and our approach resolves many of the factors
+limiting ...
+
+[paragraph on zero-inflation]The zero-inflated models for initial
+abundance and dynamics allow one to estimate and model the proportion
+of sites outside of the range of a species.  This could be especially
+useful when different factors control the range than the abundance or
+dynamics within the range.  [some examples where that is the case?]
+
+ [paragraph on random effects]
+[paragraph on Bayesian]
+In spite of the new extensions we have proposed, several aspects of
+the model could be improved. First, the precision with which the
+parameters of the state process can be estimated ultimately depends
+upon how well detection probability is estimated. When there is only a
+single survey per primary period, the information about detection
+probability comes from deviations from the parametric assumptions
+about population dynamics. Thus, without direct information about
+detection probability, the estimates will be determined by model based
+assumptions. Furthermore, there are multiple components of detection
+probability that should be accounted for to minimize bias and yield
+valid estimates of population size. [Nichols EURING paper] .
+
+Fortunately, it is easy to incorporate direct information about
+detection probability and we recommend that this be done whenever
+possible. The original method for proposed in the original paper was
+to collected replicated counts within the primary periods during a
+period in which the populations could be assumed to be closed.  a
+robust design could be used to combine multiple surveys per primary
+period to increase the precision of the estimates. We envision that
+multiple other options are available as well. For instance, there
+should be no difficulty extending the model to accommodate traditional
+capture-recapture data collected during each primary period. Removal
+or time of detection etc.. Distance-related heterogeneity in detection
+probability is another source of variation that can bias estimators of
+abundance.   Show some equations for the alternative observation
+models.
+
+[paragraph on spatially-explicit dispersal]
+ [Other possible extensions we could mention:
+	Spatially hierarchical model (so can use individual stops)
+	Other models of variability in growth or recruitment (such as negative binomial)
+	Add density dependence to survival and recruitment models]
+[Discuss causes of low p values (and other BBS results) here or in BBS
+results?]
+
+The modeling framework we described can be used to address many of the
+most pressing issue in ecology and conservation biology. For example,
+it is possible to test hypotheses about temporal and spatial
+population regulation. Furthermore, the Bayesian approach is very
+useful in that it can be used to combine multiple sources of data to
+develop mechanistic models of population dynamics. Thus one can test
+hypotheses about the effects of climate change on either explicit
+demographic parameters or in derived parameters such as population
+growth rate. Furthermore, under the Bayesian approach , population
+viability analysis is trivial because projecting populations into the
+future can be done as a component of the MCMC analysis. This allows
+for the computation of posterior distributions of parameters such as
+quasi-extinction probability.
+
 \bibliography{Dail_Mad_ext}

### 2012-08-31 Dail_Mad_ext.tex
formatting

@@
-\documentclass{article}
+\documentclass[12pt]{article}
 
-\usepackage[a4paper, hmargin={2.4cm,2.4cm}, vmargin={2.4cm,2.4cm}]{geometry}
+\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
 \usepackage[mathlines,displaymath]{lineno}
@@
 resolved using a class of hierarchical models for spatially-replicated
-time-series data recently proposed by \citet{dail_madsen:2011}. We
+time-series data recently proposed by
+\citet[Biometrics]{dail_madsen:2011}. We
 expand this class of models to accommodate classical growth
@@
 studying population dynamics while accounting for process variation
-and observation error (de Valpine and Hastings 2002, Buckland et
-al. 2004, Dennis et al. 2006). Conventional state-space models are
+and observation error \citep{devalpine_hastings:2002,
+  buckland_etal:2004, dennis_taper:1994,
+  dennis_etal:2006}. Conventional state-space models are
 simply time-series models in which the true state of the system is
@@
 applications [need to clean this up]. State-space models have been
-defined differently by different authors, but we follow Buckland et
-al. (2004) using the definition : ``xxx'' . Typically, the ecological
+defined differently by different authors, but we follow
+\citet{buckland_etal:2004} the definition : ``xxx'' . Typically, the
+ecological
 process is described by a Markovian model that includes a
@@
 observation process is typically phenomenological in the sense that
-observation error is not defined explicitly. As an example, let  be
-the abundance of a species during year  , and   be the observed data,
-which differs from  due to observation error ( ). A simple state space
-model that includes exponential growth and process variation ( ) could
-be written as
-	 	(1)
+observation error is not defined explicitly. As an example, let $N_t$ be
+the abundance of a species during year $t$, and $X_t$ be the observed data,
+which differs from  due to observation error ($\zeta$). A simple state
+space model that includes exponential growth and process variation
+($\grave{o}$) could be written as:
+\begin{gather}
+N_t = N_{t-1}\exp(r)+\grave{o}_t \\
+X_t = N_t + \zeta_t
+\end{gather}
 Here   is the intrinsic rate of increase. Typically, both random

### 2012-08-31 Dail_Mad_ext.tex
formatted intro

@@
 Here   is the intrinsic rate of increase. Typically, both random
-effects are assumed to follow normal distributions:   and   (de
-Valpine and Hastings 2002, Dennis et al. 2006). Even though this is
+effects are assumed to follow normal distributions: $\grave{o}_t \sim
+\mathrm{N}(0, \sigma)$ and $\zeta_t \sim \mathrm{N}(0, \tau)$
+\citep{devalpine_hastings:2002,dennis_etal:2006}. Even though this is
 the most widespread approach for modeling population dynamics using
@@
 random effects violate this constraint. To avoid this issue,
-ecologists often replace   with   where A is the area surveyed---i.e.,
+ecologists often replace $N_t$ with $\log(Y_t/A)$   where $A$ is the
+area surveyed---i.e.,
 abundance is replaced by the natural log of population density, which
@@
 alternatives. However, this transformation is problematic since
-abundance may be zero and thus . Generally, zeros are replaced with
+abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
+zeros are replaced with
 some small, but positive number, although the effect of this is rarely
@@
 The first term is the probability density for the first observation in
-the time series,  , yet it is intuitively not possible to estimate the
+the time series, $Y_t$, yet it is intuitively not possible to estimate the
 parameter(s) of this distribution using only a single
@@
 
-Dail and Madsen (2011) developed a model (henceforth the DM model)
+\citet{dail_madsen:2011} developed a model (henceforth the DM model)
 that resolves each of these problems with traditional state-space
@@
 only spatially and temporally replicated count data.  Their model is
-an extension of the closed-population N-mixture model (Royle
-2004)(2004), which was designed specifically to address the problem of
+an extension of the closed-population N-mixture model
+\citep{royle:2004biom}, which was designed specifically to address the
+problem of
 modeling spatial variation in abundance when detection probability is
@@
 spatially and temporally extensive datasets such as result from the
-North American Breeding Bird Survey (BBS) (Robbins et
-al. 1986). However, most of the datasets of the appropriate scale are
+North American Breeding Bird Survey (BBS)
+\citep{robbins_etal:1986}. However, most of the datasets of the
+appropriate scale are
 collected by volunteers and using protocols that are not amenable to

### 2012-08-31 Dail_Mad_ext.tex
methods

@@
 \usepackage{amsmath}
+\usepackage{amsfonts}
 \usepackage{graphicx}
@@
 \begin{gather}
-N_t = N_{t-1}\exp(r)+\grave{o}_t \\
+N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
 X_t = N_t + \zeta_t
@@
 
-\section{Methods}
+\section{Model Development}
 
-\subsection{Model Development}
+\subsection{Original Formulation}
 
@@
 conditionally related processes: the initial abundance state, apparent
-survival, recruitment, and detection probability. Dail and Madsen
-proposed two options for initial abundance (Poisson and negative
-binomial):
-(2)
-(3)
-where $N_{i1}$  is the abundance at site $i$  during year 1 and $\Lambda$ is
-the expected abundance. The negative binomial model includes the
-overdispersion parameter $\alpha$.  The survival and recruitment terms allow
-abundance to change over time:
-(4)
-where $\omega$ is the survival probability and $\gamma$ is the arrival rate (which
-can depend on Nit-1). Dail and Madsen propose three models for $\gamma$: the
-constant model, where $\gamma$ does not depend on Nit-1, and which simulates
-a ``propagule rain'' of recruitment; the autoregressive model , which
+survival, recruitment, and detection probability. \citet{dail_madsen:2011}
+proposed two distributions for initial abundance, the Poisson and negative
+binomial:
+\begin{gather}
+N_{it} \sim \mathrm{Pois}(\Lambda) \nonumber \\
+N_{it} \sim \mathrm{NB}(\Lambda, \alpha)
+\label{eq:N1}
+\end{gather}
+where $N_{i1}$ is the abundance at site $i=1,\hdots,R$ during year 1
+and $\Lambda=\mathbb{E}[N_{i1}]$ is the expected abundance. The negative
+binomial model includes the overdispersion parameter $\alpha$.  The
+survival and recruitment terms allow abundance to change over time:
+\begin{equation}
+\left.\begin{aligned}
+S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
+G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
+N_{it} &= S_{it}+G_{it}
+\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
+\label{eq:Nt}
+\end{equation}
+where $\omega$ is the survival probability and $\gamma$ is the arrival
+rate (which can depend on $N_{it-1}$). Dail and Madsen propose three models
+for $\gamma$: the constant model, where $\gamma$ does not depend on
+$N_{it-1}$, and which simulates a ``propagule rain'' of recruitment; the
+autoregressive model , which
 simulates geometric or density independent growth; and the ``no-trend''
 model, which keeps abundance fixed over time.
-The number seen is a function of detection probability:
-(5)
+
+Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model---the
+model for spatial and temporal variation in abundance. The observation
+model assumes that individuals are missed due to imperfect
+detection. The simplest model for imperfect detection is
+\begin{equation}
+  \label{eq:2}
+  n_{it} \sim \mathrm{Bin}(N_{it}, p)
+\end{equation}
+
 Initial abundance, apparent survival, recruitment, and detection can
@@
 point count data.
-Range model of initial abundance
-Dail and Madsen (2011) suggested two distributions for modeling
+
+\subsection{Range model of initial abundance}
+
+\citet{dail_madsen:2011} suggested two distributions for modeling
 initial abundance: Poisson and negative binomial.  We have extended
@@
 initial abundances can be represented as:
-(6)
+\begin{equation}
+N_{i1} \sim \left\{
+\begin{aligned}
+0 &\; \text{with probability} \; \psi \\
+\Lambda &\; \text{with probability} \; (1-\psi)
+\end{aligned} \right.
+\label{eq:ZIP}
+\end{equation}
 where $\psi$ represents the proportion of extra zeros.
+
 These models allow three sources of zero counts by observers: a
@@
 for distinguishing species' fundamental and realized niches
-(Hutchinson 1957).   The zero inflation factor can also affect
+(Hutchinson 1957).
+
+The zero inflation factor can also affect
 recruitment or population dynamics (see below).

### 2012-08-31 Dail_Mad_ext.tex
formatting

@@
 \usepackage{fancyvrb}
+\usepackage{verbatim}
 
@@
 climate.  This approach combines elements of occupancy modeling
-(MacKenzie et al. 2006) and abundance modeling, and could prove useful
-for distinguishing species' fundamental and realized niches
-(Hutchinson 1957).
+\citep{mackenzie_etal:2006} and abundance modeling.
+%, and could prove useful
+%for distinguishing species' fundamental and realized niches
+%(Hutchinson 1957).
 
@@
 abundance, but this model did not perform well in preliminary tests.
-Population growth models
+
+\subsection{Population growth models}
+
 Preliminary DM model runs for several species tended to lead to
@@
 abundance to survival and recruitment in part by making strong
-distributional assumptions (Dail and Madsen 2011).  When those
-assumptions are heavily violated the models may proportion population
-growth incorrectly into survival and recruitment, even if they
-estimates population growth accurately.
+distributional assumptions. % (Dail and Madsen 2011).
+When those assumptions are heavily violated the models may proportion
+population growth incorrectly into survival and recruitment, even if
+they estimates population growth accurately.
+
 Although partitioning population growth into survival and recruitment
@@
 Therefore, we developed a version of the DM model that estimates and
-models population growth directly.  In this model, equation 4 is
+models population growth directly.  In this model, Eq.\ref{eq:Nt} is
 simplified to:
-(7)
-where r represents the instantaneous population growth rate.  This,
+\begin{equation}
+  N_{it} \sim \text{Pois}(\exp(r)N_{it-1})
+\label{eq:exp}
+\end{equation}
+where $r$ represents the instantaneous population growth rate.  This,
 like the autoregressive version of the model, is a variant on a simple
@@
 example:
-(8)
-where K is the stable equilibrium of the population and r is the
+\begin{equation}
+  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K))
+\label{eq:rick}
+\end{equation}
+where $K$ is the stable equilibrium of the population and $r$ is the
 instantaneous population growth rate at low population densities, and
 both parameters are constrained to be positive.  This model is based
-on Ricker's (1954) discrete time version of the logistic population
-growth model.  We also implemented Gompertz's (1825) density-dependent
+on the \citet{ricker:1954} discrete time version of the logistic population
+growth model.  We also implemented the \citet{gompertz:1825} density-dependent
 model:
-(9)
-where the log of Nit-1 is only taken when Nit-1 is greater than 0
-(when Nit-1 is 0, the full expression simplifies to 0 anyway).  Here
-the interpretations of r and K are similar to in the Ricker model, but
-K is constrained to be greater than one.
+\begin{equation}
+  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1})/\log(K)))
+\label{eq:gomp}
+\end{equation}
+where the log of $N_{it-1}$ is only taken when $N_{it-1}>0$
+(when $N_{it-1} \equiv 0$, the full expression simplifies to 0 anyway).  Here
+the interpretations of $r$ and $K$ are similar to in the Ricker model, but
+$K$ is constrained to be greater than one.
+
 Because a single Poisson distribution controls the
-distribution of Nit in each of these models, the discrete
-convolution used by Dail and Madsen (2011) is not required,
+distribution of $N_{it}$ in each of these models, the discrete
+convolution used by \citet{dail_madsen:2011} is not required,
 speeding up processing time.
-Immigration and colonization models
+
+\subsection{Immigration models}
+
 The autoregressive, population growth, Ricker, and Gompertz
@@
 include both internal and external (immigration) contributions
-to population growth.  The population growth + immigration
+to population growth.  The population growth plus immigration
 model is:
-(10)
-or
-(11)
-where $\iota$ represents the immigration rate (equations 10 and 11 are
-equivalent).  This model is close to the constant DM model (equation
-4), with exp(r) instead of $\omega$ and $\iota$ instead of $\gamma$, except that the first
-process is Poisson distributed instead of binomial.  Similarly, the
-Ricker + immigration model can be represented as:
-(12)
-the Gompertz + immigration model as:
-(13)
-and the autoregressive + immigration model as:
-(14)
-These immigration models are based on the assumption that numbers
-of immigrants are not dependent on local density.  Other versions
-of these models with different assumptions are possible. The
-Ricker + immigration and Gompertz + immigration models can be
-justified by the further assumption that immigrants arrive as
+\begin{equation}
+  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1}) + \text{Poisson}(\iota)
+  \label{eq:expimm1}
+\end{equation}
+or equivalently
+\begin{equation}
+  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
+  \label{eq:expimm2}
+\end{equation}
+where $\iota$ represents the immigration rate.  This model is close to
+the constant DM model (equation 4), with $\exp(r)$ instead of $\omega$
+and $\iota$ instead of $\gamma$, except that the first
+process is Poisson distributed instead of binomial. The Ricker and
+Gompertz models can be extended to allow for immigration in the same
+way.
+\begin{comment}
+  Similarly, the Ricker + immigration model can be represented as: the
+  Gompertz + immigration model as: (13) and the autoregressive +
+  immigration model as: (14)
+\end{comment}
+
+% Save for discussion
+\begin{comment}
+  These immigration models are based on the assumption that number of
+  immigrant is not dependent on local density; however, extending
+  these models in this way is conceptually straight-forward, a Other
+  versions of these models with different assumptions are
+  possible. The Ricker + immigration and Gompertz + immigration models
+  can be justified by the further assumption that immigrants arrive as
   adults immediately before the point-count survey (so that
@@
   equilibriums are not at K and cannot be solved for analytically
-(Otto and Day 2007).  We have implemented all preceding models in
-a maximum likelihood framework by extending the unmarked package
-(Fiske and Chandler 2011) in R (R Development Core Team 2011).
-Random eEffects of oObservers
+  (Otto and Day 2007).
+\end{comment}
+
+We have implemented all preceding models in
+a maximum likelihood framework by extending the \texttt{unmarked} package
+\citep{fiske_chandler:2011} in \textbf{R} \citep{R-2012}.
+
+\subsection{Random effects of observers}
+
 Differences in observers' ability to see, hear, or identify
@@
 could not be accurately estimated).
+
 Current BBS trend estimators deal with this problem by
@@
 estimates of the fixed effects.
+
 To include random observer effects in DM models and these extensions,
@@
 interface.
-Dynamic range models
+
+\subsection{Dynamic range models}
+
 The zero-inflated Poisson distribution can be applied to not only
@@
 
-Results
+\section{Results}
 

### 2012-09-04 Dail_Mad_ext.tex
formatting

@@
 \usepackage{verbatim}
+\usepackage[colorlinks=true,citecolor=black,linkcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
 
@@
 \begin{equation}
-  \label{eq:2}
   n_{it} \sim \mathrm{Bin}(N_{it}, p)
+  \label{eq:p1}
 \end{equation}
@@
 birds has long been recognized as a potential source of error
-in avian point count surveys such as the BBS (Robbins et
-al. 1986, Sauer et al. 1994, Diefenbach et al. 2003, Alldredge
-et al. 2007, Campbell and Francis 2011).  Estimating a
-separate detection probability for each observer can be
+in avian point count surveys such as the BBS
+\citep{robbins_etal:1986,diefenbach_etal:2003,sauer_etal:1994auk,alldredge_etal:2007auk,campbell_francis:2011}.
+Estimating a separate detection probability for each observer can be
 difficult and reduces one's ability to estimate the quantities
@@
 treating observer identity as a random (as opposed to a fixed)
-effect (Link and Sauer 2002, Sauer and Link 2011).  This
-allows observer-specific differences in detection probability,
+effect \citep{link_sauer:2002,sauer_link:2011}.
+%(Link and Sauer 2002, Sauer and Link 2011).
+This allows observer-specific differences in detection probability,
 but assumes that observers are selected at random from a pool
@@
 To include random observer effects in DM models and these extensions,
-equation 5 can be modified to:
-(15)
-(16)
-where nijt is the number of bird recorded at site i by observer j in
-year t, pj is observer-specific detection probability, $\mu_p$ is the mean
+Eq.~\ref{eq:p1} can be modified to:
+\begin{gather}
+n_{ijt} \sim \mathrm{Bin}(N_{it}, p_j) \nonumber \\
+\mathrm{logit}(p_j) \sim \mathrm{Normal}(\mu_p, \sigma_p)
+\label{eq:pobs}
+\end{gather}
+where $n_{ijt}$ is the number of bird recorded at site $i$ by observer $j$ in
+year $t$, $p_j$ is observer-specific detection probability, $\mu_p$ is the mean
 detection probability (on the logit scale), and $\sigma_p$ is the standard
@@
 We have implemented random observer effects in the Bayesian framework
-using program JAGS (version 3.2.0; Plummer 2003) with the R (R
-Development Core Team 2011) package rjags (Plummer 2011) as an
+using program \textbf{JAGS} \citep[version 3.2.0]{plummer:2003}
+% (version 3.2.0; Plummer 2003)
+with the \textbf{R} (R
+Development Core Team 2011) package \texttt{rjags} \citep{plummer:2011}
+%(Plummer 2011)
 interface.
@@
 4) can be modified as follows:
-(17)
+\begin{equation}
+G_{it} \sim \left\{
+\begin{aligned}
+0 &\; \text{with probability} \; \psi \\
+\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)
+\end{aligned} \right.
+\label{eq:ZIPts}
+\end{equation}
 When both initial abundance and dynamics are modeled as zero-inflated,
@@
 models cannot recover from zero abundance).  We have implemented
-zero-inflated dynamics in the Bayesian framework using program JAGS
-(version 3.2.0; Plummer 2003)
+zero-inflated dynamics in the Bayesian framework using program
+\textrm{JAGS} \citep[version 3.2.0]{plummer:2003}.
+%(version 3.2.0; Plummer 2003)
 
@@
 detection probability.  Our first series of simulations
-assumed dynamics were exponential.  We ran 500 simulations for
-each combination of low, medium, and high $\Lambda$, $r$, and
-$p$ (Table 1; 27 total combinations).
+assumed dynamics were exponential.  We ran 1000 simulations for
+each combination of low, medium, and high $\Lambda \in \{1,5,10\}$, $r
+\in \{-0.005, 0, 0.005\}$, and
+$p \in \{0.05, 0.25, 0.5\}$. % (Table 1; 27 total combinations).
 
@@
 initial abundance (Poisson) and dynamics models as were
-simulated, implemented in the unmarked library in R.  When run
+simulated, implemented in the unmarked library in \textbf{R}.  When run
 in a maximum likelihood framework, these models require a
-maximum abundance to integrate over (Royle 2004, Dail and
-Madsen 2011); we used 150.  We report bias of estimates, root
-mean squared error, and coverage (percentage of 95% confidence
-% intervals for parameters that
-% overlap the true values).
+maximum abundance to integrate over
+\citep{royle:2004biom,dail_madsen:2011};
+%(Royle 2004, Dail and Madsen 2011);
+we used 200.  We report bias of estimates, root
+mean squared error, and coverage (percentage of 95\% confidence
+intervals for parameters that
+overlap the true values).
+
 Ideas:
@@
 Maryland and Virginia.  For our focal species, we selected ovenbirds
-(Seiurus aurocapilla), an abundant and widespread forest breeding
+(\textit{Seiurus aurocapilla}), an abundant and widespread forest breeding
 migrant with a stable or increasing trend in the region (CITE) and
-golden-winged warblers (Vermivora chrysoptera), which only breeds in
+golden-winged warblers (\textit{Vermivora chrysoptera}), which only breeds in
 the western parts of these states and has been declining in the region
@@
 Strong winds can interfere with point count observers' ability to hear
-birds (Simons et al. 2007)(cite); we tested the effects of wind speed
+birds (\citep{simons_etal:2007}); we tested the effects of wind speed
 on detection probability.  BBS volunteers record wind conditions at
-the beginning and end of each route on the Beaufort Scale (start and
-end wind; Robbins et al. 1986)(start and end wind; 0-9; cite).  When
+the beginning and end of each route on the Beaufort Scale
+\citep[start and end wind 0-9]{robbins_etal:1986}.
+%(start and end wind; Robbins et al. 1986)(start and end wind; 0-9; cite).
+When
 start or end wind was not recorded we imputed those values with the
@@
 wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind $\geq$ 3 (maximum of 3.5).
-Following Link and Sauer (2002), we also included the first time an
+Following \citet{link_sauer:2002},
+%Link and Sauer (2002),
+we also included the first time an
 observer ran a route as a predictor variable for detection
@@
 non-informative priors.  We tested for lack of convergence using X
-Markov chains for each model (Gelman and Rubin 1992).  For each chain
+Markov chains for each model \citep{gelman_rubin:1992}.
+%(Gelman and Rubin 1992).
+For each chain
 we sampled the MCMC for Y iterations, after Z tuning samples.  Latent
@@
 for both models).
+
+\begin{figure}
+  \centering
+  \includegraphics[height=8in]{figs/exp_hists}
+  \caption{Histograms of 1000 parameter estimates for each of 27 simulation cases. The verticle lines are the data-generating values.}
+\label{fig:exp_hists}
+\end{figure}
 

### 2012-09-04 Dail_Mad_ext.tex
bib, intro

@@
 \usepackage{verbatim}
-\usepackage[colorlinks=true,citecolor=black,linkcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
+\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
 
@@
 \begin{document}
+
+\title{Improved state-space models for inference about spatial and
+  temporal variation in abundance}
+
+\author{Richard B. Chandler  \\
+        \small USGS Patuxent Wildlife Research Center \\
+        \small Laurel, MD 20708
+   \and Jeffrey A. Hostetler \footnote{Contact:
+     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
+%        \small Migratory Bird Center \\
+        \small Smithsonian Conservation Biology Institute \\
+        \small National Zoological Park, MRC 5503 \\
+        \small Washington, DC 20013-7012}
+\maketitle
 
@@
 to future environmental change or management actions. State-space
-models, are widely used for these purposes, and they allow for
+models are widely used for these purposes because they allow for
 estimating parameters of classical growth models, such as exponential,
@@
 data from the North American Breeding Bird Survey. Code to fit these
-models using the R package unmarked and JAGS is also provided.
+models using the \textbf{R} package \texttt{unmarked} and
+\textbf{JAGS} is also provided.
 }
+
+\vspace{0.5cm}
 
@@
 
-Theoretical ecology requires robust models for testing hypotheses
-regarding density dependence and spatial variation in population
-dynamics. Such models are also required in applied ecological research
+Theoretical ecology requires models of population dynamics for testing
+hypotheses regarding spatial and temporal variation in abundance, for
+example to test for the existence of density-dependence population
+growth \citep{dennis_taper:1994,viljugrein_etal:2005}. Such models are
+also required in applied ecological research
 for predicting species' responses to environmental change or
-management actions. For example, many researchers are currently
-investigating the influence of weather on population dynamics so that
-predictions can be made about the effects of future climate
-change. Several complicating factors must be confronted when
+conservation actions. For example, %many researchers are currently
+models of population dynamics allow for inference about the effects of
+of climate change on population viability.
+%investigations of the influence of climate on population dynamics so that
+%predictions can be made about the effects of climate change.
+Several complicating factors must be confronted when
 addressing these questions. First, deterministic models of population
-dynamics are virtually always inadequate due to ``process variation'',
+dynamics are virtually always inadequate due to process variation,
 the inherent stochasticity in demographic parameters and environmental
@@
 of population dynamics---can rarely be observed perfectly in field
-studies because of ``observation error'', such as imperfect
-detection. Failure to account for  and heterogeneity in observation
+studies because of observation error, such as imperfect
+detection. Failure to account for process variation and observation
 error can bias the estimators of abundance and related parameters.

### 2012-09-05 Dail_Mad_ext.tex
intro and lit

@@
 \usepackage{verbatim}
-\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
+\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
 
@@
 Theoretical ecology requires models of population dynamics for testing
-hypotheses regarding spatial and temporal variation in abundance, for
-example to test for the existence of density-dependence population
-growth \citep{dennis_taper:1994,viljugrein_etal:2005}. Such models are
-also required in applied ecological research
-for predicting species' responses to environmental change or
-conservation actions. For example, %many researchers are currently
-models of population dynamics allow for inference about the effects of
-of climate change on population viability.
+hypotheses regarding spatial and temporal variation in
+abundance---hypotheses relating to the importance and existence of
+phenomenon such as density-dependent population regulation, population
+cycling, and spatial synchrony
+\citep{royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999}.
+%growth ,viljugrein_etal:2005}.
+In applied contexts, population models are used for %population viability
+%analysis, including the estimation of
+estimating extinction probabilities
+\citep{nadeem_lele:2011} and for predicting the effects of future
+environmental conditions or conservation actions on population size
+\citep{jamieson_brooks:2004}.
+%Such models are
+%also required in applied ecological research
+%for predicting species' responses to environmental change or
+%conservation actions. For example, %many researchers are currently
+%models of population dynamics allow for inference about the effects of
+%of climate change on population viability.
 %investigations of the influence of climate on population dynamics so that
 %predictions can be made about the effects of climate change.
-Several complicating factors must be confronted when
-addressing these questions. First, deterministic models of population
+In order to address these questions, several complicating factors must
+be confronted when fitting population models to data.
+First, deterministic models of population
 dynamics are virtually always inadequate due to process variation,
 the inherent stochasticity in demographic parameters and environmental
-conditions. Second, abundance---the natural state variable in studies
+conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
+Second, abundance---the natural state variable in studies
 of population dynamics---can rarely be observed perfectly in field
 studies because of observation error, such as imperfect
-detection. Failure to account for process variation and observation
-error can bias the estimators of abundance and related parameters.
+detection \citep{kery_etal:2009}. Failure to account for process
+variation and observation error can bias the estimators of abundance
+and related parameters.
 
-State-space models have become the most widespread approach for
+State-space models are a widely used approach for
 studying population dynamics while accounting for process variation
@@
   dennis_etal:2006}. Conventional state-space models are
-simply time-series models in which the true state of the system is
+simply time-series models in which the true state of the system
+(e.g. population size during each year) is
 observed imperfectly. Originally developed for xxx (cite Kalman filter

### 2012-09-05 Dail_Mad_ext.tex
intro and lit

@@
 \title{Improved state-space models for inference about spatial and
-  temporal variation in abundance}
+  temporal variation in abundance from count data}
 
@@
 cycling, and spatial synchrony
-\citep{royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999}.
-%growth ,viljugrein_etal:2005}.
-In applied contexts, population models are used for %population viability
-%analysis, including the estimation of
+\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999}.
+In applied contexts, population models are used for
 estimating extinction probabilities
-\citep{nadeem_lele:2011} and for predicting the effects of future
-environmental conditions or conservation actions on population size
-\citep{jamieson_brooks:2004}.
-%Such models are
-%also required in applied ecological research
-%for predicting species' responses to environmental change or
-%conservation actions. For example, %many researchers are currently
-%models of population dynamics allow for inference about the effects of
-%of climate change on population viability.
-%investigations of the influence of climate on population dynamics so that
-%predictions can be made about the effects of climate change.
+\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
+effects of future environmental conditions or conservation actions on
+population size \citep{jamieson_brooks:2004}.
 In order to address these questions, several complicating factors must
@@
 studies because of observation error, such as imperfect
-detection \citep{kery_etal:2009}. Failure to account for process
-variation and observation error can bias the estimators of abundance
-and related parameters.
+detection \citep{link_nichols:1994,kery_etal:2009}.
+%Failure to account for process
+%variation and observation error can bias the estimators of abundance
+%and related parameters.
 
@@
 and observation error \citep{devalpine_hastings:2002,
-  buckland_etal:2004, dennis_taper:1994,
-  dennis_etal:2006}. Conventional state-space models are
-simply time-series models in which the true state of the system
-(e.g. population size during each year) is
-observed imperfectly. Originally developed for xxx (cite Kalman filter
-papers), the ability to model both the ecological state process and
-the observation process has made them relevant in numerous ecological
-applications [need to clean this up]. State-space models have been
-defined differently by different authors, but we follow
-\citet{buckland_etal:2004} the definition : ``xxx'' . Typically, the
-ecological
-process is described by a Markovian model that includes a
-deterministic description of population change as well as two random
-sources of variation for ``process variation''. The model for the
-observation process is typically phenomenological in the sense that
-observation error is not defined explicitly. As an example, let $N_t$ be
-the abundance of a species during year $t$, and $X_t$ be the observed data,
-which differs from  due to observation error ($\zeta$). A simple state
+  buckland_etal:2004, dennis_etal:2006}. Conventional state-space
+models are simply time-series models in which the true state of the
+system (e.g. population size during each year) is observed
+imperfectly. One reason for the widespread adoption of state-space
+models in ecology is that failure to account for process variation and
+observation error can bias estimators of abundance and population
+growth parameters \citep{kery_etal:2009}. For instance, the strength of
+density dependence will be over-estimated if observation error is
+ignored \citep{link_nichols:1994,shenk_etal:1998}.
+%Originally developed for xxx (cite Kalman filter papers),
+%The ability to model both the ecological state process and
+%the observation process has made them relevant in numerous ecological
+%applications [need to clean this up]. State-space models have been
+%defined differently by different authors, but we follow
+%\citet{buckland_etal:2004} the definition : ``xxx'' .
+
+One of the simplest state-space models can be described as follows.
+% In their simplest form, state-space models include a Markovian model
+%for the ecological process of interestTypically, the
+%ecological
+%process is described by a Markovian model that includes a
+%deterministic description of population change as well as two random
+%sources of variation for ``process variation''. The model for the
+%Robservation process is typically phenomenological in the sense that
+%observation error is not defined explicitly. As an example,
+Let $N_t$ be the abundance of a species during year $t$, for
+$t=1,\hdots,T$, and $X_t$ be
+the observed data, which differs from $N_t$ due to observation error
+($\zeta$). Temporal variation in $N_t$ is modeled using a
+deterministic population growth model (e.g. exponential, logistic, etc
+\dots) and stochastic model allowing for random process variation. A
+simple state
 space model that includes exponential growth and process variation

### 2012-09-05 Dail_Mad_ext.tex
plots from Jeff, and state-space description

@@
 $t=1,\hdots,T$, and $X_t$ be
-the observed data, which differs from $N_t$ due to observation error
-($\zeta$). Temporal variation in $N_t$ is modeled using a
-deterministic population growth model (e.g. exponential, logistic, etc
-\dots) and stochastic model allowing for random process variation. A
-simple state
-space model that includes exponential growth and process variation
-($\grave{o}$) could be written as:
+the observed data, which differs from $N_t$ due to observation error,
+a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
+modeled using a deterministic population growth model
+(e.g. exponential, logistic, etc \dots) and stochastic model allowing
+for random process variation, which we denote $\grave{o}$ . As an
+example, under exponential population growth, the model could be
+written as:
 \begin{gather}
@@
 \end{gather}
-Here   is the intrinsic rate of increase. Typically, both random
+where $r$ is the intrinsic rate of increase. Typically, both random
 effects are assumed to follow normal distributions: $\grave{o}_t \sim

### 2012-09-05 Dail_Mad_ext.tex
lunch

@@
 effects of future environmental conditions or conservation actions on
-population size \citep{jamieson_brooks:2004}.
+population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
 In order to address these questions, several complicating factors must
@@
 ignored \citep{link_nichols:1994,shenk_etal:1998}.
-%Originally developed for xxx (cite Kalman filter papers),
-%The ability to model both the ecological state process and
-%the observation process has made them relevant in numerous ecological
-%applications [need to clean this up]. State-space models have been
-%defined differently by different authors, but we follow
-%\citet{buckland_etal:2004} the definition : ``xxx'' .
 
 One of the simplest state-space models can be described as follows.
-% In their simplest form, state-space models include a Markovian model
-%for the ecological process of interestTypically, the
-%ecological
-%process is described by a Markovian model that includes a
-%deterministic description of population change as well as two random
-%sources of variation for ``process variation''. The model for the
-%Robservation process is typically phenomenological in the sense that
-%observation error is not defined explicitly. As an example,
 Let $N_t$ be the abundance of a species during year $t$, for
-$t=1,\hdots,T$, and $X_t$ be
+$t=1,\hdots,T$, and let $X_t$ be
 the observed data, which differs from $N_t$ due to observation error,
@@
 modeled using a deterministic population growth model
-(e.g. exponential, logistic, etc \dots) and stochastic model allowing
-for random process variation, which we denote $\grave{o}$ . As an
+(e.g. exponential, logistic, etc \dots) and a stochastic model allowing
+for random process variation, which we denote $\grave{o}_t$ . As an
 example, under exponential population growth, the model could be
@@
 X_t = N_t + \zeta_t
+\label{eq:ss1}
 \end{gather}
-where $r$ is the intrinsic rate of increase. Typically, both random
-effects are assumed to follow normal distributions: $\grave{o}_t \sim
-\mathrm{N}(0, \sigma)$ and $\zeta_t \sim \mathrm{N}(0, \tau)$
-\citep{devalpine_hastings:2002,dennis_etal:2006}. Even though this is
+where $r$ is the intrinsic rate of increase. Typically, both error terms
+are assumed to be i.i.d random effects distrubted as
+$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
+$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
+estimation using the Kalman filter \citep{dennis_etal:2006}, which is
+much more computationally efficient than estimation methods when
+random effects are not Gaussian distributed
+\citep{devalpine_hastings:2002}.
+
+Exponential growth without density-dependence is rarely observed in
+natural populations, and thus $\exp(r)$ is often substituted for other
+population growth models such as the Ricker or Gompertz. Nonetheless,
+the basic structure shown in Eq.~\ref{eq:ss1} is common to most
+traditional state-space models. However, even though this is
 the most widespread approach for modeling population dynamics using
-time-series data, several problems are immediately evident. First, if
-abundance is a non-negative integer, as it always is, there is clearly
-a problem with this formulation because the normally distributed
-random effects violate this constraint. To avoid this issue,
-ecologists often replace $N_t$ with $\log(Y_t/A)$   where $A$ is the
-area surveyed---i.e.,
-abundance is replaced by the natural log of population density, which
-is assumed to normally distributed. This is desirable because this
-allows for the fitting of models assuming normally distributed
-residuals, which are computationally much less intensive than most
-alternatives. However, this transformation is problematic since
+time-series data, several problems are immediately evident. First,
+both $N_t$ and $X_t$ are typically non-negative integers, and hence
+the the Gaussian assumption is not justified.
+%is clearly a problem with this formulation because the normally distributed
+%random effects violate this constraint.
+One approach to circumventing this problem in hopes of adhering to the
+Gaussian assumption is to tranform abundance to
+density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
+area surveyed. However, this transformation is problematic since
 abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
-zeros are replaced with
-some small, but positive number, although the effect of this is rarely
-discussed.
+zeros are replaced with some small number, although the
+effect of this is rarely discussed.
 
-The problems mentioned thus far are minor in comparison to the more
-serious issue that the parameters of the model are often not
-identifiable. Specifically, the Markovian nature of the model implies
-the following likelihood
-\[
-LIKELIHOOD
-\]
+A more serious problem than the ones associated with the Gaussian
+assumptions is that the parameters of the simple state-space models
+such as Eq~\ref{eq:ss1} are not identifiable in many
+circumstances \citep{polansky_etal:2009}. Specifically, the Markovian
+nature of the model implies the following likelihood
+\begin{equation}
+\mathcal{L}(\Theta|\{X_t\}) = [N_1|\Theta]\prod_{t=2}^T
+[N_t|N_{t-1},\Theta]
+\label{eq:L1}
+\end{equation}
 The first term is the probability density for the first observation in
-the time series, $Y_t$, yet it is intuitively not possible to estimate the
+the time series, $X_1$, yet it is intuitively not possible to estimate the
 parameter(s) of this distribution using only a single
-observation. This problem has resulted in several papers with
-interesting names like ``multi-modal likelihoods and ridges etc...''
-The workaround here is to assume that the population is at equilibrium
+observation.
+%This problem has resulted in several papers with
+%interesting names like ``multi-modal likelihoods and ridges etc...''
+Two options have been proposed handling the identifiability
+problem. First, one can assume that the first observation in the time
+series is made without error. Second, one can assume that the population is at equilibrium
 so that that the x can be replaced with the equilibrium

### 2012-09-05 Dail_Mad_ext.tex
end of day

@@
 population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
-In order to address these questions, several complicating factors must
+In order to address these questions, two complicating factors must
 be confronted when fitting population models to data.
@@
 (e.g. exponential, logistic, etc \dots) and a stochastic model allowing
-for random process variation, which we denote $\grave{o}_t$ . As an
+for random process variation, which we denote $\grave{o}_t$. As an
 example, under exponential population growth, the model could be
 written as:
+%\begin{gather}
+%N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
+%X_t = N_t + \zeta_t
+%\label{eq:ss1}
+%\end{gather}
+\begin{subequations}
+  \label{eq:ss1}
   \begin{gather}
-N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
-X_t = N_t + \zeta_t
-\label{eq:ss1}
+    \mu_t = N_{t-1}\exp(r) \label{eq:ss1a} \\ \label{eq:ss1b}
+    N_t \sim \text{N}(\mu_t, \grave{o}_t) \\ \label{eq:ss1c}
+    X_t \sim \text{N}(N_t, \zeta_t)
   \end{gather}
-where $r$ is the intrinsic rate of increase. Typically, both error terms
-are assumed to be i.i.d random effects distrubted as
+\end{subequations}
+where $r$ is the intrinsic rate of increase. Although the exponential
+growth model shown here is often replaced with density dependent
+models, most state-space models share the common feature that the
+error terms are assumed to be i.i.d random effects distrubted as
 $\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
@@
 
-Exponential growth without density-dependence is rarely observed in
-natural populations, and thus $\exp(r)$ is often substituted for other
-population growth models such as the Ricker or Gompertz. Nonetheless,
-the basic structure shown in Eq.~\ref{eq:ss1} is common to most
-traditional state-space models. However, even though this is
-the most widespread approach for modeling population dynamics using
-time-series data, several problems are immediately evident. First,
-both $N_t$ and $X_t$ are typically non-negative integers, and hence
-the the Gaussian assumption is not justified.
+Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
+are the most widespread approach for modeling population dynamics
+using time-series data, several problems are evident. Namely, (1)
+time-series data and the underlying abundance parameters of interest
+are typically integer-valued, raising concerns about the use of
+Gaussian distribution shown in Eq~\ref{eq:ss1b}, (2) the model for
+observatoin error (Eq.~\ref{eq:ss1c}) has little biological basis, (3)
+spatial variation in abundance is not allowed, and (4) some of the
+paramters of the model are not estimable. We briefly discuss each of
+these points before describing a general approach to resolving these
+shortcomings.
+
+The fact that both abundance ($N_t$) and the observed data ($X_t$) are
+typically non-negative integers suggests that the
+Gaussian assumption in Eq.~\ref{eq:ss1b} is not justified.
 %is clearly a problem with this formulation because the normally distributed
 %random effects violate this constraint.
-One approach to circumventing this problem in hopes of adhering to the
-Gaussian assumption is to tranform abundance to
+Even so, efforts are often made to adhere to the
+Gaussian assumption by tranforming abundance to
 density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
@@
 effect of this is rarely discussed.
+
+Another problem with standard state-space models is that the
+observation error has no clear interpretation. The use of a mean zero
+Gaussian distribution implies that the observed data $X_t$ will be
+higher than $N_t$ as often as it is lower than $N_t$. It is hard
+to identify a mechanism that would cause such errors. A more likely
+form of observation error, and one that has been recognized for well
+over a century, results from failing to detect individuals that are
+present. Imperfect detection may be attributable to
+characteristics of the species under study, such as its elusiveness,
+or to the failings of the ecologist collecting the data in the field.
+Although a vast number of methods have been devised for accounting for
+this form of observation error, rarely have these methods been
+integrated into state-space models \citep[but
+see][]{buckland_etal:2004}.
 
@@
 A third problem with these models is that they do not admit
-spatial variation. This is not a fair criticism because this
-lies beyond the scope of traditional state-space models, but
-it does restrict the utility of their since it is typically
+spatial variation, which
+%. Recent work has sought to remedy this deficiencyThis is not a fair criticism because this
+%lies beyond the scope of traditional state-space models, but
+%it does
+restrict the utility of their since it is typically
 impossible assess the impacts of factors such as habitat
-fragmentation or climate change on a population without
-considering spatial variation. Indeed many populations are
+fragmentation or climate change on population dynamics without
+considering spatial variation in abundance. Indeed many populations are
 regulated by spatial processes such as source-sink dynamics.
 
+Although numerous extensions have been proposed to allow for spatial
+variation in abundance, in this paper we focus on the model of
 \citet{dail_madsen:2011} developed a model (henceforth the DM model)
-that resolves each of these problems with traditional state-space
-models. The DM model allows for inference about spatial and temporal
+that simultaneous resolves each of these problems with traditional state-space
+models. Specifically, the model (1) recognizes the discrete nature of
+count data and the underlying true abundance state, (2) has estimable
+parameters, including those describing the initial state at time
+$t=1$, and (3) allows for inference about both temporal and spatial
+variation in abundance.
+The DM model allows for inference about spatial and temporal
 dynamics in abundance while accounting for observation error using
 only spatially and temporally replicated count data.  Their model is
-an extension of the closed-population N-mixture model
+an extension of the closed-population $N$-mixture model
 \citep{royle:2004biom}, which was designed specifically to address the
 problem of
-modeling spatial variation in abundance when detection probability is
-less than unity. The DM extension relaxes the assumption of population
+modeling spatial variation in abundance in the presence of observation
+error, specifically, when detection probability is less than one. The
+DM extension relaxes the assumption of population
 closure and includes explicit parameters describing population change
@@
 
-[applications / potential etc... ] The DM model is relevant to a huge
+[applications / potential etc... ] The DM model is relevant to a vast
 number of problems confronting basic and applied ecological

### 2012-09-06 Dail_Mad_ext.tex
major revisions of intro

@@
 
-A third problem with these models is that they do not admit
+A fourth problem with these models is that they do not admit
 spatial variation, which
@@
 
-Although numerous extensions have been proposed to allow for spatial
-variation in abundance, in this paper we focus on the model of
-\citet{dail_madsen:2011} developed a model (henceforth the DM model)
-that simultaneous resolves each of these problems with traditional state-space
-models. Specifically, the model (1) recognizes the discrete nature of
-count data and the underlying true abundance state, (2) has estimable
-parameters, including those describing the initial state at time
-$t=1$, and (3) allows for inference about both temporal and spatial
-variation in abundance.
+Several extensions of state-space models have been proposed to
+overcome the limitations described
+above. \citet{devalpine_hastings:2002} described methods for fitting
+models with non-Gaussian distributions for the process and observation
+errors. Observation models with more intuitive interpretations have
+been proposed by \citet{kery_etal:2009}. \citet{lele_etal:1998} developed
+models allowing for inference about spatial and temporal variation in
+abundance, and their developments also resolved the problems of
+non-identifiability for the parameters of the initial state at time
+$t=1$. In spite of these improvements, until recently, few methods
+existed to address each issue simultaneously, and even fewer of these
+methods have been made readily available to applied ecologists.
+
+In this paper, we focus on the model of
+\citet{dail_madsen:2011} %developed a model (henceforth the DM model)
+that simultaneous resolves each of the abovementioned problems with
+traditional state-space models. Specifically, the model (1) recognizes
+the discrete nature of count data and the underlying true abundance
+state, (2) includes a realistic observation model with explicit
+detection probability parameters (3) has estimable parameters,
+including those describing the initial state at time $t=1$, and (4)
+allows for inference about both temporal and spatial variation in
+abundance. In some respects, the DM model can be viewed as a specific
+manifestation of the general state-space model formulated by
+\citep{buckland_etal:2004}. However, their model was designed to allow
+for inference about demographic parameters, which will typically
+require mark-recapture data. Although this may be viewed as the
+gold standard in many studies of wildlife populations, in this paper,
+we focus on the analysis of count data, which is the typical source of
+information used in state-state modeling applications, and one of the
+most widely available forms of ecological data---it is also the data
+required by the DM model.
+
+In the following section, we describe the DM model in its original
+form, and we explain how it resolves each of the deficieinces with
+standard state-space models. In Section~\ref{sec:ext}, we extend the
+model to accomodate classical models of population growth and to
+handle several features common to ecological
+time-series. Specifically, we describe methods for accomodating excess
+zeros and nuisance variables such as random observer effects. Both
+frquentist and Bayesian methods of inference are discussed, and code
+for fitting models is presented in the appendices. The Bayesian method
+is attractive in that it can accomodate prior information about
+detection parameters, which is available for many species, yet not
+collected as part of many existing monitoring programs. In
+Section~\ref{sec:app}, we evaluate the performance of the model
+extensions using a simulation study and by analyzing data from the
+North American Breeding Bird Survey (BBS), one of
+the most spatially and temporally extensive sets of count data on
+vertebrate populations \citep{robbins_etal:1986}. The overarching aim
+of  the paper is to provide ecologists with a flexible and accessible means of
+adressing important questions related to the variaion of abundance in
+space and time.
+believe these extensions will increase the utility of the model for
+application to existing datasets such the Breeding Bird Survey.
+
+\begin{comment}
   The DM model allows for inference about spatial and temporal
@@
   an extension of the closed-population $N$-mixture model
-\citep{royle:2004biom}, which was designed specifically to address the
-problem of
-modeling spatial variation in abundance in the presence of observation
-error, specifically, when detection probability is less than one. The
-DM extension relaxes the assumption of population
-closure and includes explicit parameters describing population change
-as a first-order Markovian process. [more details]
+  \citep{royle:2004biom}, which was designed specifically to address
+  the problem of modeling spatial variation in abundance in the
+  presence of observation error, specifically, when detection
+  probability is less than one. The DM extension relaxes the
+  assumption of population closure and includes explicit parameters
+  describing population change as a first-order Markovian
+  process. [more details]
 
-[applications / potential etc... ] The DM model is relevant to a vast
-number of problems confronting basic and applied ecological
+  [applications / potential etc... ] The DM model is relevant to a
+  vast number of problems confronting basic and applied ecological
   research. Answering questions about climate change etc... require
@@
   \citep{robbins_etal:1986}. However, most of the datasets of the
-appropriate scale are
-collected by volunteers and using protocols that are not amenable to
-traditional approaches of modeling abundance and detection probability
-based on standard capture-recapture methods. Instead they have, at
-best, simple count data for which, when it can be interpreted as
-counts of unique individuals, the DM model can be applied.. Yadi ya
+  appropriate scale are collected by volunteers and using protocols
+  that are not amenable to traditional approaches of modeling
+  abundance and detection probability based on standard
+  capture-recapture methods. Instead they have, at best, simple count
+  data for which, when it can be interpreted as counts of unique
+  individuals, the DM model can be applied.. Yadi ya
 
-Although the DM model was designed explicitly for these purposes, the
-original formulation of the model makes strict distributional
-assumptions and does not acknowledge many of the sources of variation
-inherent to existing ecological time-series data.  We propose
-variations on these models that are likely to be more realistic and
-useful in many cases.  We demonstrate the usefulness of these new
-models with simulated and real BBS data.
+  Although the DM model was designed explicitly for these purposes,
+  the original formulation of the model makes strict distributional
+  assumptions and does not acknowledge many of the sources of
+  variation inherent to existing ecological time-series data.  We
+  propose variations on these models that are likely to be more
+  realistic and useful in many cases.  We demonstrate the usefulness
+  of these new models with simulated and real BBS data.
 
-\section{Model Development}
+  In this paper, we demonstrate how the DM model resolves each of
+  these four limitations of conventional state-space models, and we
+  extend the DM model in several important ways. First, we demonstrate
+  how classical population models can be embedded within the DM
+  model. Then we present methods for handling excess zeros as well as
+  nuisance parameters such as random observer effects on detection
+  probability. A simulation study
+\end{comment}
 
-\subsection{Original Formulation}
+\section{The Dail-Madsen Model}
+\label{sec:dm}
+
+%\subsection{Original Formulation}
 
@@
 point count data.
+
+\section{Model Extensions}
+\label{sec:ext}
 
@@
 \section{Applications}
+\label{sec:app}
 

### 2012-09-06 Dail_Mad_ext.tex
expanding the description of the Dail-madsen model

@@
 
-%\subsection{Original Formulation}
+\subsection{The Data}
+
+The DM model requires count data collected at $R$ sites, each of which
+is surveyed on $T$ time periods, such as years. A site is ideally, but
+not necessarily, a
+well-defined region of a study area, such as a wetland or a patch of
+early-successional habitat. In some cases, a site may not have a clear
+biological significance, for instance in the case of randomly located
+survey plots. The timeframe of the study is arbitrary, but the general
+situation is one in which the time period is sufficiently long such
+that abundance changes at each site. We will call
+these time periods primary sampling periods to distinguish them from
+secondary sampling periods, which are repeated surveys within a
+primary period. In the case where no secondary sampling was conducted,
+let $n_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
+$i$ and primary period $t$. If $J$ secondary sampling periods were used, $n_{ijt}:
+j=1,\hdots,J$ is the count at site $i$ during secondary sampling
+occasion $j$ within year $t$. These are the observed data, which
+typically will less than the actual quantitiy of interest, abundance,
+denoted $N_{it}$. In cases where detection probability is perfect,
+abundance is observed directly such that $\{N_{it}\}$ are the data.
+
+\subsection{The Model}
 
@@
 conditionally related processes: the initial abundance state, apparent
-survival, recruitment, and detection probability. \citet{dail_madsen:2011}
-proposed two distributions for initial abundance, the Poisson and negative
-binomial:
+survival (the number of individuals that survive and do not
+emmigrate), recruitment (the number of new individuals entering
+the population), and detection probability \citet{dail_madsen:2011}.
+Taken together, the first 3 processes describe the state process---the
+variation in abundance in space and time. The forth item in the list
+is the observation process, which describes the relationship between
+the true state, abundance, and the observed count data.
+
+\subsubsection{Initial abundance}
+
+In the original formulation,
+initial abundance---abundance at site $i$ during primary period 1, was
+modeled using either the Poisson or negative binomial distributions:
 \begin{gather}
-N_{it} \sim \mathrm{Pois}(\Lambda) \nonumber \\
-N_{it} \sim \mathrm{NB}(\Lambda, \alpha)
+N_{i1} \sim \mathrm{Pois}(\Lambda_i) \nonumber \\
+N_{i1} \sim \mathrm{NB}(\Lambda_i, \alpha)
 \label{eq:N1}
 \end{gather}
-where $N_{i1}$ is the abundance at site $i=1,\hdots,R$ during year 1
-and $\Lambda=\mathbb{E}[N_{i1}]$ is the expected abundance. The negative
-binomial model includes the overdispersion parameter $\alpha$.  The
-survival and recruitment terms allow abundance to change over time:
+where
+%$N_{i1}$ is the abundance at site $i=1,\hdots,R$ during year 1 and
+$\Lambda_i$ is the expected abundance at site $i$ during year 1. The
+Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
+variance---an assumption relaxed when using the negative binomial
+distribution, which includes the overdispersion parameter $\alpha$.
+
+In either case, the model for initial abundance has two distinguising
+features. First, it allows one to model spatial variation in
+abundance. For instance, one might consider the influence of some
+environmental covariate on abundance using a simple generalized linear
+model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 + \beta^{\Lambda}_1
+x_{i}$ which is a log-linear function with an intercept and slope
+parameter for the effect of $x_i$ on the logarithm of $\Lambda$.
+
+The second thing to note is that the spatial replicates resolve the
+problem of parameter non-identifiability that are common to standard
+state-space models.
+
+\subsubsection{Abundance after $t=1$}
+
+The survival and recruitment terms allow abundance to change over time:
 \begin{equation}

### 2012-09-06 Dail_Mad_ext.tex
working on DM overview

@@
 
+The DM model is an extension of the $N$-mixture model
+\citep{royle:2004biom}, which allows for inference about spatial
+variation in abundance when individuals cannot be detected with
+certainty. To estimate both abundance parameters and parameters of the
+detection process, the model uses replicate observations at each site,
+which are collected during sufficiently short time intervals such that
+the population can safely be assumed to be closed with respect to
+births, deaths, and movement. The DM model relaxes this closure
+assumption and includes explicit parameters
+describing population change over time.
+
 \subsection{The Data}
@@
 The DM model requires count data collected at $R$ sites, each of which
-is surveyed on $T$ time periods, such as years. A site is ideally, but
-not necessarily, a
+is surveyed on $T$ time periods. A site is ideally, but not necessarily, a
 well-defined region of a study area, such as a wetland or a patch of
@@
 
-As originally formulated, the DM model includes submodels for four
-conditionally related processes: the initial abundance state, apparent
-survival (the number of individuals that survive and do not
-emmigrate), recruitment (the number of new individuals entering
-the population), and detection probability \citet{dail_madsen:2011}.
-Taken together, the first 3 processes describe the state process---the
-variation in abundance in space and time. The forth item in the list
+As originally formulated, the DM model includes submodels for three
+conditionally related processes: (1) initial abundance, i.e. the
+abundance at site $i$ during the first primary period,
+denoted $[N_{i1}]$, (2) subsequent abundance, which is dependent upon
+abundance in previous years $[N_{it}|N_{it-1}]$, and (3) the
+detection process, $[n_{it}|N_{it}]$ \citep{dail_madsen:2011}.
+The first two processes describe the state process---the
+variation in abundance in space and time. The third item in the list
 is the observation process, which describes the relationship between
@@
 
-In the original formulation,
-initial abundance---abundance at site $i$ during primary period 1, was
+In the original formulation, initial abundance was
 modeled using either the Poisson or negative binomial distributions:
@@
 N_{i1} \sim \mathrm{Pois}(\Lambda_i) \nonumber \\
+\text{or} \nonumber \\
 N_{i1} \sim \mathrm{NB}(\Lambda_i, \alpha)
@@
 Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
-variance---an assumption relaxed when using the negative binomial
-distribution, which includes the overdispersion parameter $\alpha$.
+variance, whereas the negative binomial distribution allows the
+variance to be greater than the mean with the amount of overdispersion
+determined by the parameter $\alpha$.
 
-In either case, the model for initial abundance has two distinguising
-features. First, it allows one to model spatial variation in
-abundance. For instance, one might consider the influence of some
-environmental covariate on abundance using a simple generalized linear
+Regardless of the specified distribution, the model for initial
+abundance has two distinguising features. First, it provides a
+mechanism for characterizing spatial variation in abundance. For
+instance, one might consider the influence of some environmental
+covariate ($x_i$) on abundance using a log-linear
 model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 + \beta^{\Lambda}_1
-x_{i}$ which is a log-linear function with an intercept and slope
-parameter for the effect of $x_i$ on the logarithm of $\Lambda$.
+x_{i}$. The second important point is that the spatial replicates resolve the
+problem of parameter non-identifiability that are common to standard
+state-space models because, as demonstrated by \citet{royle:2004biom},
+process variation and observation error can be estimated from
+spatially-replicated count data. Hence, the first component of the
+model addresses issues (1) and (2) identified in the introduction.
 
-The second thing to note is that the spatial replicates resolve the
-problem of parameter non-identifiability that are common to standard
-state-space models.
+\subsubsection{Abundance in subsequent time periods}
 
-\subsubsection{Abundance after $t=1$}
-
-The survival and recruitment terms allow abundance to change over time:
+The DM model assumes that abundance in time $t$ is a function of
+abundance in time $t-1$, i.e. abundance at each site evolves as a
+first order Markovian process, although
+higher order processes are also possible.  \citet{dail_madsen:2011}
+considered several models to describe the temporal dynamics; however,
+in each case they modeled $N_t$ as the sum of two random variables:
+$S_t$, the number of indivduals surviving from $t-1$ and not
+emigrating; and $G_t$ the number of new individuals entering the
+population. Their most general model was
 \begin{equation}
@@
 \end{equation}
-where $\omega$ is the survival probability and $\gamma$ is the arrival
-rate (which can depend on $N_{it-1}$). Dail and Madsen propose three models
-for $\gamma$: the constant model, where $\gamma$ does not depend on
-$N_{it-1}$, and which simulates a ``propagule rain'' of recruitment; the
-autoregressive model , which
-simulates geometric or density independent growth; and the ``no-trend''
-model, which keeps abundance fixed over time.
+where $\omega$ is the apparent survival probability and $\gamma$ is
+the recruitment rate (which can depend on $N_{it-1}$). In the absence
+of movement, $\omega$ is exactly the probability of surviving from
+year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
+in the more common scenario when immigration and emigration occur,
+$\omega$ is the probability of surviving and not emigrating, and
+$\gamma$ is product of the birth rate and the emigration rate,
+i.e. the recruitment rate. \citet{dail_madsen:2011} proposed three
+models for $\gamma$: the constant model,
+$G_t \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
+$N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
+autoregressive model, $G_t \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
+simulates geometric or density independent growth; and the
+``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
+expected abundance constant over time. As before, covariates of
+$\omega$ and $\gamma$ can be easily accomodated, for example using
+logit- and log-linear models respectively.
+
+In cases where no movement occurs, the DM model can directly estimate
+demographich parameters, and this mechanistic description of
+population change is much more appealing than classical
+population growth models such as the logistic model. However, movement
+complicates the interpretation of the the parameters, reducing the
+connection to specific population processes. In such cases, the model
+of dynamics shown in Eq~\ref{eq:Nt} can be replaced with standard
+population models as described in Sect~\ref{sec:ext}. Irrespective of
+the model for population dynamics, it is worthwhile to note that the
+discrete distributions used in the DM model avoid the issues
+associated with the Gaussian distribution used to model process
+variation in standard state-space models.
+
+\subsubsection{Observation process}
 
@@
 \end{equation}
-
-Initial abundance, apparent survival, recruitment, and detection can
-all be modeled as functions of covariates, allowing fairly
-sophisticated inference about population processes and dynamics from
-point count data.
+Surprsingly detection probability, $p$, can be estimated without
+secondary sampling occasions. However, this is only possible because
+if the parameteric assumptions of the population dynamics are met
+exactly. In practice, it is preferable to obtain direct information
+about $p$ using secondary sampling occassions or some other auxiliary
+data, as is demonstrated in section
 
@@
 \label{sec:ext}
-
-\subsection{Range model of initial abundance}
-
-\citet{dail_madsen:2011} suggested two distributions for modeling
-initial abundance: Poisson and negative binomial.  We have extended
-their model to include another distribution, the zero-inflated
-Poisson.  This could be useful when, for example, one is modeling the
-abundance of several bird species with the same set of BBS routespoint
-count surveys for all species, but this set includes routes sites
-outside the range of some of the bird species.  The distribution of
-initial abundances can be represented as:
-\begin{equation}
-N_{i1} \sim \left\{
-\begin{aligned}
-0 &\; \text{with probability} \; \psi \\
-\Lambda &\; \text{with probability} \; (1-\psi)
-\end{aligned} \right.
-\label{eq:ZIP}
-\end{equation}
-where $\psi$ represents the proportion of extra zeros.
-
-These models allow three sources of zero counts by observers: a
-species was at a route but not detected; the route was within the
-species' range but there were no birds at that site in that year; and
-the route was outside the species' range.  Furthermore, detection,
-abundance, and zero-inflation can be modeled separately as functions
-of different (or the same) covariates.  For example, detection of
-species x at site y in year z might depend on wind speed, abundance on
-forest type and weather, and zero-inflation upon elevation and
-climate.  This approach combines elements of occupancy modeling
-\citep{mackenzie_etal:2006} and abundance modeling.
-%, and could prove useful
-%for distinguishing species' fundamental and realized niches
-%(Hutchinson 1957).
-
-The zero inflation factor can also affect
-recruitment or population dynamics (see below).
-We also considered a zero-inflated negative binomial model of initial
-abundance, but this model did not perform well in preliminary tests.
 
@@
 
+%\subsection{Range model of initial abundance}
+\subsection{Excess zeros}
+
+\citet{dail_madsen:2011} suggested two distributions for modeling
+initial abundance: Poisson and negative binomial.  We have extended
+their model to include another distribution, the zero-inflated
+Poisson.  This could be useful when, for example, one is modeling the
+abundance of several bird species with the same set of BBS routespoint
+count surveys for all species, but this set includes routes sites
+outside the range of some of the bird species.  The distribution of
+initial abundances can be represented as:
+\begin{equation}
+N_{i1} \sim \left\{
+\begin{aligned}
+0 &\; \text{with probability} \; \psi \\
+\Lambda &\; \text{with probability} \; (1-\psi)
+\end{aligned} \right.
+\label{eq:ZIP}
+\end{equation}
+where $\psi$ represents the proportion of extra zeros.
+
+These models allow three sources of zero counts by observers: a
+species was at a route but not detected; the route was within the
+species' range but there were no birds at that site in that year; and
+the route was outside the species' range.  Furthermore, detection,
+abundance, and zero-inflation can be modeled separately as functions
+of different (or the same) covariates.  For example, detection of
+species x at site y in year z might depend on wind speed, abundance on
+forest type and weather, and zero-inflation upon elevation and
+climate.  This approach combines elements of occupancy modeling
+\citep{mackenzie_etal:2006} and abundance modeling.
+%, and could prove useful
+%for distinguishing species' fundamental and realized niches
+%(Hutchinson 1957).
+
+The zero inflation factor can also affect
+recruitment or population dynamics (see below).
+We also considered a zero-inflated negative binomial model of initial
+abundance, but this model did not perform well in preliminary tests.
+
+%\subsection{Dynamic range models}
+
+The zero-inflated Poisson distribution can be applied to not only
+initial abundance but also to recruitment and population growth terms.
+For example, the recruitment term of the constant DM model (equation
+4) can be modified as follows:
+\begin{equation}
+G_{it} \sim \left\{
+\begin{aligned}
+0 &\; \text{with probability} \; \psi \\
+\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)
+\end{aligned} \right.
+\label{eq:ZIPts}
+\end{equation}
+When both initial abundance and dynamics are modeled as zero-inflated,
+$\psi$ can be modeled as time-varying or not (in either case, $\psi$ can also be
+modeled as varying over space).  In the latter case, it makes sense
+for the model to "remember" which sites are inside the range of the
+species between the initial abundance stage and each time step of the
+dynamics stage (see code sample?).  Adding zero-inflation to the
+dynamics of the autoregressive, trend, Ricker, or Gompertz models
+(without immigration) should have no effect when initial abundance is
+also zero-inflated unless $\psi$ is time-varying (because these dynamic
+models cannot recover from zero abundance).  We have implemented
+zero-inflated dynamics in the Bayesian framework using program
+\textrm{JAGS} \citep[version 3.2.0]{plummer:2003}.
+%(version 3.2.0; Plummer 2003)
+
 \subsection{Random effects of observers}
@@
 
-\subsection{Dynamic range models}
+\subsection{Bayesian inference}
 
-The zero-inflated Poisson distribution can be applied to not only
-initial abundance but also to recruitment and population growth terms.
-For example, the recruitment term of the constant DM model (equation
-4) can be modified as follows:
-\begin{equation}
-G_{it} \sim \left\{
-\begin{aligned}
-0 &\; \text{with probability} \; \psi \\
-\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)
-\end{aligned} \right.
-\label{eq:ZIPts}
-\end{equation}
-When both initial abundance and dynamics are modeled as zero-inflated,
-$\psi$ can be modeled as time-varying or not (in either case, $\psi$ can also be
-modeled as varying over space).  In the latter case, it makes sense
-for the model to "remember" which sites are inside the range of the
-species between the initial abundance stage and each time step of the
-dynamics stage (see code sample?).  Adding zero-inflation to the
-dynamics of the autoregressive, trend, Ricker, or Gompertz models
-(without immigration) should have no effect when initial abundance is
-also zero-inflated unless $\psi$ is time-varying (because these dynamic
-models cannot recover from zero abundance).  We have implemented
-zero-inflated dynamics in the Bayesian framework using program
-\textrm{JAGS} \citep[version 3.2.0]{plummer:2003}.
-%(version 3.2.0; Plummer 2003)
+priors are good
 

### 2012-09-07 Dail_Mad_ext.tex
intro and overview
intro

@@
 hypotheses regarding spatial and temporal variation in
-abundance---hypotheses relating to the importance and existence of
-phenomenon such as density-dependent population regulation, population
-cycling, and spatial synchrony
-\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999}.
+abundance. For example, much theoretical work has focused on
+understanding the importance and existence of phenomenon such as
+density-dependent population regulation, population cycling, and spatial synchrony
+\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999},
+and population models are required to evaluate associated hypotheses.
 In applied contexts, population models are used for
@@
 and observation error \citep{devalpine_hastings:2002,
-  buckland_etal:2004, dennis_etal:2006}. Conventional state-space
-models are simply time-series models in which the true state of the
+  buckland_etal:2004, dennis_etal:2006}. Classical state-space
+models are time-series models in which the true state of the
 system (e.g. population size during each year) is observed
@@
 observation error can bias estimators of abundance and population
-growth parameters \citep{kery_etal:2009}. For instance, the strength of
+growth parameters.
+% \citep{kery_etal:2009}.
+For instance, the strength of
 density dependence will be over-estimated if observation error is
@@
 modeled using a deterministic population growth model
-(e.g. exponential, logistic, etc \dots) and a stochastic model allowing
+(e.g. exponential, logistic, etc \dots)~and a stochastic model allowing
 for random process variation, which we denote $\grave{o}_t$. As an
@@
 nature of the model implies the following likelihood
-\begin{equation}
-\mathcal{L}(\Theta|\{X_t\}) = [N_1|\Theta]\prod_{t=2}^T
-[N_t|N_{t-1},\Theta]
-\label{eq:L1}
-\end{equation}
+\begin{subequations}
+  \label{eq:ss1}
+  \begin{align}
+    [N_1&] \\ \label{eq:ss1b}
+    [N_t&|N_{t-1}] &\text{for} \; t=2,\hdots,T \\ \label{eq:ss1c}
+    [X_t&|N_t]  &\text{for} \; t=1,\hdots,T
+  \end{align}
+\end{subequations}
+%\begin{equation}
+%\mathcal{L}(\Theta|\{X_t\}) = [N_1|\Theta]\prod_{t=2}^T
+%[N_t|N_{t-1},\Theta]
+%\label{eq:L1}
+%\end{equation}
 The first term is the probability density for the first observation in
@@
 overcome the limitations described
-above. \citet{devalpine_hastings:2002} described methods for fitting
-models with non-Gaussian distributions for the process and observation
-errors. Observation models with more intuitive interpretations have
-been proposed by \citet{kery_etal:2009}. \citet{lele_etal:1998} developed
-models allowing for inference about spatial and temporal variation in
-abundance, and their developments also resolved the problems of
-non-identifiability for the parameters of the initial state at time
-$t=1$. In spite of these improvements, until recently, few methods
-existed to address each issue simultaneously, and even fewer of these
-methods have been made readily available to applied ecologists.
+above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
+described methods for fitting models with non-Gaussian distributions
+for the process and observation errors. Observation models with more
+intuitive interpretations, such as those that explicitly model
+detection probability, have been proposed by
+\citet{kery_etal:2009}. \citet{lele_etal:1998} and
+\citep{kery_etal:2009} developed models allowing for inference about
+spatial and temporal variation in abundance, and their developments
+also resolved the problems of non-identifiability for the parameters
+of the initial state at time $t=1$. Of these extensions, the work by
+\citet{kery_etal:2009} is unique in that it addressed multiple limitations
+simultaneously. However, their models did not include any serial
+depedence, which is a hallmark of population models. This limits the
+utility of their model for making inferences about explicit population
+processes. To do so, recent work has sought to combine multiple
+sources of information, such as counts and mark-recapture data, to
+allow for inferences about population processes such as mortality and
+survival, rather than just population growth \citep{besbeas_etal:2002,
+  buckland_etal:2004, schaub_etal:2007}. Although these integrated
+population models may be viewed as the gold standard in state-space
+modeling, ecologists are not always so fortunate to have direct
+information about vital rates, especially when inferences are required
+at large spatial scales. Rather, count data are much more common, and
+are produced by many of the largest monitoring programs in the world,
+such as the North American Breeding Bird Survey
+\citep[BBS;][]{robbins_etal:1986}.
 
@@
 that simultaneous resolves each of the abovementioned problems with
-traditional state-space models. Specifically, the model (1) recognizes
+traditional state-space models, and is designed for simple count
+data. The model (1) recognizes
 the discrete nature of count data and the underlying true abundance
@@
 allows for inference about both temporal and spatial variation in
-abundance. In some respects, the DM model can be viewed as a specific
+abundance.
+
+\begin{comment}
+  In some respects, the DM model can be viewed as a specific
   manifestation of the general state-space model formulated by
-\citep{buckland_etal:2004}. However, their model was designed to allow
-for inference about demographic parameters, which will typically
-require mark-recapture data. Although this may be viewed as the
-gold standard in many studies of wildlife populations, in this paper,
-we focus on the analysis of count data, which is the typical source of
-information used in state-state modeling applications, and one of the
-most widely available forms of ecological data---it is also the data
-required by the DM model.
+  \citep{buckland_etal:2004}. However, their model was designed to
+  allow for inference about demographic parameters, which will
+  typically require mark-recapture data. Although this may be viewed
+  as the gold standard in many studies of wildlife populations, in
+  this paper, we focus on the analysis of count data, which is the
+  typical source of information used in state-state modeling
+  applications, and one of the most widely available forms of
+  ecological data---it is also the data required by the DM model.
+\end{comment}
 
@@
 Therefore, we developed a version of the DM model that estimates and
-models population growth directly.  In this model, Eq.\ref{eq:Nt} is
+models population growth directly.  In this model, Eq.~\ref{eq:Nt} is
 simplified to:

### 2012-09-07 Dail_Mad_ext.tex
ss equations in intro

@@
 detection \citep{link_nichols:1994,kery_etal:2009}.
-%Failure to account for process
-%variation and observation error can bias the estimators of abundance
-%and related parameters.
 
@@
 observation error can bias estimators of abundance and population
-growth parameters.
-% \citep{kery_etal:2009}.
-For instance, the strength of
+growth parameters. For instance, the strength of
 density dependence will be over-estimated if observation error is
@@
 for random process variation, which we denote $\grave{o}_t$. As an
-example, under exponential population growth, the model could be
-written as:
-%\begin{gather}
-%N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
-%X_t = N_t + \zeta_t
-%\label{eq:ss1}
-%\end{gather}
-\begin{subequations}
-  \label{eq:ss1}
-  \begin{gather}
-    \mu_t = N_{t-1}\exp(r) \label{eq:ss1a} \\ \label{eq:ss1b}
-    N_t \sim \text{N}(\mu_t, \grave{o}_t) \\ \label{eq:ss1c}
-    X_t \sim \text{N}(N_t, \zeta_t)
-  \end{gather}
-\end{subequations}
-where $r$ is the intrinsic rate of increase. Although the exponential
-growth model shown here is often replaced with density dependent
-models, most state-space models share the common feature that the
-error terms are assumed to be i.i.d random effects distrubted as
+example, under density-independent exponential population growth, the
+model could be written as:
+\begin{equation}
+%  \label{eq:ss1}
+  \left.
+  \begin{aligned}
+    \mu_t &= N_{t-1}\exp(r) \label{eq:ss1a} \\ %\label{eq:ss1b}
+    N_t &\sim \text{N}(\mu_t, \grave{o}_t) \\ %\label{eq:ss1c}
+    X_t &\sim \text{N}(N_t, \zeta_t)
+  \end{aligned} \;
+  \right\} \; \text{for} \; t=2,\hdots,T
+\end{equation}
+%\begin{subequations}
+%  \label{eq:ss1}
+%  \left.
+%  \begin{gather}
+%    \mu_t = N_{t-1}\exp(r) \label{eq:ss1a} \\ \label{eq:ss1b}
+%    N_t \sim \text{Normal}(\mu_t, \grave{o}_t) \\ \label{eq:ss1c}
+%    X_t \sim \text{Normal}(N_t, \zeta_t)
+%  \end{gather}
+%  \right\} \text{for}
+%\end{subequations}
+where $r$ is the intrinsic rate of increase. Note that no model is
+presented for the initial state at $t=1$---this issue will be
+discussed subsequently. Note also that the exponential
+growth model shown in Eq.~\ref{eq:ss1a} can be easily replaced with
+any population growth model, including density-depedent models. This
+flexibility is an appealing feature of state-space models. One feature
+that is common to classical state-space models is that the
+error terms are assumed to be i.i.d random effects following Gaussian
+(Normal) distributions:
 $\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and

### 2012-09-07 Dail_Mad_ext.tex
ss equations in intro

@@
 \usepackage{amsfonts}
+\usepackage{bm}
 \usepackage{graphicx}
@@
 model could be written as:
-\begin{equation}
+% \begin{equation}
+%   \left.
+%   \begin{aligned}
+%     \mu_t &= N_{t-1}\exp(r) \\ %\label{eq:ss1b}
+%     N_t &= \mu_t + \grave{o}_t \\ %\text{N}(\mu_t, \grave{o}_t) \\ %\label{eq:ss1c}
+%     X_t &= N_t + \zeta_t %\text{N}(N_t, \zeta_t)
+%   \end{aligned} \;
+%   \right\} \; \text{for} \; t=2,\hdots,T
 %   \label{eq:ss1}
-  \left.
-  \begin{aligned}
-    \mu_t &= N_{t-1}\exp(r) \label{eq:ss1a} \\ %\label{eq:ss1b}
-    N_t &\sim \text{N}(\mu_t, \grave{o}_t) \\ %\label{eq:ss1c}
-    X_t &\sim \text{N}(N_t, \zeta_t)
-  \end{aligned} \;
-  \right\} \; \text{for} \; t=2,\hdots,T
-\end{equation}
-%\begin{subequations}
-%  \label{eq:ss1}
-%  \left.
-%  \begin{gather}
-%    \mu_t = N_{t-1}\exp(r) \label{eq:ss1a} \\ \label{eq:ss1b}
-%    N_t \sim \text{Normal}(\mu_t, \grave{o}_t) \\ \label{eq:ss1c}
-%    X_t \sim \text{Normal}(N_t, \zeta_t)
-%  \end{gather}
-%  \right\} \text{for}
-%\end{subequations}
-where $r$ is the intrinsic rate of increase. Note that no model is
-presented for the initial state at $t=1$---this issue will be
-discussed subsequently. Note also that the exponential
-growth model shown in Eq.~\ref{eq:ss1a} can be easily replaced with
-any population growth model, including density-depedent models. This
-flexibility is an appealing feature of state-space models. One feature
-that is common to classical state-space models is that the
+% \end{equation}
+\begin{subequations}
+  \begin{align}
+%    \mu_t &= N_{t-1}\exp(r) \quad \text{for} \; t=2,\hdots,T
+    N_1 &= X_1 \label{ss1a} \\
+    N_t &= \mu_t + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
+    X_t &= N_t + \zeta_t \quad\;\;\; \text{for} \; t=1,\hdots,T \label{eq:ss1c}
+  \end{align}
+  \label{eq:ss1}
+\end{subequations}
+where $\mu_t$ is a deterministic function describing population
+growth. For instance, under exponential growth, $\mu_t =
+N_{t-1}e^{r}$ where $r$ is the intrinsic rate of
+increase. Alternatively, one might consider density-dependent models,
+and.  This
+flexibility is an appealing feature of state-space models.
+Note that no model is
+presented for the initial state at $t=1$; we will return to this issue
+subsequently. Note also that the exponential
+growth model shown in Eq.~\ref{eq:ss1} can be easily replaced with
+any population growth model, including density-depedent models. One feature
+that is common to most classical state-space models is that the
 error terms are assumed to be i.i.d random effects following Gaussian
@@
 are typically integer-valued, raising concerns about the use of
-Gaussian distribution shown in Eq~\ref{eq:ss1b}, (2) the model for
-observatoin error (Eq.~\ref{eq:ss1c}) has little biological basis, (3)
+Gaussian distribution for the random effects, (2) the model for
+observatoin error has little biological basis, (3)
 spatial variation in abundance is not allowed, and (4) some of the
@@
 The fact that both abundance ($N_t$) and the observed data ($X_t$) are
-typically non-negative integers suggests that the
-Gaussian assumption in Eq.~\ref{eq:ss1b} is not justified.
-%is clearly a problem with this formulation because the normally distributed
-%random effects violate this constraint.
-Even so, efforts are often made to adhere to the
-Gaussian assumption by tranforming abundance to
+typically non-negative integers suggests that it is inappropriate to
+model process variation and observation error using Gaussian
+distributions. Nonetheless, efforts have been made to adhere to the
+Gaussian assumptions by tranforming abundance to
 density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
@@
 such as Eq~\ref{eq:ss1} are not identifiable in many
-circumstances \citep{polansky_etal:2009}. Specifically, the Markovian
-nature of the model implies the following likelihood
+circumstances \citep{polansky_etal:2009}. Recall that the
+Eq.~\ref{eq:ss1} did not specify distributions for $N_1$ or
+$X_1$; however, such distributions are needed to fully specify the
+model. Indeed, a fully-specified state-space model requires the
+specification of three probability distributions, which we denote
+using bracket notation:
 \begin{subequations}
-  \label{eq:ss1}
+  \label{eq:ss2}
   \begin{align}
-    [N_1&] \\ \label{eq:ss1b}
-    [N_t&|N_{t-1}] &\text{for} \; t=2,\hdots,T \\ \label{eq:ss1c}
-    [X_t&|N_t]  &\text{for} \; t=1,\hdots,T
+    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
+    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
+    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
   \end{align}
 \end{subequations}
-%\begin{equation}
-%\mathcal{L}(\Theta|\{X_t\}) = [N_1|\Theta]\prod_{t=2}^T
-%[N_t|N_{t-1},\Theta]
-%\label{eq:L1}
-%\end{equation}
+where $\bm{\theta}$ are the process variation parameters governing
+initial abundance, $\bm{\Theta}$ are the process variation parameters
+influencing how abundance changes over time, and $\bm{p}$ are the
+observation error parameters. In our first example, Eq.~\ref{eq:ss1},
+But what should these distributions be, and how could
+parameters of the distributions be
+estimated since there is only a single observation available? In fact,
+they cannotThe reason they were omitted is that it is not possible to
+estimate the parameters of the i
+However, a completSpecifically, the Markovian
+nature of the model implies the following likelihood
 The first term is the probability density for the first observation in

### 2012-09-07 Dail_Mad_ext.tex
intro

@@
 
-One of the simplest state-space models can be described as follows.
+A simple state-space models can be described as follows.
 Let $N_t$ be the abundance of a species during year $t$, for
@@
 a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
-modeled using a deterministic population growth model
-(e.g. exponential, logistic, etc \dots)~and a stochastic model allowing
-for random process variation, which we denote $\grave{o}_t$. As an
-example, under density-independent exponential population growth, the
-model could be written as:
-% \begin{equation}
-%   \left.
-%   \begin{aligned}
-%     \mu_t &= N_{t-1}\exp(r) \\ %\label{eq:ss1b}
-%     N_t &= \mu_t + \grave{o}_t \\ %\text{N}(\mu_t, \grave{o}_t) \\ %\label{eq:ss1c}
-%     X_t &= N_t + \zeta_t %\text{N}(N_t, \zeta_t)
-%   \end{aligned} \;
-%   \right\} \; \text{for} \; t=2,\hdots,T
-%   \label{eq:ss1}
-% \end{equation}
+modeled using a deterministic population growth model, $\mu(N_{t-1})$,
+coupled with a stochastic model allowing
+for random process variation. %, which we denote $\grave{o}_t$.
+The deterministic model may be density-dependent, for example
+logistic, or it might be density-independent, as in the case of
+exponential growth, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$
+is the intrinsic rate of increase. The full model can now be written
+as:
+%As an example, under
+%density-independent exponential population growth, the
+%model could be written as:
 \begin{subequations}
+  \label{eq:ss1}
   \begin{align}
 %    \mu_t &= N_{t-1}\exp(r) \quad \text{for} \; t=2,\hdots,T
-    N_1 &= X_1 \label{ss1a} \\
-    N_t &= \mu_t + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
-    X_t &= N_t + \zeta_t \quad\;\;\; \text{for} \; t=1,\hdots,T \label{eq:ss1c}
+    N_1 &= X_1 \label{eq:ss1a} \\
+    N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
+    X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \; t=1,\hdots,T \label{eq:ss1c}
   \end{align}
-  \label{eq:ss1}
 \end{subequations}
-where $\mu_t$ is a deterministic function describing population
-growth. For instance, under exponential growth, $\mu_t =
-N_{t-1}e^{r}$ where $r$ is the intrinsic rate of
-increase. Alternatively, one might consider density-dependent models,
-and.  This
-flexibility is an appealing feature of state-space models.
-Note that no model is
-presented for the initial state at $t=1$; we will return to this issue
-subsequently. Note also that the exponential
-growth model shown in Eq.~\ref{eq:ss1} can be easily replaced with
-any population growth model, including density-depedent models. One feature
-that is common to most classical state-space models is that the
-error terms are assumed to be i.i.d random effects following Gaussian
-(Normal) distributions:
+%{\flushleft where} $\mu(N_{t-1})$ is the deterministic population
+%growth model. %One might alternatively consider density-dependent models
+%for $\mu(N_{t-1})$, and this flexibility is an appealing feature of
+%state-space models.
+In classical state-space models, the two sets of random effects %In most classical state-space models both sets of parameters
+are assumed to be i.i.d Gaussian deviates:
 $\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
-$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
-estimation using the Kalman filter \citep{dennis_etal:2006}, which is
-much more computationally efficient than estimation methods when
-random effects are not Gaussian distributed
-\citep{devalpine_hastings:2002}.
+$\zeta_t \sim \mathrm{N}(0, \tau)$.
+Note that Eq~\ref{eq:ss1a} indicates that no process variation is
+present in year 1. This assumption is often made
+due to parameter identifiability problems, which we will describe
+shortly. First, we turn our attention to the models for the random
+effects.
 
@@
 
-The fact that both abundance ($N_t$) and the observed data ($X_t$) are
+These assumptions are motivated by
+convenience rather than biology. Specifically, the Gaussian assumption
+allows for parameter estimation using the Kalman filter
+\citep{dennis_etal:2006}, which is much more computationally efficient
+than estimation methods when random effects are not Gaussian distributed
+\citep{devalpine_hastings:2002}. The problem with this is that it
+allows for negative values of the state-variable, which is
+inconsistent with the observed data. The fact that both abundance ($N_t$) and the observed data ($X_t$) are
 typically non-negative integers suggests that it is inappropriate to

### 2012-09-07 Dail_Mad_ext.tex
new draft intro

@@
 as:
-%As an example, under
-%density-independent exponential population growth, the
-%model could be written as:
 \begin{subequations}
@@
   \begin{align}
-%    \mu_t &= N_{t-1}\exp(r) \quad \text{for} \; t=2,\hdots,T
     N_1 &= X_1 \label{eq:ss1a} \\
@@
 \end{subequations}
-%{\flushleft where} $\mu(N_{t-1})$ is the deterministic population
-%growth model. %One might alternatively consider density-dependent models
-%for $\mu(N_{t-1})$, and this flexibility is an appealing feature of
-%state-space models.
-In classical state-space models, the two sets of random effects %In most classical state-space models both sets of parameters
+where $\grave{o}_t$ is the random effect allowing for process
+variation unaccounted for by the deterministic model. In classical
+state-space models, the two sets of random effects
 are assumed to be i.i.d Gaussian deviates:
 $\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
-$\zeta_t \sim \mathrm{N}(0, \tau)$.
-Note that Eq~\ref{eq:ss1a} indicates that no process variation is
-present in year 1. This assumption is often made
-due to parameter identifiability problems, which we will describe
-shortly. First, we turn our attention to the models for the random
-effects.
+$\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
+act as though there is no process variation associated with $N_1$, as
+indicated by Eq~\ref{eq:ss1a}.
+% indicates that no process variation is
+%present in year 1. This assumption is often made
+%due to parameter identifiability problems, which we will describe
+%shortly. First, we turn our attention to the models for the random
+%effects.
 
@@
 time-series data and the underlying abundance parameters of interest
-are typically integer-valued, raising concerns about the use of
-Gaussian distribution for the random effects, (2) the model for
-observatoin error has little biological basis, (3)
+are typically integer-valued, as in the case of count data, raising
+concerns about the use of Gaussian distribution for the random
+effects, (2) the model for observation error has little biological basis, (3)
 spatial variation in abundance is not allowed, and (4) some of the
@@
 
-These assumptions are motivated by
-convenience rather than biology. Specifically, the Gaussian assumption
-allows for parameter estimation using the Kalman filter
+Use of the Gaussian distribution for modeling random process variation
+and observation error is motivated by convenience rather than
+biology. Specifically, the Gaussian assumptions
+allow for parameter estimation using the Kalman filter
 \citep{dennis_etal:2006}, which is much more computationally efficient
@@
 allows for negative values of the state-variable, which is
-inconsistent with the observed data. The fact that both abundance ($N_t$) and the observed data ($X_t$) are
-typically non-negative integers suggests that it is inappropriate to
-model process variation and observation error using Gaussian
-distributions. Nonetheless, efforts have been made to adhere to the
+inconsistent with the observed data.
+%The fact that both abundance ($N_t$) and the observed data ($X_t$) are
+%typically non-negative integers suggests that it is inappropriate to
+%model process variation and observation error using Gaussian
+%distributions.
+Nonetheless, efforts have been made to adhere to the
 Gaussian assumptions by tranforming abundance to
@@
 observation error has no clear interpretation. The use of a mean zero
-Gaussian distribution implies that the observed data $X_t$ will be
+Gaussian distribution implies that $X_t$ will be
 higher than $N_t$ as often as it is lower than $N_t$. It is hard
-to identify a mechanism that would cause such errors. A more likely
+to identify a mechanism that would cause such symmetric errors. A more likely
 form of observation error, and one that has been recognized for well
@@
 circumstances \citep{polansky_etal:2009}. Recall that the
-Eq.~\ref{eq:ss1} did not specify distributions for $N_1$ or
-$X_1$; however, such distributions are needed to fully specify the
-model. Indeed, a fully-specified state-space model requires the
-specification of three probability distributions, which we denote
+Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
+however, $N_1$ is a random variable and hence there is uncertainty
+that should be accounted for.
+%are needed to fully specify the model.
+To be more specific, a fully-specified state-space model requires at least
+three probability distributions, which we represent
 using bracket notation:
@@
 \end{subequations}
-where $\bm{\theta}$ are the process variation parameters governing
-initial abundance, $\bm{\Theta}$ are the process variation parameters
+where $\bm{\theta}$ are the process variation parameters for the
+initial state, $\bm{\Theta}$ are the process variation parameters
 influencing how abundance changes over time, and $\bm{p}$ are the
-observation error parameters. In our first example, Eq.~\ref{eq:ss1},
-But what should these distributions be, and how could
-parameters of the distributions be
-estimated since there is only a single observation available? In fact,
-they cannotThe reason they were omitted is that it is not possible to
-estimate the parameters of the i
-However, a completSpecifically, the Markovian
-nature of the model implies the following likelihood
-The first term is the probability density for the first observation in
-the time series, $X_1$, yet it is intuitively not possible to estimate the
-parameter(s) of this distribution using only a single
-observation.
-%This problem has resulted in several papers with
-%interesting names like ``multi-modal likelihoods and ridges etc...''
+observation error parameters. The three lines of general model
+described by Eq.~\ref{eq:ss2} correspond to the three equations shown
+for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
+we allow $N_1$ to be a random variable. But what should its
+distribution, $[N_1]$ be? And how could $\bm{\theta}$ be
+estimated since there is only a single observation available?
+One approach to estimating the $\bm{\theta}$ is to assume that the
+population is in equlibrium, i.e. that the expected value of $N_t$ is
+constant through time. Althought this makes the parameters
+identifiable, assuming equlibrium defeats the objective of many
+studies of population dynamics, namely determining why a population
+varies over time.
+
+\begin{comment}
+  However, a completSpecifically, the Markovian nature of the model
+  implies the following likelihood The first term is the probability
+  density for the first observation in the time series, $X_1$, yet it
+  is intuitively not possible to estimate the parameter(s) of this
+  distribution using only a single observation.
+  % This problem has resulted in several papers with interesting names
+  % like ``multi-modal likelihoods and ridges etc...''
   Two options have been proposed handling the identifiability
-problem. First, one can assume that the first observation in the time
-series is made without error. Second, one can assume that the population is at equilibrium
-so that that the x can be replaced with the equilibrium
-distribution. However, assuming equilibrium defeats the objective of
-many studies of population dynamics, namely determining if and why a
-population is at equilibrium.
+  problem. First, one can assume that the first observation in the
+  time series is made without error. Second, one can assume that the
+  population is at equilibrium so that that the x can be replaced with
+  the equilibrium distribution. However, assuming equilibrium defeats
+  the objective of many studies of population dynamics, namely
+  determining if and why a population is at equilibrium.
+\end{comment}
 
@@
 \citet{kery_etal:2009}. \citet{lele_etal:1998} and
-\citep{kery_etal:2009} developed models allowing for inference about
+\citet{kery_etal:2009} developed models allowing for inference about
 spatial and temporal variation in abundance, and their developments
@@
 \citet{kery_etal:2009} is unique in that it addressed multiple limitations
-simultaneously. However, their models did not include any serial
+simultaneously. However, their model did not include serial
 depedence, which is a hallmark of population models. This limits the
 utility of their model for making inferences about explicit population
-processes. To do so, recent work has sought to combine multiple
-sources of information, such as counts and mark-recapture data, to
-allow for inferences about population processes such as mortality and
-survival, rather than just population growth \citep{besbeas_etal:2002,
-  buckland_etal:2004, schaub_etal:2007}. Although these integrated
+processes. In contrast, recent work has sought to use state-space
+models fot inference about population processes such as mortality and
+recruitment. To do so, methods have been developed to combine multiple
+sources of information, such as counts and mark-recapture data
+\citep{besbeas_etal:2002, buckland_etal:2004,
+  schaub_etal:2007}. Although these integrated
 population models may be viewed as the gold standard in state-space
 modeling, ecologists are not always so fortunate to have direct
-information about vital rates, especially when inferences are required
-at large spatial scales. Rather, count data are much more common, and
+information about vital rates, especially at large spatial scales.
+Rather, count data are much more common, and
 are produced by many of the largest monitoring programs in the world,
@@
 traditional state-space models, and is designed for simple count
-data. The model (1) recognizes
+data.
+\begin{comment}
+  The model (1) recognizes
   the discrete nature of count data and the underlying true abundance
@@
 
-\begin{comment}
   In some respects, the DM model can be viewed as a specific
@@
 \end{comment}
-
 In the following section, we describe the DM model in its original
-form, and we explain how it resolves each of the deficieinces with
+form and explain how it resolves each of the deficieinces with
 standard state-space models. In Section~\ref{sec:ext}, we extend the
@@
 frquentist and Bayesian methods of inference are discussed, and code
-for fitting models is presented in the appendices. The Bayesian method
-is attractive in that it can accomodate prior information about
-detection parameters, which is available for many species, yet not
-collected as part of many existing monitoring programs. In
-Section~\ref{sec:app}, we evaluate the performance of the model
+for fitting models is presented in the appendices.
+%The Bayesian method
+%is attractive in that it can accomodate prior information about
+%detection parameters, which is available for many species, yet not
+%collected as part of many existing monitoring programs.
+In Section~\ref{sec:app}, we evaluate the performance of the model
 extensions using a simulation study and by analyzing data from the
@@
 vertebrate populations \citep{robbins_etal:1986}. The overarching aim
-of  the paper is to provide ecologists with a flexible and accessible means of
+of  the paper is to provide ecologists with flexible and accessible means of
 adressing important questions related to the variaion of abundance in
 space and time.
-believe these extensions will increase the utility of the model for
-application to existing datasets such the Breeding Bird Survey.
+%believe these extensions will increase the utility of the model for
+%application to existing datasets such the Breeding Bird Survey.
 
@@
 certainty. To estimate both abundance parameters and parameters of the
-detection process, the model uses replicate observations at each site,
-which are collected during sufficiently short time intervals such that
-the population can safely be assumed to be closed with respect to
-births, deaths, and movement. The DM model relaxes this closure
-assumption and includes explicit parameters
+detection process, the original $N$-mixture model uses replicate
+observations at each site, which are collected during sufficiently
+short time intervals such that the population can safely be assumed to
+be closed with respect to births, deaths, and movement. The DM model
+relaxes this closure assumption and includes explicit parameters
 describing population change over time.

### 2012-09-07 Dail_Mad_ext.tex
err

@@
 
-priors are good
+Bayesian inference has several appealing features. First, it allows
+direct probability statements to be made about a hypothesis given
+data. Second, priors are cool. We can combine data from multiple
+sources or use existing estimates of parameters as prior
+distributions. MCMC has a bad rap for being slow, but actually, in the
+case of state-space models it may be one of the few options for
+estimation, and in fact may be faster than maximizing likelihoods.
+
+Here is some BUGS code:
+\begin{verbatim}
+model {
+lambda ~ dunif(0, 5)
+r ~ dunif(-5, 5)
+p ~ dunif(0,1)
+for(i in 1:nSites) {
+  N[i,1] ~ dpois(lambda)
+  y[i,1] ~ dbin(p, N[i,1])
+  for(t in 2:nYears) {
+    mu[i,t-1] = N[i,t-1]*exp(r)
+    N[i,t] ~ dpois(mu[i,t-1])
+    n[i,t] ~ dbin(p, N[i,1])
+    }
+  }
+}
+\end{verbatim}
 
@@
 
+Our work has highlighted the limitations of classical state-space
+models as applied to ecological time series data, and we have
+demonstrated how open population $N$-mixture models overcome these
+limitations. We extended this class of models in several important
+ways, while keeping in mind the practical issues ...
+
 We made four important developments of the open population
-N-mixture model proposed by Dail and Madsen that make the
+N-mixture model proposed by \citep{dail_madsen:2011} that make the
 model more applicable to long-term data such as is collected
@@
 reconciled the objectives of traditional state-space models
-with open population N-mixture models by illustrating how
+with open population $N$-mixture models by illustrating how
 classical models of population dynamics can be embedded within
@@
 accommodated. Fourth, we presented a Bayesian analysis of the
-model, which makes it much easier to estimate parameters and
-to fine tune the model for specific purposes.
+model, which, makes it possible to incorporate prior information when
+available, and in some cases, facilitates parameter estimation.
 
@@
 [paragraph on random effects]
+
 [paragraph on Bayesian]
+The ability to fit this class of models using freely available MCMC
+software offers non-statisticians a powerful means .
+
 In spite of the new extensions we have proposed, several aspects of
@@
 probability that should be accounted for to minimize bias and yield
-valid estimates of population size. [Nichols EURING paper] .
+valid estimates of population size \citep{nichols_etal:2009}.
 
@@
 results?]
+Our emphasis was on increasing the practical utility of this class of
+models, and so we avoided several conceptually-interesting extensions
+that we belive would be computationally prohibitive in many
+cases. Nonetheless, we will discuss one---spatially-explict models of
+immigration.
 

### 2012-09-10 Dail_Mad_ext.tex
intro, methods

@@
 hypotheses about density dependence and predicting species' responses
-to future environmental change or management actions. State-space
-models are widely used for these purposes because they allow for
-estimating parameters of classical growth models, such as exponential,
-logistic, and Gompertz, while accounting for demographic stochasticity
-and observation error. Conventional state-space models, however, have
+to future environmental change or conservation actions. Accounting for
+process variation and observation error in such efforts is necessary
+to achieve unbiasedness, and doing so is possible using recently
+developed state-space models.
+%for estimators of population change to be unbiased, and state-space
+%models have been developed for this
+%models are widely used for these purposes because they allow for
+%estimating parameters of classical growth models, such as exponential,
+%logistic, and Gompertz, while accounting for demographic stochasticity
+%and observation error.
+Conventional state-space models, however, have
 three important limitations: (1) the parameters are not identifiable
@@
 resolved using a class of hierarchical models for spatially-replicated
-time-series data recently proposed by
-\citet[Biometrics]{dail_madsen:2011}. We
-expand this class of models to accommodate classical growth
+time-series data recently proposed by \citet[Biometrics]{dail_madsen:2011}.
+We expand this class of models to accommodate classical growth
 models, zero-inflation, and random effects such as observer-specific
-detection probabilities. Furthermore, we describe classical and
-Bayesian methods of parameter estimation. These new developments will
-allow researchers to apply these methods to address important
-questions regarding the factors affecting spatial and temporal
-variation in abundance. We demonstrate these developments by analyzing
-data from the North American Breeding Bird Survey. Code to fit these
-models using the \textbf{R} package \texttt{unmarked} and
+detection probabilities. We also present methods for forecasting
+populaiton size under future environmental conditions. Implementation
+of these ideas is possible using either frequentist or Bayesian
+methods, and code to fit these models using the  \textbf{R} package
+\texttt{unmarked} or
 \textbf{JAGS} is also provided.
+%is provided for doing so.
+%Furthermore, we describe classical and
+%Bayesian methods of parameter estimation, and we present methods of
+%forecasting population size under future environmental
+%conditions.
+%These new developments will
+%allow researchers to apply these methods to address important
+%questions regarding the factors affecting spatial and temporal
+%variation in abundance.
+A simulation study was used to assess bias and accurrancy of parameter
+estimators, and an analysis of %We demonstrate these developments by analyzing
+%data from the
+North American Breeding Bird Survey is presented for illustration.
 }
@@
 the inherent stochasticity in demographic parameters and environmental
-conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
+conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
 Second, abundance---the natural state variable in studies
@@
 
+\section{Statistical Inference}
+
+\subsection{Maximum likelihood}
+
+Maximum likelihood estimation of parameters in models with random
+effects requires removing the random effects
+
+\subsection{Bayesian inference}
+
+Bayesian inference has several appealing features. First, it allows
+direct probability statements to be made about a hypothesis given
+data. Second, priors are cool. We can combine data from multiple
+sources or use existing estimates of parameters as prior
+distributions. MCMC has a bad rap for being slow, but actually, in the
+case of state-space models it may be one of the few options for
+estimation, and in fact may be faster than maximizing likelihoods.
+
+Here is some BUGS code:
+\begin{verbatim}
+model {
+lambda ~ dunif(0, 5)
+r ~ dunif(-5, 5)
+p ~ dunif(0,1)
+for(i in 1:nSites) {
+  N[i,1] ~ dpois(lambda)
+  y[i,1] ~ dbin(p, N[i,1])
+  for(t in 2:nYears) {
+    mu[i,t-1] = N[i,t-1]*exp(r)
+    N[i,t] ~ dpois(mu[i,t-1])
+    n[i,t] ~ dbin(p, N[i,1])
+    }
+  }
+}
+\end{verbatim}
+
 \section{Model Extensions}
@@
 
-\subsection{Bayesian inference}
-
-Bayesian inference has several appealing features. First, it allows
-direct probability statements to be made about a hypothesis given
-data. Second, priors are cool. We can combine data from multiple
-sources or use existing estimates of parameters as prior
-distributions. MCMC has a bad rap for being slow, but actually, in the
-case of state-space models it may be one of the few options for
-estimation, and in fact may be faster than maximizing likelihoods.
-
-Here is some BUGS code:
-\begin{verbatim}
-model {
-lambda ~ dunif(0, 5)
-r ~ dunif(-5, 5)
-p ~ dunif(0,1)
-for(i in 1:nSites) {
-  N[i,1] ~ dpois(lambda)
-  y[i,1] ~ dbin(p, N[i,1])
-  for(t in 2:nYears) {
-    mu[i,t-1] = N[i,t-1]*exp(r)
-    N[i,t] ~ dpois(mu[i,t-1])
-    n[i,t] ~ dbin(p, N[i,1])
-    }
-  }
-}
-\end{verbatim}
-
 \section{Applications}

### 2012-09-10 Dail_Mad_ext.tex
break

@@
 detection probabilities. We also present methods for forecasting
-populaiton size under future environmental conditions. Implementation
+population size under future environmental conditions. Implementation
 of these ideas is possible using either frequentist or Bayesian
@@
 %variation in abundance.
-A simulation study was used to assess bias and accurrancy of parameter
-estimators, and an analysis of %We demonstrate these developments by analyzing
-%data from the
-North American Breeding Bird Survey is presented for illustration.
+A simulation study indicated that bias was negligible and coverage nominal
+and an analysis of North American Breeding Bird Survey is presented
+for illustration.
 }
@@
 a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
-modeled using a deterministic population growth model, $\mu(N_{t-1})$,
+modeled using a population growth model, $\mu(N_{t-1})$,
 coupled with a stochastic model allowing
@@
 $\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
-act as though there is no process variation associated with $N_1$, as
+ignore process variation associated with $N_1$, as
 indicated by Eq~\ref{eq:ss1a}.
-% indicates that no process variation is
-%present in year 1. This assumption is often made
-%due to parameter identifiability problems, which we will describe
-%shortly. First, we turn our attention to the models for the random
-%effects.
 
@@
 area surveyed. However, this transformation is problematic since
-abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
+local abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
 zeros are replaced with some small number, although the
@@
 we allow $N_1$ to be a random variable. But what should its
-distribution, $[N_1]$ be? And how could $\bm{\theta}$ be
+distribution, $[N_1|\bm{\theta}]$ be? And how could $\bm{\theta}$ be
 estimated since there is only a single observation available?
@@
 A fourth problem with these models is that they do not admit
-spatial variation, which
-%. Recent work has sought to remedy this deficiencyThis is not a fair criticism because this
-%lies beyond the scope of traditional state-space models, but
-%it does
-restrict the utility of their since it is typically
-impossible assess the impacts of factors such as habitat
-fragmentation or climate change on population dynamics without
-considering spatial variation in abundance. Indeed many populations are
-regulated by spatial processes such as source-sink dynamics.
+spatial variation. This reduces the scope of the inferences that can
+be drawn from the models and it ignores the importance of space in
+population regulation.
+%, which
+%restrict the utility of their since it is typically
+%impossible assess the impacts of factors such as habitat
+%fragmentation or climate change on population dynamics without
+%considering spatial variation in abundance. Indeed many populations are
+%regulated by spatial processes such as source-sink dynamics.
 
@@
 processes. In contrast, recent work has sought to use state-space
-models fot inference about population processes such as mortality and
+models for inference about population processes such as mortality and
 recruitment. To do so, methods have been developed to combine multiple
@@
 Maximum likelihood estimation of parameters in models with random
-effects requires removing the random effects
+effects requires removing the random effects by integration or
+summation.
 
@@
 
+\section{Forecasting Future Population Size}
+
 \section{Applications}

### 2012-09-14 Dail_Mad_ext.tex
added tables

@@
 \usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
+\usepackage{rotating}
 
@@
 
+\newpage
+
+% Tables
+
+\begin{table}[t]
+  \centering
+  \caption{Changed parameter values, by series of simulations.  We
+    simulated 1000 sets of data each combination of parameter values
+    for 100 sites over 40 years.  We assumed that initial abundance
+    ($\Lambda$) was Poisson distributed.  For the exponential growth
+    simulations we included all combinations of low, medium, and high
+    $\Lambda$, growth rate ($r$), and detection probability ($p$).  For the Ricker
+    model simulations we used $\Lambda$ = 10 and $p$ = 0.25, and simulated low,
+    medium, and high values of $r$ and equilibrium density ($K$).  For the
+    Ricker + immigration dynamics model we fixed all parameters the
+    same as the Ricker model (with $r$ = 0.05 and $K$ = 10) and simulated
+    low, medium, and high values of immigration rate ($\iota$).}
+  \begin{tabular}{lcccccccc}
+    \hline
+    & \multicolumn{3}{c}{Exponential Growth} && Ricker &&
+    \multicolumn{2}{c}{Ricker + Immigration} \\
+    \cline{2-4}     \cline{6-6}    \cline{8-9}
+%    & \multicolumn{3}{c}{\rule{25mm}{.1pt}} & \rule{5cm}{.5pt} & \multicolumn{2}{c}{\rule{20mm}{.1pt}} \\
+                & $\Lambda$ & $r$ & $p$ && $r$  && $K$ & $\iota$  \\
+    \hline
+    Low	        &1	&-0.01	&0.05	&&0.005	&&5	&0.005  \\
+    Med	        &5	&0	&0.25	&&0.05	&&10	&0.05   \\
+    High	&10	&0.005	&0.5	&&0.1	&&20	&0.5    \\
+    \hline
+  \end{tabular}
+\end{table}
+
+\vfill
+%\clearpage
+\newpage
+
+% Perhaps stack the species
+
+\begin{sidewaystable}
+  \centering
+  \small
+  \caption{Model selection table for ovenbirds and golden-winged
+    warblers in Maryland and Virginia, 1966-2010.  We present model name
+    and number, number of parameters (Par.), and difference in Akaike's
+    information criterion between each model and the top model of that
+    set ($\Delta$AIC).  For each species, the first section compares
+    models for initial abundance, the second for detection probability,
+    and the third for dynamics.}
+  \begin{tabular}[h]{lcclcc}
+\hline
+A. Ovenbirds	&&&		B. Golden-Winged Warblers \\
+\hline
+Model	&Par.	&$\Delta$AIC	&Model	&Par.	&$\Delta$AIC \\
+A.1. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0
+&B.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&0 \\
+A.2. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&1262.7
+&B.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0.4 \\
+A.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4
+&1264.7	&B.3. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&37.8 \\
+A.4. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
+&0	&B.4. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)	&5
+&0 \\
+A.5. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind)	&7	&0.9
+&B.5. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&3.5 \\
+A.6. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(1st)	&5	&5.0
+&B.6. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
+&5.9 \\
+A.7. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&6.4
+&B.7. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind)	&7	&9.3 \\
+A.8. NB[$\Lambda$(.)$\alpha$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
+&10	&0	&B.8. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)
+&5	&0 \\
+A.9. NB[$\Lambda$(.)$\alpha$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
+&10	&8.4
+&B.9. ZIP[$\Lambda$(.)$\psi$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(1st)
+&6	&1.7 \\
+A.10. NB[$\Lambda$(.)$\alpha$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(wind+1st)
+&9	&36.5
+&B.10. ZIP[$\Lambda$(.)$\psi$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(1st)
+&6	&2.0 \\
+A.11. NB[$\Lambda$(.)$\alpha$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
+&10	&38.6
+&B.11. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
+&7	&2.6 \\
+A.12. NB[$\Lambda$(.)$\alpha$(.)]Gompertz[$r$(.)$K$(.)]$p$(wind+1st)
+&9	&190.9
+&B.12. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz[$r$(.)$K$(.)]$p$(1st)	&6
+&3.0 \\
+A.13. NB[$\Lambda$(.)$\alpha$(.)]Ricker[$r$(.)$K$(.)]$p$(wind+1st)
+&9	&195.1
+&B.13. ZIP[$\Lambda$(.)$\psi$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
+&7	&3.7 \\
+A.14. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
+&271.3
+&B.14. ZIP[$\Lambda$(.)$\psi$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
+&7	&4.2 \\
+A.15. NB[$\Lambda$(.)$\alpha$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
+&9	&273.7
+&B.15. ZIP[$\Lambda$(.)$\psi$(.)]Ricker[$r$(.)$K$(.)]$p$(1st)	&7
+&5.2 \\
+A.16. NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
+&9	&1856.7
+&B.16. ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st)
+&6	&12.5 \\
+\hline
+\end{tabular}
+\end{sidewaystable}
+
 \end{document}

### 2012-09-21 Dail_Mad_ext.tex
first commit

--- 2012-09-10 Dail_Mad_ext.tex
@@
 process variation and observation error in such efforts is necessary
-to achieve unbiasedness, and doing so is possible using recently
+to achieve unbiasedness \hl{[may be too strange sounding a term for
+  abstract]}, and doing so is possible using recently
 developed state-space models.
@@
 three probability distributions, which we represent
-using bracket notation:
+using bracket notation: \hl{[I'm not familiar with bracket notation,
+  so these equations are confusing to me.  How widely used is this
+  notation in ecology?  Will others likely understand this?]}
 \begin{subequations}
@@
 simultaneously. However, their model did not include serial
-depedence, which is a hallmark of population models. This limits the
+dependence, which is a hallmark of population models. This limits the
 utility of their model for making inferences about explicit population
@@
 In this paper, we focus on the model of
-\citet{dail_madsen:2011} %developed a model (henceforth the DM model)
+\citet[henceforth the DM model;]{dail_madsen:2011}
+%developed a model (henceforth the DM model)
 that simultaneous resolves each of the abovementioned problems with
@@
 In the following section, we describe the DM model in its original
-form and explain how it resolves each of the deficieinces with
+form and explain how it resolves each of the deficiencies with
 standard state-space models. In Section~\ref{sec:ext}, we extend the
@@
 zeros and nuisance variables such as random observer effects. Both
-frquentist and Bayesian methods of inference are discussed, and code
+frequentist and Bayesian methods of inference are discussed, and code
 for fitting models is presented in the appendices.
@@
 of  the paper is to provide ecologists with flexible and accessible means of
-adressing important questions related to the variaion of abundance in
+adressing important questions related to the variation of abundance in
 space and time.

### 2012-09-21 supp/sim_table_exp.tex
first commit

\relax 
\providecommand\HyperFirstAtBeginDocument{\AtBeginDocument}
\HyperFirstAtBeginDocument{\ifx\hyper@anchor\@undefined
\global\let\oldcontentsline\contentsline
\gdef\contentsline#1#2#3#4{\oldcontentsline{#1}{#2}{#3}}
\global\let\oldnewlabel\newlabel
\gdef\newlabel#1#2{\newlabelxx{#1}#2}
\gdef\newlabelxx#1#2#3#4#5#6{\oldnewlabel{#1}{{#2}{#3}}}
\AtEndDocument{\ifx\hyper@anchor\@undefined
\let\contentsline\oldcontentsline
\let\newlabel\oldnewlabel
\fi}
\fi}
\global\let\hyper@last\relax 
\gdef\HyperFirstAtBeginDocument#1{#1}
\providecommand*\HyPL@Entry[1]{}
\HyPL@Entry{0<</S/D>>}
\@writefile{lot}{\contentsline {table}{\numberline {1}{\ignorespaces Simulation results from 27 cases with 1000 Monte Carlo replicates for each.}}{1}{table.1}}
\newlabel{tab:simexp}{{1}{1}{Simulation results from 27 cases with 1000 Monte Carlo replicates for each}{table.1}{}}

### 2012-09-22 Dail_Mad_ext.tex
comments and fixes

@@
   so these equations are confusing to me.  How widely used is this
-  notation in ecology?  Will others likely understand this?]}
+  notation in ecology?  Will others likely understand this?  Also, we might want
+to save space by trimming a bunch of this discussion back down.]}
 \begin{subequations}
@@
 $\omega$ is the probability of surviving and not emigrating, and
-$\gamma$ is product of the birth rate and the emigration rate,
-i.e. the recruitment rate. \citet{dail_madsen:2011} proposed three
+$\gamma$ is the sum\hl{[I think!]} of the birth rate and the emigration rate,
+i.e. the recruitment rate. \hl{[Feels a little repetitious with the material 
+before the equation, although I understand you're describing S and G 
+there and omega and gamma here.  In general, these sections 
+describing the DM models are clear and well written, but we could 
+possibly spend less time on them b/c it's already published.  
+Possibly it's better this way, though.]} \citet{dail_madsen:2011} proposed three
 models for $\gamma$: the constant model,
@@
 
-Here is some BUGS code:
+Here is some BUGS code: \hl{[This is code for a model we haven't
+presented yet.  Could we move these sections after model extensions?]}
 \begin{verbatim}
@@
 instantaneous population growth rate at low population densities, and
-both parameters are constrained to be positive.  This model is based
-on the \citet{ricker:1954} discrete time version of the logistic population
-growth model.  We also implemented the \citet{gompertz:1825} density-dependent
-model:
+both parameters are constrained to be positive.  This is a stochastic version
+of the Ricker-logistic population growth model \citep{ricker:1954}.  We also 
+implemented a modified Gompertz-logistic density-dependent
+model \citep{hart_gotelli:2011}:
 \begin{equation}
-  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1})/\log(K)))
+  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1)))
 \label{eq:gomp}
 \end{equation}
-where the log of $N_{it-1}$ is only taken when $N_{it-1}>0$
-(when $N_{it-1} \equiv 0$, the full expression simplifies to 0 anyway).  Here
-the interpretations of $r$ and $K$ are similar to in the Ricker model, but
-$K$ is constrained to be greater than one.
+Here the interpretations of $r$ and $K$ are similar to in the Ricker model.
 
@@
 \begin{equation}
-  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1}) + \text{Poisson}(\iota)
-  \label{eq:expimm1}
-\end{equation}
-or equivalently
-\begin{equation}
   N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
@@
 \end{equation}
-where $\iota$ represents the immigration rate.  This model is close to
-the constant DM model (equation 4), with $\exp(r)$ instead of $\omega$
+where $\iota$ represents the immigration rate, and is constrained to be 
+positive (this is equivalent to separate Poisson processes for growth and 
+immigration).  This model is close to
+the constant DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead of $\omega$
 and $\iota$ instead of $\gamma$, except that the first
@@
 
-The zero inflation factor can also affect
-recruitment or population dynamics (see below).
+%The zero inflation factor can also affect
+%recruitment or population dynamics (see below).
 We also considered a zero-inflated negative binomial model of initial
@@
 initial abundance but also to recruitment and population growth terms.
-For example, the recruitment term of the constant DM model (equation
-4) can be modified as follows:
+For example, the recruitment term of the constant DM model 
+(Eq.~\ref{eq:Nt}) can be modified as follows:
 \begin{equation}
@@
 \end{equation}
+\begin{comment}
 When both initial abundance and dynamics are modeled as zero-inflated,
@@
 also zero-inflated unless $\psi$ is time-varying (because these dynamic
-models cannot recover from zero abundance).  We have implemented
+models cannot recover from zero abundance).  
+\end{comment}
+We have implemented
 zero-inflated dynamics in the Bayesian framework using program
@@
 
-Our second series of simulations changed dynamics to the
-Ricker model.  We used an initial abundance of 10, a maximum
-growth rate of 0.005, and a detection probability of 0.25, and
-simulated low, medium, and high values of equilibrium
-abundance, each with 500 simulations (Table 1).  Our third
-series of simulations was based on the Ricker + immigration
-dynamics model; here we fixed all parameters the same as the
-Ricker model (with K = 10) and simulated low, medium, and high
-values of immigration rate with 500 simulation each (Table
-1).
+Our second series of simulations changed dynamics to the Ricker model.  
+We used an initial abundance of 10 and a detection probability of 0.25, 
+and simulated low, medium, and high values of equilibrium abundance 
+and maximum growth rate, each with 1000 simulations (Table 1; 
+9 total combinations).  Our third series of simulations was based on the 
+Ricker + immigration dynamics model; here we fixed all parameters the 
+same as the Ricker model (with r = 0.05 and K = 10) and simulated 
+low, medium, and high values of immigration rate with 1000 simulation 
+each (Table 1).
 
@@
 
+\begin{comment}
 Ideas:
@@
 over the wrong DID model?
+\end{comment}
 
@@
 the western parts of these states and has been declining in the region
-(CITE).  MORE INFORMATION ABOUT THE SPECIES?
-
-GENERAL ABOUT BBS DATA.  We only used data marked as acceptable for
-use in the annual BBS analysis.  We summed the total birds of each
+(CITE).  %MORE INFORMATION ABOUT THE SPECIES?
+
+%GENERAL ABOUT BBS DATA.  
+The BBS is an annual roadside survey implemented by trained observers 
+in the United States and Canada. An observer conducts 50 3-minute point counts, 
+with 400 m radiuses and 0.8 km apart from each other along a 39.4 km route. 
+We only used data marked as acceptable for use in the annual BBS analysis 
+\citep{sauer_etal:1994auk}.  We summed the total birds of each
 species seen on each route and year and used the routes (rather than
@@
 Strong winds can interfere with point count observers' ability to hear
-birds (\citep{simons_etal:2007}); we tested the effects of wind speed
+birds \citep{simons_etal:2007}; we tested the effects of wind speed
 on detection probability.  BBS volunteers record wind conditions at
@@
 
-HOW WE OBTAINED AND MANIPULATED PRISM DATA
+%HOW WE OBTAINED AND MANIPULATED PRISM DATA
 We ran a series of maximum-likelihood based models for each species,
@@
 models of initial abundance (Poisson, negative binomial, and
-zero-inflated Poisson) with population growth dynamics (equation 7)
+zero-inflated Poisson) with population growth dynamics (Eq.~\ref{eq:exp})
 and no covariates.  We selected the minimum Akaike's Information
@@
 immigration, population growth + immigration, Ricker + immigration,
-and Gompertz + immigration.  We then tested the effect of average
-minimum temperature for June and July on all dynamics parameters from
-the minimum AIC dynamics model.  When run in a maximum likelihood
-framework, these models require a maximum abundance to integrate over
-(Royle 2004, Dail and Madsen 2011); we used 600 for ovenbirds and 350
+and Gompertz + immigration.  %We then tested the effect of average
+%minimum temperature for June and July on all dynamics parameters from
+%the minimum AIC dynamics model.  
+As above, these models require a maximum abundance to integrate over
+when run in a maximum likelihood
+framework; we used 600 for ovenbirds and 350
 for GWWA.
@@
 Bayesian framework.  We added random observer effects and, where
-appropriate, zero-inflated dynamics.  WILL ADD MORE HEREWe used
-non-informative priors.  We tested for lack of convergence using X
+appropriate, zero-inflated dynamics.  We used
+non-informative priors.  We tested for lack of convergence using 5
 Markov chains for each model \citep{gelman_rubin:1992}.
@@
 For each chain
-we sampled the MCMC for Y iterations, after Z tuning samples.  Latent
-inclusion factors for temperature effects?
+we sampled the MCMC for at least 40,000 iterations, after at least 3,000 tuning samples.  
+%Latent inclusion factors for temperature effects?
 
 \section{Results}
+
+\subsection{Analysis of Breeding Bird Survey Data}
 
@@
 initial abundance over the Poisson and zero-inflated Poisson (Table
-1A, models A.1 - A.3).  Even compared to the Poisson, there was no
+2A, models A.1 - A.3).  Even compared to the Poisson, there was no
 evidence to support a zero-inflation factor for this species.  The
 best supported model for p was additive effects of wind speed and
-first run (Table 1A, model A.4).  First run and increasing wind speeds
+first run (Table 2A, model A.4).  First run and increasing wind speeds
 both decreased p.  All dynamics models with immigration were better
-supported than models without immigration (Table 1A, models A.8 -
+supported than models without immigration (Table 2A, models A.8 -
 A.16); the best supported of these was the Ricker + Immigration.  The
@@
 and the autoregressive and the constant model had the least support.
-There was little support for an effect of minimum June and July
-temperature on dynamics parameters of the Ricker + Immigration model
-(Table 1A, models A.17 - A.18).
+%There was little support for an effect of minimum June and July
+% temperature on dynamics parameters of the Ricker + Immigration model
+%(Table 2A, models A.17 - A.18).
 
@@
 the intercept for p (on the logit scale) dropped from -1.5 $\pm$ 0.1 to
--2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  [Richard, I ran
-the temperature model in the Bayesian framework (w/ and w/o random
-effect) too.  Should we include those results?  As earlier, very
-little evidence for an effect of temperature on any of the parameters,
-although I couldn't get the latent inclusion parameter to work for r.]
+-2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
+%[Richard, I ran
+%the temperature model in the Bayesian framework (w/ and w/o random
+%effect) too.  Should we include those results?  As earlier, very
+%little evidence for an effect of temperature on any of the parameters,
+%although I couldn't get the latent inclusion parameter to work for r.]
 Gelman and Rubin diagnostics and visual examination of the ? plots
@@
 abundance of GWWA than for the negative binomial; both were strongly
-supported over the Poisson (Table 1B, models B.1 - B.3).  The best
-supported model for p was an effect of first run (Table 1B, model
+supported over the Poisson (Table 2B, models B.1 - B.3).  The best
+supported model for p was an effect of first run (Table 2B, model
 B.4).  The best supported dynamics model was population growth (Table
-1B, model B.8); the estimate of r from this model was -0.058 $\pm$ 0.017
+2B, model B.8); the estimate of r from this model was -0.058 $\pm$ 0.017
 (SE).  There was no support for immigration added to the population
@@
 of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$ 0.002,
-respectively).  There was considerable support for an effect of
-average minimum June and July temperature on r in the population
-growth model (Table 1B, model B.17), with a negative slope.
-
+respectively).  %There was considerable support for an effect of
+%average minimum June and July temperature on r in the population
+%growth model (Table 1B, model B.17), with a negative slope.
 Because of the closeness of rankings for the ZIP and negative binomial
@@
 
-STILL WORKING ON JAGS GWWA RESULTS
-Estimated detection probabilities were generally low.  For OVEN, the
-estimated probability of detecting a bird with no wind and not the
-first time an observer had run a route varied between 0.046 $\pm$ 0.002
-(model A.16) and 0.192 $\pm$ 0.015 (model A.11), among models that
-accounted for both.  For GWWA, the estimated probability of detecting
-a bird by a non-first time observer varied between 0.025 $\pm$ 0.008
-(model B.18) and 0.141 $\pm$ 0.027 (model B.16), among models that
-accounted for first but not wind.
+Estimates for the top ranked model for GWWA (ZIP[Λ(.)α(.)]Trend[r(.)]p(1st)) 
+were similar when run in the Bayesian framework.  When random observer 
+effects were added, the estimate for Λ  increased dramatically 
+(from 22.4 ± 7.6 to 55.5 ± 30.9).  The estimate of the intercept for p (on the 
+logit scale) dropped from -3.5 ± 0.2 to -5.6 ± 1.1, and the estimate of σp 
+was 1.7 ± 0.7.  Adding zero-inflation to the dynamics (as well as the initial 
+abundance) has no effect for trend models, because populations that started 
+at 0 would stay at 0 in any case.  Therefore, we tested zero-inflated dynamics 
+on the Trend + Imm. model.  Adding zero-inflation to the dynamics increased 
+the estimate of ι from 0 to 0.091 ± 0.049, but had only minor effects on the 
+other parameters.  Gelman and Rubin diagnostics and visual examination of 
+the ? plots provided evidence for a lack of convergence in estimates of Λ 
+for all Bayesian models ran, and for most parameters when random effects 
+were included.
+
+Estimated detection probabilities were generally low.  The estimated probability 
+of detecting an OVEN in the top model (A.8) varied between 0.157 ± 0.014 
+and 0.184 ± 0.015, depending on wind speed and whether the observer had 
+run that route before.  The estimated probability of detecting a GWWA went 
+from 0.013 ± 0.006 for first time observers to 0.028 ± 0.008 for repeat 
+observers for the top model (B.8).  Accounting for random observer effects 
+decreased average detection probability for both species.
 
@@
 
-A unique aspect of the Dail and Madsen model as originally devised was
-that it allowed for the estimation of demographic parameters under
-strict distributional assumptions and with the assumption of
-geographic closure. Clearly, estimating demographic parameters from
-count data is a loftyn ambitious goal, and the required assumption
-will not be valid in many cases. Nonetheless, their approach is
-important because it allows for the combination of count data with
-other data types such as capture-recapture data , which offers a
-means of making inferences about demographic processes at broad
-spatial scales. Models combining count data and demographic data are
-often referred to as integrated population models, and This topic has
-received much attention lately under t
+A unique aspect of the Dail and Madsen model as originally
+devised was that it allowed for the estimation of demographic
+parameters under strict distributional assumptions and with the
+assumption of geographic closure. Clearly, estimating
+demographic parameters from count data is an ambitious goal, and
+the required assumption will not be valid in many cases. One
+advantage of the original DM models over our extensions here is
+that they model population dynamics more mechanistically, by
+focusing on survival and recruitment. As we report here, these
+models often provide unrealistic estimates of these vital rates
+and are generally less supported in model selection. Adding
+density dependence directly to one or both vital rates may solve
+both of these problems, although this might require very good
+data. Another strategy would be to use estimates of vital rates
+from demographic studies as priors in a Bayesian analysis using
+DM models. Finally, count data could be combined with other data
+types such as capture-recapture data, which offers a means of
+making inferences about demographic processes at broad spatial
+scales. Models combining count data and demographic data are
+often referred to as integrated population models (IPM; CITE),
+and we believe that a DM model would provide an effective part
+of an IPM.
 
@@
 
-[paragraph on zero-inflation]The zero-inflated models for initial
+The zero-inflated models for initial
 abundance and dynamics allow one to estimate and model the proportion
@@
 to collected replicated counts within the primary periods during a
-period in which the populations could be assumed to be closed.  a
+period in which the populations could be assumed to be closed.  A
 robust design could be used to combine multiple surveys per primary
 period to increase the precision of the estimates. We envision that
-multiple other options are available as well. For instance, there
-should be no difficulty extending the model to accommodate traditional
-capture-recapture data collected during each primary period. Removal
-or time of detection etc.. Distance-related heterogeneity in detection
-probability is another source of variation that can bias estimators of
-abundance.   Show some equations for the alternative observation
-models.
-
-[paragraph on spatially-explicit dispersal]
-[Other possible extensions we could mention:
-Spatially hierarchical model (so can use individual stops)
-Other models of variability in growth or recruitment (such as negative binomial)
-Add density dependence to survival and recruitment models]
-[Discuss causes of low p values (and other BBS results) here or in BBS
-results?]
+multiple other options are available as well, such as removal, time of 
+detection, and distance modeling ([cites]).
+%For instance, there
+%should be no difficulty extending the model to accommodate traditional
+%capture-recapture data collected during each primary period. Removal
+%or time of detection etc.. Distance-related heterogeneity in detection
+%probability is another source of variation that can bias estimators of
+%abundance.   Show some equations for the alternative observation
+%models.
+
 Our emphasis was on increasing the practical utility of this class of
@@
 immigration.
+
+In our models immigration is currently modeled as independent of
+population sizes in other sites. Several alternatives suggest
+themselves if we think about movement in a metapopulation
+context ([cite]): immigration depending on total or mean
+abundance across sites the previous time step; immigration
+depending on abundance in nearby sites only; and immigration
+depending on abundance at sites as a function of their distance
+from the receiving site (LOTS OF CITATIONS). These models would
+be based on the assumptions that either the study sites
+spatially cover the whole range of the metapopulation or the
+annual changes in the populations sampled are representative of
+a larger metapopulation. Initial tests suggest that it would be
+possible to fit some or all of these models, at least in the
+Bayesian framework.
+
+\begin{comment}
+Should we include these possible extensions as well?
+
+For data sources such as the BBS, where stops are nested within
+routes, spatially hierarchical versions of these models could be
+the best way to model data from individual stops over many
+routes. These models would allow the analyst to account for the
+potential spatial dependence of stops within a route, by having
+the route as another random effect.
+
+Other models of variability in growth or recruitment (such as
+negative binomial or generalized Poisson) could be informative.
+Stochasticity (represented as variation in the numbers of
+animals occurring in a site and year around what the
+deterministic dynamics predict) could be over or underdispersed,
+compared to the Poisson distribution. Alternatively,
+stochasticity could possibly be applied directly to population
+model parameters (such as r and K), instead of numbers of
+animals.
+\end{comment}
+
+%[paragraph on spatially-explicit dispersal]
+%[Other possible extensions we could mention:
+%Spatially hierarchical model (so can use individual stops)
+%Other models of variability in growth or recruitment (such as negative binomial)
+%Add density dependence to survival and recruitment models]
+%[Discuss causes of low p values (and other BBS results) here or in BBS
+%results?]
+
+[simulation results discussion]
+
+We have presented a test case for these models using data from
+the BBS for two species. We demonstrate that these models can be
+used to model and estimate initial abundance, detection
+probability, and density-independent and dependent functions for
+population dynamics. We tested these models in both the
+classical and Bayesian frameworks and found strengths and
+weaknesses of each approach. The Bayesian analyses permit more
+flexibility and allow the incorporation of random observer
+effects and zero-inflated dynamics. Our results suggest ignoring
+variation between observers can lead to underestimation of
+initial abundance and equilibrium population sizes. On the other
+hand, our Bayesian analyses did have convergence problems,
+particularly for the species (GWWA) with sparser data. It's
+likely that poor model fits are related to the low estimates of
+detection probability, which in turn are probably caused by the
+large 400 m radius used for BBS point counts. Again, we believe
+that direct information on detectability (such as repeated
+visits within a season) would greatly improve the fit of these
+models in both frameworks.
+\hl{[Alternatively, could put discussions of our results where
+they fit in by subject area, rather than separately. For
+instance, could discuss BBS population dynamics model selection
+in population models paragraph, BBS Bayesian vs.
+frequentist strengths and weakness in the Bayesian paragraph
+(currently not written), and GWWA ZIP results in ZIP paragraph.]}
 
@@
 demographic parameters or in derived parameters such as population
-growth rate. Furthermore, under the Bayesian approach , population
+growth rate. Furthermore, under the Bayesian approach, population
 viability analysis is trivial because projecting populations into the

### 2012-09-28 Dail_Mad_ext.tex
updated AIC tables
abstract

@@
 hypotheses about density dependence and predicting species' responses
-to future environmental change or conservation actions. Accounting for
-process variation and observation error in such efforts is necessary
-to achieve unbiasedness \hl{[may be too strange sounding a term for
-  abstract]}, and doing so is possible using recently
-developed state-space models.
-%for estimators of population change to be unbiased, and state-space
-%models have been developed for this
-%models are widely used for these purposes because they allow for
-%estimating parameters of classical growth models, such as exponential,
-%logistic, and Gompertz, while accounting for demographic stochasticity
-%and observation error.
+to future environmental change or conservation actions. Failure to
+account for observation error in such models can result in
+bias, and thus ecologists have increasingly relied on state-space
+models to directly model both the observation error and the ecological
+process of interest.
 Conventional state-space models, however, have
@@
 observation error. We demonstrate how each of these problems can be
-resolved using a class of hierarchical models for spatially-replicated
-time-series data recently proposed by \citet[Biometrics]{dail_madsen:2011}.
+resolved using a class of hierarchical models proposed by
+\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
+time-series data that attributes observation error to imperfect
+detection.
 We expand this class of models to accommodate classical growth
-models, zero-inflation, and random effects such as observer-specific
+models (e.g. exponential, Ricker, etc $\dots$), zero-inflation, and
+random effects such as observer-specific
 detection probabilities. We also present methods for forecasting
@@
 methods, and code to fit these models using the \textbf{R} package
-\texttt{unmarked} or
-\textbf{JAGS} is also provided.
-%is provided for doing so.
-%Furthermore, we describe classical and
-%Bayesian methods of parameter estimation, and we present methods of
-%forecasting population size under future environmental
-%conditions.
-%These new developments will
-%allow researchers to apply these methods to address important
-%questions regarding the factors affecting spatial and temporal
-%variation in abundance.
-A simulation study indicated that bias was negligible and coverage nominal
-and an analysis of North American Breeding Bird Survey is presented
-for illustration.
+\texttt{unmarked} or \textbf{JAGS} is also provided.
+A simulation study indicated that bias was negligible and coverage
+nominal for the proposed model extensions. An analysis of data from
+the North American Breeding Bird Survey highlighted how these methods
+can be readily applied to existing data, but it also suggested the
+that precision will be low when direct information about detection
+probability (such as is collected using distance sampling or
+replicated counts) is unavailable.
 }
@@
 % Perhaps stack the species
-
+\begin{comment}
 \begin{sidewaystable}
@@
 &B.11. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
-&7	&2.6 \\
+&7	&2.5 \\
 A.12. NB[$\Lambda$(.)$\alpha$(.)]Gompertz[$r$(.)$K$(.)]$p$(wind+1st)
-&9	&190.9
+&9	&192.8
 &B.12. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz[$r$(.)$K$(.)]$p$(1st)	&6
@@
 \end{sidewaystable}
+\end{comment}
+
+% How about separate tables?  Also updated the Gompertz results.
+\begin{table}
+  \centering
+  \small
+  \caption{Model selection table for ovenbirds in Maryland and Virginia,
+    1966-2010.  We present model name
+    and number, number of parameters (Par.), and difference in Akaike's
+    information criterion between each model and the top model of that
+    set ($\Delta$AIC).  The first section compares
+    models for initial abundance, the second for detection probability,
+    and the third for dynamics.}
+  \begin{tabular}[h]{lcc}
+\hline
+Model	&Par.	&$\Delta$AIC	\\
+\hline
+A. Initial Abundance && \\
+A.1. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0\\
+A.2. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&1262.7\\
+A.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4 &1264.7\\
+\hline
+B. Detection Probability && \\
+B.1. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
+&0	\\
+B.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind)	&7	&0.9\\
+B.3. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(1st)	&5	&5.0\\
+B.4. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&6.4\\
+\hline
+C. Dynamics && \\
+C.1. NB[$\Lambda$(.)$\alpha$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
+&10	&0	\\
+C.2. NB[$\Lambda$(.)$\alpha$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
+&10	&8.4 \\
+C.3. NB[$\Lambda$(.)$\alpha$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(wind+1st)
+&9	&36.5\\
+C.4. NB[$\Lambda$(.)$\alpha$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
+&10	&38.6\\
+C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz[$r$(.)$K$(.)]$p$(wind+1st)
+&9	&192.8\\
+C.6. NB[$\Lambda$(.)$\alpha$(.)]Ricker[$r$(.)$K$(.)]$p$(wind+1st)
+&9	&195.1\\
+C.7. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
+&271.3\\
+C.8. NB[$\Lambda$(.)$\alpha$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
+&9	&273.7\\
+C.9. NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
+&9	&1856.7\\
+\hline
+\end{tabular}
+\end{table}
+
+\vfill
+%\clearpage
+\newpage
+
+\begin{table}
+  \centering
+  \small
+  \caption{Model selection table for golden-winged
+    warblers in Maryland and Virginia, 1966-2010.  We present model name
+    and number, number of parameters (Par.), and difference in Akaike's
+    information criterion between each model and the top model of that
+    set ($\Delta$AIC).  The first section compares
+    models for initial abundance, the second for detection probability,
+    and the third for dynamics.}
+  \begin{tabular}[h]{lcc}
+\hline
+Model	&Par.	&$\Delta$AIC\\
+\hline
+A. Initial Abundance && \\
+A.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&0 \\
+A.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0.4 \\
+A.3. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&37.8 \\
+\hline
+B. Detection Probability && \\
+B.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)	&5
+&0 \\
+B.2. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&3.5 \\
+B.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
+&5.9 \\
+B.4. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind)	&7	&9.3 \\
+\hline
+C. Dynamics && \\
+C.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)
+&5	&0 \\
+C.2. ZIP[$\Lambda$(.)$\psi$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(1st)
+&6	&1.7 \\
+C.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(1st)
+&6	&2.0 \\
+C.4. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
+&7	&2.5 \\
+C.5. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz[$r$(.)$K$(.)]$p$(1st)	&6
+&3.0 \\
+C.6. ZIP[$\Lambda$(.)$\psi$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
+&7	&3.7 \\
+C.7. ZIP[$\Lambda$(.)$\psi$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
+&7	&4.2 \\
+C.8. ZIP[$\Lambda$(.)$\psi$(.)]Ricker[$r$(.)$K$(.)]$p$(1st)	&7
+&5.2 \\
+C.9. ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st)
+&6	&12.5 \\
+\hline
+\end{tabular}
+\end{table}
 

### 2012-09-28 Dail_Mad_ext.tex
intro

@@
 logistic, or it might be density-independent, as in the case of
-exponential growth, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$
-is the intrinsic rate of increase. The full model can now be written
-as:
+exponential growth.
+%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
+The full model can now be written as:
 \begin{subequations}
@@
 state-space models, the two sets of random effects
-are assumed to be i.i.d Gaussian deviates:
+are assumed to be i.i.d. Gaussian deviates:
 $\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
@@
 ignore process variation associated with $N_1$, as
-indicated by Eq~\ref{eq:ss1a}.
+indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
+assumed to be at equilibrium such that the $N_1$ can be regarded as an
+outcom of the equilibrium distribution.
 
@@
 \citep{devalpine_hastings:2002}. The problem with this is that it
-allows for negative values of the state-variable, which is
-inconsistent with the observed data.
+allows for negative and non-integer values of the state-variable, which is
+inconsistent with the observed data, and may result in implausible predictions.
 %The fact that both abundance ($N_t$) and the observed data ($X_t$) are
@@
 %distributions.
-Nonetheless, efforts have been made to adhere to the
-Gaussian assumptions by tranforming abundance to
-density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
-area surveyed. However, this transformation is problematic since
-local abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
-zeros are replaced with some small number, although the
-effect of this is rarely discussed.
+Transforming the count data and the underlying abundance variables do
+not fully resolve these issues \citep{ohara_kotze:2010}.
+%Nonetheless, efforts have been made to adhere to the
+%Gaussian assumptions by tranforming abundance to
+%density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
+%area surveyed. However, this transformation is problematic since
+%local abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
+%zeros are replaced with some small number, although the
+%effect of this is rarely discussed.
 

### 2012-09-28 Dail_Mad_ext.tex
intro

@@
 a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
-modeled using a population growth model, $\mu(N_{t-1})$,
-coupled with a stochastic model allowing
-for random process variation. %, which we denote $\grave{o}_t$.
-The deterministic model may be density-dependent, for example
-logistic, or it might be density-independent, as in the case of
-exponential growth.
+modeled using a population growth model, $\mu(N_{t-1})$
+and random process variation denoted $\grave{o}_t$.
+The growth model may be density-dependent, as in the case of the
+Ricker model, or it might be density-independent, such as when growth
+is exponential.
 %, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
@@
 Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
-are the most widespread approach for modeling population dynamics
+are among the most widespread approaches for modeling population dynamics
 using time-series data, several problems are evident. Namely, (1)
@@
 inconsistent with the observed data, and may result in implausible predictions.
-%The fact that both abundance ($N_t$) and the observed data ($X_t$) are
-%typically non-negative integers suggests that it is inappropriate to
-%model process variation and observation error using Gaussian
-%distributions.
 Transforming the count data and the underlying abundance variables do
-not fully resolve these issues \citep{ohara_kotze:2010}.
-%Nonetheless, efforts have been made to adhere to the
-%Gaussian assumptions by tranforming abundance to
-%density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
-%area surveyed. However, this transformation is problematic since
-%local abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
-%zeros are replaced with some small number, although the
-%effect of this is rarely discussed.
+little to resolve these issues \citep{ohara_kotze:2010}.
 
@@
 three probability distributions, which we represent
-using bracket notation: \hl{[I'm not familiar with bracket notation,
-  so these equations are confusing to me.  How widely used is this
-  notation in ecology?  Will others likely understand this?  Also, we might want
-to save space by trimming a bunch of this discussion back down.]}
+using bracket notation (i.e., $[Y|\Omega]$ is interpretted
+as the probability distribution of the random variable $Y$ given the
+parameter $\Omega$). The three probability distributions required for a
+state-space model are:
+%\hl{[I'm not familiar with bracket notation,
+%  so these equations are confusing to me.  How widely used is this
+%  notation in ecology?  Will others likely understand this?  Also, we might want
+%to save space by trimming a bunch of this discussion back down.]}
 \begin{subequations}
@@
 influencing how abundance changes over time, and $\bm{p}$ are the
-observation error parameters. The three lines of general model
-described by Eq.~\ref{eq:ss2} correspond to the three equations shown
-for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
-we allow $N_1$ to be a random variable. But what should its
-distribution, $[N_1|\bm{\theta}]$ be? And how could $\bm{\theta}$ be
-estimated since there is only a single observation available?
-One approach to estimating the $\bm{\theta}$ is to assume that the
-population is in equlibrium, i.e. that the expected value of $N_t$ is
-constant through time. Althought this makes the parameters
-identifiable, assuming equlibrium defeats the objective of many
-studies of population dynamics, namely determining why a population
-varies over time.
-
-\begin{comment}
-  However, a completSpecifically, the Markovian nature of the model
-  implies the following likelihood The first term is the probability
-  density for the first observation in the time series, $X_1$, yet it
-  is intuitively not possible to estimate the parameter(s) of this
-  distribution using only a single observation.
-  % This problem has resulted in several papers with interesting names
-  % like ``multi-modal likelihoods and ridges etc...''
-  Two options have been proposed handling the identifiability
-  problem. First, one can assume that the first observation in the
-  time series is made without error. Second, one can assume that the
-  population is at equilibrium so that that the x can be replaced with
-  the equilibrium distribution. However, assuming equilibrium defeats
-  the objective of many studies of population dynamics, namely
-  determining if and why a population is at equilibrium.
-\end{comment}
+observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
+and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
+models, but what should be the distribution for Eq.~\ref{eq:ss2}?
+%The three lines of general model
+%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
+%for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
+%we allow $N_1$ to be a random variable. But what should its
+%istribution, $[N_1|\bm{\theta}]$ be?
+And how could $\bm{\theta}$ be
+estimated given that there is only a single observation available?
+Often, these problems are simply assumed away. For
+example, sometimes researchers assume that there is no process variation in
+the first year, although this seems unjustified. Alternatively, the
+population may be assumed to be at equilibrium such that the expected
+value of $N_t$ is constant through time. Although this
+makes the parameters $\bm{\theta}$ identifiable, assuming equlibrium
+defeats the objective of many studies of population dynamics, namely
+determining why a population varies over time.
 
@@
 utility of their model for making inferences about explicit population
-processes. In contrast, recent work has sought to use state-space
-models for inference about population processes such as mortality and
-recruitment. To do so, methods have been developed to combine multiple
-sources of information, such as counts and mark-recapture data
-\citep{besbeas_etal:2002, buckland_etal:2004,
-  schaub_etal:2007}. Although these integrated
-population models may be viewed as the gold standard in state-space
-modeling, ecologists are not always so fortunate to have direct
-information about vital rates, especially at large spatial scales.
-Rather, count data are much more common, and
-are produced by many of the largest monitoring programs in the world,
-such as the North American Breeding Bird Survey
-\citep[BBS;][]{robbins_etal:1986}.
+processes.
 
@@
 Should we include these possible extensions as well?
+
+In contrast, recent work has sought to use state-space
+models for inference about population processes such as mortality and
+recruitment. To do so, methods have been developed to combine multiple
+sources of information, such as counts and mark-recapture data
+\citep{besbeas_etal:2002, buckland_etal:2004,
+  schaub_etal:2007}. Although these integrated
+population models may be viewed as the gold standard in state-space
+modeling, ecologists are not always so fortunate to have direct
+information about vital rates, especially at large spatial scales.
+Rather, count data are much more common, and
+are produced by many of the largest monitoring programs in the world,
+such as the North American Breeding Bird Survey
+\citep[BBS;][]{robbins_etal:1986}.
 

### 2012-09-28 Dail_Mad_ext.tex
data

@@
 state-space model are:
+%%%% JEFF, does this extra sentence help? This notation comes from
+%%%% several of the original state-space papers and is widespread in
+%%%% books like Andy's and Bill's.
 %\hl{[I'm not familiar with bracket notation,
@@
   \begin{align}
-    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
+    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
     [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
@@
 and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
-models, but what should be the distribution for Eq.~\ref{eq:ss2}?
+models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
 %The three lines of general model
@@
 population regulation.
-%, which
-%restrict the utility of their since it is typically
-%impossible assess the impacts of factors such as habitat
-%fragmentation or climate change on population dynamics without
-%considering spatial variation in abundance. Indeed many populations are
-%regulated by spatial processes such as source-sink dynamics.
 
@@
 also resolved the problems of non-identifiability for the parameters
-of the initial state at time $t=1$. Of these extensions, the work by
-\citet{kery_etal:2009} is unique in that it addressed multiple limitations
+of the initial state at time $t=1$. Of these extensions, only the work by
+\citet{kery_etal:2009} addressed several of these limitations
 simultaneously. However, their model did not include serial
@@
 \citet[henceforth the DM model;]{dail_madsen:2011}
-%developed a model (henceforth the DM model)
 that simultaneous resolves each of the abovementioned problems with
@@
 data.
-\begin{comment}
-  The model (1) recognizes
-  the discrete nature of count data and the underlying true abundance
-  state, (2) includes a realistic observation model with explicit
-  detection probability parameters (3) has estimable parameters,
-  including those describing the initial state at time $t=1$, and (4)
-  allows for inference about both temporal and spatial variation in
-  abundance.
-
-  In some respects, the DM model can be viewed as a specific
-  manifestation of the general state-space model formulated by
-  \citep{buckland_etal:2004}. However, their model was designed to
-  allow for inference about demographic parameters, which will
-  typically require mark-recapture data. Although this may be viewed
-  as the gold standard in many studies of wildlife populations, in
-  this paper, we focus on the analysis of count data, which is the
-  typical source of information used in state-state modeling
-  applications, and one of the most widely available forms of
-  ecological data---it is also the data required by the DM model.
-\end{comment}
 In the following section, we describe the DM model in its original
@@
 The DM model requires count data collected at $R$ sites, each of which
-is surveyed on $T$ time periods. A site is ideally, but not necessarily, a
+is surveyed on $T$ primary sampling periods. A site is ideally a
 well-defined region of a study area, such as a wetland or a patch of
-early-successional habitat. In some cases, a site may not have a clear
-biological significance, for instance in the case of randomly located
-survey plots. The timeframe of the study is arbitrary, but the general
-situation is one in which the time period is sufficiently long such
-that abundance changes at each site. We will call
-these time periods primary sampling periods to distinguish them from
-secondary sampling periods, which are repeated surveys within a
-primary period. In the case where no secondary sampling was conducted,
+early-successional habitat, although it could be an arbitrarily
+defined region such as a randomly located
+survey plots. The timeframe of the study is arbitrary, but in general
+it will be sufficiently long to allow for temporal variation in
+abundance. For instance, a primary sampling period could be a one
+month long breeding season, which could be surveyed once per year for
+$T$ years. If the a site is surveyed on $J$ occassions during a
+primary period, these are called secondary sampling occasions.
+In the case where no secondary sampling was conducted, i.e. $J=1$,
 let $n_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
-$i$ and primary period $t$. If $J$ secondary sampling periods were used, $n_{ijt}:
-j=1,\hdots,J$ is the count at site $i$ during secondary sampling
-occasion $j$ within year $t$. These are the observed data, which
-typically will less than the actual quantitiy of interest, abundance,
+$i$ and primary period $t$. If secondary sampling periods were used,
+an extra dimension is added so tha twe have $n_{ijt}:
+j=1,\hdots,J$. Typically, we $n_{it}$ will be lower than the actual
+quantitiy of interest, abundance,
 denoted $N_{it}$. In cases where detection probability is perfect,
-abundance is observed directly such that $\{N_{it}\}$ are the data.
+abundance is observed directly such that $n_{it} \equiv N_{it}$.
 
@@
 \section{Discussion}
+
+\begin{comment}
+  The model (1) recognizes
+  the discrete nature of count data and the underlying true abundance
+  state, (2) includes a realistic observation model with explicit
+  detection probability parameters (3) has estimable parameters,
+  including those describing the initial state at time $t=1$, and (4)
+  allows for inference about both temporal and spatial variation in
+  abundance.
+
+  In some respects, the DM model can be viewed as a specific
+  manifestation of the general state-space model formulated by
+  \citep{buckland_etal:2004}. However, their model was designed to
+  allow for inference about demographic parameters, which will
+  typically require mark-recapture data. Although this may be viewed
+  as the gold standard in many studies of wildlife populations, in
+  this paper, we focus on the analysis of count data, which is the
+  typical source of information used in state-state modeling
+  applications, and one of the most widely available forms of
+  ecological data---it is also the data required by the DM model.
+\end{comment}
 

### 2012-09-28 Dail_Mad_ext.tex
original formulation

@@
 
-As originally formulated, the DM model includes submodels for three
-conditionally related processes: (1) initial abundance, i.e. the
+As with traditional state-space models, the DM model includes the
+three conditionally related processes shown in Eq.~\ref{eq:ss2}, which
+correspond to (1) initial abundance, i.e. the
 abundance at site $i$ during the first primary period,
-denoted $[N_{i1}]$, (2) subsequent abundance, which is dependent upon
-abundance in previous years $[N_{it}|N_{it-1}]$, and (3) the
-detection process, $[n_{it}|N_{it}]$ \citep{dail_madsen:2011}.
+denoted (2) abundance at time $t$ (for $t>1$) which depends upon
+abundance at $t-1$, and (3) the
+detection process \citep{dail_madsen:2011}.
 The first two processes describe the state process---the
-variation in abundance in space and time. The third item in the list
-is the observation process, which describes the relationship between
-the true state, abundance, and the observed count data.
+variation in abundance in space and time. The third process simply
+describes the the relationship between
+abundance and the observed count data.
 
@@
 In the original formulation, initial abundance was
-modeled using either the Poisson or negative binomial distributions:
-\begin{gather}
-N_{i1} \sim \mathrm{Pois}(\Lambda_i) \nonumber \\
-\text{or} \nonumber \\
-N_{i1} \sim \mathrm{NB}(\Lambda_i, \alpha)
-\label{eq:N1}
-\end{gather}
-where
-%$N_{i1}$ is the abundance at site $i=1,\hdots,R$ during year 1 and
-$\Lambda_i$ is the expected abundance at site $i$ during year 1. The
-Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
-variance, whereas the negative binomial distribution allows the
+modeled using either the Poisson or negative binomial
+distributions. Hence, in the Poisson case, we have
+\begin{equation}
+  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
+  \label{eq:N1br}
+\end{equation}
+where $\Lambda_i$ is the expected abundance at site $i$ during year 1.
+From hereout, we will reserve bracket notation for arbitrary
+probability distributions, and for simplicity, use the equivalent notation $N_{i1}
+\sim \text{Pois}(\Lambda_i)$ when referencing specific distributions.
+%\begin{gather}
+%N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
+%\text{or} \nonumber \\
+%N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
+%\label{eq:N1}
+%\end{gather}
+The Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
+variance, whereas the negative binomial distribution, $N_{i1} \sim
+\text{NB}(\Lambda, \alpha)$, allows the
 variance to be greater than the mean with the amount of overdispersion
@@
 spatially-replicated count data. Hence, the first component of the
-model addresses issues (1) and (2) identified in the introduction.
+model addresses both the issues of spatial inference and parameter
+identifiability discussed previously.
 
@@
 in each case they modeled $N_t$ as the sum of two random variables:
-$S_t$, the number of indivduals surviving from $t-1$ and not
-emigrating; and $G_t$ the number of new individuals entering the
+$S_{it}$, the number of indivduals surviving from $t-1$ and not
+emigrating; and $G_{it}$ the number of new individuals entering the
 population. Their most general model was
@@
 where $\omega$ is the apparent survival probability and $\gamma$ is
-the recruitment rate (which can depend on $N_{it-1}$). In the absence
-of movement, $\omega$ is exactly the probability of surviving from
-year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
-in the more common scenario when immigration and emigration occur,
-$\omega$ is the probability of surviving and not emigrating, and
-$\gamma$ is the sum\hl{[I think!]} of the birth rate and the emigration rate,
-i.e. the recruitment rate. \hl{[Feels a little repetitious with the material
-before the equation, although I understand you're describing S and G
-there and omega and gamma here.  In general, these sections
-describing the DM models are clear and well written, but we could
-possibly spend less time on them b/c it's already published.
-Possibly it's better this way, though.]} \citet{dail_madsen:2011} proposed three
+the recruitment rate (which can depend on $N_{it-1}$).
+%%%% I agree with your comment, but I think we need to distinguish
+%%%% between things like actual and apparent survival probs.
+%\hl{[Feels a little repetitious with the material
+%before the equation, although I understand you're describing S and G
+%there and omega and gamma here.  In general, these sections
+%describing the DM models are clear and well written, but we could
+%possibly spend less time on them b/c it's already published.
+%Possibly it's better this way, though.]}
+\citet{dail_madsen:2011} proposed three
 models for $\gamma$: the constant model,
-$G_t \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
+$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
 $N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
-autoregressive model, $G_t \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
+autoregressive model, $G_{it} \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
 simulates geometric or density independent growth; and the
@@
 
-In cases where no movement occurs, the DM model can directly estimate
+In the absence
+of movement, $\omega$ is exactly the probability of surviving from
+year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
+in the more common scenario when immigration and emigration occur,
+$\omega$ is the probability of surviving and not emigrating, and
+$\gamma$ is the sum %\hl{[I think!]}
+of the birth rate and the immigration rate,
+i.e. the recruitment rate. In cases where no movement occurs, the DM model can directly estimate
 demographich parameters, and this mechanistic description of

### 2012-09-28 Dail_Mad_ext.tex
likelihood

@@
 the recruitment rate (which can depend on $N_{it-1}$).
-%%%% I agree with your comment, but I think we need to distinguish
-%%%% between things like actual and apparent survival probs.
-%\hl{[Feels a little repetitious with the material
-%before the equation, although I understand you're describing S and G
-%there and omega and gamma here.  In general, these sections
-%describing the DM models are clear and well written, but we could
-%possibly spend less time on them b/c it's already published.
-%Possibly it's better this way, though.]}
 \citet{dail_madsen:2011} proposed three
@@
 of the birth rate and the immigration rate,
-i.e. the recruitment rate. In cases where no movement occurs, the DM model can directly estimate
+i.e. the recruitment rate. In the former case where no movement
+occurs, the DM model can directly estimate
 demographich parameters, and this mechanistic description of
@@
 connection to specific population processes. In such cases, the model
-of dynamics shown in Eq~\ref{eq:Nt} can be replaced with standard
+of dynamics shown in Eq.~\ref{eq:Nt} can be replaced with standard
 population models as described in Sect~\ref{sec:ext}. Irrespective of
@@
 model assumes that individuals are missed due to imperfect
-detection. The simplest model for imperfect detection is
+detection. The simplest model for imperfect detection, and the one
+considered in the original formulation, is
 \begin{equation}
@@
 Surprsingly detection probability, $p$, can be estimated without
-secondary sampling occasions. However, this is only possible because
+secondary sampling occasions. However, this is only possible
 if the parameteric assumptions of the population dynamics are met
@@
 about $p$ using secondary sampling occassions or some other auxiliary
-data, as is demonstrated in section
+data, \hl{as is demonstrated in section....}
 
@@
 effects requires removing the random effects by integration or
-summation.
+summation. For Markovian models, this can be done recursively, making
+the likelihood tractible. A general expression for the likelihood of
+the DM model is:
+\begin{equation}
+  \mathcal{L}(\Lambda, \Phi, \mathbf{p} | \{n_{it}\}) =
+  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
+  [n_{i1}|N_{i1}][N_{i1}|\Lambda] \left\{ \prod_{t=T}^2 \left\{ \sum_{N_{it}=0}^{\infty}
+  [n_{it}|N_{it}][N_{it}|N_{it-1}] \right\} \right\} \right\}
+  \label{eq:like}
+\end{equation}
 
@@
 
-Here is some BUGS code: \hl{[This is code for a model we haven't
-presented yet.  Could we move these sections after model extensions?]}
+Here is some BUGS code:
+
 \begin{verbatim}
@@
   for(t in 2:nYears) {
-    mu[i,t-1] = N[i,t-1]*exp(r)
-    N[i,t] ~ dpois(mu[i,t-1])
+    S[i,t-1] ~ dbin(omega, N[i,t-1])
+    G[i,t-1] ~ dpois(gamma)
+    N[i,t] <- S[i,t-1] + G[i,t-1]
     n[i,t] ~ dbin(p, N[i,1])
@@
 speeding up processing time.
+
+\begin{verbatim}
+model {
+lambda ~ dunif(0, 5)
+r ~ dunif(-5, 5)
+p ~ dunif(0,1)
+for(i in 1:nSites) {
+  N[i,1] ~ dpois(lambda)
+  y[i,1] ~ dbin(p, N[i,1])
+  for(t in 2:nYears) {
+    mu[i,t-1] = N[i,t-1]*exp(r)
+    N[i,t] ~ dpois(mu[i,t-1])
+    n[i,t] ~ dbin(p, N[i,1])
+    }
+  }
+}
+\end{verbatim}
 

### 2012-09-28 Dail_Mad_ext.tex
Bayes

@@
 summation. For Markovian models, this can be done recursively, making
-the likelihood tractible. A general expression for the likelihood of
-the DM model is:
+the likelihood tractible, the general expression of which can be written:
 \begin{equation}
@@
 \end{equation}
+This likelihood can be maximized numerically using the \textbf{R}
+package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples are
+given in the Appendix.
 
@@
 direct probability statements to be made about a hypothesis given
-data. Second, priors are cool. We can combine data from multiple
+data. [EXPLAIN POSTERIORS AND PRIORS]
+Second, Bayesian methods offer straight-forward appraoches for
+combine data from multiple
 sources or use existing estimates of parameters as prior
-distributions. MCMC has a bad rap for being slow, but actually, in the
-case of state-space models it may be one of the few options for
-estimation, and in fact may be faster than maximizing likelihoods.
-
-Here is some BUGS code:
+distributions. In practice, simulation methods such as Markov chain
+Monte Carlo (MCMC) are used to simulate the posterior
+distributions. Although MCMC can be slow, our experience is that it
+may actually converge faster than the amount of time required to
+maximize the likelihood shown in Eq.\ref{eq:like}, especially when the
+upper bound of the summation must be high, e.g. $K>200$. Furthermore,
+MCMC may be the only possible option for estimating parameters using
+some of the extensions described below, such as allowing for
+additional random effects.
+
+Writeing custom MCMC algorthims is often tedious and foreign to
+ecologists, but software packages such as \textbf{WinBUGS} and
+\textbf{JAGS} overcome the technical difficulties by allowing users to
+specify the model using a simple symbolic descriptions. For example,
+here is \textbf{BUGS} code for the ``constant'' model described above.
 
@@
 lambda ~ dunif(0, 5)
-r ~ dunif(-5, 5)
+omega ~ dunif(0,1)
+gamma ~ dunif(0, 10)
 p ~ dunif(0,1)

### 2012-10-02 Dail_Mad_ext.tex
initial abundance

@@
 
-In the original formulation, initial abundance was
-modeled using either the Poisson or negative binomial
-distributions. Hence, in the Poisson case, we have
-\begin{equation}
-  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
-  \label{eq:N1br}
-\end{equation}
-where $\Lambda_i$ is the expected abundance at site $i$ during year 1.
-From hereout, we will reserve bracket notation for arbitrary
-probability distributions, and for simplicity, use the equivalent notation $N_{i1}
-\sim \text{Pois}(\Lambda_i)$ when referencing specific distributions.
-%\begin{gather}
-%N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
-%\text{or} \nonumber \\
-%N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
-%\label{eq:N1}
-%\end{gather}
+Recall that conventional state-space models assume that the
+distribution for the initial time period,
+$[N_1 | \bm{\theta}]$, is either the equilibrium
+distribution, or has zero variance.
+In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
+as either a Poisson or negative binomial random variable:
+\begin{gather}
+N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
+\text{or} \nonumber \\
+N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
+\label{eq:N1}
+\end{gather}
+%\begin{equation}
+%  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
+%  \label{eq:N1br}
+%\end{equation}
+where $\Lambda_i$ is the expected abundance at site $i$ during year
+1. In the Poisson case, the eqivalent expression of
+Eq.~\ref{eq:ss2a} is $[N_{i1} | \Lambda] =
+\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
+%From hereout, we will reserve bracket notation for arbitrary
+%probability distributions, and for simplicity, use the equivalent notation $N_{i1}
+%\sim \text{Pois}(\Lambda_i)$ when referencing specific distributions.
 The Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
@@
 determined by the parameter $\alpha$.
+Note that $N_{i1} \sim \text{Pois}(\Lambda)$ is equivalent to
+$[N_{i1} | ] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
 
@@
   \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
-  [n_{i1}|N_{i1}][N_{i1}|\Lambda] \left\{ \prod_{t=T}^2 \left\{ \sum_{N_{it}=0}^{\infty}
-  [n_{it}|N_{it}][N_{it}|N_{it-1}] \right\} \right\} \right\}
+  [n_{i1}|N_{i1}][N_{i1}|\Lambda] \left\{
+%  \prod_{t=T}^2 \left\{
+  \sum_{N_{i1}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
+  [n_{it}|N_{it}][N_{it}|N_{it-1}] %\right\}
+  \right\} \right\}
   \label{eq:like}

### 2012-10-02 Dail_Mad_ext.tex
transition probs

@@
 \end{gather}
-%\begin{equation}
-%  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
-%  \label{eq:N1br}
-%\end{equation}
 where $\Lambda_i$ is the expected abundance at site $i$ during year
-1. In the Poisson case, the eqivalent expression of
-Eq.~\ref{eq:ss2a} is $[N_{i1} | \Lambda] =
-\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
-%From hereout, we will reserve bracket notation for arbitrary
-%probability distributions, and for simplicity, use the equivalent notation $N_{i1}
-%\sim \text{Pois}(\Lambda_i)$ when referencing specific distributions.
+1.
 The Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
-variance, whereas the negative binomial distribution, $N_{i1} \sim
-\text{NB}(\Lambda, \alpha)$, allows the
+variance, whereas the negative binomial distribution allows the
 variance to be greater than the mean with the amount of overdispersion
 determined by the parameter $\alpha$.
-Note that $N_{i1} \sim \text{Pois}(\Lambda)$ is equivalent to
-$[N_{i1} | ] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
+In the Poisson case, the eqivalent expression of
+$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
+%$[N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
+\begin{equation}
+  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
+  \label{eq:N1br}
+\end{equation}
 
@@
 
+To make the connection between conventional state-space models and the
+DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
+Eq.~\ref{eq:ss2b}, with an expression derived from
+Eq.~\ref{eq:Nt}. This requires summing over all possible values of $S$
+and $G$, which is accomplished using the discrete convolution:
+\begin{equation}
+[N_{it}|N_{it-1}, \omega, \gamma] =
+  \sum_{S_{it-1}=0}^{\min(N_{it},N_{it-1})}
+  \binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}} \times
+  \frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}
+  \label{eq:P}
+\end{equation}
+where here we assumed no dependenced of $\gamma$ on $N_{it-1}$.
+
 In the absence
 of movement, $\omega$ is exactly the probability of surviving from
-year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
+year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate (under
+the geometric growth model). This ability to directly estimate
+demographic parameters from simple count data is one of the DM model's
+most appearling features. However,
 in the more common scenario when immigration and emigration occur,
-$\omega$ is the probability of surviving and not emigrating, and
-$\gamma$ is the sum %\hl{[I think!]}
-of the birth rate and the immigration rate,
-i.e. the recruitment rate. In the former case where no movement
-occurs, the DM model can directly estimate
-demographich parameters, and this mechanistic description of
-population change is much more appealing than classical
-population growth models such as the logistic model. However, movement
-complicates the interpretation of the the parameters, reducing the
-connection to specific population processes. In such cases, the model
-of dynamics shown in Eq.~\ref{eq:Nt} can be replaced with standard
-population models as described in Sect~\ref{sec:ext}. Irrespective of
-the model for population dynamics, it is worthwhile to note that the
-discrete distributions used in the DM model avoid the issues
-associated with the Gaussian distribution used to model process
-variation in standard state-space models.
+$\omega$ and $\gamma$ no longer can be intepretted as vital
+rates. Rather, $\omega$ is the probability of surviving and not
+emigrating, and
+$\gamma$ is the sum of the birth rate and the immigration rate,
+i.e. the recruitment rate. In such cases, we propose
+replacing Eq.~\ref{eq:Nt} with standard
+population models.
+
+\begin{comment}
+  Irrespective of the model for population dynamics, it is worthwhile
+  to note that the discrete distributions used in the DM model avoid
+  the issues associated with the Gaussian distribution used to model
+  process variation in standard state-space models.
+\end{comment}
 

### 2012-10-02 Dail_Mad_ext.tex
on to extensions

@@
 In the case where no secondary sampling was conducted, i.e. $J=1$,
-let $n_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
+let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
 $i$ and primary period $t$. If secondary sampling periods were used,
-an extra dimension is added so tha twe have $n_{ijt}:
-j=1,\hdots,J$. Typically, we $n_{it}$ will be lower than the actual
+an extra dimension is added so tha twe have $X_{ijt}:
+j=1,\hdots,J$. Typically, we $X_{it}$ will be lower than the actual
 quantitiy of interest, abundance,
-denoted $N_{it}$. In cases where detection probability is perfect,
-abundance is observed directly such that $n_{it} \equiv N_{it}$.
-
-\subsection{The Model}
+denoted $X_{it}$. In cases where detection probability is perfect,
+abundance is observed directly such that $X_{it} \equiv N_{it}$.
+
+\subsection{The Original Formulation of the Model}
 
@@
 
-Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model---the
-model for spatial and temporal variation in abundance. The observation
+Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model, i.e. the
+model for abundance dynamics. This state
+model has allowed us to replace the initial abundance distribution and
+transition models of conventional state-space models, with
+alternatives that (1) allow for modeling spatial variation in
+abundance, (2) respect the discrete nature of the counts, and (3) have
+parameters that are estimable. The one
+remaining task is to replace the observation model
+$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be more
+easily interpretted. The observation
 model assumes that individuals are missed due to imperfect
@@
 \begin{equation}
-  n_{it} \sim \mathrm{Bin}(N_{it}, p)
+  X_{it} \sim \mathrm{Bin}(N_{it}, p)
   \label{eq:p1}
 \end{equation}
-Surprsingly detection probability, $p$, can be estimated without
-secondary sampling occasions. However, this is only possible
-if the parameteric assumptions of the population dynamics are met
-exactly. In practice, it is preferable to obtain direct information
-about $p$ using secondary sampling occassions or some other auxiliary
-data, \hl{as is demonstrated in section....}
+or equivalently,
+\begin{equation}
+  [X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}} \bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
+  \label{eq:1}
+\end{equation}
+Under this model, $\bm{p}$ are interpretted as detection probability
+parameters...
+
+%%%% Save for discussion?
+%Surprsingly detection probability, $p$, can be estimated without
+%secondary sampling occasions. However, this is only possible
+%if the parameteric assumptions of the population dynamics are met
+%exactly. In practice, it is preferable to obtain direct information
+%about $p$ using secondary sampling occassions or some other auxiliary
+%data, \hl{as is demonstrated in section....}
+
+The entire model can now be written:
+\begin{gather}
+  N_{i1} \sim \text{Pois}(\Lambda) \nonumber \\
+  S_{it}|N_{it-1} \sim \mathrm{Bin}(N_{it-1}, \omega) \nonumber \\
+  G_{it}|N_{it-1}\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
+  N_{it} = S_{it}+G_{it} \nonumber \\
+  X_{it} \sim \text{Bin}(N_{it}, p) \nonumber
+  \label{eq:hm}
+\end{gather}
 
@@
 
-Maximum likelihood estimation of parameters in models with random
-effects requires removing the random effects by integration or
+Maximum likelihood estimation of parameters in models with discrete random
+effects requires removing them by
 summation. For Markovian models, this can be done recursively, making
-the likelihood tractible, the general expression of which can be written:
-\begin{equation}
-  \mathcal{L}(\Lambda, \Phi, \mathbf{p} | \{n_{it}\}) =
+the likelihood tractible. A general expression of the likelihood can be written:
+\begin{multline}
+  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
   \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
-  [n_{i1}|N_{i1}][N_{i1}|\Lambda] \left\{
+  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
 %  \prod_{t=T}^2 \left\{
-  \sum_{N_{i1}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
-  [n_{it}|N_{it}][N_{it}|N_{it-1}] %\right\}
+  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
+  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
   \right\} \right\}
   \label{eq:like}
-\end{equation}
-This likelihood can be maximized numerically using the \textbf{R}
+\end{multline}
+where $\infty$ is replaced by a finite upper bound $K$, which should
+be high enough such that the MLEs are not affected by increasint it
+further. This likelihood can be maximized numerically using the \textbf{R}
 package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples are
@@
 specify the model using a simple symbolic descriptions. For example,
-here is \textbf{BUGS} code for the ``constant'' model described above.
-
+here is \textbf{BUGS} code for the ``geometric'' model described shown
+in Eq.~\ref{eq:hm}:
 \begin{verbatim}
@@
     S[i,t-1] ~ dbin(omega, N[i,t-1])
-    G[i,t-1] ~ dpois(gamma)
+    G[i,t-1] ~ dpois(gamma*N[i,t-1])
     N[i,t] <- S[i,t-1] + G[i,t-1]

### 2012-10-02 Dail_Mad_ext.tex
hhhh

@@
 
-Preliminary DM model runs for several species tended to lead to
-estimates of survival that were unrealistically high and recruitment
-that were unrealistically low, or the reverse (compared to independent
-demographic analyses).  The DM models are able to partition changes in
-abundance to survival and recruitment in part by making strong
-distributional assumptions. % (Dail and Madsen 2011).
-When those assumptions are heavily violated the models may proportion
-population growth incorrectly into survival and recruitment, even if
-they estimates population growth accurately.
+% Preliminary DM model runs for several species tended to lead to
+% estimates of survival that were unrealistically high and recruitment
+% that were unrealistically low, or the reverse (compared to independent
+% demographic analyses).  The DM models are able to partition changes in
+% abundance to survival and recruitment in part by making strong
+% distributional assumptions. % (Dail and Madsen 2011).
+% When those assumptions are heavily violated the models may proportion
+% population growth incorrectly into survival and recruitment, even if
+% they estimates population growth accurately.
 
 Although partitioning population growth into survival and recruitment
-is informative, it is not needed for all applications.  Furthermore, a
-simpler model would have several merits: faster running time, fewer
-total model combinations, and possibly more realistic estimates.
-Therefore, we developed a version of the DM model that estimates and
-models population growth directly.  In this model, Eq.~\ref{eq:Nt} is
-simplified to:
+is ideal in terms of providing a mechanistic description of population
+dynamics, it is not always possible using simple count data,
+especially when the sites are not closed with respect to immigration
+and emigration. %Furthermore, a
+%simpler model would have several merits: faster running time, fewer
+%total model combinations, and possibly more realistic estimates.
+If the mechanistic model must be sacrificed, we suggest replacing it
+with classical population growth models that have a long history in
+ecology. The simplest of which is the exponential growth
+model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic rate of
+increase. Although demographic and environmental stochasticity in $r$
+could be modeled as functions of covariates or random effects, a
+simpler way of allow for stocasticity is
+to regard $N_{it}$ as a Poisson random variable, simplifying
+Eq.~\ref{eq:Nt} to:
 \begin{equation}
-  N_{it} \sim \text{Pois}(\exp(r)N_{it-1})
+  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
 \label{eq:exp}
 \end{equation}
-where $r$ represents the instantaneous population growth rate.  This,
+This,
 like the autoregressive version of the model, is a variant on a simple
@@
 %recruitment or population dynamics (see below).
-We also considered a zero-inflated negative binomial model of initial
-abundance, but this model did not perform well in preliminary tests.
+%We also considered a zero-inflated negative binomial model of initial
+%abundance, but this model did not perform well in preliminary tests.
 
@@
 
+\subsection{Environmental and demographic stochasticity}
+
+Even though the DM model is conceptually very simple, even in its most
+stripped down form it includes many random effects (the $N$'s) making
+it challenging to compute the likelihood or implement MCMC
+algorithms. Nonetheless, in some cases additional random effects may
+be of interest. For example, in conventional state-space models,
+demographic stochasticity is often modeled as $r_t \sim
+\text{Norm}(\mu_r, \sigma_r)$.
+
 \subsection{Random effects of observers}
 
-Differences in observers' ability to see, hear, or identify
+Variation in detection probability among observers is another source
+of random variation that may need to be accounted for applying the DM
+model. For example, differences in observers' ability to see, hear, or identify
 birds has long been recognized as a potential source of error
@@
 \begin{gather}
-n_{ijt} \sim \mathrm{Bin}(N_{it}, p_j) \nonumber \\
+X_{ijt} \sim \mathrm{Bin}(N_{it}, p_j) \nonumber \\
 \mathrm{logit}(p_j) \sim \mathrm{Normal}(\mu_p, \sigma_p)
@@
 \end{gather}
-where $n_{ijt}$ is the number of bird recorded at site $i$ by observer $j$ in
+where $X_{ijt}$ is the number of bird recorded at site $i$ by observer $j$ in
 year $t$, $p_j$ is observer-specific detection probability, $\mu_p$ is the mean
@@
 % (version 3.2.0; Plummer 2003)
-with the \textbf{R} (R
-Development Core Team 2011) package \texttt{rjags} \citep{plummer:2011}
-%(Plummer 2011)
+with the \textbf{R} package \texttt{rjags} \citep{plummer:2011, R-2012}
 interface.
 
-\section{Forecasting Future Population Size}
+\subsection{Alternative Observation Models}
+
+It would be cool to add something on this, but maybe it's asking too much
+
+\subsection{Forecasting Future Population Size}
+
+Ditto
 
@@
 for both models).
-
-\begin{figure}
-  \centering
-  \includegraphics[height=8in]{figs/exp_hists}
-  \caption{Histograms of 1000 parameter estimates for each of 27 simulation cases. The verticle lines are the data-generating values.}
-\label{fig:exp_hists}
-\end{figure}
 
@@
 
+\newpage
+
+%%% Figures
+
+\begin{figure}
+  \centering
+  \includegraphics[height=8in]{figs/exp_hists}
+  \caption{Histograms of 1000 parameter estimates for each of 27
+    simulation cases. The verticle lines are the data-generating
+    values.}
+\label{fig:exp_hists}
+\end{figure}
+
 \end{document}

### 2012-10-02 Dail_Mad_ext.tex
discussion

@@
 
-  In some respects, the DM model can be viewed as a specific
-  manifestation of the general state-space model formulated by
-  \citep{buckland_etal:2004}. However, their model was designed to
-  allow for inference about demographic parameters, which will
-  typically require mark-recapture data. Although this may be viewed
-  as the gold standard in many studies of wildlife populations, in
-  this paper, we focus on the analysis of count data, which is the
-  typical source of information used in state-state modeling
-  applications, and one of the most widely available forms of
-  ecological data---it is also the data required by the DM model.
 \end{comment}
@@
 limitations. We extended this class of models in several important
-ways, while keeping in mind the practical issues ...
+ways that should make them more suitable to typical ecological
+datasets that are characterized by sparse counts, zero-inflation, and
+observation error due to imperfect detection.
 
@@
 
+In contrast, recent work has sought to use state-space
+models for inference about population processes such as mortality and
+recruitment. To do so, methods have been developed to combine multiple
+sources of information, such as counts and mark-recapture data
+\citep{besbeas_etal:2002, buckland_etal:2004,
+  schaub_etal:2007}. Although these integrated
+population models may be viewed as the gold standard in state-space
+modeling, ecologists are not always so fortunate to have direct
+information about vital rates, especially at large spatial scales.
+Rather, count data are much more common, and
+are produced by many of the largest monitoring programs in the world,
+such as the North American Breeding Bird Survey
+\citep[BBS;][]{robbins_etal:1986}.
+
+In the context of IPMs, the DM model can be viewed as a specific
+manifestation of the general state-space model formulated by
+\citep{buckland_etal:2004}. However, their model was designed to
+allow for inference about demographic parameters, which will
+typically require mark-recapture data. Although this may be viewed
+as the gold standard in many studies of wildlife populations, in
+this paper, we focus on the analysis of count data, which is the
+typical source of information used in state-state modeling
+applications, and one of the most widely available forms of
+ecological data---it is also the data required by the DM model.
+
 In the absence of direct information about demographic parameters, and
@@
 
-The zero-inflated models for initial
-abundance and dynamics allow one to estimate and model the proportion
-of sites outside of the range of a species.  This could be especially
-useful when different factors control the range than the abundance or
-dynamics within the range.  [some examples where that is the case?]
-
+\begin{comment}
+  The zero-inflated models for initial abundance and dynamics allow
+  one to estimate and model the proportion of sites outside of the
+  range of a species.  This could be especially useful when different
+  factors control the range than the abundance or dynamics within the
+  range.  [some examples where that is the case?]
+\end{comment}
 [paragraph on random effects]
@@
 \begin{comment}
-Should we include these possible extensions as well?
-
-In contrast, recent work has sought to use state-space
-models for inference about population processes such as mortality and
-recruitment. To do so, methods have been developed to combine multiple
-sources of information, such as counts and mark-recapture data
-\citep{besbeas_etal:2002, buckland_etal:2004,
-  schaub_etal:2007}. Although these integrated
-population models may be viewed as the gold standard in state-space
-modeling, ecologists are not always so fortunate to have direct
-information about vital rates, especially at large spatial scales.
-Rather, count data are much more common, and
-are produced by many of the largest monitoring programs in the world,
-such as the North American Breeding Bird Survey
-\citep[BBS;][]{robbins_etal:1986}.
 
@@
 quasi-extinction probability.
+
+\section{Acknowledgements}
+
+We thank J. Andrew Royle and T. Scott Sillett for helpful comments
+throughout this process.
 

### 2012-10-02 Dail_Mad_ext.tex
discussion

@@
   abundance.
-
 \end{comment}
@@
 models as applied to ecological time series data, and we have
-demonstrated how open population $N$-mixture models overcome these
+demonstrated how a class of open population $N$-mixture models
+proposed by \citet{dail_madsen:2011} overcomes these
 limitations. We extended this class of models in several important
-ways that should make them more suitable to typical ecological
-datasets that are characterized by sparse counts, zero-inflation, and
-observation error due to imperfect detection.
-
-We made four important developments of the open population
-N-mixture model proposed by \citep{dail_madsen:2011} that make the
-model more applicable to long-term data such as is collected
-by the North American Breeding Bird Survey. First, we have
-reconciled the objectives of traditional state-space models
-with open population $N$-mixture models by illustrating how
-classical models of population dynamics can be embedded within
-the framework. Second, we have demonstrated methods of
-accounting for zero-inflation in the time-series. Third, we
-have illustrated how additional random effects such as
-observer-specific detection probabilities can be
-accommodated. Fourth, we presented a Bayesian analysis of the
-model, which, makes it possible to incorporate prior information when
-available, and in some cases, facilitates parameter estimation.
-
-A unique aspect of the Dail and Madsen model as originally
-devised was that it allowed for the estimation of demographic
-parameters under strict distributional assumptions and with the
-assumption of geographic closure. Clearly, estimating
+ways to accomodate common features of ecological
+datasets, including sparse counts, zero-inflation, demographic
+stochasticity, and random variation in observation error.
+%We made four important developments of the open population
+%N-mixture model proposed by \citet{dail_madsen:2011} that make the
+%model more applicable to long-term data such as is collected
+%by the North American Breeding Bird Survey.
+%First, we have
+%reconciled the objectives of traditional state-space models
+%with open population $N$-mixture models by illustrating how
+%classical models of population dynamics can be embedded within
+%the framework. Second, we have demonstrated methods of
+%accounting for zero-inflation in the time-series. Third, we
+%have illustrated how additional random effects such as
+%observer-specific detection probabilities can be
+%accommodated. Fourth, we presented a Bayesian analysis of the
+%model, which, makes it possible to incorporate prior information when
+%available, and in some cases, facilitates parameter estimation.
+
+We also demonstrated how many of the objectives of classical
+state-space models can be met using this. For example, one of the
+primary aims of our paper was to demonstrate how classical
+population growth models can be embedded in the DM model.
+In some ways, such population growth models are more phenomenological
+than mechanistic, and this runs counter to the motivation for the DM
+model. However,
+%A unique aspect of the Dail and Madsen model as originally
+%devised was that it allowed for the estimation of demographic
+%parameters under strict distributional assumptions and with the
+%assumption of geographic closure. Clearly,
+estimating
 demographic parameters from count data is an ambitious goal, and
-the required assumption will not be valid in many cases. One
-advantage of the original DM models over our extensions here is
-that they model population dynamics more mechanistically, by
-focusing on survival and recruitment. As we report here, these
-models often provide unrealistic estimates of these vital rates
-and are generally less supported in model selection. Adding
-density dependence directly to one or both vital rates may solve
-both of these problems, although this might require very good
-data. Another strategy would be to use estimates of vital rates
-from demographic studies as priors in a Bayesian analysis using
-DM models. Finally, count data could be combined with other data
-types such as capture-recapture data, which offers a means of
+the required assumption will not be valid in many cases, especially
+when immigration and emigration occur. This was supported by the
+unrealistic estimates of vital rates reported here and by the lack of
+support for the mechanistic models relative to the population growth
+models.
+%One
+%advantage of the original DM models over our extensions here is
+%that they model population dynamics more mechanistically, by
+%focusing on survival and recruitment. As we report here, these
+%models often provide unrealistic estimates of these vital rates
+%and are generally less supported in model selection.
+%Adding
+%density dependence directly to one or both vital rates may solve
+%both of these problems, although this might require very good
+%data.
+One strategy for adhereing to the explicit model would be to use
+estimates of vital rates from demographic studies as priors in a
+Bayesian analysis. Alternatively, count data could be combined with
+other data types such as capture-recapture data, which offers a means of
 making inferences about demographic processes at broad spatial
-scales. Models combining count data and demographic data are
-often referred to as integrated population models (IPM; CITE),
-and we believe that a DM model would provide an effective part
-of an IPM.
-
-In contrast, recent work has sought to use state-space
-models for inference about population processes such as mortality and
-recruitment. To do so, methods have been developed to combine multiple
-sources of information, such as counts and mark-recapture data
-\citep{besbeas_etal:2002, buckland_etal:2004,
-  schaub_etal:2007}. Although these integrated
+scales. Models that are fitted to count data and demographic data are
+often referred to as integrated population models
+\citep[][IPM]{besbeas_etal:2002, buckland_etal:2004,schaub_etal:2007}
+and we believe that a DM model would provide an framework for an
+IPM. However, we caution that when count data are affected by movement
+and vital rates, IPMs would need to model
+
+%In contrast, recent work has sought to use state-space
+%models for inference about population processes such as mortality and
+%recruitment. To do so, methods have been developed to combine multiple
+%sources of information, such as counts and mark-recapture data
+%\citep{besbeas_etal:2002, buckland_etal:2004,
+%  schaub_etal:2007}.
+
+Although these integrated
 population models may be viewed as the gold standard in state-space

### 2012-10-02 Dail_Mad_ext.tex
discussion

@@
 proposed by \citet{dail_madsen:2011} overcomes these
-limitations. We extended this class of models in several important
+limitations. We extended their model in several important
 ways to accomodate common features of ecological
@@
 
-We also demonstrated how many of the objectives of classical
-state-space models can be met using this. For example, one of the
+We also demonstrated how many of the objectives of conventional
+state-space modeling can be accomplished within this expanded
+framework. For example, one of the
 primary aims of our paper was to demonstrate how classical
-population growth models can be embedded in the DM model.
-In some ways, such population growth models are more phenomenological
+population growth models can be embedded in the DM model. Although we
+view this as an important connection to make between the two different
+classes of state-space models, we believe that
+%In some ways,
+such population growth models are more phenomenological
 than mechanistic, and this runs counter to the motivation for the DM
-model. However,
-%A unique aspect of the Dail and Madsen model as originally
-%devised was that it allowed for the estimation of demographic
-%parameters under strict distributional assumptions and with the
-%assumption of geographic closure. Clearly,
-estimating
+model. However, estimating
 demographic parameters from count data is an ambitious goal, and
@@
 models.
-%One
-%advantage of the original DM models over our extensions here is
-%that they model population dynamics more mechanistically, by
-%focusing on survival and recruitment. As we report here, these
-%models often provide unrealistic estimates of these vital rates
-%and are generally less supported in model selection.
-%Adding
-%density dependence directly to one or both vital rates may solve
-%both of these problems, although this might require very good
-%data.
-One strategy for adhereing to the explicit model would be to use
-estimates of vital rates from demographic studies as priors in a
-Bayesian analysis. Alternatively, count data could be combined with
-other data types such as capture-recapture data, which offers a means of
-making inferences about demographic processes at broad spatial
-scales. Models that are fitted to count data and demographic data are
+
+One strategy for adhereing to the original formulation of the model
+that included survival and recruitment parameters would be to use
+either prior information or additional data. For example, in a
+Bayesian analysis existing estimates of demographic parameters could
+be used as informative priors, which might make it possible to
+seperate the effects of movement versus births and deaths.
+Alternatively, ecologists could combine count data with data from, for example,
+capture-recapture data. Models that are fitted to count data and demographic data are
 often referred to as integrated population models
-\citep[][IPM]{besbeas_etal:2002, buckland_etal:2004,schaub_etal:2007}
-and we believe that a DM model would provide an framework for an
-IPM. However, we caution that when count data are affected by movement
-and vital rates, IPMs would need to model
+\citep[IPM;][]{besbeas_etal:2002, buckland_etal:2004,schaub_etal:2007}
+and we believe that a DM model could work well in this context.
+In fact, the DM model could be viewed as a specific case of the
+IPM model proposed by \citet{buckland_etal:2004}.
+However, we caution that when count data are affected by movement
+and vital rates, it would be unwise to ignore the movement processes.
 
@@
 \citep[BBS;][]{robbins_etal:1986}.
-
-In the context of IPMs, the DM model can be viewed as a specific
-manifestation of the general state-space model formulated by
-\citep{buckland_etal:2004}. However, their model was designed to
-allow for inference about demographic parameters, which will
-typically require mark-recapture data. Although this may be viewed
-as the gold standard in many studies of wildlife populations, in
-this paper, we focus on the analysis of count data, which is the
-typical source of information used in state-state modeling
-applications, and one of the most widely available forms of
-ecological data---it is also the data required by the DM model.
-
 In the absence of direct information about demographic parameters, and
-when the original assumptions of the model do not hold, we have
-offered extensions of the model to estimate derived parameters such as
-population growth rate. This has been the traditional emphasis of
-state-space models, and our approach resolves many of the factors
-limiting ...
-
-\begin{comment}
-  The zero-inflated models for initial abundance and dynamics allow
-  one to estimate and model the proportion of sites outside of the
-  range of a species.  This could be especially useful when different
-  factors control the range than the abundance or dynamics within the
-  range.  [some examples where that is the case?]
-\end{comment}
+when the original assumptions of the model do not hold, the model can
+still be used for important purposes such as predicting future
+population size and changes in species distributions.
+
 [paragraph on random effects]
 
-[paragraph on Bayesian]
-The ability to fit this class of models using freely available MCMC
-software offers non-statisticians a powerful means .
+Another important benefit of this class of models is that it can be
+readily implemented using frequentist or Bayesian methods by
+ecologists lacking skills in computer programming and advanced
+statistics. We have included an appendix with extensive code to
+simulate and fit many of the extensions discussed in this paper.
 
@@
 about population dynamics. Thus, without direct information about
-detection probability, the estimates will be determined by model based
-assumptions. Furthermore, there are multiple components of detection
-probability that should be accounted for to minimize bias and yield
-valid estimates of population size \citep{nichols_etal:2009}.
-
-Fortunately, it is easy to incorporate direct information about
+detection probability, the estimates will be determined by the model's
+parametric assumptions. Furthermore, properly modeling detection
+probability can involve account for nuances such as difference among
+individuals and probabilities of being available for detection
+\citep{nichols_etal:2009}.
+
+Fortunately, it is straight-forward to incorporate direct information about
 detection probability and we recommend that this be done whenever
@@
 multiple other options are available as well, such as removal, time of
-detection, and distance modeling ([cites]).
-%For instance, there
-%should be no difficulty extending the model to accommodate traditional
-%capture-recapture data collected during each primary period. Removal
-%or time of detection etc.. Distance-related heterogeneity in detection
-%probability is another source of variation that can bias estimators of
-%abundance.   Show some equations for the alternative observation
-%models.
+detection, and distance modeling ([cites]). These could be
+accomplished by extending the open population $N$-mixture model in
+exactly the same way as the closed population version has been
+extended \citep[e.g.,][]{royle_etal:2004}.
 

### 2012-10-02 Dail_Mad_ext.tex
discussion

@@
 immigration.
-
 In our models immigration is currently modeled as independent of
@@
 
-The modeling framework we described can be used to address many of the
-most pressing issue in ecology and conservation biology. For example,
-it is possible to test hypotheses about temporal and spatial
-population regulation. Furthermore, the Bayesian approach is very
-useful in that it can be used to combine multiple sources of data to
-develop mechanistic models of population dynamics. Thus one can test
-hypotheses about the effects of climate change on either explicit
+The modeling framework we described can be used to address many
+pressing issues in ecology and conservation biology. For example,
+%it is possible to test hypotheses about temporal and spatial
+%population regulation. Furthermore, the Bayesian approach is very
+%useful in that it can be used to combine multiple sources of data to
+%develop mechanistic models of population dynamics. Thus one can test
+it allows for inference %hypotheses
+about the effects of climate change on either explicit
 demographic parameters or in derived parameters such as population
-growth rate. Furthermore, under the Bayesian approach, population
-viability analysis is trivial because projecting populations into the
-future can be done as a component of the MCMC analysis. This allows
-for the computation of posterior distributions of parameters such as
-quasi-extinction probability.
+growth rate. Furthermore, %under the Bayesian approach, population
+in addition to assessing past climate change on population parameters,
+projecting populations under future climate scenarios is also
+possible. Thus, we expect this modeling methodology can be used to
+make quantiative assessments of species vulnerability to climate
+change and other important human-induced drivers of population dynamics.
 
@@
 We thank J. Andrew Royle and T. Scott Sillett for helpful comments
-throughout this process.
+and suggestions.
 

### 2012-10-02 Dail_Mad_ext.tex
future

@@
 
-\subsection{Alternative Observation Models}
-
-It would be cool to add something on this, but maybe it's asking too much
-
 \subsection{Forecasting Future Population Size}
 
-Ditto
+State-space models allow for inference about future population size,
+a critical component of population viability analysis as well
+as conservaiton planning. Let $N^+_{t}$ denote population size at some
+future point in time and $[N^+_t]$ denote the probability
+distribution for this random variable. If we can specify this
+distribution, we can make statements about the
+probability that $N^+_t = k$ where $k$ is any possible population size
+including zero.
+
+The classical approach to computing $[N^+_t]$ is based upon Empirical
+Bayes methods. Note that neither Empiriis not actually a Bayesian
+method, it is used to estim.
+We take the parameter estimates and plug them back in.... For example
+to obtain the posterior distribution for $[N_{i1}]$
+\begin{equation}
+  [\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1}, \hat{p}]}{\sum_{N_{i1}=0}^K [N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
+  \label{eq:3}
+\end{equation}
+Once we have estimated the posterior distribution for the first time
+period, we can then proceed to estimate the posterior distribution for
+subsequent time periods using recursion.
+
+Note that when $X_{it}$ is not observed, including for future time
+periods, the distribution $[X_{it}|N_{it}]$ is taken to be uniform
+such that it has no influence on the posterior. In other words, the
+posterior for years with no observed data is determined entirely by
+the posterior for the previous year (and the population dynamics
+parameters).
+
+If covariates of the dynamics parameters are present, the problem
+becomes much harder because, to fully account for uncertainty, we need
+a model for the future values of the covariates as well.
+
+In this case, the Bayesian approach is the most direct and easy to
+implement because population size at any future point in time is
+regarded as an unknown parameter like any other. Thus, one can simply
+compute the posterior distribution of $N^+_t$, which is called the
+posterior predictive distribution. In practical terms, this simply
+means adding ``missing data'' to the observed count data, and allowing
+the MCMC algorithm to sample the latent abundance variables
+conditional only on abundance in each previous time period (and the
+parameters governing population dynamics). An example of this is
+provided in the appendix.
 

### 2012-10-02 Dail_Mad_ext.tex
future

@@
   [\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1}, \hat{p}]}{\sum_{N_{i1}=0}^K [N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
-  \label{eq:3}
+  \label{eq:eb1}
 \end{equation}
@@
 subsequent time periods using recursion.
+\begin{equation}
+  [\hat{N}_{it}] =
+  \frac{[N_{it}|N_{it-1},\hat{\bm{\Theta}}][X_{it}|N_{it},
+    \hat{p}]}{\sum_{N_{it}=0}^K [N_{it}|N_{it-1}, \hat{\Theta}][X_{it}|N_{it},\hat{p}]}
+  \label{eq:eb2}
+\end{equation}
 
@@
 posterior for years with no observed data is determined entirely by
-the posterior for the previous year (and the population dynamics
-parameters).
+the estimated posterior from the previous year and the population dynamics
+parameters.
 
@@
 becomes much harder because, to fully account for uncertainty, we need
-a model for the future values of the covariates as well.
-
-In this case, the Bayesian approach is the most direct and easy to
-implement because population size at any future point in time is
-regarded as an unknown parameter like any other. Thus, one can simply
-compute the posterior distribution of $N^+_t$, which is called the
-posterior predictive distribution. In practical terms, this simply
-means adding ``missing data'' to the observed count data, and allowing
-the MCMC algorithm to sample the latent abundance variables
-conditional only on abundance in each previous time period (and the
-parameters governing population dynamics). An example of this is
-provided in the appendix.
+a model for the future values of the covariates as well. When the
+covariates are continuous, this requires integrating over their
+possible values, which becomes very challenging
+computationally. Indeed, simulation methods such as MCMC are the only
+viable options when numerous covariates are present, which will often
+be the case in many ecological studies.
+
+Relative to the computational challenges associated with the empirical
+Bayes methods, the Bayesian approach is much more easy to implement,
+especially since MCMC will be generally be relied upon in the first
+place. Futhermore, because population size at any future point in time is
+regarded as an unknown parameter like any other, computing the
+posterior distribution of $N^+_t$ requires no additional steps in MCMC
+algorithm, other than formally stating that the count data are ``missing''
+in future years.  An example of this is provided in the appendix.
 

### 2012-10-14 Dail_Mad_ext.tex
rough simulation results plus

@@
 
-\title{Improved state-space models for inference about spatial and
-  temporal variation in abundance from count data}
+\title{Improved state-space models for inference about spatial
+and temporal variation in abundance from count data}
 
@@
 \citet[Biometrics]{dail_madsen:2011} for spatially-replicated
-time-series data that attributes observation error to imperfect
-detection.
+time-series data that attributes observation error to imperfect detection.
 We expand this class of models to accommodate classical growth
-models (e.g. exponential, Ricker, etc $\dots$), zero-inflation, and
-random effects such as observer-specific
+models (e.g. exponential, Ricker-logistic, etc $\dots$), zero-inflation,
+and random effects such as observer-specific
 detection probabilities. We also present methods for forecasting
@@
 \textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
-Gompertz model, immigration, open population point count models,
-random observer effects, range, Ricker model, zero-inflated
+Gompertz-logistic model, immigration, open population point count models,
+random observer effects, range, Ricker-logistic model, zero-inflated
 
@@
 and population models are required to evaluate associated hypotheses.
-In applied contexts, population models are used for
-estimating extinction probabilities
+In applied contexts, population models are used for estimating extinction probabilities
 \citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
@@
 be confronted when fitting population models to data. 
-First, deterministic models of population
-dynamics are virtually always inadequate due to process variation,
-the inherent stochasticity in demographic parameters and environmental
+First, deterministic models of population dynamics are virtually always inadequate due to process
+variation, the inherent stochasticity in demographic parameters and environmental
 conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
@@
 system (e.g. population size during each year) is observed
-imperfectly. One reason for the widespread adoption of state-space
-models in ecology is that failure to account for process variation and
-observation error can bias estimators of abundance and population
-growth parameters. For instance, the strength of
+imperfectly. One reason for the widespread adoption of
+state-space models in ecology is that failure to account for process
+variation and observation error can bias estimators of abundance and
+population growth parameters. For instance, the strength of
 density dependence will be over-estimated if observation error is
@@
 The growth model may be density-dependent, as in the case of the 
-Ricker model, or it might be density-independent, such as when growth
+Ricker-logistic model, or it might be density-independent, such as when growth
 is exponential.
@@
     N_1 &= X_1 \label{eq:ss1a} \\
-    N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
-    X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \; t=1,\hdots,T \label{eq:ss1c}
+N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \;
+t=2,\hdots,T \label{eq:ss1b} \\
+X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
+t=1,\hdots,T \label{eq:ss1c}
   \end{align}
@@
 assumed to be at equilibrium such that the $N_1$ can be regarded as an
-outcom of the equilibrium distribution.
+outcome of the equilibrium distribution.
 
@@
 spatial variation in abundance is not allowed, and (4) some of the
-paramters of the model are not estimable. We briefly discuss each of
+parameters of the model are not estimable. We briefly discuss each of
 these points before describing a general approach to resolving these
@@
 Gaussian distribution implies that $X_t$ will be
-higher than $N_t$ as often as it is lower than $N_t$. It is hard
-to identify a mechanism that would cause such symmetric errors. A more likely
-form of observation error, and one that has been recognized for well
+higher than $N_t$ as often as it is lower than $N_t$. 
+It is hard to identify a mechanism that would cause such symmetric errors.
+A more likely form of observation error, and one that has been recognized for well
 over a century, results from failing to detect individuals that are
@@
 three probability distributions, which we represent
-using bracket notation (i.e., $[Y|\Omega]$ is interpretted
+using bracket notation (i.e., $[Y|\Omega]$ is interpreted
 as the probability distribution of the random variable $Y$ given the
@@
     [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
-    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
+[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
+\label{eq:ss2c}
     [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
@@
 %istribution, $[N_1|\bm{\theta}]$ be?
-And how could $\bm{\theta}$ be
-estimated given that there is only a single observation available?
-Often, these problems are simply assumed away. For
+And how could $\bm{\theta}$ be estimated given that there is only a single observation
+available? Often, these problems are simply assumed away. For
 example, sometimes researchers assume that there is no process variation in
@@
 value of $N_t$ is constant through time. Although this 
-makes the parameters $\bm{\theta}$ identifiable, assuming equlibrium
+makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
 defeats the objective of many studies of population dynamics, namely
@@
 Several extensions of state-space models have been proposed to
-overcome the limitations described
-above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
-described methods for fitting models with non-Gaussian distributions
-for the process and observation errors. Observation models with more
-intuitive interpretations, such as those that explicitly model
-detection probability, have been proposed by
-\citet{kery_etal:2009}. \citet{lele_etal:1998} and
+overcome the limitations described above. \citet{devalpine_hastings:2002} and
+\citet{kery_etal:2009} described methods for fitting models with non-Gaussian
+distributions for the process and observation errors. Observation models with
+more intuitive interpretations, such as those that explicitly model
+detection probability, have been proposed by \citet{kery_etal:2009}. \citet{lele_etal:1998} and
 \citet{kery_etal:2009} developed models allowing for inference about
@@
 also resolved the problems of non-identifiability for the parameters
-of the initial state at time $t=1$. Of these extensions, only the work by
-\citet{kery_etal:2009} addressed several of these limitations
+of the initial state at time $t=1$. Of these extensions, only
+the work by \citet{kery_etal:2009} addressed several of these limitations
 simultaneously. However, their model did not include serial
-dependence, which is a hallmark of population models. This limits the
-utility of their model for making inferences about explicit population
-processes.
-
-In this paper, we focus on the model of
-\citet[henceforth the DM model;]{dail_madsen:2011}
-that simultaneous resolves each of the abovementioned problems with
-traditional state-space models, and is designed for simple count
-data.
+dependence, which is a hallmark of population models. This
+limits the utility of their model for making inferences about explicit
+population processes.
+
+In this paper, we focus on the model of \citet[henceforth the DM model;]{dail_madsen:2011}
+that simultaneous resolves each of the aforementioned problems with
+traditional state-space models, and is designed for simple countdata.
 In the following section, we describe the DM model in its original
@@
 standard state-space models. In Section~\ref{sec:ext}, we extend the
-model to accomodate classical models of population growth and to
-handle several features common to ecological
-time-series. Specifically, we describe methods for accomodating excess
-zeros and nuisance variables such as random observer effects. Both
-frequentist and Bayesian methods of inference are discussed, and code
-for fitting models is presented in the appendices.
+model to accommodate classical models of population growth and
+to handle several features common to ecological
+time-series. Specifically, we describe methods for accommodating
+excess zeros and nuisance variables such as random observer effects.
+Both frequentist and Bayesian methods of inference are discussed, and
+code for fitting models is presented in the appendices.
 %The Bayesian method
-%is attractive in that it can accomodate prior information about
+%is attractive in that it can accommodate prior information about
 %detection parameters, which is available for many species, yet not
@@
 of the paper is to provide ecologists with flexible and accessible means of
-adressing important questions related to the variation of abundance in
+addressing important questions related to the variation of abundance in
 space and time.
@@
 %application to existing datasets such the Breeding Bird Survey.
-
 \begin{comment}
   The DM model allows for inference about spatial and temporal
-  dynamics in abundance while accounting for observation error using
-  only spatially and temporally replicated count data.  Their model is
-  an extension of the closed-population $N$-mixture model
-  \citep{royle:2004biom}, which was designed specifically to address
-  the problem of modeling spatial variation in abundance in the
+dynamics in abundance while accounting for observation error
+using only spatially and temporally replicated count data. Their model
+is an extension of the closed-population $N$-mixture model
+\citep{royle:2004biom}, which was designed specifically to
+address the problem of modeling spatial variation in abundance in the
   presence of observation error, specifically, when detection
   probability is less than one. The DM extension relaxes the
-  assumption of population closure and includes explicit parameters
-  describing population change as a first-order Markovian
+assumption of population closure and includes explicit
+parameters describing population change as a first-order Markovian
   process. [more details]
@@
 [applications / potential etc... ] The DM model is relevant to a
-  vast number of problems confronting basic and applied ecological
-  research. Answering questions about climate change etc... require
-  spatially and temporally extensive datasets such as result from the
+vast number of problems confronting basic and applied ecologicalresearch. Answering questions about climate change etc...
+require
+spatially and temporally extensive datasets such as result from
+the
   North American Breeding Bird Survey (BBS)
-  \citep{robbins_etal:1986}. However, most of the datasets of the
-  appropriate scale are collected by volunteers and using protocols
+\citep{robbins_etal:1986}. However, most of the datasets of theappropriate scale are collected by volunteers and using
+protocols
   that are not amenable to traditional approaches of modeling
   abundance and detection probability based on standard
-  capture-recapture methods. Instead they have, at best, simple count
-  data for which, when it can be interpreted as counts of unique
-  individuals, the DM model can be applied.. Yadi ya
-
-  Although the DM model was designed explicitly for these purposes,
-  the original formulation of the model makes strict distributional
+capture-recapture methods. Instead they have, at best, simple
+count
+data for which, when it can be interpreted as counts of unique  individuals, the DM model can be applied.. Yadi ya
+
+Although the DM model was designed explicitly for these
+purposes,
+the original formulation of the model makes strict
+distributional
   assumptions and does not acknowledge many of the sources of
-  variation inherent to existing ecological time-series data.  We
-  propose variations on these models that are likely to be more
-  realistic and useful in many cases.  We demonstrate the usefulness
+variation inherent to existing ecological time-series data.  We  propose variations on these models that are likely to be more
+realistic and useful in many cases. We demonstrate the
+usefulness
   of these new models with simulated and real BBS data.
 
-  In this paper, we demonstrate how the DM model resolves each of
-  these four limitations of conventional state-space models, and we
-  extend the DM model in several important ways. First, we demonstrate
+In this paper, we demonstrate how the DM model resolves each ofthese four limitations of conventional state-space models, and
+we
+extend the DM model in several important ways. First, we
+demonstrate
   how classical population models can be embedded within the DM
-  model. Then we present methods for handling excess zeros as well as
-  nuisance parameters such as random observer effects on detection
-  probability. A simulation study
+model. Then we present methods for handling excess zeros as well
+as
+nuisance parameters such as random observer effects on detection  probability. A simulation study
 \end{comment}
@@
 variation in abundance when individuals cannot be detected with
-certainty. To estimate both abundance parameters and parameters of the
-detection process, the original $N$-mixture model uses replicate
+certainty. To estimate both abundance parameters and parameters
+of the detection process, the original $N$-mixture model uses replicate
 observations at each site, which are collected during sufficiently
-short time intervals such that the population can safely be assumed to
-be closed with respect to births, deaths, and movement. The DM model
-relaxes this closure assumption and includes explicit parameters
+short time intervals such that the population can safely be
+assumed to be closed with respect to births, deaths, and movement. The DM
+model relaxes this closure assumption and includes explicit parameters
 describing population change over time.
@@
 
-The DM model requires count data collected at $R$ sites, each of which
-is surveyed on $T$ primary sampling periods. A site is ideally a
-well-defined region of a study area, such as a wetland or a patch of
-early-successional habitat, although it could be an arbitrarily
-defined region such as a randomly located
-survey plots. The timeframe of the study is arbitrary, but in general
+The DM model requires count data collected at $R$ sites, each of
+which is surveyed on $T$ primary sampling periods. A site is ideally a
+well-defined region of a study area, such as a wetland or a
+patch of early-successional habitat, although it could be an arbitrarily
+defined region such as a randomly located survey plots. The time
+frame of the study is arbitrary, but in general
 it will be sufficiently long to allow for temporal variation in
-abundance. For instance, a primary sampling period could be a one
-month long breeding season, which could be surveyed once per year for
-$T$ years. If the a site is surveyed on $J$ occassions during a
+abundance. For instance, a primary sampling period could be a
+one month long breeding season, which could be surveyed once per
+year for $T$ years. If the a site is surveyed on $J$ occasions during a
 primary period, these are called secondary sampling occasions.
-In the case where no secondary sampling was conducted, i.e. $J=1$,
-let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
-$i$ and primary period $t$. If secondary sampling periods were used,
-an extra dimension is added so tha twe have $X_{ijt}:
-j=1,\hdots,J$. Typically, we $X_{it}$ will be lower than the actual
-quantitiy of interest, abundance,
-denoted $X_{it}$. In cases where detection probability is perfect,
+In the case where no secondary sampling was conducted, i.e.
+$J=1$, let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
+at site $i$ and primary period $t$. If secondary sampling periods were
+used, an extra dimension is added so that we have $X_{ijt}:
+j=1,\hdots,J$. Typically, we $X_{it}$ will be lower than the
+actual quantity of interest, abundance,
+denoted $N_{it}$. In cases where detection probability is perfect,
 abundance is observed directly such that $X_{it} \equiv N_{it}$.
@@
 As with traditional state-space models, the DM model includes the
-three conditionally related processes shown in Eq.~\ref{eq:ss2}, which
-correspond to (1) initial abundance, i.e. the
+three conditionally related processes shown in Eq.~\ref{eq:ss2},
+which correspond to (1) initial abundance, i.e. the
 abundance at site $i$ during the first primary period,
@@
 The first two processes describe the state process---the
-variation in abundance in space and time. The third process simply
-describes the the relationship between
+variation in abundance in space and time. The third process
+simply describes the the relationship between
 abundance and the observed count data.
@@
 \end{gather}
-where $\Lambda_i$ is the expected abundance at site $i$ during year
-1.
-The Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
-variance, whereas the negative binomial distribution allows the
-variance to be greater than the mean with the amount of overdispersion
-determined by the parameter $\alpha$.
-In the Poisson case, the eqivalent expression of
+where $\Lambda_i$ is the expected abundance at site $i$ during
+year 1.
+The Poisson distribution assumes that the mean of $N_{i1}$ is
+equal to its variance, whereas the negative binomial distribution allows the
+variance to be greater than the mean with the amount of
+overdispersion determined by the parameter $\alpha$.
+In the Poisson case, the equivalent expression of
 $[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
-%$[N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
+%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
 \begin{equation}
-  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
+[N_{i1} | \Lambda] =
+\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
   \label{eq:N1br}
@@
 Regardless of the specified distribution, the model for initial
-abundance has two distinguising features. First, it provides a
+abundance has two distinguishing features. First, it provides a
 mechanism for characterizing spatial variation in abundance. For
@@
 covariate ($x_i$) on abundance using a log-linear
-model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 + \beta^{\Lambda}_1
-x_{i}$. The second important point is that the spatial replicates resolve the
-problem of parameter non-identifiability that are common to standard
-state-space models because, as demonstrated by \citet{royle:2004biom},
+model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 +
+\beta^{\Lambda}_1
+x_{i}$. The second important point is that the spatial
+replicates resolve the
+problem of parameter non-identifiability that are common to
+standard state-space models because, as demonstrated by
+\citet{royle:2004biom},
 process variation and observation error can be estimated from
-spatially-replicated count data. Hence, the first component of the
-model addresses both the issues of spatial inference and parameter
-identifiability discussed previously.
+spatially-replicated count data. Hence, the first component of
+the model addresses both the issues of spatial inference and
+parameter identifiability discussed previously.
 
@@
 abundance in time $t-1$, i.e. abundance at each site evolves as a
-first order Markovian process, although
-higher order processes are also possible.  \citet{dail_madsen:2011}
-considered several models to describe the temporal dynamics; however,
-in each case they modeled $N_t$ as the sum of two random variables:
+first order Markovian process, although higher order processes are also possible.
+\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
+however, in each case they modeled $N_t$ as the sum of two random variables:
 $S_{it}$, the number of indivduals surviving from $t-1$ and not
-emigrating; and $G_{it}$ the number of new individuals entering the
-population. Their most general model was
+emigrating; and $G_{it}$ the number of new individuals entering
+the population. Their most general model was
 \begin{equation}
@@
 \end{equation}
-where $\omega$ is the apparent survival probability and $\gamma$ is
-the recruitment rate (which can depend on $N_{it-1}$).
+where $\omega$ is the apparent survival probability and $\gamma$
+is the recruitment rate (which can depend on $N_{it-1}$).
 \citet{dail_madsen:2011} proposed three
 models for $\gamma$: the constant model,
-$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
-$N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
-autoregressive model, $G_{it} \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
+$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
+depend on $N_{it-1}$, and which simulates a ``propagule rain'' of new
+individuals; the autoregressive model, $G_{it} \sim
+\mathrm{Pois}(\gamma(N_{it-1}))$, which
 simulates geometric or density independent growth; and the
@@
 expected abundance constant over time. As before, covariates of
-$\omega$ and $\gamma$ can be easily accomodated, for example using
-logit- and log-linear models respectively.
-
-To make the connection between conventional state-space models and the
-DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
+$\omega$ and $\gamma$ can be easily accommodated, for example
+using logit- and log-linear models respectively.
+
+To make the connection between conventional state-space models
+and the DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
 Eq.~\ref{eq:ss2b}, with an expression derived from
-Eq.~\ref{eq:Nt}. This requires summing over all possible values of $S$
-and $G$, which is accomplished using the discrete convolution:
+Eq.~\ref{eq:Nt}. This requires summing over all possible values
+of $S$ and $G$, which is accomplished using the discrete convolution:
 \begin{equation}
@@
   \sum_{S_{it-1}=0}^{\min(N_{it},N_{it-1})}
-  \binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}} \times
-  \frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}
-  \label{eq:P}
+\binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}}
+\times
+\frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}  \label{eq:P}
 \end{equation}
-where here we assumed no dependenced of $\gamma$ on $N_{it-1}$.
-
-In the absence
-of movement, $\omega$ is exactly the probability of surviving from
-year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate (under
-the geometric growth model). This ability to directly estimate
-demographic parameters from simple count data is one of the DM model's
-most appearling features. However,
-in the more common scenario when immigration and emigration occur,
-$\omega$ and $\gamma$ no longer can be intepretted as vital
+where here we assumed no dependence of $\gamma$ on $N_{it-1}$.
+
+In the absence of movement, $\omega$ is exactly the probability of surviving
+from year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate
+(under the geometric growth model). This ability to directly estimate
+demographic parameters from simple count data is one of the DM
+model's most appearling features. However,
+in the more common scenario when immigration and emigration
+occur, $\omega$ and $\gamma$ no longer can be interpreted as vital
 rates. Rather, $\omega$ is the probability of surviving and not
-emigrating, and
-$\gamma$ is the sum of the birth rate and the immigration rate,
+emigrating, and $\gamma$ is the sum of the birth rate and the immigration rate,
 i.e. the recruitment rate. In such cases, we propose
-replacing Eq.~\ref{eq:Nt} with standard
-population models.
+replacing Eq.~\ref{eq:Nt} with standard population models.
 
 \begin{comment}
-  Irrespective of the model for population dynamics, it is worthwhile
-  to note that the discrete distributions used in the DM model avoid
-  the issues associated with the Gaussian distribution used to model
+Irrespective of the model for population dynamics, it is
+worthwhile
+to note that the discrete distributions used in the DM model
+avoid
+the issues associated with the Gaussian distribution used to
+model
   process variation in standard state-space models.
@@
 
-Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model, i.e. the
-model for abundance dynamics. This state
-model has allowed us to replace the initial abundance distribution and
-transition models of conventional state-space models, with
+Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
+i.e. the model for abundance dynamics. This state
+model has allowed us to replace the initial abundance
+distribution and transition models of conventional state-space models, with
 alternatives that (1) allow for modeling spatial variation in
-abundance, (2) respect the discrete nature of the counts, and (3) have
-parameters that are estimable. The one
+abundance, (2) respect the discrete nature of the counts, and
+(3) have parameters that are estimable. The one
 remaining task is to replace the observation model
-$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be more
-easily interpretted. The observation
+$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be
+more easily interpreted. The observation
 model assumes that individuals are missed due to imperfect
-detection. The simplest model for imperfect detection, and the one
-considered in the original formulation, is
+detection. The simplest model for imperfect detection, and the
+one considered in the original formulation, is
 \begin{equation}
@@
 \begin{equation}
-  [X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}} \bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
+[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
+\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
   \label{eq:1}
 \end{equation}
-Under this model, $\bm{p}$ are interpretted as detection probability
+Under this model, $\bm{p}$ are interpretted as detection
+probability
 parameters...
@@
 %%%% Save for discussion?
-%Surprsingly detection probability, $p$, can be estimated without
+%Surprisingly detection probability, $p$, can be estimated without
 %secondary sampling occasions. However, this is only possible
-%if the parameteric assumptions of the population dynamics are met
+%if the parametric assumptions of the population dynamics are met
 %exactly. In practice, it is preferable to obtain direct information
-%about $p$ using secondary sampling occassions or some other auxiliary
+%about $p$ using secondary sampling occasions or some other auxiliary
 %data, \hl{as is demonstrated in section....}
@@
 
-Maximum likelihood estimation of parameters in models with discrete random
-effects requires removing them by
-summation. For Markovian models, this can be done recursively, making
-the likelihood tractible. A general expression of the likelihood can be written:
+Maximum likelihood estimation of parameters in models with
+discrete random effects requires removing them by
+summation. For Markovian models, this can be done recursively,
+making the likelihood tractable. A general expression of the likelihood
+can be written:
 \begin{multline}
@@
 \end{multline}
-where $\infty$ is replaced by a finite upper bound $K$, which should
-be high enough such that the MLEs are not affected by increasint it
-further. This likelihood can be maximized numerically using the \textbf{R}
-package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples are
-given in the Appendix.
+where $\infty$ is replaced by a finite upper bound $K$, which
+%%% Can we rename the upper bound?  We're also using K
+%%% for equilibrium abundance.
+should be high enough such that the MLEs are not affected by increasing
+it further. This likelihood can be maximized numerically using the
+\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
+are given in the Appendix.
 
@@
 
-Bayesian inference has several appealing features. First, it allows
-direct probability statements to be made about a hypothesis given
-data. [EXPLAIN POSTERIORS AND PRIORS]
-Second, Bayesian methods offer straight-forward appraoches for
-combine data from multiple
-sources or use existing estimates of parameters as prior
-distributions. In practice, simulation methods such as Markov chain
-Monte Carlo (MCMC) are used to simulate the posterior
+Bayesian inference has several appealing features. First, it
+allows direct probability statements to be made about a hypothesis
+given data. [EXPLAIN POSTERIORS AND PRIORS]
+Second, Bayesian methods offer straight-forward approaches for
+combine data from multiple sources or use existing estimates of parameters as prior
+distributions. In practice, simulation methods such as Markov
+chain Monte Carlo (MCMC) are used to simulate the posterior
 distributions. Although MCMC can be slow, our experience is that it
 may actually converge faster than the amount of time required to
-maximize the likelihood shown in Eq.\ref{eq:like}, especially when the
-upper bound of the summation must be high, e.g. $K>200$. Furthermore,
-MCMC may be the only possible option for estimating parameters using
-some of the extensions described below, such as allowing for
+maximize the likelihood shown in Eq.\ref{eq:like}, especially
+when the upper bound of the summation must be high, e.g. $K>200$.
+Furthermore, MCMC may be the only possible option for estimating parameters
+using some of the extensions described below, such as allowing for
 additional random effects.
@@
 ecologists, but software packages such as \textbf{WinBUGS} and
-\textbf{JAGS} overcome the technical difficulties by allowing users to
-specify the model using a simple symbolic descriptions. For example,
-here is \textbf{BUGS} code for the ``geometric'' model described shown
+\textbf{JAGS} overcome the technical difficulties by allowing
+users to specify the model using a simple symbolic descriptions. For
+example here is \textbf{BUGS} code for the ``geometric'' model described
+shown
 in Eq.~\ref{eq:hm}:
@@
 ecology. The simplest of which is the exponential growth
-model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic rate of
-increase. Although demographic and environmental stochasticity in $r$
-could be modeled as functions of covariates or random effects, a
-simpler way of allow for stocasticity is
-to regard $N_{it}$ as a Poisson random variable, simplifying
+model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
+rate of increase. Although demographic and environmental stochasticity
+in $r$ could be modeled as functions of covariates or random effects, a
+simpler way of allow for stochasticity is to regard $N_{it}$ as a Poisson random variable, simplifying
 Eq.~\ref{eq:Nt} to:
@@
 \end{equation}
-where $K$ is the stable equilibrium of the population and $r$ is the
-instantaneous population growth rate at low population densities, and
-both parameters are constrained to be positive.  This is a stochastic version
-of the Ricker-logistic population growth model \citep{ricker:1954}.  We also
+where $K$ is the stable equilibrium of the population and $r$ is
+the
+instantaneous population growth rate at low population
+densities, and
+both parameters are constrained to be positive. This is a
+stochastic version
+of the Ricker-logistic population growth model
+\citep{ricker:1954}. We also
 implemented a modified Gompertz-logistic density-dependent
@@
 \begin{equation}
-  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1)))
+N_{it} \sim
+\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1)))
 \label{eq:gomp}
 \end{equation}
-Here the interpretations of $r$ and $K$ are similar to in the Ricker model.
+Here the interpretations of $r$ and $K$ are similar to in the
+Ricker-logistic model.
 
@@
 
+%% Is this here on purpose?  We haven't introduced it, and I was referring to 
+%% the speed-up these models provide in the maximum-likelihood framework
+%% (even the classic DM models don't require the convolution in the Bayesian framework, right?).
 \begin{verbatim}
@@
 
-The autoregressive, population growth, Ricker, and Gompertz
+The autoregressive, exponential, Ricker-logistic, and Gompertz-logistic
 versions of the DM models all share a common feature (or bug):
@@
 include both internal and external (immigration) contributions
-to population growth.  The population growth plus immigration
+to population growth.  The exponential plus immigration
 model is:
@@
 \end{equation}
-where $\iota$ represents the immigration rate, and is constrained to be
-positive (this is equivalent to separate Poisson processes for growth and
-immigration).  This model is close to
-the constant DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead of $\omega$
-and $\iota$ instead of $\gamma$, except that the first
-process is Poisson distributed instead of binomial. The Ricker and
-Gompertz models can be extended to allow for immigration in the same
-way.
+where $\iota$ represents the immigration rate, and is
+constrained to be
+positive (this is equivalent to separate Poisson processes for
+growth and immigration).  This model is close to
+the constant DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
+of $\omega$ and $\iota$ instead of $\gamma$, except that the first
+process is Poisson distributed instead of binomial. The autoregressive, Ricker-logistic,
+and Gompertz-logistic models can be extended to allow for immigration in the
+same way.
 \begin{comment}
-  Similarly, the Ricker + immigration model can be represented as: the
-  Gompertz + immigration model as: (13) and the autoregressive +
-  immigration model as: (14)
+Similarly, the Ricker-logistic + immigration model can be represented as:
+the
+Gompertz-logistic + immigration model as: (13) and the autoregressive +  immigration model as: (14)
 \end{comment}
@@
 \begin{comment}
-  These immigration models are based on the assumption that number of
-  immigrant is not dependent on local density; however, extending
-  these models in this way is conceptually straight-forward, a Other
+These immigration models are based on the assumption that number
+of
+immigrant is not dependent on local density; however, extendingthese models in this way is conceptually straight-forward, a
+Other
   versions of these models with different assumptions are
-  possible. The Ricker + immigration and Gompertz + immigration models
-  can be justified by the further assumption that immigrants arrive as
+possible. The Ricker-logistic + immigration and Gompertz-logistic + immigration
+models
+can be justified by the further assumption that immigrants
+arrive as
   adults immediately before the point-count survey (so that
-  density-dependent processes do not have a chance to act on the
-  numbers of immigrants; Otto and Day 2007 p. 161).  Deterministic
-  versions of these models do reach stable equilibriums, but the
-  equilibriums are not at K and cannot be solved for analytically
-  (Otto and Day 2007).
+density-dependent processes do not have a chance to act on thenumbers of immigrants; Otto and Day 2007 p.
+161).
+Deterministicversions of these models do reach stable equilibriums, but theequilibriums are not at K and cannot be solved for analytically  (Otto and Day 2007).
 \end{comment}
@@
 We have implemented all preceding models in
-a maximum likelihood framework by extending the \texttt{unmarked} package
+a maximum likelihood framework by extending the
+\texttt{unmarked} package
 \citep{fiske_chandler:2011} in \textbf{R} \citep{R-2012}.
@@
 
-\citet{dail_madsen:2011} suggested two distributions for modeling
-initial abundance: Poisson and negative binomial.  We have extended
+\citet{dail_madsen:2011} suggested two distributions for
+modeling
+initial abundance: Poisson and negative binomial. We have
+extended
 their model to include another distribution, the zero-inflated
-Poisson.  This could be useful when, for example, one is modeling the
-abundance of several bird species with the same set of BBS routespoint
-count surveys for all species, but this set includes routes sites
-outside the range of some of the bird species.  The distribution of
+Poisson. This could be useful when, for example, one is modeling
+the
+abundance of several bird species with the same set of BBS
+routespoint
+count surveys for all species, but this set includes routes
+sites
+outside the range of some of the bird species. The distribution
+of
 initial abundances can be represented as:
@@
 species was at a route but not detected; the route was within the
-species' range but there were no birds at that site in that year; and
-the route was outside the species' range.  Furthermore, detection,
-abundance, and zero-inflation can be modeled separately as functions
-of different (or the same) covariates.  For example, detection of
-species x at site y in year z might depend on wind speed, abundance on
-forest type and weather, and zero-inflation upon elevation and
+species' range but there were no birds at that site in that
+year; and the route was outside the species' range. Furthermore,
+detection, abundance, and zero-inflation can be modeled separately as
+functions of different (or the same) covariates. For example, detection of
+species x at site y in year z might depend on wind speed,
+abundance on forest type and weather, and zero-inflation upon elevation and
 climate.  This approach combines elements of occupancy modeling
@@
 The zero-inflated Poisson distribution can be applied to not only
-initial abundance but also to recruitment and population growth terms.
-For example, the recruitment term of the constant DM model
+initial abundance but also to recruitment and population growth
+terms. For example, the recruitment term of the constant DM model
 (Eq.~\ref{eq:Nt}) can be modified as follows:
@@
 0 &\; \text{with probability} \; \psi \\
-\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)
-\end{aligned} \right.
+\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
 \label{eq:ZIPts}
@@
 \begin{comment}
-When both initial abundance and dynamics are modeled as zero-inflated,
-$\psi$ can be modeled as time-varying or not (in either case, $\psi$ can also be
-modeled as varying over space).  In the latter case, it makes sense
-for the model to "remember" which sites are inside the range of the
-species between the initial abundance stage and each time step of the
+When both initial abundance and dynamics are modeled as
+zero-inflated,
+$\psi$ can be modeled as time-varying or not (in either case,
+$\psi$ can also be
+modeled as varying over space). In the latter case, it makes
+sense
+for the model to "remember" which sites are inside the range of
+the
+species between the initial abundance stage and each time step
+of the
 dynamics stage (see code sample?).  Adding zero-inflation to the
-dynamics of the autoregressive, trend, Ricker, or Gompertz models
-(without immigration) should have no effect when initial abundance is
-also zero-inflated unless $\psi$ is time-varying (because these dynamic
+dynamics of the autoregressive, exponential, Ricker-logistic, or Gompertz-logistic
+models
+(without immigration) should have no effect when initial
+abundance is
+also zero-inflated unless $\psi$ is time-varying (because these
+dynamic
 models cannot recover from zero abundance).
@@
 
-Even though the DM model is conceptually very simple, even in its most
-stripped down form it includes many random effects (the $N$'s) making
-it challenging to compute the likelihood or implement MCMC
-algorithms. Nonetheless, in some cases additional random effects may
-be of interest. For example, in conventional state-space models,
+Even though the DM model is conceptually very simple, even in
+its most stripped down form it includes many random effects (the $N$'s)
+making it challenging to compute the likelihood or implement MCMC
+algorithms. Nonetheless, in some cases additional random effects
+may be of interest. For example, in conventional state-space models,
 demographic stochasticity is often modeled as $r_t \sim
@@
 
-Variation in detection probability among observers is another source
-of random variation that may need to be accounted for applying the DM
-model. For example, differences in observers' ability to see, hear, or identify
-birds has long been recognized as a potential source of error
+Variation in detection probability among observers is another
+source of random variation that may need to be accounted for applying
+the DM model. For example, differences in observers' ability to see,
+hear, or identify birds has long been recognized as a potential source of error
 in avian point count surveys such as the BBS 
 \citep{robbins_etal:1986,diefenbach_etal:2003,sauer_etal:1994auk,alldredge_etal:2007auk,campbell_francis:2011}.
-Estimating a separate detection probability for each observer can be
-difficult and reduces one's ability to estimate the quantities
+Estimating a separate detection probability for each observer
+can be difficult and reduces one's ability to estimate the quantities
 of interest.  This problem is compounded by the fact that
@@
 
-To include random observer effects in DM models and these extensions,
-Eq.~\ref{eq:p1} can be modified to:
+To include random observer effects in DM models and these
+extensions, Eq.~\ref{eq:p1} can be modified to:
 \begin{gather}
@@
 \end{gather}
-where $X_{ijt}$ is the number of bird recorded at site $i$ by observer $j$ in
-year $t$, $p_j$ is observer-specific detection probability, $\mu_p$ is the mean
-detection probability (on the logit scale), and $\sigma_p$ is the standard
-deviation of the random observer effects (also on the logit scale).
-We have implemented random observer effects in the Bayesian framework
-using program \textbf{JAGS} \citep[version 3.2.0]{plummer:2003}
+where $X_{ijt}$ is the number of bird recorded at site $i$ by
+observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
+$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
+the standard deviation of the random observer effects (also on the logit scale). 
+We have implemented random observer effects in the Bayesian
+framework using program \textbf{JAGS} \citep[version 3.2.0]{plummer:2003}
 % (version 3.2.0; Plummer 2003)
-with the \textbf{R} package \texttt{rjags} \citep{plummer:2011, R-2012}
-interface.
+with the \textbf{R} package \texttt{rjags} \citep{plummer:2011,R-2012} interface.
 
@@
 a critical component of population viability analysis as well
-as conservaiton planning. Let $N^+_{t}$ denote population size at some
+as conservation planning. Let $N^+_{t}$ denote population size at some
 future point in time and $[N^+_t]$ denote the probability
@@
 distribution, we can make statements about the
-probability that $N^+_t = k$ where $k$ is any possible population size
-including zero.
-
-The classical approach to computing $[N^+_t]$ is based upon Empirical
-Bayes methods. Note that neither Empiriis not actually a Bayesian
-method, it is used to estim.
-We take the parameter estimates and plug them back in.... For example
-to obtain the posterior distribution for $[N_{i1}]$
+probability that $N^+_t = k$ where $k$ is any possible
+population size including zero.
+
+The classical approach to computing $[N^+_t]$ is based upon
+Empirical Bayes methods. Note that neither Empiriis not actually a
+Bayesian method, it is used to estim.
+We take the parameter estimates and plug them back in.... For
+example to obtain the posterior distribution for $[N_{i1}]$
 \begin{equation}
-  [\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1}, \hat{p}]}{\sum_{N_{i1}=0}^K [N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
+[\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},
+\hat{p}]}{\sum_{N_{i1}=0}^K
+[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
   \label{eq:eb1}
 \end{equation}
-Once we have estimated the posterior distribution for the first time
-period, we can then proceed to estimate the posterior distribution for
-subsequent time periods using recursion.
+Once we have estimated the posterior distribution for the first
+time period, we can then proceed to estimate the posterior
+distribution for subsequent time periods using recursion.
 \begin{equation}
@@
   \frac{[N_{it}|N_{it-1},\hat{\bm{\Theta}}][X_{it}|N_{it},
-    \hat{p}]}{\sum_{N_{it}=0}^K [N_{it}|N_{it-1}, \hat{\Theta}][X_{it}|N_{it},\hat{p}]}
+\hat{p}]}{\sum_{N_{it}=0}^K [N_{it}|N_{it-1},
+\hat{\Theta}][X_{it}|N_{it},\hat{p}]}
   \label{eq:eb2}
@@
 
-Note that when $X_{it}$ is not observed, including for future time
-periods, the distribution $[X_{it}|N_{it}]$ is taken to be uniform
-such that it has no influence on the posterior. In other words, the
-posterior for years with no observed data is determined entirely by
-the estimated posterior from the previous year and the population dynamics
-parameters.
+Note that when $X_{it}$ is not observed, including for future
+time periods, the distribution $[X_{it}|N_{it}]$ is taken to be
+uniform such that it has no influence on the posterior. In other words,
+the posterior for years with no observed data is determined entirely
+by the estimated posterior from the previous year and the
+population dynamics parameters.
 
 If covariates of the dynamics parameters are present, the problem
-becomes much harder because, to fully account for uncertainty, we need
-a model for the future values of the covariates as well. When the
-covariates are continuous, this requires integrating over their
+becomes much harder because, to fully account for uncertainty,
+we need a model for the future values of the covariates as well. When
+the covariates are continuous, this requires integrating over their
 possible values, which becomes very challenging
-computationally. Indeed, simulation methods such as MCMC are the only
-viable options when numerous covariates are present, which will often
-be the case in many ecological studies.
-
-Relative to the computational challenges associated with the empirical
-Bayes methods, the Bayesian approach is much more easy to implement,
-especially since MCMC will be generally be relied upon in the first
-place. Futhermore, because population size at any future point in time is
-regarded as an unknown parameter like any other, computing the
-posterior distribution of $N^+_t$ requires no additional steps in MCMC
-algorithm, other than formally stating that the count data are ``missing''
-in future years.  An example of this is provided in the appendix.
+computationally. Indeed, simulation methods such as MCMC are the
+only viable options when numerous covariates are present, which will
+often be the case in many ecological studies.
+
+Relative to the computational challenges associated with the
+empirical Bayes methods, the Bayesian approach is much more easy to
+implement, especially since MCMC will be generally be relied upon in the
+first place. Furthermore, because population size at any future point
+in time is regarded as an unknown parameter like any other, computing the
+posterior distribution of $N^+_t$ requires no additional steps
+in MCMC algorithm, other than formally stating that the count data are
+``missing'' in future years. An example of this is provided in the appendix.
 
@@
 assumed dynamics were exponential.  We ran 1000 simulations for
-each combination of low, medium, and high $\Lambda \in \{1,5,10\}$, $r
-\in \{-0.005, 0, 0.005\}$, and
+each combination of low, medium, and high $\Lambda \in
+\{1,5,10\}$, $r \in \{-0.005, 0, 0.005\}$, and
 $p \in \{0.05, 0.25, 0.5\}$. % (Table 1; 27 total combinations).
-
-Our second series of simulations changed dynamics to the Ricker model.
-We used an initial abundance of 10 and a detection probability of 0.25,
-and simulated low, medium, and high values of equilibrium abundance
-and maximum growth rate, each with 1000 simulations (Table 1;
-9 total combinations).  Our third series of simulations was based on the
-Ricker + immigration dynamics model; here we fixed all parameters the
-same as the Ricker model (with r = 0.05 and K = 10) and simulated
-low, medium, and high values of immigration rate with 1000 simulation
-each (Table 1).
+Our second series of simulations changed dynamics to the Ricker-logistic
+model. We used an initial abundance of 10 and a detection probability
+of 0.25, and simulated low, medium, and high values of equilibrium
+abundance and maximum growth rate, each with 1000 simulations (Table 1;
+9 total combinations). Our third series of simulations was based
+on the Ricker-logistic + immigration dynamics model; here we fixed all
+parameters the same as the Ricker-logistic model (with r = 0.05 and K = 10) and
+simulated low, medium, and high values of immigration rate with 1000
+simulation each (Table 1).
 
@@
 initial abundance (Poisson) and dynamics models as were
-simulated, implemented in the unmarked library in \textbf{R}.  When run
-in a maximum likelihood framework, these models require a
+simulated, implemented in the unmarked library in \textbf{R}.
+When run in a maximum likelihood framework, these models require a
 maximum abundance to integrate over
@@
 mean squared error, and coverage (percentage of 95\% confidence
-intervals for parameters that
-overlap the true values).
+intervals for parameters that overlap the true values).
 
@@
 Ideas:
-I.	Test importance of random effects modeling, and other differences between JAGS and unmarked
+I. Test importance of random effects modeling, and other
+differences between JAGS and unmarked
 II.	Model with low p values, or a range of p values
-III.	Ability to distinguish correct initial abundance and dynamics models
+III. Ability to distinguish correct initial abundance and
+dynamics models
 IV.	When can you detect DD if it exists (power)?
 V.	Building on the last two: what if your model set doesn't
-include the correct DD model?  When will the wrong DD model be favored
+include the correct DD model? When will the wrong DD model be
+favored
 over the wrong DID model?
@@
 
-We applied these models to North American Breeding Bird Survey (BBS)
-data from 1966-2010 for two species in the bordering US states
-Maryland and Virginia.  For our focal species, we selected ovenbirds
-(\textit{Seiurus aurocapilla}), an abundant and widespread forest breeding
-migrant with a stable or increasing trend in the region (CITE) and
-golden-winged warblers (\textit{Vermivora chrysoptera}), which only breeds in
-the western parts of these states and has been declining in the region
-(CITE).  %MORE INFORMATION ABOUT THE SPECIES?
+We applied these models to North American Breeding Bird Survey
+(BBS) data from 1966-2010 for two species in the bordering US states
+Maryland and Virginia. For our focal species, we selected
+ovenbirds (\textit{Seiurus aurocapilla}), an abundant and widespread
+forest breeding migrant with a stable or increasing trend in the region (CITE)
+and golden-winged warblers (\textit{Vermivora chrysoptera}), which
+only breeds in the western parts of these states and has been declining in the
+region (CITE).  %MORE INFORMATION ABOUT THE SPECIES?
 
 %GENERAL ABOUT BBS DATA.
-The BBS is an annual roadside survey implemented by trained observers
-in the United States and Canada. An observer conducts 50 3-minute point counts,
-with 400 m radiuses and 0.8 km apart from each other along a 39.4 km route.
-We only used data marked as acceptable for use in the annual BBS analysis
-\citep{sauer_etal:1994auk}.  We summed the total birds of each
-species seen on each route and year and used the routes (rather than
-the individual stops) as our sites.
-Strong winds can interfere with point count observers' ability to hear
-birds \citep{simons_etal:2007}; we tested the effects of wind speed
-on detection probability.  BBS volunteers record wind conditions at
-the beginning and end of each route on the Beaufort Scale
-\citep[start and end wind 0-9]{robbins_etal:1986}.
-%(start and end wind; Robbins et al. 1986)(start and end wind; 0-9; cite).
-When
-start or end wind was not recorded we imputed those values with the
-start or end mean.  We took the mean of start and end wind for each
-route and put this average wind scale value into four categories: 0 $\leq$
-wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind $\geq$ 3 (maximum of 3.5).
+The BBS is an annual roadside survey implemented by trained
+observers in the United States and Canada. An observer conducts 50
+3-minute point counts, with 400 m radiuses and 0.8 km apart from each other along a
+39.4 km route. We only used data marked as acceptable for use in the annual BBS
+analysis \citep{sauer_etal:1994auk}.  We summed the total birds of each
+species seen on each route and year and used the routes (rather
+than the individual stops) as our sites.
+Strong winds can interfere with point count observers' ability
+to hear birds \citep{simons_etal:2007}; we tested the effects of wind
+speed on detection probability. BBS volunteers record wind conditions
+at the beginning and end of each route on the Beaufort Scale
+\citep[start and end wind]{robbins_etal:1986}.
+%(start and end wind; Robbins et al. 1986)(start and end wind;0-9; cite).
+When start or end wind was not recorded we imputed those values with
+the start or end mean. We took the mean of start and end wind for
+each route and put this average wind scale value into four
+categories: 0 $\leq$ wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind
+$\geq$ 3 (maximum of 3.5).
 Following \citet{link_sauer:2002},
 %Link and Sauer (2002),
-we also included the first time an
-observer ran a route as a predictor variable for detection
+we also included the first time an observer ran a route as a predictor variable for detection
 probability.
@@
 %HOW WE OBTAINED AND MANIPULATED PRISM DATA
-We ran a series of maximum-likelihood based models for each species,
-and then a series of Bayesian models.  We started by testing three
-models of initial abundance (Poisson, negative binomial, and
-zero-inflated Poisson) with population growth dynamics (Eq.~\ref{eq:exp})
-and no covariates.  We selected the minimum Akaike's Information
-Criterion (AIC) model from that set to test three additional models
-for p: wind, first, and wind + first.   We selected the minimum AIC
-model from that set to test eight additional models of dynamics:
-constant, autoregressive, Ricker, Gompertz, autoregressive +
-immigration, population growth + immigration, Ricker + immigration,
-and Gompertz + immigration.  %We then tested the effect of average
+We ran a series of maximum-likelihood based models for each
+species, and then a series of Bayesian models. We started by testing
+three models of initial abundance (Poisson, negative binomial, and
+zero-inflated Poisson) with exponential dynamics
+(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
+Criterion (AIC) model from that set to test three additional
+models for p: wind, first, and wind + first. We selected the minimum
+AIC model from that set to test eight additional models of dynamics:
+constant, autoregressive, Ricker-logistic, Gompertz-logistic, autoregressive +
+immigration, exponential + immigration, Ricker-logistic +
+immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
 %minimum temperature for June and July on all dynamics parameters from
 %the minimum AIC dynamics model.
-As above, these models require a maximum abundance to integrate over
-when run in a maximum likelihood
-framework; we used 600 for ovenbirds and 350
-for GWWA.
-
-We ran the top ranked models from the maximum likelihood analyses in a
-Bayesian framework.  We added random observer effects and, where
-appropriate, zero-inflated dynamics.  We used
-non-informative priors.  We tested for lack of convergence using 5
-Markov chains for each model \citep{gelman_rubin:1992}.
+As above, these models require a maximum abundance to integrate
+over when run in a maximum likelihood
+framework; we used 600 for ovenbirds and 350 for golden-winged warblers.
+
+We ran the top ranked models from the maximum likelihood
+analyses in a Bayesian framework.  We added random observer 
+effects and, where appropriate, zero-inflated dynamics.  We used
+non-informative priors. We tested for lack of convergence using
+5 Markov chains for each model \citep{gelman_rubin:1992}.
 %(Gelman and Rubin 1992).
-For each chain
-we sampled the MCMC for at least 40,000 iterations, after at least 3,000 tuning samples.
+For each chain we sampled the MCMC for at least 40,000 iterations, after at
+least 3,000 tuning samples.
 %Latent inclusion factors for temperature effects?
@@
 
+\subsection{Simulation Study}
+
+Generally, estimator bias and error were low with the exponential model, and 
+coverage was nominal (Fig. ~\ref{fig:exp_hists}; Appendix ?). However, estimates of $r$ were 
+imprecise and somewhat biased when the true value of $\Lambda$ was low (1). 
+Simulations of the Ricker-logistic model also generally performed well (Figure 2; 
+Appendix ?), except when the true value of $r$ was low (0.005). In this case the 
+estimates of $K$ were imprecise and biased, especially when the true value of $K$ 
+was high.  All three cases simulated for the Ricker-logistic + immigration model 
+performed well (Figure 3; Appendix ?).
+
 \subsection{Analysis of Breeding Bird Survey Data}
@@
 The negative binomial distribution was strongly supported for ovenbird
-initial abundance over the Poisson and zero-inflated Poisson (Table
-2A, models A.1 - A.3).  Even compared to the Poisson, there was no
-evidence to support a zero-inflation factor for this species.  The
-best supported model for p was additive effects of wind speed and
-first run (Table 2A, model A.4).  First run and increasing wind speeds
-both decreased p.  All dynamics models with immigration were better
-supported than models without immigration (Table 2A, models A.8 -
-A.16); the best supported of these was the Ricker + Immigration.  The
-best supported model of those without immigration was the Gompertz,
-and the autoregressive and the constant model had the least support.
+initial abundance over the Poisson and zero-inflated Poisson
+(Table 2A). Even compared to the Poisson, there was
+no evidence to support a zero-inflation factor for this species.
+The best supported model for p was additive effects of wind speed
+and first run (Table 2B, model B.1). First run and increasing wind
+speeds both decreased p. All dynamics models with immigration were
+better supported than models without immigration (Table 2C); 
+the best supported of these was the Ricker-logistic + immigration.
+The best supported model of those without immigration was the
+Gompertz-logistic, and the autoregressive and the constant model had the least
+support.
 %There was little support for an effect of minimum June and July
@@
 
-Estimates for the top ranked model for OVEN
-(NB[$\Lambda (.) \alpha (.)$]Ricker+Imm.[$r(.)K(.) \iota (.)$]p(wind+1st)) were similar when
-run in the Bayesian framework, except that the estimate of r more than
-halved (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for Bayesian model
-estimates we present mean and SD]).  When random observer effects were
-added, estimates for $\Lambda$ and K increased dramatically (from 31.6 $\pm$ 4.9
-to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1, respectively), and
-the estimate for r was intermediate (0.018 $\pm$ 0.006).  The estimate of
-the intercept for p (on the logit scale) dropped from -1.5 $\pm$ 0.1 to
--2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.
-%[Richard, I ran
-%the temperature model in the Bayesian framework (w/ and w/o random
-%effect) too.  Should we include those results?  As earlier, very
-%little evidence for an effect of temperature on any of the parameters,
-%although I couldn't get the latent inclusion parameter to work for r.]
-Gelman and Rubin diagnostics and visual examination of the ? plots
+Estimates for the top ranked model for ovenbird
+(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota
+(.)$]p(wind+1st)) were similar when
+run in the Bayesian framework, except that the estimate of r
+more than halved (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
+Bayesian model estimates we present mean and SD]). When random observer effects
+were added, estimates for $\Lambda$ and K increased dramatically
+(from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
+respectively), and the estimate for r was intermediate (0.018 $\pm$ 0.006). The
+estimate of the intercept for p (on the logit scale) dropped from -1.5 $\pm$
+0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.
+
+Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
 provided some evidence for a lack of convergence in estimates of the
@@
 There was slightly more support for the ZIP distribution for initial
-abundance of GWWA than for the negative binomial; both were strongly
-supported over the Poisson (Table 2B, models B.1 - B.3).  The best
-supported model for p was an effect of first run (Table 2B, model
-B.4).  The best supported dynamics model was population growth (Table
-2B, model B.8); the estimate of r from this model was -0.058 $\pm$ 0.017
-(SE).  There was no support for immigration added to the population
-growth or autoregressive models (Table 1B, models B.10 and B.13).
-However, the Gompertz + immigration and Ricker + immigration models
-were supported over the corresponding models without immigration
-(Table 1B, models B.11, B.12, B.14, and B.15), despite low estimates
-of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$ 0.002,
-respectively).  %There was considerable support for an effect of
+abundance of golden-winged warbler than for the negative binomial; both were strongly
+supported over the Poisson (Table 3A). The best
+supported model for p was an effect of first run (Table 3B, model
+B.1). The best supported dynamics model was exponential
+(Table 3C, model C.1); the estimate of r from this model was -0.058
+$\pm$ 0.017 (SE). There was no support for immigration added to the
+exponential or autoregressive models (Table 3C, models C.4 and
+C.6).
+However, the Gompertz-logistic + immigration and Ricker-logistic + immigration
+models were supported over the corresponding models without immigration
+(Table 3C, models C.4, C.5, C.7, and C.8), despite low estimates
+of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$
+0.002, respectively).  
+%There was considerable support for an effect of
 %average minimum June and July temperature on r in the population
@@
 Because of the closeness of rankings for the ZIP and negative binomial
-distributions for GWWA, we also ran the subsequent models with the
-negative binomial initial abundance.  Rankings of subsequent models
-were similar, but with a p covariate (first) the negative binomial
-distribution outranked the ZIP.
-
-Estimates for the top ranked model for GWWA (ZIP[Λ(.)α(.)]Trend[r(.)]p(1st))
-were similar when run in the Bayesian framework.  When random observer
-effects were added, the estimate for Λ  increased dramatically
-(from 22.4 ± 7.6 to 55.5 ± 30.9).  The estimate of the intercept for p (on the
-logit scale) dropped from -3.5 ± 0.2 to -5.6 ± 1.1, and the estimate of σp
-was 1.7 ± 0.7.  Adding zero-inflation to the dynamics (as well as the initial
-abundance) has no effect for trend models, because populations that started
-at 0 would stay at 0 in any case.  Therefore, we tested zero-inflated dynamics
-on the Trend + Imm. model.  Adding zero-inflation to the dynamics increased
-the estimate of ι from 0 to 0.091 ± 0.049, but had only minor effects on the
-other parameters.  Gelman and Rubin diagnostics and visual examination of
-the ? plots provided evidence for a lack of convergence in estimates of Λ
-for all Bayesian models ran, and for most parameters when random effects
-were included.
+distributions for golden-winged warbler, we also ran the subsequent models with the
+negative binomial initial abundance. Rankings of subsequent
+models were similar, but with a p covariate (first) the negative
+binomial distribution outranked the ZIP.
+
+Estimates for the top ranked model for golden-winged warbler
+(ZIP[Λ(.)α(.)]Exponential[r(.)]p(1st))
+were similar when run in the Bayesian framework. When random
+observer effects were added, the estimate for $\Lambda$ increased dramatically
+(from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
+for p (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
+estimate of σ was 1.7 $\pm$ 0.7. Adding zero-inflation to the dynamics (as well as
+the initial abundance) has no effect for exponential models, because populations
+that started at 0 would stay at 0 in any case. Therefore, we tested
+zero-inflated dynamics on the Exponential + Immigration model. Adding zero-inflation to the dynamics
+increased the estimate of ι from 0 to 0.091 $\pm$ 0.049, but had only minor
+effects on the other parameters. Gelman and Rubin diagnostics and visual
+examination of the chain trajectory and density plots provided evidence for a lack of convergence in
+estimates of $\Lambda$ for all Bayesian models ran, and for most parameters when random
+effects were included.
 
 Estimated detection probabilities were generally low. The estimated probability
-of detecting an OVEN in the top model (A.8) varied between 0.157 ± 0.014
-and 0.184 ± 0.015, depending on wind speed and whether the observer had
-run that route before.  The estimated probability of detecting a GWWA went
-from 0.013 ± 0.006 for first time observers to 0.028 ± 0.008 for repeat
-observers for the top model (B.8).  Accounting for random observer effects
-decreased average detection probability for both species.
+of detecting an ovenbird in the top model (Table 2, model C.1) varied between 0.157
+$\pm$ 0.014 and 0.184 $\pm$ 0.015, depending on wind speed and whether the
+observer had run that route before. The estimated probability of detecting a
+golden-winged warbler went from 0.013 $\pm$ 0.006 for first time observers to 0.028 $\pm$ 0.008 for
+repeat observers for the top model (Table 3, model C.1). Accounting for random
+observer effects decreased average detection probability for both species.
 
 For both species, constant dynamics models estimated very high
-survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05 for OVEN and 0.935 $\pm$ 0.011 for
-GWWA).   The autoregressive and autoregressive + immigration models
-estimated very low survival probabilities for OVEN ($\omega$ = 0.058 $\pm$ 0.046
-and 0.026 $\pm$ 0.060, respectively), but not for GWWA ($\omega$ = 0.715 $\pm$ 0.288
-for both models).
+survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05 for ovenbird and
+0.935 $\pm$ 0.011 for
+golden-winged warbler). The autoregressive and autoregressive + immigration
+models estimated very low survival probabilities for ovenbird ($\omega$ =
+0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively), 
+but not for golden-winged warbler ($\omega$ = 0.715 $\pm$ 0.288 for both models).
 
@@
   The model (1) recognizes
-  the discrete nature of count data and the underlying true abundance
-  state, (2) includes a realistic observation model with explicit
-  detection probability parameters (3) has estimable parameters,
-  including those describing the initial state at time $t=1$, and (4)
-  allows for inference about both temporal and spatial variation in
+the discrete nature of count data and the underlying true
+abundance
+state, (2) includes a realistic observation model with explicitdetection probability parameters (3) has estimable parameters,including those describing the initial state at time $t=1$, and
+(4)
+allows for inference about both temporal and spatial variation
+in
   abundance.
@@
 %available, and in some cases, facilitates parameter estimation.
-
 We also demonstrated how many of the objectives of conventional
 state-space modeling can be accomplished within this expanded
-framework. For example, one of the
-primary aims of our paper was to demonstrate how classical
-population growth models can be embedded in the DM model. Although we
-view this as an important connection to make between the two different
-classes of state-space models, we believe that
+framework. For example, one of the primary aims of our paper was to demonstrate how classical
+population growth models can be embedded in the DM model.
+Although we view this as an important connection to make between the two
+different classes of state-space models, we believe that
 %In some ways,
 such population growth models are more phenomenological
-than mechanistic, and this runs counter to the motivation for the DM
-model. However, estimating
+than mechanistic, and this runs counter to the motivation for
+the DM model. However, estimating
 demographic parameters from count data is an ambitious goal, and
-the required assumption will not be valid in many cases, especially
-when immigration and emigration occur. This was supported by the
-unrealistic estimates of vital rates reported here and by the lack of
-support for the mechanistic models relative to the population growth
-models.
-
-One strategy for adhereing to the original formulation of the model
-that included survival and recruitment parameters would be to use
-either prior information or additional data. For example, in a
-Bayesian analysis existing estimates of demographic parameters could
-be used as informative priors, which might make it possible to
+the required assumption will not be valid in many cases,
+especially when immigration and emigration occur. This was supported by the
+unrealistic estimates of vital rates reported here and by the
+lack of support for the mechanistic models relative to the population
+growth models.
+
+One strategy for adhereing to the original formulation of the
+model that included survival and recruitment parameters would be to
+use either prior information or additional data. For example, in a
+Bayesian analysis existing estimates of demographic parameters
+could be used as informative priors, which might make it possible to
 seperate the effects of movement versus births and deaths.
-Alternatively, ecologists could combine count data with data from, for example,
-capture-recapture data. Models that are fitted to count data and demographic data are
-often referred to as integrated population models
-\citep[IPM;][]{besbeas_etal:2002, buckland_etal:2004,schaub_etal:2007}
+Alternatively, ecologists could combine count data with data
+from, for example,
+capture-recapture data. Models that are fitted to count data and
+demographic data are often referred to as integrated population models
+\citep[IPM;][]{besbeas_etal:2002,
+buckland_etal:2004,schaub_etal:2007}
 and we believe that a DM model could work well in this context.
@@
 IPM model proposed by \citet{buckland_etal:2004}.
-However, we caution that when count data are affected by movement
-and vital rates, it would be unwise to ignore the movement processes.
+However, we caution that when count data are affected by
+movement and vital rates, it would be unwise to ignore the movement
+processes.
 
@@
 
-Although these integrated
-population models may be viewed as the gold standard in state-space
-modeling, ecologists are not always so fortunate to have direct
+Although these integrated population models may be viewed as the gold standard in
+state-space modeling, ecologists are not always so fortunate to have direct
 information about vital rates, especially at large spatial scales.
 Rather, count data are much more common, and
-are produced by many of the largest monitoring programs in the world,
-such as the North American Breeding Bird Survey
+are produced by many of the largest monitoring programs in the
+world, such as the North American Breeding Bird Survey
 \citep[BBS;][]{robbins_etal:1986}.
 In the absence of direct information about demographic parameters, and
-when the original assumptions of the model do not hold, the model can
-still be used for important purposes such as predicting future
+when the original assumptions of the model do not hold, the
+model can still be used for important purposes such as predicting future
 population size and changes in species distributions.
@@
 simulate and fit many of the extensions discussed in this paper.
-
 In spite of the new extensions we have proposed, several aspects of
@@
 parameters of the state process can be estimated ultimately depends
-upon how well detection probability is estimated. When there is only a
-single survey per primary period, the information about detection
-probability comes from deviations from the parametric assumptions
-about population dynamics. Thus, without direct information about
-detection probability, the estimates will be determined by the model's
-parametric assumptions. Furthermore, properly modeling detection
-probability can involve account for nuances such as difference among
-individuals and probabilities of being available for detection
+upon how well detection probability is estimated. When there is
+only a single survey per primary period, the information about
+detection probability comes from deviations from the parametric
+assumptions about population dynamics. Thus, without direct information
+about detection probability, the estimates will be determined by the
+model's parametric assumptions. Furthermore, properly modeling detection
+probability can involve account for nuances such as difference
+among individuals and probabilities of being available for detection
 \citep{nichols_etal:2009}.
@@
 detection probability and we recommend that this be done whenever
-possible. The original method for proposed in the original paper was
-to collected replicated counts within the primary periods during a
-period in which the populations could be assumed to be closed.  A
-robust design could be used to combine multiple surveys per primary
-period to increase the precision of the estimates. We envision that
-multiple other options are available as well, such as removal, time of
-detection, and distance modeling ([cites]). These could be
-accomplished by extending the open population $N$-mixture model in
-exactly the same way as the closed population version has been
+possible. The original method for proposed in the original paper
+was to collected replicated counts within the primary periods during
+a period in which the populations could be assumed to be closed. A
+robust design could be used to combine multiple surveys per
+primary period to increase the precision of the estimates. We envision
+that multiple other options are available as well, such as removal,
+time of detection, and distance modeling ([cites]). These could be
+accomplished by extending the open population $N$-mixture model
+in exactly the same way as the closed population version has been
 extended \citep[e.g.,][]{royle_etal:2004}.
 
-Our emphasis was on increasing the practical utility of this class of
-models, and so we avoided several conceptually-interesting extensions
-that we belive would be computationally prohibitive in many
+Our emphasis was on increasing the practical utility of this
+class of models, and so we avoided several conceptually-interesting
+extensions that we belive would be computationally prohibitive in many
 cases. Nonetheless, we will discuss one---spatially-explict models of
@@
 hand, our Bayesian analyses did have convergence problems,
-particularly for the species (GWWA) with sparser data. It's
+particularly for the species (golden-winged warbler) with sparser data. It's
 likely that poor model fits are related to the low estimates of
@@
 frequentist strengths and weakness in the Bayesian paragraph
-(currently not written), and GWWA ZIP results in ZIP paragraph.]}
+(currently not written), and golden-winged warbler ZIP results in ZIP
+paragraph.]}
 
 The modeling framework we described can be used to address many
-pressing issues in ecology and conservation biology. For example,
+pressing issues in ecology and conservation biology. For
+example,
 %it is possible to test hypotheses about temporal and spatial
@@
 about the effects of climate change on either explicit
-demographic parameters or in derived parameters such as population
-growth rate. Furthermore, %under the Bayesian approach, population
-in addition to assessing past climate change on population parameters,
-projecting populations under future climate scenarios is also
-possible. Thus, we expect this modeling methodology can be used to
-make quantiative assessments of species vulnerability to climate
-change and other important human-induced drivers of population dynamics.
+demographic parameters or in derived parameters such as
+population growth rate. Furthermore, %under the Bayesian approach, population
+in addition to assessing past climate change on population
+parameters, projecting populations under future climate scenarios is also
+possible. Thus, we expect this modeling methodology can be used
+to make quantiative assessments of species vulnerability to climate
+change and other important human-induced drivers of population
+dynamics.
 
@@
 
-We thank J. Andrew Royle and T. Scott Sillett for helpful comments
-and suggestions.
+We thank J. Andrew Royle and T. Scott Sillett for helpful
+comments and suggestions. We gratefully acknowledge National Park 
+Service, Smithsonian Institution, and [more] for funding support.
 
@@
 for 100 sites over 40 years.  We assumed that initial abundance
-    ($\Lambda$) was Poisson distributed.  For the exponential growth
-    simulations we included all combinations of low, medium, and high
-    $\Lambda$, growth rate ($r$), and detection probability ($p$).  For the Ricker
-    model simulations we used $\Lambda$ = 10 and $p$ = 0.25, and simulated low,
-    medium, and high values of $r$ and equilibrium density ($K$).  For the
-    Ricker + immigration dynamics model we fixed all parameters the
-    same as the Ricker model (with $r$ = 0.05 and $K$ = 10) and simulated
-    low, medium, and high values of immigration rate ($\iota$).}
+($\Lambda$) was Poisson distributed.  For the exponential 
+simulations we included all combinations of low, medium, and
+high $\Lambda$, growth rate ($r$), and detection probability ($p$).
+For the Ricker-logistic model simulations we used $\Lambda$ = 10 and $p$ = 0.25, and
+simulated low, medium, and high values of $r$ and equilibrium density ($K$).
+For the Ricker-logistic + immigration dynamics model we fixed all parameters the
+same as the Ricker-logistic model (with $r$ = 0.05 and $K$ = 10) and
+simulated low, medium, and high values of immigration rate ($\iota$).}  
 \begin{tabular}{lcccccccc}
     \hline
-    & \multicolumn{3}{c}{Exponential Growth} && Ricker &&
-    \multicolumn{2}{c}{Ricker + Immigration} \\
-    \cline{2-4}     \cline{6-6}    \cline{8-9}
+    & \multicolumn{3}{c}{Exponential Growth} && \multicolumn{2}{c}{Ricker-logistic} &&
+    Ricker-logistic + Immigration \\
+    \cline{2-4}     \cline{6-7}    \cline{9-9}
 % & \multicolumn{3}{c}{\rule{25mm}{.1pt}} & \rule{5cm}{.5pt} & \multicolumn{2}{c}{\rule{20mm}{.1pt}} \\
-                & $\Lambda$ & $r$ & $p$ && $r$  && $K$ & $\iota$  \\
+& $\Lambda$ & $r$ & $p$ && $r$  & $K$ && $\iota$  \\    
 \hline
-    Low	        &1	&-0.01	&0.05	&&0.005	&&5	&0.005  \\
-    Med	        &5	&0	&0.25	&&0.05	&&10	&0.05   \\
-    High	&10	&0.005	&0.5	&&0.1	&&20	&0.5    \\
+    Low	        &1	&-0.01	&0.05	&&0.005	 &5	&&0.005  \\
+    Med	        &5	&0	&0.25	&&0.05	&10	&&0.05   \\
+    High		 &10 &0.005	&0.5	&&0.1	&20	&&0.5    \\
     \hline
@@
 \caption{Model selection table for ovenbirds and golden-winged
-    warblers in Maryland and Virginia, 1966-2010.  We present model name
-    and number, number of parameters (Par.), and difference in Akaike's
-    information criterion between each model and the top model of that
-    set ($\Delta$AIC).  For each species, the first section compares
-    models for initial abundance, the second for detection probability,
+warblers in Maryland and Virginia, 1966-2010. We present model
+name
+and number, number of parameters (Par.), and difference in
+Akaike's
+information criterion between each model and the top model of
+that
+set ($\Delta$AIC).  For each species, the first section comparesmodels for initial abundance, the second for detection
+probability,
     and the third for dynamics.}
@@
 Model	&Par.	&$\Delta$AIC	&Model	&Par.	&$\Delta$AIC \\
-A.1. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0
+A.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0
 &B.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&0 \\
 A.2. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&1262.7
-&B.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0.4 \\
-A.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4
+&B.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0.4 \\A.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4
 &1264.7	&B.3. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&37.8 \\
@@
 A.7. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&6.4
-&B.7. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind)	&7	&9.3 \\
-A.8. NB[$\Lambda$(.)$\alpha$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
+&B.7. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind) &7 &9.3
+\\
+A.8.
+NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
 &10	&0	&B.8. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)
 &5	&0 \\
-A.9. NB[$\Lambda$(.)$\alpha$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
+A.9.
+NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
 &10	&8.4
-&B.9. ZIP[$\Lambda$(.)$\psi$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(1st)
+&B.9.
+ZIP[$\Lambda$(.)$\psi$(.)]Autoregressive[$\gamma$(.)$\omega$(.)]$p$(1st)
 &6	&1.7 \\
-A.10. NB[$\Lambda$(.)$\alpha$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(wind+1st)
+A.10.
+NB[$\Lambda$(.)$\alpha$(.)]Growth+Immigration[$r$(.)$\iota$(.)]$p$(wind+1st)
 &9	&36.5
-&B.10. ZIP[$\Lambda$(.)$\psi$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(1st)
+&B.10.
+ZIP[$\Lambda$(.)$\psi$(.)]Growth+Immigration[$r$(.)$\iota$(.)]$p$(1st)
 &6	&2.0 \\
-A.11. NB[$\Lambda$(.)$\alpha$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
+A.11.
+NB[$\Lambda$(.)$\alpha$(.)]Autoregressive+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
 &10	&38.6
-&B.11. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
+&B.11.
+ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
 &7	&2.5 \\
-A.12. NB[$\Lambda$(.)$\alpha$(.)]Gompertz[$r$(.)$K$(.)]$p$(wind+1st)
+A.12.
+NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(wind+1st)
 &9	&192.8
-&B.12. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz[$r$(.)$K$(.)]$p$(1st)	&6
+&B.12. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(1st)
+&6
 &3.0 \\
-A.13. NB[$\Lambda$(.)$\alpha$(.)]Ricker[$r$(.)$K$(.)]$p$(wind+1st)
+A.13.
+NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(wind+1st)
 &9	&195.1
-&B.13. ZIP[$\Lambda$(.)$\psi$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
+&B.13.
+ZIP[$\Lambda$(.)$\psi$(.)]Autoregressive+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
 &7	&3.7 \\
@@
 &271.3
-&B.14. ZIP[$\Lambda$(.)$\psi$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
+&B.14.
+ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
 &7	&4.2 \\
-A.15. NB[$\Lambda$(.)$\alpha$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
+A.15.
+NB[$\Lambda$(.)$\alpha$(.)]Autoregressive[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
 &9	&273.7
-&B.15. ZIP[$\Lambda$(.)$\psi$(.)]Ricker[$r$(.)$K$(.)]$p$(1st)	&7
-&5.2 \\
-A.16. NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
+&B.15. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(1st)	&7&5.2 \\
+A.16.
+NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
 &9	&1856.7
-&B.16. ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st)
+&B.16.
+ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st)
 &6	&12.5 \\
@@
 
-% How about separate tables?  Also updated the Gompertz results.
+% How about separate tables?  Also updated the Gompertz-logistic results.
 \begin{table}
@@
 \caption{Model selection table for ovenbirds in Maryland and Virginia,
-    1966-2010.  We present model name
-    and number, number of parameters (Par.), and difference in Akaike's
-    information criterion between each model and the top model of that
-    set ($\Delta$AIC).  The first section compares
-    models for initial abundance, the second for detection probability,
-    and the third for dynamics.}
+    1966-2010.  We present model name and number, number of 
+parameters (Par.), and difference in Akaike's
+information criterion between each model and the top model of
+that set ($\Delta$AIC).  The first section compares
+models for initial abundance, the second for detection
+probability, and the third for dynamics.}
   \begin{tabular}[h]{lcc}
@@
 A. Initial Abundance && \\
-A.1. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0\\
-A.2. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&1262.7\\
-A.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4 &1264.7\\
+A.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0\\
+A.2. P[$\Lambda$(.)]Exponential[$r$(.)]$p$(.)	&3	&1262.7\\
+A.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4 &1264.7\\
 \hline
 B. Detection Probability && \\
-B.1. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
-&0	\\
-B.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind)	&7	&0.9\\
-B.3. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(1st)	&5	&5.0\\
-B.4. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&6.4\\
+B.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &0	\\
+B.2. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind) &7 &0.9\\
+B.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(1st)	&5	&5.0
+\\B.4. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&6.4\\
 \hline
 C. Dynamics && \\
-C.1. NB[$\Lambda$(.)$\alpha$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
-&10	&0	\\
-C.2. NB[$\Lambda$(.)$\alpha$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
-&10	&8.4 \\
-C.3. NB[$\Lambda$(.)$\alpha$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(wind+1st)
-&9	&36.5\\
-C.4. NB[$\Lambda$(.)$\alpha$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
-&10	&38.6\\
-C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz[$r$(.)$K$(.)]$p$(wind+1st)
-&9	&192.8\\
-C.6. NB[$\Lambda$(.)$\alpha$(.)]Ricker[$r$(.)$K$(.)]$p$(wind+1st)
-&9	&195.1\\
-C.7. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
-&271.3\\
-C.8. NB[$\Lambda$(.)$\alpha$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
-&9	&273.7\\
-C.9. NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
-&9	&1856.7\\
+C.1. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&0	\\
+C.2. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&8.4 \\
+C.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(wind+1st) &9	&36.5\\
+C.4. NB[$\Lambda$(.)$\alpha$(.)]Autoregressive+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st) &10	&38.6\\
+C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&192.8\\
+C.6. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&195.1\\
+C.7. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &271.3\\
+C.8. NB[$\Lambda$(.)$\alpha$(.)]Autoregressive[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&273.7\\
+C.9. NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&1856.7\\
 \hline
@@
   \caption{Model selection table for golden-winged
-    warblers in Maryland and Virginia, 1966-2010.  We present model name
-    and number, number of parameters (Par.), and difference in Akaike's
-    information criterion between each model and the top model of that
-    set ($\Delta$AIC).  The first section compares
-    models for initial abundance, the second for detection probability,
-    and the third for dynamics.}
+warblers in Maryland and Virginia, 1966-2010. We present model
+name and number, number of parameters (Par.), and difference in
+Akaike's information criterion between each model and the top model of
+that set ($\Delta$AIC).  The first section compares
+models for initial abundance, the second for detection
+probability, and the third for dynamics.}
   \begin{tabular}[h]{lcc}
@@
 A. Initial Abundance && \\
-A.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&0 \\
-A.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0.4 \\
-A.3. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&37.8 \\
+A.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4	&0 \\
+A.2. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0.4 \\
+A.3. P[$\Lambda$(.)]Exponential[$r$(.)]$p$(.)	&3	&37.8 \\
 \hline
 B. Detection Probability && \\
-B.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)	&5
-&0 \\
-B.2. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&3.5 \\
-B.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
-&5.9 \\
-B.4. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind)	&7	&9.3 \\
+B.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(1st)	&5 &0 \\
+B.2. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4	&3.5 \\
+B.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &5.9 \\
+B.4. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(wind) &7 &9.3\\
 \hline
 C. Dynamics && \\
-C.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)
-&5	&0 \\
-C.2. ZIP[$\Lambda$(.)$\psi$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(1st)
-&6	&1.7 \\
-C.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(1st)
-&6	&2.0 \\
-C.4. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
-&7	&2.5 \\
-C.5. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz[$r$(.)$K$(.)]$p$(1st)	&6
-&3.0 \\
-C.6. ZIP[$\Lambda$(.)$\psi$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
-&7	&3.7 \\
-C.7. ZIP[$\Lambda$(.)$\psi$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
-&7	&4.2 \\
-C.8. ZIP[$\Lambda$(.)$\psi$(.)]Ricker[$r$(.)$K$(.)]$p$(1st)	&7
-&5.2 \\
-C.9. ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st)
-&6	&12.5 \\
+C.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(1st) &5	&0 \\
+C.2. ZIP[$\Lambda$(.)$\psi$(.)]Autoregressive[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&1.7 \\
+C.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(1st) &6	&2.0 \\
+C.4. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st) &7	&2.5 \\
+C.5. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(1st)	&6&3.0 \\
+C.6. ZIP[$\Lambda$(.)$\psi$(.)]Autoregressive+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st) &7	&3.7 \\
+C.7. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st) &7	&4.2 \\
+C.8. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(1st)	&7 &5.2 \\
+C.9. ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&12.5 \\
 \hline
@@
 \caption{Histograms of 1000 parameter estimates for each of 27
-    simulation cases. The verticle lines are the data-generating
-    values.}
+simulation cases with exponential dynamics. The vertical lines are the 
+data-generating values.}
 \label{fig:exp_hists}

### 2012-10-14 supp/sim_table_exp.tex
rough simulation results plus

@@
 \HyPL@Entry{0<</S/D>>}
-\@writefile{lot}{\contentsline {table}{\numberline {1}{\ignorespaces Simulation results from 27 cases with 1000 Monte Carlo replicates for each.}}{1}{table.1}}
-\newlabel{tab:simexp}{{1}{1}{Simulation results from 27 cases with 1000 Monte Carlo replicates for each}{table.1}{}}
+\@writefile{lot}{\contentsline {table}{\numberline {1}{\ignorespaces Simulation results from 27 cases of trend dynamics with 1000 Monte Carlo replicates for each.}}{1}{table.1}}
+\newlabel{tab:simexp}{{1}{1}{Simulation results from 27 cases of trend dynamics with 1000 Monte Carlo replicates for each}{table.1}{}}

### 2012-10-19 Dail_Mad_ext.tex
more simulation results

@@
 Generally, estimator bias and error were low with the exponential model, and 
-coverage was nominal (Fig. ~\ref{fig:exp_hists}; Appendix ?). However, estimates of $r$ were 
-imprecise and somewhat biased when the true value of $\Lambda$ was low (1). 
-Simulations of the Ricker-logistic model also generally performed well (Figure 2; 
+coverage was nominal (93.4\% - 97.0\%; Fig.~\ref{fig:exp_hists}; Appendix ?). 
+However, estimates of $r$ were 
+imprecise (RMSE between 0.020 and 0.025; relative RMSE between 413\% and 
+482\% for non-zero true $r$) and biased low (between -0.008 and -0.005; relative bias between
+-151\% and -101\% for non-zero true $r$) when the true value of $\Lambda$ was low (1). 
+Simulations of the Ricker-logistic model also generally performed well (Fig.~\ref{fig:rick_hists}; 
 Appendix ?), except when the true value of $r$ was low (0.005). In this case the 
-estimates of $K$ were imprecise and biased, especially when the true value of $K$ 
-was high.  All three cases simulated for the Ricker-logistic + immigration model 
-performed well (Figure 3; Appendix ?).
+estimates of $r$ and $K$ were inaccurate (relative RMSE between 87.4\% and 2091.8\%),
+biased high (between 65.4\% and 198.9\%), and had low coverage (between 77.6\% and 94.3\%).
+In other cases for the Ricker-logistic model, relative RMSE was between 6.7\%
+and 55\%, bias was between -6.4\% and 5.8\%, and coverage was between
+91.9\% and 96.4\%.  All three cases simulated for the Ricker-logistic + immigration model 
+performed fairly well (Fig.~\ref{fig:ricki_hists}; Appendix ?; relative RMSE between 7.2\% 
+and 62.9\%, bias between -3.5\% and 3.3\%, and coverage between
+93.2\% and 96.4\%).
 
@@
 Estimates for the top ranked model for ovenbird
-(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota
-(.)$]p(wind+1st)) were similar when
+%(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
+were similar when
 run in the Bayesian framework, except that the estimate of r
@@
 0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.
-
 Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
@@
 Estimates for the top ranked model for golden-winged warbler
-(ZIP[Λ(.)α(.)]Exponential[r(.)]p(1st))
+%(ZIP[Λ(.)α(.)]Exponential[r(.)]p(1st))
 were similar when run in the Bayesian framework. When random
@@
 (from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
-for p (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
-estimate of σ was 1.7 $\pm$ 0.7. Adding zero-inflation to the dynamics (as well as
+for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
+estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. Adding zero-inflation to the dynamics (as well as
 the initial abundance) has no effect for exponential models, because populations
@@
 
+\begin{sidewaysfigure}
+  \centering
+  \includegraphics{figs/rick_hists}
+\caption{Histograms of 1000 parameter estimates for each of 9
+simulation cases with Ricker-logistic dynamics. The vertical lines are the 
+data-generating values.}
+\label{fig:rick_hists}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+  \centering
+  \includegraphics[width=8.5in]{figs/ricki_hists}
+\caption{Histograms of 1000 parameter estimates for each of 3
+simulation cases with Ricker-logistic + immigration dynamics. The vertical lines are the 
+data-generating values.}
+\label{fig:ricki_hists}
+\end{sidewaysfigure}
+
 \end{document}

### 2012-11-08 Dail_Mad_ext.tex
Map figure and simulation discussion

@@
 In applied contexts, population models are used for estimating extinction probabilities
-\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
+\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
 effects of future environmental conditions or conservation actions on
@@
 least 3,000 tuning samples.
+
+We also used these models to obtain route and 
+year specific abundance estimates.  Because of the memory requirements involved,
+we obtained these estimates from a single chain of 5,000 iterations, after 35,000
+tuning samples.  We present these estimates for ovenbirds only.
 %Latent inclusion factors for temperature effects?
@@
 intercept for p for all Bayesian models ran.
+
+We estimated ovenbird abundance by route and year (and averaged over routes) 
+using the Bayesian model including random observer effects (Fig.~\ref{fig:oven_N}).  
+Estimated average abundance oscillated over time with a slowly increasing trend.  Estimated ovenbird
+abundance was highest in western and eastern Maryland in 1970 and 1980, but became
+more evenly distributed by 2010.  Coefficient of variation (CV) for route and year specific
+abundance estimates varied between 0.11 and 2.31 (mean = 0.46).  By comparison, CV
+for the estimate of mean initial abundance ($\Lambda$) was 0.15.
 
@@
 
-[simulation results discussion]
+In our analysis of simulated data, we found that our models generally estimated parameters well.
+Two exceptions were the estimates of instantaneous growth rate in exponential model when the 
+true value of initial abundance was low and the estimates of equilibrium abundance and maximum
+growth rate in the Ricker model when the true value of maximum growth rate was low.  
+The failure of the first case was likely due to a large number of sites starting at abundance 0 or reaching it early in the simulation.  
+%[Richard, maybe come up with a summary statistic for this based on the simulation N values?]
+Since a site that hits abundance 0 in this simulation stayed at 0, this provided little information with
+which to estimate growth rate.  The failure of the second case is also reasonable: with very
+slow movement toward the equilibrium abundance comes little information for estimating
+that abundance or the rate of growth.  In other cases, bias was low and coverage nominal.
 
@@
 visits within a season) would greatly improve the fit of these
-models in both frameworks.
+models in both frameworks.  We also were able to obtain year
+and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
+although with generally lower precision than more global estimates.  These sort of estimates
+may be more reliable when abundance and dynamics are modeled 
+as functions of habitat covariates, which we did not do here.
 \hl{[Alternatively, could put discussions of our results where
@@
 in population models paragraph, BBS Bayesian vs.
-frequentist strengths and weakness in the Bayesian paragraph
-(currently not written), and golden-winged warbler ZIP results in ZIP
+frequentist strengths and weakness in the Bayesian paragraph, 
+and golden-winged warbler ZIP results in ZIP
 paragraph.]}
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year2}
+\caption{Maps of BBS route and year specific estimated abundances for ovenbirds, for the years 1970,
+1980, 1990, 2000, and 2010.  Lower right panel depicts estimated mean route abundance by year
+(mean and 95\% credible interval).}
+\label{fig:oven_N}
+\end{figure}
+
 \end{document}

### 2012-12-14 Dail_Mad_ext.tex
Stochasticity section and random effect paragraph

@@
 \begin{equation}
-  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K))
+  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K)))
 \label{eq:rick}
@@
 N_{it} \sim
-\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1)))
+\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1))))
 \label{eq:gomp}
@@
 the
-abundance of several bird species with the same set of BBS
-routespoint
+abundance of several species with the same set of point
 count surveys for all species, but this set includes routes
 sites
-outside the range of some of the bird species. The distribution
+outside the range of some of the species. The distribution
 of
@@
 These models allow three sources of zero counts by observers: a
-species was at a route but not detected; the route was within the
-species' range but there were no birds at that site in that
-year; and the route was outside the species' range. Furthermore,
+species was at a site but not detected; the site was within the
+species' range but there were no animals at that site in that
+year; and the site was outside the species' range. Furthermore,
 detection, abundance, and zero-inflation can be modeled separately as
-functions of different (or the same) covariates. For example, detection of
-species x at site y in year z might depend on wind speed,
+functions of different (or the same) covariates. For example, detection of a
+species might depend on wind speed,
 abundance on forest type and weather, and zero-inflation upon elevation and
-climate.  This approach combines elements of occupancy modeling
-\citep{mackenzie_etal:2006} and abundance modeling.
+climate.  
+%This approach combines elements of occupancy modeling
+%\citep{mackenzie_etal:2006} and abundance modeling.
 %, and could prove useful
@@
 \subsection{Environmental and demographic stochasticity}
-
-Even though the DM model is conceptually very simple, even in
+\begin{comment}
+Even though the DM model is conceptually simple, even in
 its most stripped down form it includes many random effects (the $N$'s)
@@
 may be of interest. For example, in conventional state-space models,
-demographic stochasticity is often modeled as $r_t \sim
+environmental stochasticity is often modeled as $r_t \sim
 \text{Norm}(\mu_r, \sigma_r)$.
+\end{comment}
+The original DM models include demographic stochasticity (variation in population growth 
+rate due to the randomness of birth and death processes) through random 
+binomial draws for survival and Poisson draws for recruitment.  Our population
+growth extensions use a single Poisson draw for abundance: $N_{it} \sim 
+\text{Pois}(f(N_{it-1}))$.  This is also a commonly used (and mathematically equivalent) 
+method of modeling demographic stochasticity 
+\citep{bonsall_hastings:2004,melbourne_hastings:2008}.
+
+One way to model environmental stochasticity (variation in population growth
+rate due to stochastic environmental conditions) is by explicitly
+modeling dynamic parameters as functions of environmental covariates, such
+as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
+function of temperature and $K$ a function of rainfall:
+\begin{gather}
+r_{it} = \exp(r_{0} + r_{temp}T_{it}) \nonumber \\
+K_{it} = \exp(K_{0} + K_{rain}R_{it})  \\
+N_{it} \sim
+\text{Pois}(N_{it-1}\exp(r_{it-1}(1-\log(N_{it-1}+1)/\log(K_{it-1}+1)))) \nonumber
+\label{eq:weather}
+\end{gather}
+where $T_{it}$ is site and year specific temperature, $R_{it}$ is site and year specific rainfall,
+and the errors for the coefficients $r_{0}$, $r_{temp}$, $K_{0}$, and $K_{rain}$ are 
+normally distributed.
+
+Another way to model environmental stochasticity is as lognormal variation in observed local population 
+growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  This could be applied to both density-independent and density-dependent models.
+For example, for the Ricker-logistic + immigration model:
+\begin{equation}
+N_{it} \sim
+\text{Pois}(N_{it-1}\exp(\nu_{it} + r(1-N_{it-1}/K)) + \iota)
+\label{eq:nuRand}
+\end{equation}
+where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
+Three variations suggest themselves.  The first has independent environmental stochasticity 
+between sites, and might be appropriate where sites are widely dispersed.  In the second 
+all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{it} = \nu_{t}$),
+which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
+In cases where sites can be close together but the range is broad, a multivariate normal distribution for $\nu_{it}$ with correlation coefficients
+between sites in a year dependent on distance (geographic or otherwise) might be appropriate.
 
@@
 ovenbirds (\textit{Seiurus aurocapilla}), an abundant and widespread
-forest breeding migrant with a stable or increasing trend in the region (CITE)
+forest breeding migrant with a stable or increasing trend in the region \citep{porneluzi_etal:2011}
 and golden-winged warblers (\textit{Vermivora chrysoptera}), which
 only breeds in the western parts of these states and has been declining in the
-region (CITE).  %MORE INFORMATION ABOUT THE SPECIES?
+region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?
 
@@
 
-[paragraph on random effects]
+Even though the DM models and our extentions are conceptually simple, even in
+their most stripped down form they include many random effects (the $N$'s, which
+incorporate demographic stochasticity), making it challenging to compute the likelihood 
+or implement MCMC algorithms. Nonetheless, in some cases additional random effects
+may be of interest.  We have shown a few approaches for incorporating some of these
+(environmental stochasticity and random observer effects).  Other approaches to modeling
+each of these may also be worth trying.  Incorporating these random effects can increase
+the difficulty of achieving model convergence, but they are certainly important.
+[Something about how much including each of these random effects changes ovenbird estimates.]
 
@@
 that multiple other options are available as well, such as removal,
-time of detection, and distance modeling ([cites]). These could be
+multiple observer, and distance modeling \citep{williams_etal:2002}. These could be
 accomplished by extending the open population $N$-mixture model
@@
 themselves if we think about movement in a metapopulation
-context ([cite]): immigration depending on total or mean
+context \citep{hanski:1998}: immigration depending on total or mean
 abundance across sites the previous time step; immigration
@@
 depending on abundance at sites as a function of their distance
-from the receiving site (LOTS OF CITATIONS). These models would
+from the receiving site \citep{hastings:1991,hanski:1998}. These models would
 be based on the assumptions that either the study sites
@@
 Bayesian framework.
-
+% Could use a better reference than Hastings 1991 for the total abundance 
+% immigration models, but haven't been able to find one.  And haven't found
+% any references for nearby site immigration (maybe that's because it's not a great idea).
 \begin{comment}
@@
 as functions of habitat covariates, which we did not do here.
-\hl{[Alternatively, could put discussions of our results where
+\begin{comment}
+[Alternatively, could put discussions of our results where
 they fit in by subject area, rather than separately. For
@@
 and golden-winged warbler ZIP results in ZIP
-paragraph.]}
+paragraph.]
+\end{comment}
 

### 2013-02-14 Dail_Mad_ext.tex
Several minor edits

@@
 We expand this class of models to accommodate classical growth
-models (e.g. exponential, Ricker-logistic, etc $\dots$), zero-inflation,
+models (e.g. exponential and Ricker-logistic), zero-inflation,
 and random effects such as observer-specific
@@
 
-In this paper, we focus on the model of \citet[henceforth the DM model;]{dail_madsen:2011}
+In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
 that simultaneous resolves each of the aforementioned problems with
-traditional state-space models, and is designed for simple countdata.
+traditional state-space models, and is designed for simple count data.
 In the following section, we describe the DM model in its original
@@
 used, an extra dimension is added so that we have $X_{ijt}:
-j=1,\hdots,J$. Typically, we $X_{it}$ will be lower than the
+j=1,\hdots,J$. Typically, we know $X_{it}$ will be lower than the
 actual quantity of interest, abundance,
@@
 \end{equation}
+% Maybe we don't need the Poisson distriubtion equation? - Jeff
 
@@
 model assumes that individuals are missed due to imperfect
-detection. The simplest model for imperfect detection, and the
-one considered in the original formulation, is
+detection. The simplest model for imperfect detection is
 \begin{equation}
@@
 \end{equation}
-Under this model, $\bm{p}$ are interpretted as detection
-probability
-parameters...
+Under this model, $\bm{p}$ are interpreted as detection
+probability parameters.
 
@@
 example here is \textbf{BUGS} code for the ``geometric'' model described
-shown
 in Eq.~\ref{eq:hm}:
@@
     N[i,t] <- S[i,t-1] + G[i,t-1]
-    n[i,t] ~ dbin(p, N[i,1])
+    n[i,t] ~ dbin(p, N[i,t])
     }
@@
 
-%% Is this here on purpose?  We haven't introduced it, and I was referring to 
-%% the speed-up these models provide in the maximum-likelihood framework
-%% (even the classic DM models don't require the convolution in the Bayesian framework, right?).
-\begin{verbatim}
-model {
-lambda ~ dunif(0, 5)
-r ~ dunif(-5, 5)
-p ~ dunif(0,1)
-for(i in 1:nSites) {
-  N[i,1] ~ dpois(lambda)
-  y[i,1] ~ dbin(p, N[i,1])
-  for(t in 2:nYears) {
-    mu[i,t-1] = N[i,t-1]*exp(r)
-    N[i,t] ~ dpois(mu[i,t-1])
-    n[i,t] ~ dbin(p, N[i,1])
-    }
-  }
-}
-\end{verbatim}
-
 \subsection{Immigration models}
@@
 were similar when
-run in the Bayesian framework, except that the estimate of r
+run in the Bayesian framework, except that the estimate of $r$
 more than halved (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
 Bayesian model estimates we present mean and SD]). When random observer effects
-were added, estimates for $\Lambda$ and K increased dramatically
+were added, estimates for $\Lambda$ and $K$ increased dramatically
 (from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
 respectively), and the estimate for r was intermediate (0.018 $\pm$ 0.006). The
-estimate of the intercept for p (on the logit scale) dropped from -1.5 $\pm$
-0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.
+estimate of the intercept for $p$ (on the logit scale) dropped from -1.5 $\pm$
+0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  When
+environmental stochasticity was added, the estimate for $K$ increased (to 112 $\pm$ 58.8) and $r$ was
+again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
+on estimate of $p$ or $\Lambda$. The highest estimates of $\Lambda$ and $K$ came 
+from the model that included environmental stochasticity and random observer effects (43.8 $\pm$
+7.9 and 141.5 $\pm$ 69.5, respectively).  
 Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
 provided some evidence for a lack of convergence in estimates of the
-intercept for p for all Bayesian models ran.
+intercept for $p$ for all Bayesian models ran.
 
@@
 for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
-estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. Adding zero-inflation to the dynamics (as well as
+estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. Adding environmental stochasticity
+also increased the estimate of $\Lambda$ (to 42.9 $\pm$ 17.9) and the model with
+both random effects had the highest estimate of $\Lambda$ (74.6 $\pm$ 32.9). 
+Adding zero-inflation to the dynamics (as well as
 the initial abundance) has no effect for exponential models, because populations
@@
 zero-inflated dynamics on the Exponential + Immigration model. Adding zero-inflation to the dynamics
-increased the estimate of ι from 0 to 0.091 $\pm$ 0.049, but had only minor
+increased the estimate of $\iota$ from 0 to 0.091 $\pm$ 0.049, but had only minor
 effects on the other parameters. Gelman and Rubin diagnostics and visual
 examination of the chain trajectory and density plots provided evidence for a lack of convergence in
-estimates of $\Lambda$ for all Bayesian models ran, and for most parameters when random
+estimates of $\Lambda$ for all Bayesian models ran, and for most parameters when random observer
 effects were included.
@@
 the difficulty of achieving model convergence, but they are certainly important.
-[Something about how much including each of these random effects changes ovenbird estimates.]
+For example, the estimate of carrying capacity for the ovenbird increased by more than 150\%
+when environmental stochasticity and random observer effects were included.
 

### 2013-02-26 Dail_Mad_ext.tex
Several minor edits2

@@
 
-\author{Richard B. Chandler  \\
-        \small USGS Patuxent Wildlife Research Center \\
-        \small Laurel, MD 20708
-   \and Jeffrey A. Hostetler \footnote{Contact:
-     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
+\author{Jeffrey A. Hostetler\\ %\footnote{Contact:
+%     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
 %        \small Migratory Bird Center \\
@@
         \small National Zoological Park, MRC 5503 \\
-        \small Washington, DC 20013-7012}
+        \small Washington, DC 20013-7012\\
+   \and Richard B. Chandler \footnote{Contact:
+     \url{RChandler@usgs.gov}, +001-301-497-5856} \\
+        \small USGS Patuxent Wildlife Research Center \\
+        \small Laurel, MD 20708}
 \maketitle
@@
 methods, and code to fit these models using the \textbf{R} package
-\texttt{unmarked} or \textbf{JAGS} is also provided.
+\texttt{unmarked} and \textbf{JAGS} is also provided.
 A simulation study indicated that bias was negligible and coverage
@@
 overdispersion determined by the parameter $\alpha$.
-In the Poisson case, the equivalent expression of
-$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
+%In the Poisson case, the equivalent expression of
+%$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
 %$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
-\begin{equation}
-[N_{i1} | \Lambda] =
-\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
-  \label{eq:N1br}
-\end{equation}
+%\begin{equation}
+%[N_{i1} | \Lambda] =
+%\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
+%  \label{eq:N1br}
+%\end{equation}
 % Maybe we don't need the Poisson distriubtion equation? - Jeff
@@
 \end{equation}
-or equivalently,
-\begin{equation}
-[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
-\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
-  \label{eq:1}
-\end{equation}
-Under this model, $\bm{p}$ are interpreted as detection
-probability parameters.
+%or equivalently,
+%\begin{equation}
+%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
+%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
+%  \label{eq:1}
+%\end{equation}
+Under this model, $\bm{p}$ is interpreted as the detection
+probability parameter.
 
@@
 \end{multline}
-where $\infty$ is replaced by a finite upper bound $K$, which
+where $\infty$ is replaced by a finite upper bound $N_{max}$, which
 %%% Can we rename the upper bound?  We're also using K
@@
 maximize the likelihood shown in Eq.\ref{eq:like}, especially
-when the upper bound of the summation must be high, e.g. $K>200$.
+when the upper bound of the summation must be high, e.g. $N_{max}>200$.
 Furthermore, MCMC may be the only possible option for estimating parameters
@@
 \textbf{JAGS} overcome the technical difficulties by allowing
-users to specify the model using a simple symbolic descriptions. For
-example here is \textbf{BUGS} code for the ``geometric'' model described
-in Eq.~\ref{eq:hm}:
-\begin{verbatim}
-model {
-lambda ~ dunif(0, 5)
-omega ~ dunif(0,1)
-gamma ~ dunif(0, 10)
-p ~ dunif(0,1)
-for(i in 1:nSites) {
-  N[i,1] ~ dpois(lambda)
-  y[i,1] ~ dbin(p, N[i,1])
-  for(t in 2:nYears) {
-    S[i,t-1] ~ dbin(omega, N[i,t-1])
-    G[i,t-1] ~ dpois(gamma*N[i,t-1])
-    N[i,t] <- S[i,t-1] + G[i,t-1]
-    n[i,t] ~ dbin(p, N[i,t])
-    }
-  }
-}
-\end{verbatim}
+users to specify the model using a simple symbolic descriptions.
+
+%For example here is \textbf{BUGS} code for the ``geometric'' model described
+%in Eq.~\ref{eq:hm}:
+%\begin{verbatim}
+%model {
+%lambda ~ dunif(0, 5)
+%omega ~ dunif(0,1)
+%gamma ~ dunif(0, 10)
+%p ~ dunif(0,1)
+%for(i in 1:nSites) {
+%  N[i,1] ~ dpois(lambda)
+%  y[i,1] ~ dbin(p, N[i,1])
+%  for(t in 2:nYears) {
+%    S[i,t-1] ~ dbin(omega, N[i,t-1])
+%    G[i,t-1] ~ dpois(gamma*N[i,t-1])
+%    N[i,t] <- S[i,t-1] + G[i,t-1]
+%    n[i,t] ~ dbin(p, N[i,t])
+%    }
+%  }
+%}
+%\end{verbatim}
 
@@
 in avian point count surveys such as the BBS 
-\citep{robbins_etal:1986,diefenbach_etal:2003,sauer_etal:1994auk,alldredge_etal:2007auk,campbell_francis:2011}.
-Estimating a separate detection probability for each observer
-can be difficult and reduces one's ability to estimate the quantities
-of interest.  This problem is compounded by the fact that
-observers differ greatly in the number of surveys they have
-run (so that many observers' separate detection probabilities
-could not be accurately estimated).
+\citep{robbins_etal:1986,sauer_etal:1994auk,diefenbach_etal:2003,campbell_francis:2011}.%,alldredge_etal:2007auk
+%Estimating a separate detection probability for each observer
+%can be difficult and reduces one's ability to estimate the quantities
+%of interest.  This problem is compounded by the fact that
+%observers differ greatly in the number of surveys they have
+%run (so that many observers' separate detection probabilities
+%could not be accurately estimated).
 
 Current BBS trend estimators deal with this problem by
-treating observer identity as a random (as opposed to a fixed)
+treating observer identity as a random %(as opposed to a fixed)
 effect \citep{link_sauer:2002,sauer_link:2011}.
 %(Link and Sauer 2002, Sauer and Link 2011).
-This allows observer-specific differences in detection probability,
-but assumes that observers are selected at random from a pool
-of potential observers.  Models that contain both random and
-fixed effects are referred to as mixed models.  Often in mixed
-models (as in this case), the random effect is modeled not
-because it of interest in itself, but to avoid bias in the
-estimates of the fixed effects.
-
+%This allows observer-specific differences in detection probability,
+%but assumes that observers are selected at random from a pool
+%of potential observers.  Models that contain both random and
+%fixed effects are referred to as mixed models.  Often in mixed
+%models (as in this case), the random effect is modeled not
+%because it of interest in itself, but to avoid bias in the
+%estimates of the fixed effects.
+%
 To include random observer effects in DM models and these
@@
 We have implemented random observer effects in the Bayesian
-framework using program \textbf{JAGS} \citep[version 3.2.0]{plummer:2003}
+framework using program \textbf{JAGS} \citep[version 3.2.0, see Appendix]{plummer:2003}.
 % (version 3.2.0; Plummer 2003)
-with the \textbf{R} package \texttt{rjags} \citep{plummer:2011,R-2012} interface.
+%with the \textbf{R} package \texttt{rjags} \citep{plummer:2011,R-2012} interface.
 
@@
 The classical approach to computing $[N^+_t]$ is based upon
-Empirical Bayes methods. Note that neither Empiriis not actually a
-Bayesian method, it is used to estim.
-We take the parameter estimates and plug them back in.... For
-example to obtain the posterior distribution for $[N_{i1}]$
-\begin{equation}
-[\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},
-\hat{p}]}{\sum_{N_{i1}=0}^K
-[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
-  \label{eq:eb1}
-\end{equation}
-Once we have estimated the posterior distribution for the first
-time period, we can then proceed to estimate the posterior
-distribution for subsequent time periods using recursion.
-\begin{equation}
-  [\hat{N}_{it}] =
-  \frac{[N_{it}|N_{it-1},\hat{\bm{\Theta}}][X_{it}|N_{it},
-\hat{p}]}{\sum_{N_{it}=0}^K [N_{it}|N_{it-1},
-\hat{\Theta}][X_{it}|N_{it},\hat{p}]}
-  \label{eq:eb2}
-\end{equation}
-
-Note that when $X_{it}$ is not observed, including for future
-time periods, the distribution $[X_{it}|N_{it}]$ is taken to be
-uniform such that it has no influence on the posterior. In other words,
-the posterior for years with no observed data is determined entirely
-by the estimated posterior from the previous year and the
-population dynamics parameters.
-
+Empirical Bayes methods [CITE]. 
+%Note that neither Empiriis not actually a
+%Bayesian method, it is used to estim.
+%We take the parameter estimates and plug them back in.... For
+%example to obtain the posterior distribution for $[N_{i1}]$
+%\begin{equation}
+%[\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},
+%\hat{p}]}{\sum_{N_{i1}=0}^K
+%[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
+%  \label{eq:eb1}
+%\end{equation}
+%Once we have estimated the posterior distribution for the first
+%time period, we can then proceed to estimate the posterior
+%distribution for subsequent time periods using recursion.
+%\begin{equation}
+%  [\hat{N}_{it}] =
+%  \frac{[N_{it}|N_{it-1},\hat{\bm{\Theta}}][X_{it}|N_{it},
+%\hat{p}]}{\sum_{N_{it}=0}^K [N_{it}|N_{it-1},
+%\hat{\Theta}][X_{it}|N_{it},\hat{p}]}
+%  \label{eq:eb2}
+%\end{equation}
+%
+%Note that when $X_{it}$ is not observed, including for future
+%time periods, the distribution $[X_{it}|N_{it}]$ is taken to be
+%uniform such that it has no influence on the posterior. In other words,
+%the posterior for years with no observed data is determined entirely
+%by the estimated posterior from the previous year and the
+%population dynamics parameters.
+%
 If covariates of the dynamics parameters are present, the problem
@@
 When run in a maximum likelihood framework, these models require a
-maximum abundance to integrate over
+a finite upper bound $N_{max}$,
 \citep{royle:2004biom,dail_madsen:2011};
@@
 %the minimum AIC dynamics model.
-As above, these models require a maximum abundance to integrate
+As above, these models require a a finite upper bound ($N_{max}$) to integrate
 over when run in a maximum likelihood
@@
 analyses in a Bayesian framework.  We added random observer 
-effects and, where appropriate, zero-inflated dynamics.  We used
+effects and, where appropriate, environmental stochasticity or zero-inflated dynamics.  We used
 non-informative priors. We tested for lack of convergence using
@@
 population dynamics. We tested these models in both the
-classical and Bayesian frameworks and found strengths and
+frequentist and Bayesian frameworks and found strengths and
 weaknesses of each approach. The Bayesian analyses permit more

### 2013-02-26 supp/code_samples.tex
Several minor edits3

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[12pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

%%% END Article customizations

%%% The "real" document content comes below...

\title{Code Samples}
%\author{The Author}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle

\section{\texttt{unmarked} code}

\subsection{Exponential model}
\subsection{Ricker-logistic model}
\subsection{Gompertz-logistic model}
\subsection{Exponential plus immigration model}
\subsection{Ricker-logistic plus immigration model}
\subsection{Gompertz-logistic plus immigration model}
\subsection{Autoregressive plus immigration model}
\subsection{Zero-inflated Poisson initial abundance model}
\subsection{Simulation study code}
\subsection{BBS analysis code}

\section{\textbf{JAGS} code}

\subsection{Constant model}
\subsection{Exponential model}
\subsection{Ricker-logistic model}
\subsection{Gompertz-logistic model}
\subsection{Exponential plus immigration model}
\subsection{Ricker-logistic plus immigration model}
\subsection{Gompertz-logistic plus immigration model}
\subsection{Autoregressive plus immigration model}
\subsection{Zero-inflated Poisson dynamics model}
\subsection{Environmental stochasticity models}
\subsection{Forecast model}
\subsection{Simulation study code}
\subsection{BBS analysis code}

\end{document}

### 2013-03-26 Dail_Mad_ext.tex
Some trimming

@@
 The DM model requires count data collected at $R$ sites, each of
-which is surveyed on $T$ primary sampling periods. A site is ideally a
-well-defined region of a study area, such as a wetland or a
-patch of early-successional habitat, although it could be an arbitrarily
-defined region such as a randomly located survey plots. The time
-frame of the study is arbitrary, but in general
-it will be sufficiently long to allow for temporal variation in
-abundance. For instance, a primary sampling period could be a
-one month long breeding season, which could be surveyed once per
-year for $T$ years. If the a site is surveyed on $J$ occasions during a
-primary period, these are called secondary sampling occasions.
-In the case where no secondary sampling was conducted, i.e.
-$J=1$, let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
-at site $i$ and primary period $t$. If secondary sampling periods were
-used, an extra dimension is added so that we have $X_{ijt}:
-j=1,\hdots,J$. Typically, we know $X_{it}$ will be lower than the
+which is surveyed on $T$ primary sampling periods. 
+%A site is ideally a
+%well-defined region of a study area, such as a wetland or a
+%patch of early-successional habitat, although it could be an arbitrarily
+%defined region such as a randomly located survey plots. The time
+%frame of the study is arbitrary, but in general
+%it will be sufficiently long to allow for temporal variation in
+%abundance. For instance, a primary sampling period could be a
+%one month long breeding season, which could be surveyed once per
+%year for $T$ years. If the a site is surveyed on $J$ occasions during a
+%primary period, these are called secondary sampling occasions.
+%In the case where no secondary sampling was conducted, i.e. $J=1$, 
+Let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
+at site $i$ and primary period $t$. 
+%If secondary sampling periods were
+%used, an extra dimension is added so that we have $X_{ijt}:
+%j=1,\hdots,J$. 
+Typically, we know $X_{it}$ will be lower than the
 actual quantity of interest, abundance,
@@
 The first two processes describe the state process---the
-variation in abundance in space and time. The third process
-simply describes the the relationship between
+variation in abundance in space and time. The third process %simply 
+describes the the relationship between
 abundance and the observed count data.
@@
 abundance in time $t-1$, i.e. abundance at each site evolves as a
-first order Markovian process, although higher order processes are also possible.
+first order Markovian process. %, although higher order processes are also possible.
 \citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
@@
 
-To make the connection between conventional state-space models
-and the DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
-Eq.~\ref{eq:ss2b}, with an expression derived from
-Eq.~\ref{eq:Nt}. This requires summing over all possible values
-of $S$ and $G$, which is accomplished using the discrete convolution:
-\begin{equation}
-[N_{it}|N_{it-1}, \omega, \gamma] =
-  \sum_{S_{it-1}=0}^{\min(N_{it},N_{it-1})}
-\binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}}
-\times
-\frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}  \label{eq:P}
-\end{equation}
-where here we assumed no dependence of $\gamma$ on $N_{it-1}$.
-
-In the absence of movement, $\omega$ is exactly the probability of surviving
-from year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate
-(under the geometric growth model). This ability to directly estimate
-demographic parameters from simple count data is one of the DM
-model's most appearling features. However,
-in the more common scenario when immigration and emigration
-occur, $\omega$ and $\gamma$ no longer can be interpreted as vital
-rates. Rather, $\omega$ is the probability of surviving and not
-emigrating, and $\gamma$ is the sum of the birth rate and the immigration rate,
-i.e. the recruitment rate. In such cases, we propose
-replacing Eq.~\ref{eq:Nt} with standard population models.
+%To make the connection between conventional state-space models
+%and the DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
+%Eq.~\ref{eq:ss2b}, with an expression derived from
+%Eq.~\ref{eq:Nt}. This requires summing over all possible values
+%of $S$ and $G$, which is accomplished using the discrete convolution:
+%\begin{equation}
+%[N_{it}|N_{it-1}, \omega, \gamma] =
+%  \sum_{S_{it-1}=0}^{\min(N_{it},N_{it-1})}
+%\binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}}
+%\times
+%\frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}  \label{eq:P}
+%\end{equation}
+%where here we assumed no dependence of $\gamma$ on $N_{it-1}$.
+%
+%In the absence of movement, $\omega$ is exactly the probability of surviving
+%from year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate
+%(under the geometric growth model). This ability to directly estimate
+%demographic parameters from simple count data is one of the DM
+%model's most appearling features. However,
+%in the more common scenario when immigration and emigration
+%occur, $\omega$ and $\gamma$ no longer can be interpreted as vital
+%rates. Rather, $\omega$ is the probability of surviving and not
+%emigrating, and $\gamma$ is the sum of the birth rate and the immigration rate,
+%i.e. the recruitment rate. In such cases, we propose
+%replacing Eq.~\ref{eq:Nt} with standard population models.
 
@@
 
-Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
-i.e. the model for abundance dynamics. This state
-model has allowed us to replace the initial abundance
-distribution and transition models of conventional state-space models, with
-alternatives that (1) allow for modeling spatial variation in
-abundance, (2) respect the discrete nature of the counts, and
-(3) have parameters that are estimable. The one
+%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
+%i.e. the model for abundance dynamics. This state
+%model has allowed us to replace the initial abundance
+%distribution and transition models of conventional state-space models, with
+%alternatives that (1) allow for modeling spatial variation in
+%abundance, (2) respect the discrete nature of the counts, and
+%(3) have parameters that are estimable.
+The one
 remaining task is to replace the observation model
@@
 
-The entire model can now be written:
-\begin{gather}
-  N_{i1} \sim \text{Pois}(\Lambda) \nonumber \\
-S_{it}|N_{it-1} \sim \mathrm{Bin}(N_{it-1}, \omega) \nonumber \\  
-G_{it}|N_{it-1}\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
-  N_{it} = S_{it}+G_{it} \nonumber \\
-  X_{it} \sim \text{Bin}(N_{it}, p) \nonumber
-  \label{eq:hm}
-\end{gather}
-
-\section{Statistical Inference}
-
-\subsection{Maximum likelihood}
-
-Maximum likelihood estimation of parameters in models with
-discrete random effects requires removing them by
-summation. For Markovian models, this can be done recursively,
-making the likelihood tractable. A general expression of the likelihood
-can be written:
-\begin{multline}
-  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
-  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
-  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
-%  \prod_{t=T}^2 \left\{
-  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
-  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
-  \right\} \right\}
-  \label{eq:like}
-\end{multline}
-where $\infty$ is replaced by a finite upper bound $N_{max}$, which
+%The entire model can now be written:
+%\begin{gather}
+%  N_{i1} \sim \text{Pois}(\Lambda) \nonumber \\
+%S_{it}|N_{it-1} \sim \mathrm{Bin}(N_{it-1}, \omega) \nonumber \\  
+%G_{it}|N_{it-1}\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
+%  N_{it} = S_{it}+G_{it} \nonumber \\
+%  X_{it} \sim \text{Bin}(N_{it}, p) \nonumber
+%  \label{eq:hm}
+%\end{gather}
+
+%\subsection{Statistical Inference}
+
+%\subsubsection{Maximum likelihood}
+
+%Maximum likelihood estimation of parameters in models with
+%discrete random effects requires removing them by
+%summation. For Markovian models, this can be done recursively,
+%making the likelihood tractable \citep{dail_madsen:2011}. 
+%A general expression of the likelihood can be written:
+%\begin{multline}
+%  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
+%  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
+%  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
+%%  \prod_{t=T}^2 \left\{
+%  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
+%  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
+%  \right\} \right\}
+%  \label{eq:like}
+%\end{multline}
+%where $\infty$ is replaced by a finite upper bound $N_{max}$, which
 %%% Can we rename the upper bound?  We're also using K
 %%% for equilibrium abundance.
-should be high enough such that the MLEs are not affected by increasing
-it further. This likelihood can be maximized numerically using the
+%should be high enough such that the MLEs are not affected by increasing
+%it further. 
+%The likelihood can be maximized numerically using the
+%\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
+%are given in the Appendix.
+
+\subsection{Bayesian inference}
+
+\citet{dail_madsen:2011} demonstrated how to fit these models using maximum 
+likelihood estimation. The likelihood can be maximized numerically using the
 \textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
@@
 
-\subsection{Bayesian inference}
-
 Bayesian inference has several appealing features. First, it
 allows direct probability statements to be made about a hypothesis
-given data. [EXPLAIN POSTERIORS AND PRIORS]
+given data \citep{link_barker:2010}.    
 Second, Bayesian methods offer straight-forward approaches for
@@
 chain Monte Carlo (MCMC) are used to simulate the posterior
-distributions. Although MCMC can be slow, our experience is that it
-may actually converge faster than the amount of time required to
-maximize the likelihood shown in Eq.\ref{eq:like}, especially
-when the upper bound of the summation must be high, e.g. $N_{max}>200$.
-Furthermore, MCMC may be the only possible option for estimating parameters
+distributions. %Although MCMC can be slow, our experience is that it
+%may actually converge faster than the amount of time required to
+%maximize the likelihood shown in Eq.\ref{eq:like}, especially
+%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
+%Furthermore, 
+MCMC may be the only possible option for estimating parameters
 using some of the extensions described below, such as allowing for
@@
 
-Writeing custom MCMC algorthims is often tedious and foreign to
-ecologists, but software packages such as \textbf{WinBUGS} and
-\textbf{JAGS} overcome the technical difficulties by allowing
-users to specify the model using a simple symbolic descriptions.
+Writing custom MCMC algorthims is often tedious and foreign to
+ecologists, but software packages such as %\textbf{WinBUGS} and
+\textbf{JAGS} \citep{plummer:2003} overcome the technical 
+difficulties by allowing users to specify the model using a simple symbolic 
+descriptions. Examples are given in the Appendix.
 
@@
 This,
-like the autoregressive version of the model, is a variant on a simple
+like the geometric-recruitment version of the model, is a variant on a simple
 density-independent exponential model of population growth.
@@
 
-The autoregressive, exponential, Ricker-logistic, and Gompertz-logistic
+The geometric-recruitment, exponential, Ricker-logistic, and Gompertz-logistic
 versions of the DM models all share a common feature (or bug):
@@
 growth and immigration).  This model is close to
-the constant DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
+the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
 of $\omega$ and $\iota$ instead of $\gamma$, except that the first
-process is Poisson distributed instead of binomial. The autoregressive, Ricker-logistic,
+process is Poisson distributed instead of binomial. The geometric-recruitment, Ricker-logistic,
 and Gompertz-logistic models can be extended to allow for immigration in the
-same way.
+same way.   Deterministic versions of the Ricker-logistic + immigration and the 
+Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
+refer to $K$ in these models as the semi-equilibrium abundance.
 \begin{comment}
@@
 the
-Gompertz-logistic + immigration model as: (13) and the autoregressive +  immigration model as: (14)
+Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
 \end{comment}
@@
 initial abundance but also to recruitment and population growth
-terms. For example, the recruitment term of the constant DM model
+terms. For example, the recruitment term of the constant-recruitment DM model
 (Eq.~\ref{eq:Nt}) can be modified as follows:
@@
 dynamics stage (see code sample?).  Adding zero-inflation to the
-dynamics of the autoregressive, exponential, Ricker-logistic, or Gompertz-logistic
+dynamics of the geometric-recruitment, exponential, Ricker-logistic, or Gompertz-logistic
 models
@@
 zero-inflated dynamics in the Bayesian framework using program
-\textrm{JAGS} \citep[version 3.2.0]{plummer:2003}.
+\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
 %(version 3.2.0; Plummer 2003)
@@
 as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
-function of temperature and $K$ a function of rainfall:
+function of temperature ($T$) and $K$ a function of rainfall ($R$):
 \begin{gather}
-r_{it} = \exp(r_{0} + r_{temp}T_{it}) \nonumber \\
-K_{it} = \exp(K_{0} + K_{rain}R_{it})  \\
-N_{it} \sim
-\text{Pois}(N_{it-1}\exp(r_{it-1}(1-\log(N_{it-1}+1)/\log(K_{it-1}+1)))) \nonumber
+\log(r_{it}) = \beta^{r}_{0} + \beta^{r}_{1}T_{it} \nonumber \\
+\log(K_{it}) = \beta^{K}_{0} + \beta^{K}_{1}R_{it}  
+%N_{it} \sim
+%\text{Pois}(N_{it-1}\exp(r_{it-1}(1-\log(N_{it-1}+1)/\log(K_{it-1}+1)))) \nonumber
 \label{eq:weather}
 \end{gather}
-where $T_{it}$ is site and year specific temperature, $R_{it}$ is site and year specific rainfall,
-and the errors for the coefficients $r_{0}$, $r_{temp}$, $K_{0}$, and $K_{rain}$ are 
-normally distributed.
 
@@
 State-space models allow for inference about future population size,
-a critical component of population viability analysis as well
+a critical component of population viability analysis (PVA) as well
 as conservation planning. Let $N^+_{t}$ denote population size at some
@@
 
-The classical approach to computing $[N^+_t]$ is based upon
-Empirical Bayes methods [CITE]. 
+%The classical approach to computing $[N^+_t]$ is based upon
+%Empirical Bayes methods [CITE]. 
 %Note that neither Empiriis not actually a
@@
 %
-If covariates of the dynamics parameters are present, the problem
-becomes much harder because, to fully account for uncertainty,
-we need a model for the future values of the covariates as well. When
-the covariates are continuous, this requires integrating over their
-possible values, which becomes very challenging
-computationally. Indeed, simulation methods such as MCMC are the
-only viable options when numerous covariates are present, which will
-often be the case in many ecological studies.
-
-Relative to the computational challenges associated with the
-empirical Bayes methods, the Bayesian approach is much more easy to
+%If covariates of the dynamics parameters are present, the problem
+%becomes much harder because, to fully account for uncertainty,
+%we need a model for the future values of the covariates as well. When
+%the covariates are continuous, this requires integrating over their
+%possible values, which becomes very challenging
+%computationally. Indeed, simulation methods such as MCMC are the
+%only viable options when numerous covariates are present, which will
+%often be the case in many ecological studies.
+
+%% COULD INCLUDE THIS PARAGRAPH, OR NOT.
+%A standard approach to PVA uses previously derived estimates of current 
+%population size, vital rate or population growth parameters, and environmental
+%stochasticity to stochastically project the population forward and calculate probabilities
+%of extinction or other measures of risk.  Estimates from DM models and the extensions
+%given here can be used in this context.  Uncertainty in extinction or quasi-extinction probabilities
+%can be estimated by either drawing values for parameters from their posterior distributions
+%(if estimated using MCMC) or randomly drawing values for parameters using their means and standard
+%errors (if estimated using maximum likelihood), and running a number of simulations with each
+%set of parameter values \citep{white:2000,hostetler_etal:2012}.
+
+%Relative to the computational challenges associated with the
+%empirical Bayes methods, 
+A fully Bayesian approach to PVA is straightforward to
 implement, especially since MCMC will be generally be relied upon in the
-first place. Furthermore, because population size at any future point
+first place. Because population size at any future point %Furthermore, b
 in time is regarded as an unknown parameter like any other, computing the
@@
 AIC model from that set to test eight additional models of dynamics:
-constant, autoregressive, Ricker-logistic, Gompertz-logistic, autoregressive +
+constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, geometric-recruitment +
 immigration, exponential + immigration, Ricker-logistic +
@@
 analyses in a Bayesian framework.  We added random observer 
-effects and, where appropriate, environmental stochasticity or zero-inflated dynamics.  We used
+effects and environmental stochasticity (modeled as lognormal
+variation in regional population growth rate, Eq.~\ref{eq:nuRand}) and, where appropriate, zero-inflated dynamics.  We used
 non-informative priors. We tested for lack of convergence using
@@
 For each chain we sampled the MCMC for at least 40,000 iterations, after at
-least 3,000 tuning samples.
+least 3,000 adaptation and tuning samples.
 
@@
 year specific abundance estimates.  Because of the memory requirements involved,
-we obtained these estimates from a single chain of 5,000 iterations, after 35,000
-tuning samples.  We present these estimates for ovenbirds only.
+we obtained these estimates from a single chain.  We present these estimates for ovenbirds only.
 %Latent inclusion factors for temperature effects?
@@
 initial abundance over the Poisson and zero-inflated Poisson
-(Table 2A). Even compared to the Poisson, there was
-no evidence to support a zero-inflation factor for this species.
-The best supported model for p was additive effects of wind speed
+(Table 2A). %Even compared to the Poisson, there was
+%no evidence to support a zero-inflation factor for this species.
+The best supported model for $p$ was additive effects of wind speed
 and first run (Table 2B, model B.1). First run and increasing wind
-speeds both decreased p. All dynamics models with immigration were
+speeds both decreased $p$. All dynamics models with immigration were
 better supported than models without immigration (Table 2C); 
 the best supported of these was the Ricker-logistic + immigration.
-The best supported model of those without immigration was the
-Gompertz-logistic, and the autoregressive and the constant model had the least
-support.
+%The best supported model of those without immigration was the
+%Gompertz-logistic, and the geometric-recruitment and the constant model had the least
+%support.
 %There was little support for an effect of minimum June and July
@@
 run in the Bayesian framework, except that the estimate of $r$
-more than halved (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
-Bayesian model estimates we present mean and SD]). When random observer effects
-were added, estimates for $\Lambda$ and $K$ increased dramatically
-(from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
-respectively), and the estimate for r was intermediate (0.018 $\pm$ 0.006). The
-estimate of the intercept for $p$ (on the logit scale) dropped from -1.5 $\pm$
-0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  When
-environmental stochasticity was added, the estimate for $K$ increased (to 112 $\pm$ 58.8) and $r$ was
-again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
-on estimate of $p$ or $\Lambda$. The highest estimates of $\Lambda$ and $K$ came 
-from the model that included environmental stochasticity and random observer effects (43.8 $\pm$
-7.9 and 141.5 $\pm$ 69.5, respectively).  
+more than halved.
+% (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
+%Bayesian model estimates we present mean and SD]). 
+When random observer effects
+were added, estimates for $\Lambda$ and $K$ increased dramatically,
+%(from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
+%respectively)
+and the estimate for r was intermediate. The %(0.018 $\pm$ 0.006)
+estimate of the intercept for $p$ (on the logit scale) dropped. % from -1.5 $\pm$
+%0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
+When environmental stochasticity was added, the estimate for $K$ increased. %(to 112 $\pm$ 58.8) and $r$ was
+%again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
+%on estimate of $p$ or $\Lambda$. 
+The highest estimates of $\Lambda$ and $K$ came 
+from the model that included environmental stochasticity and random observer effects %(43.8 $\pm$
+%7.9 and 141.5 $\pm$ 69.5, respectively).  
 Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
@@
 using the Bayesian model including random observer effects (Fig.~\ref{fig:oven_N}).  
-Estimated average abundance oscillated over time with a slowly increasing trend.  Estimated ovenbird
+Estimated average abundance over time alternated between growing and stable periods.  
+Estimated ovenbird
 abundance was highest in western and eastern Maryland in 1970 and 1980, but became
-more evenly distributed by 2010.  Coefficient of variation (CV) for route and year specific
-abundance estimates varied between 0.11 and 2.31 (mean = 0.46).  By comparison, CV
-for the estimate of mean initial abundance ($\Lambda$) was 0.15.
+more evenly distributed by 2010.  
+%These aren't up to date, and it doesn't look like people use CV as a measure of 
+%uncertainty in MCMC estimates anyway.
+%Coefficient of variation (CV) for route and year specific
+%abundance estimates varied between 0.11 and 2.31 (mean = 0.46).  By comparison, CV
+%for the estimate of mean initial abundance ($\Lambda$) was 0.15.
+As expected, uncertainty in estimates of route and year specific abundance was high.
 
@@
 supported over the Poisson (Table 3A). The best
-supported model for p was an effect of first run (Table 3B, model
+supported model for $p$ was an effect of first run (Table 3B, model
 B.1). The best supported dynamics model was exponential
-(Table 3C, model C.1); the estimate of r from this model was -0.058
-$\pm$ 0.017 (SE). There was no support for immigration added to the
-exponential or autoregressive models (Table 3C, models C.4 and
-C.6).
-However, the Gompertz-logistic + immigration and Ricker-logistic + immigration
-models were supported over the corresponding models without immigration
-(Table 3C, models C.4, C.5, C.7, and C.8), despite low estimates
-of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$
-0.002, respectively).  
+(Table 3C, model C.1). %; the estimate of r from this model was -0.058
+%$\pm$ 0.017 (SE). There was no support for immigration added to the
+%exponential or geometric-recruitment models (Table 3C, models C.4 and
+%C.6).
+%However, the Gompertz-logistic + immigration and Ricker-logistic + immigration
+%models were supported over the corresponding models without immigration
+%(Table 3C, models C.4, C.5, C.7, and C.8), despite low estimates
+%of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$
+%0.002, respectively).  
 %There was considerable support for an effect of
@@
 distributions for golden-winged warbler, we also ran the subsequent models with the
-negative binomial initial abundance. Rankings of subsequent
-models were similar, but with a p covariate (first) the negative
-binomial distribution outranked the ZIP.
+negative binomial initial abundance; rankings of subsequent
+models were similar. %, but with a $p$ covariate (first) the negative
+%binomial distribution outranked the ZIP.
 
@@
 were similar when run in the Bayesian framework. When random
-observer effects were added, the estimate for $\Lambda$ increased dramatically
-(from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
-for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
-estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. Adding environmental stochasticity
-also increased the estimate of $\Lambda$ (to 42.9 $\pm$ 17.9) and the model with
-both random effects had the highest estimate of $\Lambda$ (74.6 $\pm$ 32.9). 
+observer effects were added, the estimate for $\Lambda$ increased dramatically,
+and the estimate of the intercept for $p$ dropped.
+%(from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
+%for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
+%estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. 
+Adding environmental stochasticity also increased the estimate of $\Lambda$ 
+%(to 42.9 $\pm$ 17.9) 
+and the model with both random effects had the highest estimate of $\Lambda$.
+% (74.6 $\pm$ 32.9). 
 Adding zero-inflation to the dynamics (as well as
@@
 zero-inflated dynamics on the Exponential + Immigration model. Adding zero-inflation to the dynamics
-increased the estimate of $\iota$ from 0 to 0.091 $\pm$ 0.049, but had only minor
+increased the estimate of $\iota$, but had only minor % from 0 to 0.091 $\pm$ 0.049
 effects on the other parameters. Gelman and Rubin diagnostics and visual
@@
 
-Estimated detection probabilities were generally low. The estimated probability
-of detecting an ovenbird in the top model (Table 2, model C.1) varied between 0.157
-$\pm$ 0.014 and 0.184 $\pm$ 0.015, depending on wind speed and whether the
-observer had run that route before. The estimated probability of detecting a
-golden-winged warbler went from 0.013 $\pm$ 0.006 for first time observers to 0.028 $\pm$ 0.008 for
-repeat observers for the top model (Table 3, model C.1). Accounting for random
-observer effects decreased average detection probability for both species.
-
-For both species, constant dynamics models estimated very high
+%Estimated detection probabilities were generally low. The estimated probability
+%of detecting an ovenbird in the top model (Table 2, model C.1) varied between 0.157
+%$\pm$ 0.014 and 0.184 $\pm$ 0.015, depending on wind speed and whether the
+%observer had run that route before. The estimated probability of detecting a
+%golden-winged warbler went from 0.013 $\pm$ 0.006 for first time observers to 0.028 $\pm$ 0.008 for
+%repeat observers for the top model (Table 3, model C.1). Accounting for random
+%observer effects decreased average detection probability for both species.
+
+For both species, constant-recruitment dynamics models estimated very high
 survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05 for ovenbird and
 0.935 $\pm$ 0.011 for
-golden-winged warbler). The autoregressive and autoregressive + immigration
+golden-winged warbler). The geometric-recruitment and geometric-recruitment + immigration
 models estimated very low survival probabilities for ovenbird ($\omega$ =
@@
 
-\vfill
+%\vfill
 %\clearpage
@@
 &B.9.
-ZIP[$\Lambda$(.)$\psi$(.)]Autoregressive[$\gamma$(.)$\omega$(.)]$p$(1st)
+ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st)
 &6	&1.7 \\
@@
 A.11.
-NB[$\Lambda$(.)$\alpha$(.)]Autoregressive+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
+NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
 &10	&38.6
@@
 &B.13.
-ZIP[$\Lambda$(.)$\psi$(.)]Autoregressive+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
+ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
 &7	&3.7 \\
@@
 A.15.
-NB[$\Lambda$(.)$\alpha$(.)]Autoregressive[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
+NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
 &9	&273.7
@@
 A.16.
-NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
+NB[$\Lambda$(.)$\alpha$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
 &9	&1856.7
 &B.16.
-ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st)
+ZIP[$\Lambda$(.)$\psi$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st)
 &6	&12.5 \\
@@
 C.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(wind+1st) &9	&36.5\\
-C.4. NB[$\Lambda$(.)$\alpha$(.)]Autoregressive+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st) &10	&38.6\\
+C.4. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st) &10	&38.6\\
 C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&192.8\\
@@
 C.7. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &271.3\\
-C.8. NB[$\Lambda$(.)$\alpha$(.)]Autoregressive[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&273.7\\
-C.9. NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&1856.7\\
+C.8. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&273.7\\
+C.9. NB[$\Lambda$(.)$\alpha$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&1856.7\\
 \hline
@@
 C.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(1st) &5	&0 \\
-C.2. ZIP[$\Lambda$(.)$\psi$(.)]Autoregressive[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&1.7 \\
+C.2. ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&1.7 \\
 C.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(1st) &6	&2.0 \\
@@
 C.5. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(1st)	&6&3.0 \\
-C.6. ZIP[$\Lambda$(.)$\psi$(.)]Autoregressive+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st) &7	&3.7 \\
+C.6. ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st) &7	&3.7 \\
 C.7. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st) &7	&4.2 \\
 C.8. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(1st)	&7 &5.2 \\
-C.9. ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&12.5 \\
+C.9. ZIP[$\Lambda$(.)$\psi$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&12.5 \\
 \hline
@@
   \centering
-  \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year2}
+  \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year4}
 \caption{Maps of BBS route and year specific estimated abundances for ovenbirds, for the years 1970,

### 2013-03-26 Dail_Mad_ext.tex
Trim intro and remove GWWA

@@
 \bibliographystyle{ecology}
+\hyphenpenalty=100000
 
@@
 \textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
-Gompertz-logistic model, immigration, open population point count models,
-random observer effects, range, Ricker-logistic model, zero-inflated
+detectability, Gompertz-logistic model, immigration, open population point 
+count models, random observer effects, Ricker-logistic model, zero-inflated
 
@@
 
-A more serious problem than the ones associated with the Gaussian
-assumptions is that the parameters of the simple state-space models
+A more serious problem %than the ones associated with the Gaussian assumptions 
+is that the parameters of the simple state-space models
 such as Eq~\ref{eq:ss1} are not identifiable in many
@@
 %are needed to fully specify the model.
-To be more specific, a fully-specified state-space model requires at least
-three probability distributions, which we represent
-using bracket notation (i.e., $[Y|\Omega]$ is interpreted
-as the probability distribution of the random variable $Y$ given the
-parameter $\Omega$). The three probability distributions required for a
-state-space model are:
+%To be more specific, a fully-specified state-space model requires at least
+%three probability distributions, which we represent
+%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
+%as the probability distribution of the random variable $Y$ given the
+%parameter $\Omega$). The three probability distributions required for a
+%state-space model are:
 %%%% JEFF, does this extra sentence help? This notation comes from
@@
 %to save space by trimming a bunch of this discussion back down.]}
-\begin{subequations}
-  \label{eq:ss2}
-  \begin{align}
-    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
-[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
-\label{eq:ss2c}
-    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
-  \end{align}
-\end{subequations}
-where $\bm{\theta}$ are the process variation parameters for the
-initial state, $\bm{\Theta}$ are the process variation parameters
-influencing how abundance changes over time, and $\bm{p}$ are the
-observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
-and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
-models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
+%\begin{subequations}
+%  \label{eq:ss2}
+%  \begin{align}
+%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
+%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
+%\label{eq:ss2c}
+%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
+%  \end{align}
+%\end{subequations}
+%where $\bm{\theta}$ are the process variation parameters for the
+%initial state, $\bm{\Theta}$ are the process variation parameters
+%influencing how abundance changes over time, and $\bm{p}$ are the
+%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
+%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
+%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
 %The three lines of general model
@@
 %istribution, $[N_1|\bm{\theta}]$ be?
-And how could $\bm{\theta}$ be estimated given that there is only a single observation
-available? Often, these problems are simply assumed away. For
-example, sometimes researchers assume that there is no process variation in
-the first year, although this seems unjustified. Alternatively, the
-population may be assumed to be at equilibrium such that the expected
-value of $N_t$ is constant through time. Although this 
-makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
-defeats the objective of many studies of population dynamics, namely
-determining why a population varies over time.
+%And how could $\bm{\theta}$ be estimated given that there is only a single observation
+%available? Often, these problems are simply assumed away. For
+%example, sometimes researchers assume that there is no process variation in
+%the first year, although this seems unjustified. Alternatively, the
+%population may be assumed to be at equilibrium such that the expected
+%value of $N_t$ is constant through time. Although this 
+%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
+%defeats the objective of many studies of population dynamics, namely
+%determining why a population varies over time.
 
@@
 As with traditional state-space models, the DM model includes the
-three conditionally related processes shown in Eq.~\ref{eq:ss2},
+three conditionally related processes shown in Eq.~\ref{eq:ss1},
 which correspond to (1) initial abundance, i.e. the
@@
 remaining task is to replace the observation model
-$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be
+$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
 more easily interpreted. The observation
@@
 We applied these models to North American Breeding Bird Survey
-(BBS) data from 1966-2010 for two species in the bordering US states
+(BBS) data from 1966-2010 in the bordering US states
 Maryland and Virginia. For our focal species, we selected
 ovenbirds (\textit{Seiurus aurocapilla}), an abundant and widespread
-forest breeding migrant with a stable or increasing trend in the region \citep{porneluzi_etal:2011}
-and golden-winged warblers (\textit{Vermivora chrysoptera}), which
-only breeds in the western parts of these states and has been declining in the
-region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?
+forest breeding migrant with a stable or increasing trend in the region \citep{porneluzi_etal:2011}.
+%and golden-winged warblers (\textit{Vermivora chrysoptera}), which
+%only breeds in the western parts of these states and has been declining in the
+%region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?
 
@@
 %HOW WE OBTAINED AND MANIPULATED PRISM DATA
-We ran a series of maximum-likelihood based models for each
-species, and then a series of Bayesian models. We started by testing
+We ran a series of maximum-likelihood based models, 
+and then a series of Bayesian models. We started by testing
 three models of initial abundance (Poisson, negative binomial, and
@@
 over when run in a maximum likelihood
-framework; we used 600 for ovenbirds and 350 for golden-winged warblers.
+framework; we used 600. % for ovenbirds and 350 for golden-winged warblers.
 
@@
 effects and environmental stochasticity (modeled as lognormal
-variation in regional population growth rate, Eq.~\ref{eq:nuRand}) and, where appropriate, zero-inflated dynamics.  We used
+variation in regional population growth rate, Eq.~\ref{eq:nuRand}). % and, where appropriate, zero-inflated dynamics.  
+We used
 non-informative priors. We tested for lack of convergence using
@@
 year specific abundance estimates.  Because of the memory requirements involved,
-we obtained these estimates from a single chain.  We present these estimates for ovenbirds only.
+we obtained these estimates from a single chain.  %We present these estimates for ovenbirds only.
 %Latent inclusion factors for temperature effects?
@@
 The highest estimates of $\Lambda$ and $K$ came 
-from the model that included environmental stochasticity and random observer effects %(43.8 $\pm$
+from the model that included environmental stochasticity and random observer effects. %(43.8 $\pm$
 %7.9 and 141.5 $\pm$ 69.5, respectively).  
@@
 
-There was slightly more support for the ZIP distribution for initial
-abundance of golden-winged warbler than for the negative binomial; both were strongly
-supported over the Poisson (Table 3A). The best
-supported model for $p$ was an effect of first run (Table 3B, model
-B.1). The best supported dynamics model was exponential
-(Table 3C, model C.1). %; the estimate of r from this model was -0.058
+%There was slightly more support for the ZIP distribution for initial
+%abundance of golden-winged warbler than for the negative binomial; both were strongly
+%supported over the Poisson (Table 3A). The best
+%supported model for $p$ was an effect of first run (Table 3B, model
+%B.1). The best supported dynamics model was exponential
+%(Table 3C, model C.1). %; the estimate of r from this model was -0.058
 %$\pm$ 0.017 (SE). There was no support for immigration added to the
@@
 %growth model (Table 1B, model B.17), with a negative slope.
-Because of the closeness of rankings for the ZIP and negative binomial
-distributions for golden-winged warbler, we also ran the subsequent models with the
-negative binomial initial abundance; rankings of subsequent
-models were similar. %, but with a $p$ covariate (first) the negative
+%Because of the closeness of rankings for the ZIP and negative binomial
+%distributions for golden-winged warbler, we also ran the subsequent models with the
+%negative binomial initial abundance; rankings of subsequent
+%models were similar. %, but with a $p$ covariate (first) the negative
 %binomial distribution outranked the ZIP.
 
-Estimates for the top ranked model for golden-winged warbler
-%(ZIP[Λ(.)α(.)]Exponential[r(.)]p(1st))
-were similar when run in the Bayesian framework. When random
-observer effects were added, the estimate for $\Lambda$ increased dramatically,
-and the estimate of the intercept for $p$ dropped.
-%(from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
-%for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
-%estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. 
-Adding environmental stochasticity also increased the estimate of $\Lambda$ 
-%(to 42.9 $\pm$ 17.9) 
-and the model with both random effects had the highest estimate of $\Lambda$.
-% (74.6 $\pm$ 32.9). 
-Adding zero-inflation to the dynamics (as well as
-the initial abundance) has no effect for exponential models, because populations
-that started at 0 would stay at 0 in any case. Therefore, we tested
-zero-inflated dynamics on the Exponential + Immigration model. Adding zero-inflation to the dynamics
-increased the estimate of $\iota$, but had only minor % from 0 to 0.091 $\pm$ 0.049
-effects on the other parameters. Gelman and Rubin diagnostics and visual
-examination of the chain trajectory and density plots provided evidence for a lack of convergence in
-estimates of $\Lambda$ for all Bayesian models ran, and for most parameters when random observer
-effects were included.
+%Estimates for the top ranked model for golden-winged warbler
+%%(ZIP[Λ(.)α(.)]Exponential[r(.)]p(1st))
+%were similar when run in the Bayesian framework. When random
+%observer effects were added, the estimate for $\Lambda$ increased dramatically,
+%and the estimate of the intercept for $p$ dropped.
+%%(from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
+%%for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
+%%estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. 
+%Adding environmental stochasticity also increased the estimate of $\Lambda$ 
+%%(to 42.9 $\pm$ 17.9) 
+%and the model with both random effects had the highest estimate of $\Lambda$.
+%% (74.6 $\pm$ 32.9). 
+%Adding zero-inflation to the dynamics (as well as
+%the initial abundance) has no effect for exponential models, because populations
+%that started at 0 would stay at 0 in any case. Therefore, we tested
+%zero-inflated dynamics on the Exponential + Immigration model. Adding zero-inflation to the dynamics
+%increased the estimate of $\iota$, but had only minor % from 0 to 0.091 $\pm$ 0.049
+%effects on the other parameters. Gelman and Rubin diagnostics and visual
+%examination of the chain trajectory and density plots provided evidence for a lack of convergence in
+%estimates of $\Lambda$ for all Bayesian models ran, and for most parameters when random observer
+%effects were included.
 
@@
 
-For both species, constant-recruitment dynamics models estimated very high
-survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05 for ovenbird and
-0.935 $\pm$ 0.011 for
-golden-winged warbler). The geometric-recruitment and geometric-recruitment + immigration
-models estimated very low survival probabilities for ovenbird ($\omega$ =
-0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively), 
-but not for golden-winged warbler ($\omega$ = 0.715 $\pm$ 0.288 for both models).
+%For both species, 
+Constant-recruitment dynamics models estimated very high
+survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05). % for ovenbird and
+%0.935 $\pm$ 0.011 for
+%golden-winged warbler). 
+The geometric-recruitment and geometric-recruitment + immigration
+models estimated very low survival probabilities %for ovenbird 
+($\omega$ =
+0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively). 
+%but not for golden-winged warbler ($\omega$ = 0.715 $\pm$ 0.288 for both models).
 
@@
 ways to accomodate common features of ecological
-datasets, including sparse counts, zero-inflation, demographic
+datasets, including sparse counts, zero-inflation, environmental
 stochasticity, and random variation in observation error.
@@
 In fact, the DM model could be viewed as a specific case of the
-IPM model proposed by \citet{buckland_etal:2004}.
+IPM proposed by \citet{buckland_etal:2004}.
 However, we caution that when count data are affected by
@@
 the difficulty of achieving model convergence, but they are certainly important.
-For example, the estimate of carrying capacity for the ovenbird increased by more than 150\%
+For example, the estimate of semi-equilibrium abundance for the ovenbird increased by more than 150\%
 when environmental stochasticity and random observer effects were included.
@@
 We have presented a test case for these models using data from
-the BBS for two species. We demonstrate that these models can be
+the BBS for ovenbirds. We demonstrate that these models can be
 used to model and estimate initial abundance, detection
@@
 flexibility and allow the incorporation of random observer
-effects and zero-inflated dynamics. Our results suggest ignoring
+effects, environmental stochasticity, and zero-inflated dynamics. 
+Our results suggest ignoring
 variation between observers can lead to underestimation of
 initial abundance and equilibrium population sizes. On the other
-hand, our Bayesian analyses did have convergence problems,
-particularly for the species (golden-winged warbler) with sparser data. It's
-likely that poor model fits are related to the low estimates of
+hand, our Bayesian analyses did have some convergence problems.
+%particularly for the species (golden-winged warbler) with sparser data. 
+It's likely that poor model fits are related to the low estimates of
 detection probability, which in turn are probably caused by the
@@
 
-\begin{table}
-  \centering
-  \small
-  \caption{Model selection table for golden-winged
-warblers in Maryland and Virginia, 1966-2010. We present model
-name and number, number of parameters (Par.), and difference in
-Akaike's information criterion between each model and the top model of
-that set ($\Delta$AIC).  The first section compares
-models for initial abundance, the second for detection
-probability, and the third for dynamics.}
-  \begin{tabular}[h]{lcc}
-\hline
-Model	&Par.	&$\Delta$AIC\\
-\hline
-A. Initial Abundance && \\
-A.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4	&0 \\
-A.2. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0.4 \\
-A.3. P[$\Lambda$(.)]Exponential[$r$(.)]$p$(.)	&3	&37.8 \\
-\hline
-B. Detection Probability && \\
-B.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(1st)	&5 &0 \\
-B.2. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4	&3.5 \\
-B.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &5.9 \\
-B.4. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(wind) &7 &9.3\\
-\hline
-C. Dynamics && \\
-C.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(1st) &5	&0 \\
-C.2. ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&1.7 \\
-C.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(1st) &6	&2.0 \\
-C.4. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st) &7	&2.5 \\
-C.5. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(1st)	&6&3.0 \\
-C.6. ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st) &7	&3.7 \\
-C.7. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st) &7	&4.2 \\
-C.8. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(1st)	&7 &5.2 \\
-C.9. ZIP[$\Lambda$(.)$\psi$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&12.5 \\
-\hline
-\end{tabular}
-\end{table}
-
-\newpage
+%\begin{table}
+%  \centering
+%  \small
+%  \caption{Model selection table for golden-winged
+%warblers in Maryland and Virginia, 1966-2010. We present model
+%name and number, number of parameters (Par.), and difference in
+%Akaike's information criterion between each model and the top model of
+%that set ($\Delta$AIC).  The first section compares
+%models for initial abundance, the second for detection
+%probability, and the third for dynamics.}
+%  \begin{tabular}[h]{lcc}
+%\hline
+%Model	&Par.	&$\Delta$AIC\\
+%\hline
+%A. Initial Abundance && \\
+%A.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4	&0 \\
+%A.2. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0.4 \\
+%A.3. P[$\Lambda$(.)]Exponential[$r$(.)]$p$(.)	&3	&37.8 \\
+%\hline
+%B. Detection Probability && \\
+%B.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(1st)	&5 &0 \\
+%B.2. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4	&3.5 \\
+%B.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &5.9 \\
+%B.4. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(wind) &7 &9.3\\
+%\hline
+%C. Dynamics && \\
+%C.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(1st) &5	&0 \\
+%C.2. ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&1.7 \\
+%C.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(1st) &6	&2.0 \\
+%C.4. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st) &7	&2.5 \\
+%C.5. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(1st)	&6&3.0 \\
+%C.6. ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st) &7	&3.7 \\
+%C.7. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st) &7	&4.2 \\
+%C.8. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(1st)	&7 &5.2 \\
+%C.9. ZIP[$\Lambda$(.)$\psi$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&12.5 \\
+%\hline
+%\end{tabular}
+%\end{table}
+
+%\newpage
 

### 2013-03-29 Dail_Mad_ext.tex
Formatting and further trims

@@
 \documentclass[12pt]{article}
-
+\renewcommand*\abstractname{Summary}
 \usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
@@
 \bibliographystyle{ecology}
-\hyphenpenalty=100000
+%\hyphenpenalty=100000
 
@@
 \abstract{
-Models of population dynamics play a central role in theoretical and
-applied ecology where they are used for purposes such as testing 
+\noindent
+\begin{enumerate}
+\item Models of population dynamics play a central role in theoretical and
+applied ecology, where they are used for purposes such as testing 
 hypotheses about density dependence and predicting species' responses 
@@
 process of interest.
-Conventional state-space models, however, have
+\item Conventional state-space models, however, have
 three important limitations: (1) the parameters are not identifiable
@@
 time-series data that attributes observation error to imperfect detection.
-We expand this class of models to accommodate classical growth
+\item We expand this class of models to accommodate classical growth
 models (e.g. exponential and Ricker-logistic), zero-inflation,
@@
 \texttt{unmarked} and \textbf{JAGS} is also provided.
-A simulation study indicated that bias was negligible and coverage
+\item A simulation study indicated that bias was negligible and coverage
 nominal for the proposed model extensions. An analysis of data from
@@
 replicated counts) is unavailable.
+\end{enumerate}
 }
@@
 
-Bayesian inference has several appealing features. First, it
+However, Bayesian inference has several appealing features. First, it
 allows direct probability statements to be made about a hypothesis
@@
 model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
-rate of increase. Although demographic and environmental stochasticity
-in $r$ could be modeled as functions of covariates or random effects, a
-simpler way of allow for stochasticity is to regard $N_{it}$ as a Poisson random variable, simplifying
-Eq.~\ref{eq:Nt} to:
+rate of increase. %Although demographic and environmental stochasticity
+%in $r$ could be modeled as functions of covariates or random effects, a
+%simpler way of 
+To allow for demographic stochasticity, we can regard $N_{it}$ as a Poisson 
+random variable, simplifying Eq.~\ref{eq:Nt} to:
 \begin{equation}
@@
 positive (this is equivalent to separate Poisson processes for
-growth and immigration).  This model is close to
-the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
-of $\omega$ and $\iota$ instead of $\gamma$, except that the first
-process is Poisson distributed instead of binomial. The geometric-recruitment, Ricker-logistic,
+growth and immigration).  %This model is close to
+%the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
+%of $\omega$ and $\iota$ instead of $\gamma$, except that the first
+%process is Poisson distributed instead of binomial. 
+The geometric-recruitment, Ricker-logistic,
 and Gompertz-logistic models can be extended to allow for immigration in the
 same way.   Deterministic versions of the Ricker-logistic + immigration and the 
-Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
+Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
+are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
 refer to $K$ in these models as the semi-equilibrium abundance.
@@
 in avian point count surveys such as the BBS 
-\citep{robbins_etal:1986,sauer_etal:1994auk,diefenbach_etal:2003,campbell_francis:2011}.%,alldredge_etal:2007auk
+\citep{robbins_etal:1986,sauer_etal:1994auk,campbell_francis:2011}.%,alldredge_etal:2007auk %diefenbach_etal:2003,
 %Estimating a separate detection probability for each observer
@@
 to hear birds \citep{simons_etal:2007}; we tested the effects of wind
-speed on detection probability. BBS volunteers record wind conditions
-at the beginning and end of each route on the Beaufort Scale
-\citep[start and end wind]{robbins_etal:1986}.
-%(start and end wind; Robbins et al. 1986)(start and end wind;0-9; cite).
-When start or end wind was not recorded we imputed those values with
-the start or end mean. We took the mean of start and end wind for
-each route and put this average wind scale value into four
-categories: 0 $\leq$ wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind
-$\geq$ 3 (maximum of 3.5).
+speed on detection probability, using the truncated mean of wind conditions at the start and end of
+each route on the Beaufort scale as a categorical predictor. 
+%BBS volunteers record wind conditions
+%at the beginning and end of each route on the Beaufort Scale
+%\citep[start and end wind]{robbins_etal:1986}.
+%%(start and end wind; Robbins et al. 1986)(start and end wind;0-9; cite).
+%When start or end wind was not recorded we imputed those values with
+%the start or end mean. We took the mean of start and end wind for
+%each route and put this average wind scale value into four
+%categories: 0 $\leq$ wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind
+%$\geq$ 3 (maximum of 3.5).
 Following \citet{link_sauer:2002},
@@
 
-%HOW WE OBTAINED AND MANIPULATED PRISM DATA
-We ran a series of maximum-likelihood based models, 
+We ran a series of frequentist models, 
 and then a series of Bayesian models. We started by testing
@@
 AIC model from that set to test eight additional models of dynamics:
-constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, geometric-recruitment +
-immigration, exponential + immigration, Ricker-logistic +
-immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
+constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, 
+the previous three with immigration, and exponential + immigration.
+%geometric-recruitment +
+%immigration, exponential + immigration, Ricker-logistic +
+%immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
 %minimum temperature for June and July on all dynamics parameters from
@@
 least 3,000 adaptation and tuning samples.  
-
 We also used these models to obtain route and 
@@
 %(Table 2A, models A.17 - A.18).
-
-Estimates for the top ranked model for ovenbird
+%For both species, 
+Constant-recruitment dynamics models estimated very high
+survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05). % for ovenbird and
+%0.935 $\pm$ 0.011 for
+%golden-winged warbler). 
+The geometric-recruitment and geometric-recruitment + immigration
+models estimated very low survival probabilities %for ovenbird 
+($\omega$ =
+0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively). 
+%but not for golden-winged warbler ($\omega$ = 0.715 $\pm$ 0.288 for both models).
+
+Estimates for the top ranked model %for ovenbird
 %(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
@@
 %observer effects decreased average detection probability for both species.
-
-%For both species, 
-Constant-recruitment dynamics models estimated very high
-survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05). % for ovenbird and
-%0.935 $\pm$ 0.011 for
-%golden-winged warbler). 
-The geometric-recruitment and geometric-recruitment + immigration
-models estimated very low survival probabilities %for ovenbird 
-($\omega$ =
-0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively). 
-%but not for golden-winged warbler ($\omega$ = 0.715 $\pm$ 0.288 for both models).
 
@@
 demographic parameters from count data is an ambitious goal, and
-the required assumption will not be valid in many cases,
+the required assumptions will not be valid in many cases,
 especially when immigration and emigration occur. This was supported by the
@@
 are produced by many of the largest monitoring programs in the
-world, such as the North American Breeding Bird Survey
-\citep[BBS;][]{robbins_etal:1986}.
+world, such as the BBS. %North American Breeding Bird Survey
+%\citep[BBS;][]{robbins_etal:1986}.
 In the absence of direct information about demographic parameters, and
@@
 
-We have presented a test case for these models using data from
-the BBS for ovenbirds. We demonstrate that these models can be
-used to model and estimate initial abundance, detection
-probability, and density-independent and dependent functions for
-population dynamics. We tested these models in both the
-frequentist and Bayesian frameworks and found strengths and
-weaknesses of each approach. The Bayesian analyses permit more
-flexibility and allow the incorporation of random observer
-effects, environmental stochasticity, and zero-inflated dynamics. 
-Our results suggest ignoring
-variation between observers can lead to underestimation of
-initial abundance and equilibrium population sizes. On the other
-hand, our Bayesian analyses did have some convergence problems.
-%particularly for the species (golden-winged warbler) with sparser data. 
-It's likely that poor model fits are related to the low estimates of
-detection probability, which in turn are probably caused by the
-large 400 m radius used for BBS point counts. Again, we believe
-that direct information on detectability (such as repeated
-visits within a season) would greatly improve the fit of these
-models in both frameworks.  We also were able to obtain year
-and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
-although with generally lower precision than more global estimates.  These sort of estimates
-may be more reliable when abundance and dynamics are modeled 
-as functions of habitat covariates, which we did not do here.
+%We have presented a test case for these models using data from
+%the BBS for ovenbirds. We demonstrate that these models can be
+%used to model and estimate initial abundance, detection
+%probability, and density-independent and dependent functions for
+%population dynamics. We tested these models in both the
+%frequentist and Bayesian frameworks and found strengths and
+%weaknesses of each approach. The Bayesian analyses permit more
+%flexibility and allow the incorporation of random observer
+%effects, environmental stochasticity, and zero-inflated dynamics. 
+%Our results suggest ignoring
+%variation between observers can lead to underestimation of
+%initial abundance and equilibrium population sizes. On the other
+%hand, our Bayesian analyses did have some convergence problems.
+%%particularly for the species (golden-winged warbler) with sparser data. 
+%It's likely that poor model fits are related to the low estimates of
+%detection probability, which in turn are probably caused by the
+%large 400 m radius used for BBS point counts. Again, we believe
+%that direct information on detectability (such as repeated
+%visits within a season) would greatly improve the fit of these
+%models in both frameworks.  We also were able to obtain year
+%and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
+%although with generally lower precision than more global estimates.  These sort of estimates
+%may be more reliable when abundance and dynamics are modeled 
+%as functions of habitat covariates, which we did not do here.
 \begin{comment}

### 2013-09-18 Dail_Mad_ext.tex
Edits from Scott plus appendices updated

@@
 %     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
-%        \small Migratory Bird Center \\
+        \small Migratory Bird Center \\
         \small Smithsonian Conservation Biology Institute \\
@@
    \and Richard B. Chandler \footnote{Contact:
-     \url{RChandler@usgs.gov}, +001-301-497-5856} \\
+     \url{rbchan@uga.edu}, +001-706-542-5815} \\
         \small USGS Patuxent Wildlife Research Center \\
@@
 account for observation error in such models can result in 
-bias, and thus ecologists have increasingly relied on state-space
+bias. Therefore, ecologists have increasingly relied on state-space
 models to directly model both the observation error and the ecological
@@
 in many common situations, (2) they do not admit spatial variation in
-population dynamics, and (3) there is no clear interpretation of the
+population dynamics, and (3) we have no clear interpretation of the
 observation error. We demonstrate how each of these problems can be
@@
 of these ideas is possible using either frequentist or Bayesian
-methods, and code to fit these models using the \textbf{R} package
+methods. We provide code to fit these models using the \textbf{R} package
 \texttt{unmarked} and \textbf{JAGS} is also provided.
@@
 the North American Breeding Bird Survey highlighted how these methods
-can be readily applied to existing data, but it also suggested the
-that precision will be low when direct information about detection
+can be readily applied to existing data, but it also suggested 
+that precision would be low when we lack direct information about detection
 probability (such as is collected using distance sampling or
-replicated counts) is unavailable.
+replicated counts).
 \end{enumerate}
@@
 
-\newpage
+%\newpage
 
@@
 density-dependent population regulation, population cycling, and spatial synchrony
-\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999},
+\citep{may:1975,turchin:1990,bjornstad_etal:1999}, %,royama:1977,dennis_taper:1994
 and population models are required to evaluate associated hypotheses.
@@
 $\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
-$\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
+$\zeta_t \sim \mathrm{N}(0, \tau)$. Standard practice is to
 ignore process variation associated with $N_1$, as
@@
 are among the most widespread approaches for modeling population dynamics
-using time-series data, several problems are evident. Namely, (1)
+using time-series data, several problems exist. First,
 time-series data and the underlying abundance parameters of interest
@@
 concerns about the use of Gaussian distribution for the random
-effects, (2) the model for observation error has little biological basis, (3)
-spatial variation in abundance is not allowed, and (4) some of the
+effects. Second, the model for observation error has little biological basis. 
+Third, the model does not account for spatial variation in abundance. Lastly, some of the
 parameters of the model are not estimable. We briefly discuss each of
@@
 than estimation methods when random effects are not Gaussian distributed
-\citep{devalpine_hastings:2002}. The problem with this is that it
-allows for negative and non-integer values of the state-variable, which is
+\citep{devalpine_hastings:2002}. %The problem with this is that it
+This approach allows for negative and non-integer values of the state-variable, which is
 inconsistent with the observed data, and may result in implausible predictions.
@@
 higher than $N_t$ as often as it is lower than $N_t$. 
-It is hard to identify a mechanism that would cause such symmetric errors.
+Identifying a biological mechanism that would cause such symmetric errors
+is difficult.
 A more likely form of observation error, and one that has been recognized for well
@@
 characteristics of the species under study, such as its elusiveness,
-or to the failings of the ecologist collecting the data in the field.
+or to individuals being missed by the ecologist collecting the data in the field.
 Although a vast number of methods have been devised for accounting for
@@
 
+Common formulations of state space models do not admit
+spatial variation. This reduces the scope of the inferences that can
+be drawn from the models. Moreover, space is an important factor in the
+regulation of natural populations \citep{murdoch:1994,
+pulliam:1988, gill_etal:2001}.
+
 A more serious problem %than the ones associated with the Gaussian assumptions 
@@
 circumstances \citep{polansky_etal:2009}. Recall that the
-Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
-however, $N_1$ is a random variable and hence there is uncertainty
-that should be accounted for.  
+Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
+which $N_1$ is a random variable. Thus, uncertainty exists
+that should be accounted for in the model.  
 %are needed to fully specify the model.
@@
 
-A fourth problem with these models is that they do not admit
-spatial variation. This reduces the scope of the inferences that can
-be drawn from the models and it ignores the importance of space in
-population regulation.
-
 Several extensions of state-space models have been proposed to
@@
 variation in abundance when individuals cannot be detected with
-certainty. To estimate both abundance parameters and parameters
-of the detection process, the original $N$-mixture model uses replicate
-observations at each site, which are collected during sufficiently
+certainty. To estimate parameters associated with both abundance 
+and the detection process, the original $N$-mixture model uses replicate
+observations at each site. The observations are collected during sufficiently
 short time intervals such that the population can safely be
@@
 
-As with traditional state-space models, the DM model includes the
+Like traditional state-space models, the DM model includes the
 three conditionally related processes shown in Eq.~\ref{eq:ss1},
@@
 ``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
-expected abundance constant over time. As before, covariates of
+expected abundance constant over time. Covariates of
 $\omega$ and $\gamma$ can be easily accommodated, for example
@@
 
-However, Bayesian inference has several appealing features. First, it
+Bayesian inference, however, has several appealing features. First, it
 allows direct probability statements to be made about a hypothesis
@@
 %    N[i,t] <- S[i,t-1] + G[i,t-1]
-%    n[i,t] ~ dbin(p, N[i,t])
+%    y[i,t] ~ dbin(p, N[i,t])
 %    }
@@
 
-Although partitioning population growth into survival and recruitment
+Partitioning population growth into survival and recruitment
 is ideal in terms of providing a mechanistic description of population
-dynamics, it is not always possible using simple count data,
+dynamics, but this is not always possible using simple count data,
 especially when the sites are not closed with respect to immigration
@@
 with classical population growth models that have a long history in
-ecology. The simplest of which is the exponential growth
-model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
+ecology. The exponential growth model is the simplest: 
+$f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
 rate of increase. %Although demographic and environmental stochasticity
@@
 like the geometric-recruitment version of the model, is a variant on a simple
-density-independent exponential model of population growth.
+density-independent, exponential model of population growth.
 Density-dependent versions of the model are also possible.  For
@@
 Ricker-logistic model.
-
 Because a single Poisson distribution controls the
@@
 versions of the DM models all share a common feature (or bug):
-once the population at a site reaches 0, it must remain at 0.
-This is because all contributions to population growth are
-local in these models.  We generalized each these models that
+once the population at a site reaches 0, it must remain at 0
+because all contributions to population growth are assumed to be
+local.  We therefore generalized these models to
 include both internal and external (immigration) contributions
@@
 
-\citet{dail_madsen:2011} suggested two distributions for
-modeling
+\citet{dail_madsen:2011} suggested two distributions for modeling
 initial abundance: Poisson and negative binomial. We have
-extended
-their model to include another distribution, the zero-inflated
+extended their model to include another distribution, the zero-inflated
 Poisson. This could be useful when, for example, one is modeling
-the
-abundance of several species with the same set of point
-count surveys for all species, but this set includes routes
-sites
+the abundance of several species with the same set of point
+count surveys for all species, but some survey sites are
 outside the range of some of the species. The distribution
-of
-initial abundances can be represented as:
+of initial abundances can be represented as:
 \begin{equation}
@@
 species was at a site but not detected; the site was within the
-species' range but there were no animals at that site in that
+species' range but no individuals occured at that site in that
 year; and the site was outside the species' range. Furthermore,
@@
 
-The zero-inflated Poisson distribution can be applied to not only
-initial abundance but also to recruitment and population growth
-terms. For example, the recruitment term of the constant-recruitment DM model
+The zero-inflated Poisson distribution can be applied to recruitment and 
+population growth terms as well as initial abundance. For example, 
+the recruitment term of the constant-recruitment DM model
 (Eq.~\ref{eq:Nt}) can be modified as follows:
@@
 \end{comment}
-The original DM models include demographic stochasticity (variation in population growth 
-rate due to the randomness of birth and death processes) through random 
+The original DM models include demographic stochasticity --- variation in population growth 
+rate due to the randomness of birth and death processes --- through random 
 binomial draws for survival and Poisson draws for recruitment.  Our population
@@
 
-One way to model environmental stochasticity (variation in population growth
-rate due to stochastic environmental conditions) is by explicitly
+One way to model environmental stochasticity --- variation in population growth
+rate due to stochastic environmental conditions --- is by explicitly
 modeling dynamic parameters as functions of environmental covariates, such
@@
 Three variations suggest themselves.  The first has independent environmental stochasticity 
-between sites, and might be appropriate where sites are widely dispersed.  In the second 
+between sites, and might be appropriate where sites are widely dispersed.  In the second, 
 all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{it} = \nu_{t}$),
 which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
-In cases where sites can be close together but the range is broad, a multivariate normal distribution for $\nu_{it}$ with correlation coefficients
-between sites in a year dependent on distance (geographic or otherwise) might be appropriate.
+Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
+for $\nu_{it}$ with distance-dependent (geographic or otherwise) correlation coefficients
+between sites in a year might be appropriate.
 
@@
 Variation in detection probability among observers is another
-source of random variation that may need to be accounted for applying
+source of random variation that may need to be considered when applying
 the DM model. For example, differences in observers' ability to see,
@@
 A fully Bayesian approach to PVA is straightforward to
-implement, especially since MCMC will be generally be relied upon in the
-first place. Because population size at any future point %Furthermore, b
-in time is regarded as an unknown parameter like any other, computing the
+implement with MCMC. Because population size at any future point %Furthermore, b
+in time is regarded as a standard, unknown parameter, computing the
 posterior distribution of $N^+_t$ requires no additional steps
-in MCMC algorithm, other than formally stating that the count data are
-``missing'' in future years. An example of this is provided in the appendix.
+in the MCMC algorithm, other than formally stating that the count data are
+``missing'' in future years. An example of this is provided in the Appendix.
 
@@
 
-We estimated the parameters for each simulation using the same
-initial abundance (Poisson) and dynamics models as were
-simulated, implemented in the unmarked library in \textbf{R}.
+We used the \texttt{unmarked} library in \textbf{R} to estimate the parameters 
+for each simulation with the same
+initial abundance (Poisson) and dynamics models as were simulated.
 When run in a maximum likelihood framework, these models require a
@@
 (BBS) data from 1966-2010 in the bordering US states
-Maryland and Virginia. For our focal species, we selected
-ovenbirds (\textit{Seiurus aurocapilla}), an abundant and widespread
-forest breeding migrant with a stable or increasing trend in the region \citep{porneluzi_etal:2011}.
+Maryland and Virginia. For our focal species, we selected the
+ovenbird (\textit{Seiurus aurocapilla}), an abundant and widespread,
+forest-breeding migratory songbird with a stable or increasing trend in the region \citep{porneluzi_etal:2011}.
 %and golden-winged warblers (\textit{Vermivora chrysoptera}), which
@@
 observers in the United States and Canada. An observer conducts 50
-3-minute point counts, with 400 m radiuses and 0.8 km apart from each other along a
+3-minute point counts, with 400 m radii and 0.8 km apart from each other along a
 39.4 km route. We only used data marked as acceptable for use in the annual BBS
-analysis \citep{sauer_etal:1994auk}.  We summed the total birds of each
-species seen on each route and year and used the routes (rather
+analysis \citep{sauer_etal:1994auk}.  We summed the number of ovenbirds
+seen on each route and year, and used the routes (rather
 than the individual stops) as our sites.
@@
 We ran the top ranked models from the maximum likelihood
-analyses in a Bayesian framework.  We added random observer 
-effects and environmental stochasticity (modeled as lognormal
-variation in regional population growth rate, Eq.~\ref{eq:nuRand}). % and, where appropriate, zero-inflated dynamics.  
-We used
-non-informative priors. We tested for lack of convergence using
-5 Markov chains for each model \citep{gelman_rubin:1992}.
-%(Gelman and Rubin 1992).
+analyses in a Bayesian framework with non-informative priors.  We added random observer 
+effects and environmental stochasticity, modeled as lognormal
+variation in regional population growth rate, Eq.~\ref{eq:nuRand}. % and, where appropriate, zero-inflated dynamics.  
+We tested for lack of convergence using
+five Markov chains for each model \citep{gelman_rubin:1992}.
 For each chain we sampled the MCMC for at least 40,000 iterations, after at
@@
 we obtained these estimates from a single chain.  %We present these estimates for ovenbirds only.
-%Latent inclusion factors for temperature effects?
 
@@
 
-Generally, estimator bias and error were low with the exponential model, and 
-coverage was nominal (93.4\% - 97.0\%; Fig.~\ref{fig:exp_hists}; Appendix ?). 
+Estimator bias and error were generally low with the exponential model, and 
+coverage was nominal (93.4\% - 97.0\%; Fig.~\ref{fig:exp_hists}; Table S2.1 in Appendix 2). 
 However, estimates of $r$ were 
@@
 Simulations of the Ricker-logistic model also generally performed well (Fig.~\ref{fig:rick_hists}; 
-Appendix ?), except when the true value of $r$ was low (0.005). In this case the 
+Table S2.2 in Appendix 2), except when the true value of $r$ was low (0.005). In this case the 
 estimates of $r$ and $K$ were inaccurate (relative RMSE between 87.4\% and 2091.8\%),
@@
 91.9\% and 96.4\%.  All three cases simulated for the Ricker-logistic + immigration model 
-performed fairly well (Fig.~\ref{fig:ricki_hists}; Appendix ?; relative RMSE between 7.2\% 
+performed fairly well (Fig.~\ref{fig:ricki_hists}; Table S2.3 in Appendix 2; relative RMSE between 7.2\% 
 and 62.9\%, bias between -3.5\% and 3.3\%, and coverage between
@@
 initial abundance over the Poisson and zero-inflated Poisson
-(Table 2A). %Even compared to the Poisson, there was
+(Table S3.1A in Appendix 3). %Even compared to the Poisson, there was
 %no evidence to support a zero-inflation factor for this species.
-The best supported model for $p$ was additive effects of wind speed
-and first run (Table 2B, model B.1). First run and increasing wind
-speeds both decreased $p$. All dynamics models with immigration were
-better supported than models without immigration (Table 2C); 
-the best supported of these was the Ricker-logistic + immigration.
+The best supported model for $p$ included additive effects of wind speed
+and first run of a route by an observer (Table S3.1B, model B.1). First run 
+and increasing wind speeds both decreased $p$. All dynamics models 
+with immigration were better supported than models without immigration 
+(Table S3.1C); the best supported of these was the Ricker-logistic + immigration 
+(see Appendix 3, Table S3.2 OR Figure S3.1 for estimates).
+%This model estimated $\Lambda$ at 33.4 $\pm$ 4.7, $r$ at 0.026 $\pm$ 0.006, 
+%$K$ at  56.9 $\pm$ 9.4, and $\iota$ at 0.283 $\pm$ 0.043.
 %The best supported model of those without immigration was the
@@
 %For both species, 
-Constant-recruitment dynamics models estimated very high
-survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05). % for ovenbird and
+Constant-recruitment dynamics models estimated unrealisticly high
+survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05), whereas % for ovenbird and
 %0.935 $\pm$ 0.011 for
 %golden-winged warbler). 
-The geometric-recruitment and geometric-recruitment + immigration
-models estimated very low survival probabilities %for ovenbird 
+the geometric-recruitment and geometric-recruitment + immigration
+models estimated nonviably low survival probabilities %for ovenbird 
 ($\omega$ =
@@
 run in the Bayesian framework, except that the estimate of $r$
-more than halved.
-% (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
+more than halved (Table S3.2 OR Figure S3.1, Appendix 3). %to 0.011 $\pm$ 0.005 [for
 %Bayesian model estimates we present mean and SD]). 
 When random observer effects
-were added, estimates for $\Lambda$ and $K$ increased dramatically,
-%(from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
-%respectively)
+were added, estimates for $\Lambda$ and $K$ increased dramatically
+%(to 42.5 $\pm$ 7.3 and 101.8 $\pm$ 21.1,
+%respectively),
 and the estimate for r was intermediate. The %(0.018 $\pm$ 0.006)
-estimate of the intercept for $p$ (on the logit scale) dropped. % from -1.5 $\pm$
-%0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
+estimate of the intercept for $p$ dropped. % from -1.5 $\pm$
+%0.1 to -2.0 $\pm$ 0.1 (on the logit scale), and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
 When environmental stochasticity was added, the estimate for $K$ increased. %(to 112 $\pm$ 58.8) and $r$ was
@@
 Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
-provided some evidence for a lack of convergence in estimates of the
-intercept for $p$ for all Bayesian models ran.
+provided little evidence of lack of convergence for any of the Bayesian models ran.
 
 We estimated ovenbird abundance by route and year (and averaged over routes) 
-using the Bayesian model including random observer effects (Fig.~\ref{fig:oven_N}).  
+using the Bayesian model with random observer effects (Fig.~\ref{fig:oven_N}).  
 Estimated average abundance over time alternated between growing and stable periods.  
@@
 \citep[IPM;][]{besbeas_etal:2002,
-buckland_etal:2004,schaub_etal:2007}
-and we believe that a DM model could work well in this context.
-In fact, the DM model could be viewed as a specific case of the
+buckland_etal:2004,schaub_etal:2007}.
+We believe that a DM modeling framework could work well in this context,
+and that the DM model could be viewed as a specific case of the
 IPM proposed by \citet{buckland_etal:2004}.
@@
 
-Although these integrated population models may be viewed as the gold standard in
-state-space modeling, ecologists are not always so fortunate to have direct
+Integrated population models may be viewed as the gold standard in
+state-space modeling, but ecologists do not always have direct
 information about vital rates, especially at large spatial scales.
-Rather, count data are much more common, and
+Count data are much more common, relatively inexpensive to obtain, and
 are produced by many of the largest monitoring programs in the
@@
 In the absence of direct information about demographic parameters, and
-when the original assumptions of the model do not hold, the
-model can still be used for important purposes such as predicting future
+when the original assumptions of the model do not hold, our extensions to the
+DM model can still be used for important purposes such as predicting future
 population size and changes in species distributions.
 
-Even though the DM models and our extentions are conceptually simple, even in
-their most stripped down form they include many random effects (the $N$'s, which
+DM models and our extensions are conceptually simple, but even in
+their most basic form they include many random effects (the $N$'s, which
 incorporate demographic stochasticity), making it challenging to compute the likelihood 
@@
 may be of interest.  We have shown a few approaches for incorporating some of these
-(environmental stochasticity and random observer effects).  Other approaches to modeling
-each of these may also be worth trying.  Incorporating these random effects can increase
-the difficulty of achieving model convergence, but they are certainly important.
+(environmental stochasticity and random observer effects).  %Other approaches to modeling
+%each of these may also be worth trying.  
+Although incorporating these random effects can increase
+the difficulty of achieving model convergence and decrease estimate precision, they are certainly important.
 For example, the estimate of semi-equilibrium abundance for the ovenbird increased by more than 150\%
@@
 ecologists lacking skills in computer programming and advanced
-statistics. We have included an appendix with extensive code to
+statistics. We have included an appendix with code to
 simulate and fit many of the extensions discussed in this paper.
@@
 parameters of the state process can be estimated ultimately depends
-upon how well detection probability is estimated. When there is
+upon how well detection probability is estimated. With
 only a single survey per primary period, the information about
@@
 model's parametric assumptions. Furthermore, properly modeling detection
-probability can involve account for nuances such as difference
+probability can involve accounting for nuances such as difference
 among individuals and probabilities of being available for detection
@@
 
-Fortunately, it is straight-forward to incorporate direct information about
-detection probability and we recommend that this be done whenever
-possible. The original method for proposed in the original paper
-was to collected replicated counts within the primary periods during
-a period in which the populations could be assumed to be closed. A
-robust design could be used to combine multiple surveys per
+Fortunately, incorporating direct information about detection probability 
+is straightforward and we recommend that this be done whenever
+possible. \citet{dail_madsen:2011} proposed that
+replicated counts be conducted within an interval in which the populations can 
+be assumed to be closed. A robust design \citep{pollock:1982} could be 
+used to combine multiple surveys per
 primary period to increase the precision of the estimates. We envision
@@
 Our emphasis was on increasing the practical utility of this
-class of models, and so we avoided several conceptually-interesting
-extensions that we belive would be computationally prohibitive in many
+class of models, and so we avoided several conceptually interesting
+extensions that we believe would be computationally prohibitive in many
 cases. Nonetheless, we will discuss one---spatially-explict models of
 immigration. 
-In our models immigration is currently modeled as independent of
+In our models, immigration is currently modeled as independent of
 population sizes in other sites. Several alternatives suggest
 themselves if we think about movement in a metapopulation
-context \citep{hanski:1998}: immigration depending on total or mean
+context \citep{hanski:1998}: immigration conditional on total or mean
 abundance across sites the previous time step; immigration
-depending on abundance in nearby sites only; and immigration
-depending on abundance at sites as a function of their distance
+conditional on abundance in nearby sites only; and immigration
+conditional on abundance at sites as a function of their distance
 from the receiving site \citep{hastings:1991,hanski:1998}. These models would
 be based on the assumptions that either the study sites
-spatially cover the whole range of the metapopulation or the
+encompass the whole range of the metapopulation or that the
 annual changes in the populations sampled are representative of
@@
 
-In our analysis of simulated data, we found that our models generally estimated parameters well.
+Our models generally estimated parameters well in analysis of simulated data.
 Two exceptions were the estimates of instantaneous growth rate in exponential model when the 
-true value of initial abundance was low and the estimates of equilibrium abundance and maximum
-growth rate in the Ricker model when the true value of maximum growth rate was low.  
+true value of initial abundance was low, and the estimates of equilibrium abundance and maximum
+growth rate in the Ricker-logistic model when the true value of maximum growth rate was low.  
 The failure of the first case was likely due to a large number of sites starting at abundance 0 or reaching it early in the simulation.  
 %[Richard, maybe come up with a summary statistic for this based on the simulation N values?]
-Since a site that hits abundance 0 in this simulation stayed at 0, this provided little information with
+Because a site that hits abundance 0 in this simulation stayed at 0, this provided little information with
 which to estimate growth rate.  The failure of the second case is also reasonable: with very
@@
 
-The modeling framework we described can be used to address many
+The modeling framework we described can be used and extended to address many
 pressing issues in ecology and conservation biology. For
@@
 We thank J. Andrew Royle and T. Scott Sillett for helpful
-comments and suggestions. We gratefully acknowledge National Park 
-Service, Smithsonian Institution, and [more] for funding support.
+comments and suggestions. We gratefully acknowledge the U. S. National Park 
+Service, the Smithsonian Institution, and [Richard's funding sources] for funding support.
 
@@
 % How about separate tables?  Also updated the Gompertz-logistic results.
-\begin{table}
-  \centering
-  \small
-\caption{Model selection table for ovenbirds in Maryland and Virginia,
-    1966-2010.  We present model name and number, number of 
-parameters (Par.), and difference in Akaike's
-information criterion between each model and the top model of
-that set ($\Delta$AIC).  The first section compares
-models for initial abundance, the second for detection
-probability, and the third for dynamics.}
-  \begin{tabular}[h]{lcc}
-\hline
-Model	&Par.	&$\Delta$AIC	\\
-\hline
-A. Initial Abundance && \\
-A.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0\\
-A.2. P[$\Lambda$(.)]Exponential[$r$(.)]$p$(.)	&3	&1262.7\\
-A.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4 &1264.7\\
-\hline
-B. Detection Probability && \\
-B.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &0	\\
-B.2. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind) &7 &0.9\\
-B.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(1st)	&5	&5.0
-\\B.4. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&6.4\\
-\hline
-C. Dynamics && \\
-C.1. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&0	\\
-C.2. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&8.4 \\
-C.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(wind+1st) &9	&36.5\\
-C.4. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st) &10	&38.6\\
-C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&192.8\\
-C.6. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&195.1\\
-C.7. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &271.3\\
-C.8. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&273.7\\
-C.9. NB[$\Lambda$(.)$\alpha$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&1856.7\\
-\hline
-\end{tabular}
-\end{table}
+%\begin{table}
+%  \centering
+%  \small
+%\caption{Model selection table for ovenbirds in Maryland and Virginia,
+%    1966-2010.  We present model name and number, number of 
+%parameters (Par.), and difference in Akaike's
+%information criterion between each model and the top model of
+%that set ($\Delta$AIC).  The first section compares
+%models for initial abundance, the second for detection
+%probability, and the third for dynamics.}
+%  \begin{tabular}[h]{lcc}
+%\hline
+%Model	&Par.	&$\Delta$AIC	\\
+%\hline
+%A. Initial Abundance && \\
+%A.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0\\
+%A.2. P[$\Lambda$(.)]Exponential[$r$(.)]$p$(.)	&3	&1262.7\\
+%A.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4 &1264.7\\
+%\hline
+%B. Detection Probability && \\
+%B.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &0	\\
+%B.2. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind) &7 &0.9\\
+%B.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(1st)	&5	&5.0
+%\\B.4. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&6.4\\
+%\hline
+%C. Dynamics && \\
+%C.1. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&0	\\
+%C.2. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&8.4 \\
+%C.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(wind+1st) &9	&36.5\\
+%C.4. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st) &10	&38.6\\
+%C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&192.8\\
+%C.6. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&195.1\\
+%C.7. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &271.3\\
+%C.8. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&273.7\\
+%C.9. NB[$\Lambda$(.)$\alpha$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&1856.7\\
+%\hline
+%\end{tabular}
+%\end{table}
 
@@
   \centering
-  \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year4}
+  \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year6}
 \caption{Maps of BBS route and year specific estimated abundances for ovenbirds, for the years 1970,

### 2013-09-18 supp/code_samples.tex
Edits from Scott plus appendices updated

@@
-% !TEX TS-program = pdflatex
-% !TEX encoding = UTF-8 Unicode
-
-% This is a simple template for a LaTeX document using the "article" class.
-% See "book", "report", "letter" for other types of document.
-
-\documentclass[12pt]{article} % use larger type; default would be 10pt
-
-\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
-
-%%% Examples of Article customizations
-% These packages are optional, depending whether you want the features they provide.
-% See the LaTeX Companion or other references for full information.
-
-%%% PAGE DIMENSIONS
-\usepackage{geometry} % to change the page dimensions
-\geometry{a4paper} % or letterpaper (US) or a5paper or....
-% \geometry{margin=2in} % for example, change the margins to 2 inches all round
-% \geometry{landscape} % set up the page for landscape
-%   read geometry.pdf for detailed page layout information
-
-\usepackage{graphicx} % support the \includegraphics command and options
-
-% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
-
-%%% PACKAGES
-\usepackage{booktabs} % for much better looking tables
-\usepackage{array} % for better arrays (eg matrices) in maths
-\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
-\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
-\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
-% These packages are all incorporated in the memoir class to one degree or another...
-
-%%% HEADERS & FOOTERS
-\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
-\pagestyle{fancy} % options: empty , plain , fancy
-\renewcommand{\headrulewidth}{0pt} % customise the layout...
-\lhead{}\chead{}\rhead{}
-\lfoot{}\cfoot{\thepage}\rfoot{}
-
-%%% SECTION TITLE APPEARANCE
-\usepackage{sectsty}
-\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
-% (This matches ConTeXt defaults)
-
-%%% ToC (table of contents) APPEARANCE
-\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
-\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
-\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
-\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!
+\documentclass[12pt]{article}
+\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
+\usepackage[mathlines,displaymath]{lineno}
+\usepackage{setspace}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{bm}
+\usepackage{graphicx}
+\usepackage[round,colon,authoryear]{natbib}
+\usepackage{color, soul}
+\usepackage{fancyvrb}
+\usepackage{verbatim}
+\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
+\usepackage{rotating}
 
@@
 
-\title{Code Samples}
-%\author{The Author}
+\begin{document}
+\title{Appendix 1: Code Samples}
+\maketitle
 %\date{} % Activate to display a given date or no date (if empty),
          % otherwise the current date is printed 
-
-\begin{document}
-\maketitle
-
 \section{\texttt{unmarked} code}
 
-\subsection{Exponential model}
-\subsection{Ricker-logistic model}
-\subsection{Gompertz-logistic model}
-\subsection{Exponential plus immigration model}
-\subsection{Ricker-logistic plus immigration model}
-\subsection{Gompertz-logistic plus immigration model}
-\subsection{Autoregressive plus immigration model}
-\subsection{Zero-inflated Poisson initial abundance model}
+%\subsection{Exponential model}
+%\subsection{Ricker-logistic model}
+%\subsection{Gompertz-logistic model}
+%\subsection{Exponential plus immigration model}
+%\subsection{Ricker-logistic plus immigration model}
+%\subsection{Gompertz-logistic plus immigration model}
+%\subsection{Autoregressive plus immigration model}
+%\subsection{Zero-inflated Poisson initial abundance model}
 \subsection{Simulation study code}
+
+File sim\_exp.R:
+\begin{verbatim}
+# Simulate data under exponential growth model
+sim.exp <- function(lambda, r, p, nSites=100, nYears=45) {
+    y <- N <- matrix(NA, nSites, nYears)
+    N[,1] <- rpois(nSites, lambda)
+    for(t in 2:nYears) {
+        mu <- N[,t-1]*exp(r)
+        N[,t] <- rpois(nSites, mu)
+    }
+    y[] <- rbinom(nSites*nYears, N, p)
+    return(list(N=N, y=y, lambda=lambda, r=r, p=p,
+                seed=.Random.seed, seed.kind=RNGkind()))
+}
+\end{verbatim}
+
+File sim\_ricker.r:
+\begin{verbatim}
+# Simulate data under Ricker model
+sim.ricker <- function(lambda, r, K, p, nSites=100, nYears=45) {
+    y <- N <- matrix(NA, nSites, nYears)
+    N[,1] <- rpois(nSites, lambda)
+    for(t in 2:nYears) {
+        mu <- N[,t-1]*exp(r*(1-N[,t-1]/K))
+        N[,t] <- rpois(nSites, mu)
+    }
+    y[] <- rbinom(nSites*nYears, N, p)
+    return(list(N=N, y=y, lambda=lambda, r=r, p=p,
+                seed=.Random.seed, seed.kind=RNGkind()))
+}
+\end{verbatim}
+
+File sim\_ricki.r:
+\begin{verbatim}
+# Simulate data under Ricker + immigration model
+sim.ricki <- function(lambda, r, K, iota, p, nSites=100, nYears=45) {
+    y <- N <- matrix(NA, nSites, nYears)
+    N[,1] <- rpois(nSites, lambda)
+    for(t in 2:nYears) {
+        mu <- N[,t-1]*exp(r*(1-N[,t-1]/K)) + iota
+        N[,t] <- rpois(nSites, mu)
+    }
+    y[] <- rbinom(nSites*nYears, N, p)
+    return(list(N=N, y=y, lambda=lambda, r=r, K=K, iota=iota, p=p,
+                seed=.Random.seed, seed.kind=RNGkind()))
+}
+\end{verbatim}
+
+File ricker\_r0.1K5.r:
+\begin{verbatim}
+# Simulate data under Ricker model
+
+# TO MODIFY
+# (1) Change simulator... source(SOMETHING ELSE)
+# (2) Change stub
+# (3) Change simout
+# (4) Change simulation parameters
+# (5) Change pcountOpen arguments: "dynamics" and "K"
+# (6) Results
+
+ls()
+stub <- "ricker_r0.1K5"  # (2)
+setwd(paste('simulated', stub, sep='/'))
+source("../sim_ricker.r")  # (1)
+ls()
+
+library(unmarked)
+
+nsim <- 200
+simout.mle <- simout.cover <- matrix(NA, nsim, 4)  # (3)
+colnames(simout.mle) <- colnames(simout.cover) <- c("Lambda", "r", "K", "p") # (3)
+
+# Simulate
+set.seed(345489)
+lambda <- 10 # (4)
+r <- 0.1     # (4)
+K <- 5       # (4)
+p <- 0.25    # (4)
+for(i in 1:nsim) {
+    cat("\ndoing", i, "at", format(Sys.time()), "\n")
+    data.name <- paste("data_", stub, "_", i, sep="")
+    data.file <- paste("data/", data.name, ".gzip", sep="")
+    sim.i <- sim.ricker(lambda=lambda, r=r, K=K, p=p, nSites=100, nYears=40) # (1)
+    cat("   max(N)=", max(sim.i$N), "\n", sep="")
+    assign(data.name, sim.i)
+    save(list=data.name, file=data.file) # save simulated data
+    umf.i <- unmarkedFramePCO(y=sim.i$y, numPrimary=ncol(sim.i$y))
+    fm.i <- pcountOpen(~1, ~1, ~1, ~1, umf.i, K=200, dynamics="ricker",  # (5)
+                       starts=c(log(lambda), log(r), log(K), qlogis(p))) # (5)
+    model.name <- paste("fm_", stub, "_", i, sep="")
+    model.file <- paste("fm/", model.name, ".gzip", sep="")
+    assign(model.name, fm.i)
+    save(list=model.name, file=model.file) # save fitted models
+    mle.i <- coef(fm.i)
+    simout.mle[i,] <- c(exp(mle.i[1]), exp(mle.i[2]), exp(mle.i[3]), plogis(mle.i[4])) # (6)
+    cat("   mle=", round(simout.mle[i,], 4), "\n")
+    write.csv(simout.mle, paste(stub, "_mle.csv", sep=""),
+              row.names=FALSE)
+    CI.lam <- confint(fm.i, type="lambda") # (6)
+    CI.r <- confint(fm.i, type="gamma")    # (6)
+    CI.K <- confint(fm.i, type="omega")    # (6)
+    CI.p <- confint(fm.i, type="det")      # (6)
+    simout.cover[i,1] <- log(lambda) >= CI.lam[1] & log(lambda) <= CI.lam[2] # (6)
+    simout.cover[i,2] <- log(r) >= CI.r[1] & log(r) <= CI.r[2]               # (6)
+    simout.cover[i,3] <- log(K) >= CI.K[1] & log(K) <= CI.K[2]               # (6)
+    simout.cover[i,4] <- qlogis(p) >= CI.p[1] & qlogis(p) <= CI.p[2]         # (6)
+    write.csv(simout.cover, paste(stub, "_cover.csv", sep=""),
+              row.names=FALSE)
+    rm(list=c(data.name, model.name))
+}
+\end{verbatim}
+
+File ricki\_r0.05K10iota0.5.r:
+\begin{verbatim}
+# Simulate data under Ricker + immigration model
+
+# TO MODIFY
+# (1) Change simulator... source(SOMETHING ELSE)
+# (2) Change stub
+# (3) Change simout
+# (4) Change simulation parameters
+# (5) Change pcountOpen arguments: "dynamics" and "K"
+# (6) Results
+
+stub <- "ricki_r0.05K10iota0.5"  # (2)
+setwd(paste('simulated', stub, sep='/'))
+ls()
+source("../sim_ricki.r")  # (1)
+ls()
+
+library(unmarked)
+
+nsim <- 1000
+simout.mle <- simout.cover <- matrix(NA, nsim, 5)  # (3)
+colnames(simout.mle) <- colnames(simout.cover) <-
+    c("Lambda", "r", "K", "iota", "p") # (3)
+
+# Simulate
+set.seed(345489)
+lambda <- 10 # (4)
+r <- 0.05    # (4)
+K <- 10      # (4)
+iota <- 0.5  # (4)
+p <- 0.25    # (4)
+for(i in 1:nsim) {
+    cat("\ndoing", i, "at", format(Sys.time()), "\n")
+    data.name <- paste("data_", stub, "_", i, sep="")
+    data.file <- paste("data/", data.name, ".gzip", sep="")
+    # Create simulated data
+    sim.i <- sim.ricki(lambda=lambda, r=r, K=K, iota=iota, p=p, nSites=100, nYears=40) # (1)
+    cat("   max(N)=", max(sim.i$N), "\n", sep="")
+    assign(data.name, sim.i)
+    # Save simulated data
+    save(list=data.name, file=data.file)
+    # Run pcountOpen model 
+    umf.i <- unmarkedFramePCO(y=sim.i$y, numPrimary=ncol(sim.i$y))
+    fm.i <- pcountOpen(~1, ~1, ~1, ~1, umf.i, K=200, dynamics="ricker", # (5)
+      immigration=TRUE, starts=c(log(lambda), log(r), log(K), qlogis(p), log(iota))) # (5)
+    # Save fitted models
+    model.name <- paste("fm_", stub, "_", i, sep="")
+    model.file <- paste("fm/", model.name, ".gzip", sep="")
+    assign(model.name, fm.i)
+    save(list=model.name, file=model.file) 
+    # Save maximum likelihood estimates
+    mle.i <- coef(fm.i)
+    simout.mle[i,] <- c(exp(mle.i[1]), exp(mle.i[2]), exp(mle.i[3]), exp(mle.i[5]), plogis(mle.i[4])) # (6)
+    cat("   mle=", round(simout.mle[i,], 4), "\n")
+    write.csv(simout.mle, paste(stub, "_mle.csv", sep=""),
+              row.names=FALSE)
+    # Save coverage
+    CI.lam <- confint(fm.i, type="lambda") # (6)
+    CI.r <- confint(fm.i, type="gamma")    # (6)
+    CI.K <- confint(fm.i, type="omega")    # (6)
+    CI.p <- confint(fm.i, type="det")      # (6)
+    CI.iota <- confint(fm.i, type="iota")  # (6)
+    simout.cover[i,1] <- log(lambda) >= CI.lam[1] & log(lambda) <= CI.lam[2] # (6)
+    simout.cover[i,2] <- log(r) >= CI.r[1] & log(r) <= CI.r[2]               # (6)
+    simout.cover[i,3] <- log(K) >= CI.K[1] & log(K) <= CI.K[2]               # (6)
+    simout.cover[i,4] <- log(iota) >= CI.iota[1] & log(iota) <= CI.iota[2]   # (6)
+    simout.cover[i,5] <- qlogis(p) >= CI.p[1] & qlogis(p) <= CI.p[2]         # (6)
+    write.csv(simout.cover, paste(stub, "_cover.csv", sep=""),
+              row.names=FALSE)
+    # Clean up
+    rm(list=c(data.name, model.name))
+}
+\end{verbatim}
+
 \subsection{BBS analysis code}
+\begin{verbatim}
+library(unmarked)
+
+# Read in input files
+ydata <- as.matrix(read.csv('oven3a.csv', row.names=1))
+observers <- as.matrix(read.csv('observers.csv', row.names=1))
+first.run <- as.matrix(read.csv('first.run.csv', row.names=1))
+wind.factor <- as.matrix(read.csv('wind.factor.csv', row.names=1))
+# Fix formatting
+wind.factor[wind.factor==" 0"] <- "0"
+wind.factor[wind.factor==" 1"] <- "1"
+wind.factor[wind.factor==" 2"] <- "2"
+# Combine obsCovs together into list
+obs4 <- list(observers = observers, first.run = first.run, wind.factor = 
+  wind.factor)
+
+# Basic settings and labels
+year1 <- 66
+aou <- 'oven'
+gam.model <- list(~1)[[1]] 
+om.model <- list(~1)[[1]] 
+lam.model <- list(~1)[[1]] 
+K <- 600
+iota.model <- list(~1) [[1]]
+
+# Input data for pcountOpen
+bird.frame <- unmarkedFramePCO(ydata, obsCovs=obs4, numPrimary=45)
+
+# Run mixture models for initial abundance  
+mods1 <- rep('', 3)
+outputs1 = list()
+for (i in 1:3) {
+  mixture <- c('P','NB','ZIP')[i]
+  dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[3]
+  p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[1]]
+  immigration <- FALSE
+  mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+    paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+    substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+    'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+    ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+    'K', sep='.')))), K, sep="")
+  mods1[i] <- mod.name
+  file.name <- paste(mod.name, 'gzip', sep='.')
+  print(file.name)
+  fm <- pcountOpen(data=bird.frame, mixture=mixture, gammaformula=gam.model, 
+    lambdaformula=lam.model, omegaformula=om.model, pformula=p.model, K=K, 
+    dynamics=dynamics, immigration=immigration, iotaformula=iota.model)
+  outputs1[[i]] <- fm
+  print(summary(fm))
+  save(fm, file=file.name)
+}
+names(outputs1) <- mods1
+
+AIC.table <- function(mods, sort=T) {
+  tab <- data.frame(model = names(mods), npar = NA, AIC = NA, delta = NA, 
+    weight = NA)
+  for (i in 1:length(mods)) {
+    tab$npar[i] <- length(coef(mods[[i]]))
+    tab$AIC[i] <- mods[[i]]@AIC
+  }
+  tab$delta <- tab$AIC - min(tab$AIC)
+  tab$weight <- exp(-tab$delta/2)/sum(exp(-tab$delta/2))
+  if (sort)
+    tab <- tab[order(tab$AIC), ]
+  return(tab)
+}
+
+# Compare these models
+mix.aic <- AIC.table(outputs1)
+mix.aic
+mix.num <- as.integer(rownames(mix.aic)[1])
+mix.num
+
+# Run models for detection probability  
+mods2 <- c(mods1[mix.num], rep('', 3))
+outputs2 <- list(outputs1[[mix.num]])
+for (i in 2:4) {
+  mixture <- c('P','NB','ZIP')[mix.num]
+  dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[3]
+  p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[i]]
+  immigration <- FALSE
+  mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+    paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+    substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+    'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+    ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+    'K', sep='.')))), K, sep="")
+  mods2[i] <- mod.name
+  file.name <- paste(mod.name, 'gzip', sep='.')
+  print(file.name)
+  fm <- pcountOpen(data=bird.frame, mixture=mixture, gammaformula=gam.model, 
+    lambdaformula=lam.model, omegaformula=om.model, pformula=p.model, K=K, 
+    dynamics=dynamics, immigration=immigration, iotaformula=iota.model)
+  outputs2[[i]] <- fm
+  print(summary(fm))
+  save(fm, file=file.name)
+}
+names(outputs2) <- mods2
+
+# Compare these models
+p.aic <- AIC.table(outputs2)
+p.aic
+p.num <- as.integer(rownames(p.aic)[1])
+p.num
+
+# Run models for population dynamics  
+mods3 <- c(mods2[p.num], rep('', 8))
+outputs3 <- list(outputs2[[p.num]])
+# Constant dynamics
+mixture <- c('P','NB','ZIP')[mix.num]
+dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[1]
+p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[p.num]]
+immigration <- FALSE
+mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+  paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+  'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'K', sep='.')))), K, sep="")
+mods3[2] <- mod.name
+file.name <- paste(mod.name, 'gzip', sep='.')
+print(file.name)
+fm <- pcountOpen(data=bird.frame, mixture=mixture, gammaformula=gam.model, 
+  lambdaformula=lam.model, omegaformula=om.model, pformula=p.model, K=K, 
+  dynamics=dynamics, immigration=immigration, iotaformula=iota.model)
+outputs3[[2]] <- fm
+print(summary(fm))
+save(fm, file=file.name)
+
+# Autoregressive dynamics
+mixture <- c('P','NB','ZIP')[mix.num]
+dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[2]
+p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[p.num]]
+immigration <- FALSE
+mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+  paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+  'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'K', sep='.')))), K, sep="")
+mods3[3] <- mod.name
+file.name <- paste(mod.name, 'gzip', sep='.')
+print(file.name)
+fm <- pcountOpen(data=bird.frame, mixture=mixture, gammaformula=gam.model, 
+  lambdaformula=lam.model, omegaformula=om.model, pformula=p.model, K=K, 
+  dynamics=dynamics, immigration=immigration, iotaformula=iota.model)
+outputs3[[3]] <- fm
+print(summary(fm))
+save(fm, file=file.name)
+
+# Ricker dynamics
+mixture <- c('P','NB','ZIP')[mix.num]
+dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[4]
+p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[p.num]]
+immigration <- FALSE
+mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+  paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+  'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'K', sep='.')))), K, sep="")
+mods3[4] <- mod.name
+file.name <- paste(mod.name, 'gzip', sep='.')
+print(file.name)
+# Ricker and Gompertz generally seem to fail with default starting values, 
+# so we set them
+starts <- coef(outputs2[[p.num]])
+starts <- c(starts[1], -2.35, starts[1]+0.3, starts[3:length(starts)])
+fm <- pcountOpen(data=bird.frame, mixture=mixture, gammaformula=gam.model, 
+  lambdaformula=lam.model, omegaformula=om.model, pformula=p.model, K=K, 
+  dynamics=dynamics, immigration=immigration, iotaformula=iota.model, 
+  starts=starts)
+outputs3[[4]] <- fm
+print(summary(fm))
+save(fm, file=file.name)
+
+# Gompertz dynamics
+mixture <- c('P','NB','ZIP')[mix.num]
+dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[5]
+p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[p.num]]
+immigration <- FALSE
+mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+  paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+  'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'K', sep='.')))), K, sep="")
+mods3[5] <- mod.name
+file.name <- paste(mod.name, 'gzip', sep='.')
+print(file.name)
+# Ricker and Gompertz generally seem to fail with default starting values, so we set them
+starts <- coef(outputs2[[p.num]])
+starts <- c(starts[1], -2.35, starts[1]+0.3, starts[3:length(starts)])
+fm <- pcountOpen(data=bird.frame, gammaformula=gam.model, 
+  lambdaformula=lam.model, omegaformula=om.model, pformula=p.model, K=K, 
+  mixture=mixture, dynamics=dynamics, immigration=immigration, 
+  iotaformula=iota.model, starts=starts)
+outputs3[[5]] <- fm
+print(summary(fm))
+save(fm, file=file.name)
+
+# Trend + immigration dynamics
+mixture <- c('P','NB','ZIP')[mix.num]
+dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[3]
+p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[p.num]]
+immigration <- TRUE
+mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+  paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+  'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'K', sep='.')))), K, sep="")
+mods3[6] <- mod.name
+file.name <- paste(mod.name, 'gzip', sep='.')
+print(file.name)
+fm <- pcountOpen(data=bird.frame, mixture=mixture, gammaformula=gam.model, 
+  lambdaformula=lam.model, omegaformula=om.model, pformula=p.model, K=K, 
+  dynamics=dynamics, immigration=immigration, iotaformula=iota.model)
+outputs3[[6]] <- fm
+print(summary(fm))
+save(fm, file=file.name)
+
+# Autoregressive + immigration dynamics
+mixture <- c('P','NB','ZIP')[mix.num]
+dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[2]
+p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[p.num]]
+immigration <- TRUE
+mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+  paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+  'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'K', sep='.')))), K, sep="")
+mods3[7] <- mod.name
+file.name <- paste(mod.name, 'gzip', sep='.')
+print(file.name)
+fm <- pcountOpen(data=bird.frame, mixture=mixture, gammaformula=gam.model, 
+  lambdaformula=lam.model, omegaformula=om.model, pformula=p.model, K=K, 
+  dynamics=dynamics, immigration=immigration, iotaformula=iota.model)
+outputs3[[7]] <- fm
+print(summary(fm))
+save(fm, file=file.name)
+
+# Ricker + immigration dynamics
+mixture <- c('P','NB','ZIP')[mix.num]
+dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[4]
+p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[p.num]]
+immigration <- TRUE
+mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+  paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+  'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'K', sep='.')))), K, sep="")
+mods3[8] <- mod.name
+file.name <- paste(mod.name, 'gzip', sep='.')
+print(file.name)
+# Ricker and Gompertz generally seem to fail with default starting values, 
+# so we set them
+starts <- coef(outputs2[[p.num]])
+starts <- c(starts[1], -2.35, starts[1]+0.3, starts[3:(length(starts)-1)], 
+  -1.5, starts[length(starts)])
+fm <- pcountOpen(data=bird.frame, gammaformula=gam.model, 
+  lambdaformula=lam.model, omegaformula=om.model, pformula=p.model, K=K, 
+  mixture=mixture, dynamics=dynamics, immigration=immigration, 
+  iotaformula=iota.model, starts=starts)
+outputs3[[8]] <- fm
+print(summary(fm))
+save(fm, file=file.name)
+
+# Gompertz + immigration dynamics
+mixture <- c('P','NB','ZIP')[mix.num]
+dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[5]
+p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[p.num]]
+immigration <- TRUE
+mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+  paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+  'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'K', sep='.')))), K, sep="")
+mods3[9] <- mod.name
+file.name <- paste(mod.name, 'gzip', sep='.')
+print(file.name)
+# Ricker and Gompertz generally seem to fail with default starting values, 
+# so we set them
+starts <- coef(outputs2[[p.num]])
+starts <- c(starts[1], -2.35, starts[1]+0.3, starts[3:(length(starts)-1)], 
+  -1.5, starts[length(starts)])
+fm <- pcountOpen(data=bird.frame, gammaformula=gam.model, 
+  lambdaformula=lam.model, omegaformula=om.model, pformula=p.model, K=K, 
+  mixture=mixture, dynamics=dynamics, immigration=immigration, 
+  iotaformula=iota.model, starts=starts)
+outputs3[[9]] <- fm
+print(summary(fm))
+save(fm, file=file.name)
+
+names(outputs3) <- mods3
+
+# Compare these models
+dyn.aic <- AIC.table(outputs3)
+dyn.aic
+dyn.num <- as.integer(rownames(dyn.aic)[1])
+dyn.num
+\end{verbatim}
 
@@
 
-\subsection{Constant model}
+\subsection{Geometric-recruitment (autoregressive) model}
+\begin{verbatim}
+model {
+lambda ~ dunif(0, 5)
+omega ~ dunif(0,1)
+gamma ~ dunif(0, 10)
+p ~ dunif(0,1)
+for(i in 1:nSites) {
+  N[i,1] ~ dpois(lambda)
+  y[i,1] ~ dbin(p, N[i,1])
+  for(t in 2:nYears) {
+    S[i,t-1] ~ dbin(omega, N[i,t-1])
+    G[i,t-1] ~ dpois(gamma * N[i,t-1])
+    N[i,t] <- S[i,t-1] + G[i,t-1]
+    y[i,t] ~ dbin(p, N[i,t])
+    }
+  }
+}
+\end{verbatim}
+
 \subsection{Exponential model}
+\begin{verbatim}
+model {
+lambda ~ dunif(0, 5)
+r ~ dnorm(0,0.01)
+p ~ dunif(0,1)
+for(i in 1:nSites) {
+  N[i,1] ~ dpois(lambda)
+  y[i,1] ~ dbin(p, N[i,1])
+  for(t in 2:nYears) {
+    N[i,t] ~ dpois(exp(r) * N[i,t-1])
+    y[i,t] ~ dbin(p, N[i,t])
+    }
+  }
+}
+\end{verbatim}
+
 \subsection{Ricker-logistic model}
+\begin{verbatim}
+model {
+lambda ~ dunif(0, 50)
+r ~ dunif(0,5)
+K ~ dunif(0, 50)
+p ~ dunif(0,1)
+for(i in 1:nSites) {
+  N[i,1] ~ dpois(lambda)
+  y[i,1] ~ dbin(p, N[i,1])
+  for(t in 2:nYears) {
+    muN[i,t-1] <- N[i,t-1] * exp(r * (1 - N[i,t-1] / K))
+    N[i,t] ~ dpois(muN[i,t-1])
+    y[i,t] ~ dbin(p, N[i,t])
+    }
+  }
+}
+\end{verbatim}
+
 \subsection{Gompertz-logistic model}
+\begin{verbatim}
+model {
+lambda ~ dunif(0, 50)
+r ~ dunif(0,5)
+K ~ dunif(0, 50)
+p ~ dunif(0,1)
+for(i in 1:nSites) {
+  N[i,1] ~ dpois(lambda)
+  y[i,1] ~ dbin(p, N[i,1])
+  for(t in 2:nYears) {
+    muN[i,t-1] <- N[i,t-1] * exp(r * (1 - log(N[i,t-1] + 1) / log(K + 1)))
+    N[i,t] ~ dpois(muN[i,t-1])
+    y[i,t] ~ dbin(p, N[i,t])
+    }
+  }
+}
+\end{verbatim}
+
 \subsection{Exponential plus immigration model}
-\subsection{Ricker-logistic plus immigration model}
-\subsection{Gompertz-logistic plus immigration model}
-\subsection{Autoregressive plus immigration model}
+\begin{verbatim}
+model {
+lambda ~ dunif(0, 5)
+r ~ dnorm(0,0.01)
+p ~ dunif(0,1)
+iota ~ dunif(0, 5)
+for(i in 1:nSites) {
+  N[i,1] ~ dpois(lambda)
+  y[i,1] ~ dbin(p, N[i,1])
+  for(t in 2:nYears) {
+    muN[i,t-1] <- N[i,t-1] * exp(r) + iota
+    N[i,t] ~ dpois(muN[i,t-1])
+    y[i,t] ~ dbin(p, N[i,t])
+    }
+  }
+}
+\end{verbatim}
+
 \subsection{Zero-inflated Poisson dynamics model}
+\begin{verbatim}
+model {
+lambda ~ dunif(0, 200)
+psi ~ dunif(0, 1)
+# Trend + immigration
+r ~ dnorm(0, 0.01)
+iota ~ dunif(0, 20)
+p ~ dunif(0,1)
+
+for(i in 1:nSites) {
+  range[i] ~ dbern(1-psi)
+  N[i,1] ~ dpois(lambda*range[i])
+  for(t in 2:nYears) {
+      muN[i,t-1] <- (N[i,t-1]*exp(r)+iota)*range[i]
+      N[i,t] ~ dpois(muN[i,t-1])
+    }
+  for(t in 1:nYears) {
+    y[i,t] ~ dbin(p, N[i,t])
+    }
+  }
+}
+\end{verbatim}
+
 \subsection{Environmental stochasticity models}
+\begin{verbatim}
+model {
+# Explicitly modeling dynamic parameters as functions of environmental covariates
+# Negative binomial initial abundance, Gompertz-logistic dynamics
+lambda ~ dunif(0, 400)
+alpha ~ dunif(0, 400)
+P <- alpha/(alpha + lambda)
+K0 ~ dnorm(0, 0.01)
+Krain ~ dnorm(0, 0.01)
+r0 ~ dnorm(0, 0.01)
+rTemp ~ dnorm(0, 0.01)
+p ~ dunif(0,1)
+
+for(i in 1:nSites) {
+  N[i,1] ~ dnegbin(P, alpha)
+  for(t in 2:nYears) {
+      K[i,t-1] <- exp(K0 + Krain*rain[i,t-1]) 
+      r[i,t-1] <- exp(r0 + rTemp*temp[i,t-1]) 
+      muN[i,t-1] <- N[i,t-1] * exp(r[i,t-1] * (1 - log(N[i,t-1] + 1) / log(K[i,t-1] + 1)))
+      N[i,t] ~ dpois(muN[i,t-1])
+  }
+  for(t in 1:nYears) {
+    y[i,t] ~ dbin(p, N[i,t])
+  }
+}
+}
+\end{verbatim}
+
+\begin{verbatim}
+model {
+# Independent environmental stochasticity between sites
+# Negative binomial initial abundance, Ricker-logistic dynamics + immigration
+lambda ~ dunif(0, 400)
+alpha ~ dunif(0, 400)
+P <- alpha/(alpha + lambda)
+K ~ dunif(0, 400)
+r ~ dunif(0, 5)
+iota ~ dunif(0, 10)
+sigma.nu ~ dunif(0, 2)
+tau.nu <- 1 / (sigma.nu * sigma.nu)
+p ~ dunif(0, 1)
+
+for(i in 1:nSites) {
+  N[i,1] ~ dnegbin(P, alpha)
+  for(t in 2:nYears) {
+      nu[i,t-1] ~ dnorm(0, tau.nu) 
+      muN[i,t-1] <- N[i, t-1] * exp(nu[i, t-1] + r * (1 - log(N[i, t-1] + 1) / log(K + 1))) + iota
+      N[i,t] ~ dpois(muN[i, t-1])
+  }
+  for(t in 1:nYears) {
+    y[i,t] ~ dbin(p, N[i,t])
+  }
+}
+}
+\end{verbatim}
+
+\begin{verbatim}
+model {
+# Regional stochasticity 
+# Negative binomial initial abundance, Ricker-logistic dynamics + immigration
+lambda ~ dunif(0, 400)
+alpha ~ dunif(0, 400)
+P <- alpha/(alpha + lambda)
+K ~ dunif(0, 400)
+r ~ dunif(0, 5)
+iota ~ dunif(0, 10)
+sigma.nu ~ dunif(0, 2)
+tau.nu <- 1 / (sigma.nu * sigma.nu)
+p ~ dunif(0, 1)
+
+for(t in 2:nYears) {
+  nu[t-1] ~ dnorm(0, tau.nu)
+}
+
+for(i in 1:nSites) {
+  N[i,1] ~ dnegbin(P, alpha)
+  for(t in 2:nYears) {
+      muN[i,t-1] <- N[i, t-1] * exp(nu[t-1] + r * (1 - log(N[i, t-1] + 1) / log(K + 1))) + iota
+      N[i,t] ~ dpois(muN[i, t-1])
+  }
+  for(t in 1:nYears) {
+    y[i,t] ~ dbin(p, N[i,t])
+  }
+}
+}
+\end{verbatim}
+
+% Don't have this working yet!  Should we leave it out or try to get it working?
+%\begin{verbatim}
+%model {
+%# Correlated environmental stochasticity between sites (exponential decay with distance)
+%# Poisson initial abundance, exponential dynamics 
+%lambda ~ dunif(0, 100)
+%#alpha ~ dunif(0, 100)
+%#P <- alpha/(alpha + lambda)
+%#K ~ dunif(0, 400)
+%r ~ dnorm(0, 0.01)
+%#iota ~ dunif(0, 10)
+%sigma.nu ~ dunif(0, 2)
+%l.nu ~ dunif(0, 1000)
+%for(i in 1:nSites) {
+%  cor.mat[i,i] <- 1
+%#  cov.mat[i,i] <- sigma.nu * sigma.nu
+%  for(j in 1:(i-1)) {
+%    cor.mat[i,j] <- exp(-((dist.mat[i,j]/(2*l.nu))^2))
+%    cor.mat[j,i] <- cor.mat[i,j]
+%  }
+%}
+%cov.mat[1:nSites,1:nSites] <- sigma.nu * cor.mat[1:nSites,1:nSites] * sigma.nu
+%tau.nu <- inverse(cov.mat)
+%p ~ dunif(0, 1)
+%
+%
+%for(t in 2:nYears) {
+%  nu[1:nSites,t-1] ~ dmnorm(mu.nu, tau.nu)
+%}
+%
+%for(i in 1:nSites) {
+%  N[i,1] ~ dpois(lambda)
+%  for(t in 2:nYears) {
+%      muN[i,t-1] <- N[i, t-1] * exp(nu[i, t-1] + r) #+ iota
+%      N[i,t] ~ dpois(muN[i, t-1])
+%  }
+%  for(t in 1:nYears) {
+%    y[i,t] ~ dbin(p, N[i,t])
+%  }
+%}
+%}
+%
+%\end{verbatim}
+
+\subsection{Random effects of observers}
+\begin{verbatim}
+model {
+lambda ~ dunif(0, 50)
+r ~ dunif(0,5)
+K ~ dunif(0, 50)
+p0 ~ dnorm(0, 0.01)
+sigma.p ~ dunif(0,10)
+tau.p <- 1 / (sigma.p * sigma.p)
+p1st ~ dnorm(0, 0.01)
+
+for (o in 1:nObs) {
+  logitpObs[o] ~ dnorm(p0,tau.p)
+}
+
+for(i in 1:nSites) {
+  N[i,1] ~ dpois(lambda)
+  logit(p[i,1]) <- logitpObs[obsMat[i,1]] 
+  y[i,1] ~ dbin(p[i,1], N[i,1])
+  for(t in 2:nYears) {
+    muN[i,t-1] <- N[i,t-1] * exp(r * (1 - log(N[i,t-1] + 1) / log(K + 1)))
+    N[i,t] ~ dpois(muN[i,t-1])
+    logit(p[i,t]) <- logitpObs[obsMat[i,t]] 
+    y[i,t] ~ dbin(p[i,t], N[i,t])
+    }
+  }
+}
+\end{verbatim}
+
 \subsection{Forecast model}
-\subsection{Simulation study code}
+Since the forecast model can be implemented identically to a purely retrospective model
+as far as JAGS code goes, we are including R code as well to show the differences in implementation.
+\begin{verbatim}
+library(rjags)
+
+# Settings
+nRun <- 100000 # Number of times to run the MCMC, after burnin
+nThin <- 25 # Thinning rate (only store 1 iteration out of every nThin)
+nChain <- 3 # Number of separate chains to run MCMC (for testing convergence)
+nBurn <- 10000 # Number of times to run the MCMC before getting output
+nAdapt <- 1000 # Number of times to run the MCMC for adaptation at beginning
+year1 <- 1966 # First year of estimation data
+year2 <- 2010 # Last year of estimation data/first year of projection
+year3 <- 2100 # Last year of projection
+thresholds <- c(1,10,50) # Quasi-extirpation levels to monitor
+
+# Read in input files
+ydata <- as.matrix(read.csv('oven3a.csv', row.names=1))
+observers <- as.matrix(read.csv('observers.csv', row.names=1))
+wind.factor <- as.matrix(read.csv('wind.factor.csv', row.names=1))
+# Fix formatting
+wind.factor[wind.factor==" 0"] <- "0"
+wind.factor[wind.factor==" 1"] <- "1"
+wind.factor[wind.factor==" 2"] <- "2"
+
+# More settings
+species = 'oven'
+aou <- 6740
+nSites <- nrow(ydata)
+nYears <- year3 - year1 + 1
+nPast <- year2 - year1 + 1
+nFuture <- year3 - year2
+
+# File names to save
+mod.name <- paste(species, 'sample.proj.JAGS', sep='.')
+file.name <- paste(mod.name, 'RData', sep='.')
+file.name2 <- paste(mod.name, 'gzip', sep='.')
+file.name2
+
+# For projection, fill in the future counts with all NA
+ydata <- cbind(ydata, matrix(NA, nrow=nSites, ncol=nFuture))
+
+fakeN <- ydata + 10 # Try to set up plausible initial values for N
+fillNAs <- function(N, default) { # Need to replace NAs in initial N with values
+  tMax <- ncol(N)
+  defaults <- which(is.na(N[,1]) & is.na(N[,2]))
+  N[defaults,1] <- default # Assign missing values from the 1st year which also have missing values for the 2nd year with default supplied
+  subsequent <- which(is.na(N[,1]))
+  N[subsequent,1] <- N[subsequent,2] # Assign missing values from the 1st year which don't have missing values for the 2nd year with the 2nd year's value
+  for (t in 2:(tMax-1)) {
+    # Assign missing values from this year which also have missing values for the next year with the values from the previous year
+    previous <- which(is.na(N[,t]) & is.na(N[,t+1]))
+    N[previous,t] <- N[previous,t-1] 
+    # Assign missing values from this year which don't have missing values for the next year with the mean of the previous and next year
+    means <- which(is.na(N[,t]))
+    N[means,t] <- ceiling((N[means, t-1] + N[means, t+1])/2)
+  }
+  # Assign missing values from the last year with the values from the previous year
+  previous <- which(is.na(N[,tMax]))
+  N[previous, tMax] <- N[previous, tMax - 1]
+  return(N)
+}
+fakeN <- fillNAs(fakeN, 24)
+head(fakeN)
+
+# Replace missing observer numbers with a new value, calculate number of observers
+obs.nr5 <- cbind(observers, matrix(NA, nrow=nSites, ncol=nFuture))
+nObs <- max(as.vector(obs.nr5), na.rm=T) + 1
+obs.nr5[is.na(obs.nr5)] = nObs
+# Set wind speed category values, counting routes not run as wind speed 0 
+wind1 <- cbind(matrix(as.integer(!is.na(wind.factor) & wind.factor=="1"), nSites, nPast),
+  matrix(0, nSites, nFuture))
+wind2 <- cbind(matrix(as.integer(!is.na(wind.factor) & wind.factor=="2"), nSites, nPast),
+  matrix(0, nSites, nFuture))
+wind3 <- cbind(matrix(as.integer(!is.na(wind.factor) & wind.factor=="3+"), nSites, nPast),
+  matrix(0, nSites, nFuture))
+
+# Ricker + immigration model, NB initial abundance
+# Logit-normal random observer effect
+# Wind (factor) effects on p
+# Projection model included
+sink(file="dm.proj.oven1.txt")
+cat("model {
+lambda ~ dunif(0, 200) # Average initial abundance
+alpha ~ dunif(0, 200) # Initial abundance overdispersion
+# JAGS doesn't allow lambda based parameterization of NB, so reparameterize
+P <- alpha/(alpha + lambda)
+# Carrying capacity (K) 
+K ~ dunif(0, 200)
+r ~ dunif(0, 2) # Instantaneous growth rate (return to equil. abundance)
+iota ~ dunif(0, 10) # Immigration rate
+# Detection probability parameters
+p0 ~ dnorm(0, 0.01) # Intercept
+sigma.p ~ dunif(0,10) # SD due to random observer effects
+tau.p <- 1 / (sigma.p * sigma.p) # Convert SD to precision
+p1 ~ dnorm(0, 0.01) # Effect of wind speed 1 on detection probability
+p2 ~ dnorm(0, 0.01) # Effect of wind speed 2 on detection probability
+p3 ~ dnorm(0, 0.01) # Effect of wind speed 3 or higher on detection probability
+
+# Random observer effects, by observer
+for (o in 1:nObs) {
+  logitpObs[o] ~ dnorm(p0,tau.p)
+}
+
+for(i in 1:nSites) {
+  N[i,1] ~ dnegbin(P, alpha) #Initial abundance
+  # Detection probability 
+  logit(p[i,1]) <- logitpObs[obsMat[i,1]] + p1 * wind1[i,1] + p2 * wind2[i,1] + 
+    p3 * wind3[i,1] 
+  # Observed counts, based on true abundance and detection probabilities
+  y[i,1] ~ dbin(p[i,1], N[i,1])
+  for(t in 2:nYears) {
+    # Expected abundance this time step (Ricker + immigration model)
+    muN[i,t-1] <- N[i,t-1] * exp(r * (1 - N[i,t - 1] / K)) + iota
+    # Due to demographic stochasticity, actual abundance differs from expected
+    N[i,t] ~ dpois(muN[i,t-1])
+    # Detection probability 
+    logit(p[i,t]) <- logitpObs[obsMat[i,t]] + p1 * wind1[i,t] + 
+      p2 * wind2[i,t] + p3 * wind3[i,t]
+    # Observed counts, based on true abundance and detection probabilities 
+    y[i,t] ~ dbin(p[i,t], N[i,t])
+  }
+}
+
+# If don't have enough memory to just trace N, instead trace derived variables 
+# of interest
+for(t in 1:nYears) { 
+  meanN[t] <- mean(N[,t]) # Average by year
+  for(j in 1:nThresholds) { 
+  # Proportion of sites (quasi-)extirpated by year (not cumulative over time)
+    PQE[j,t] <- sum(N[,t]<thresholds[j])/nSites 
+  } 
+}
+
+# If you're also interested in spatial patterns, 
+# snapshots for three years, but all sites
+N1 <- N[,1] 
+N2 <- N[,nPast] 
+N3 <- N[,nYears]
+
+}", fill=TRUE)
+sink()
+
+# Data to feed into JAGS
+dat.proj.oven1 <- list(nSites=nSites, nYears=nYears, nPast=nPast, 
+  nObs=nObs, y=ydata, thresholds=thresholds, nThresholds=length(thresholds),
+  obsMat = obs.nr5, wind1=wind1, wind2=wind2, wind3=wind3)
+# Initial values for parameters
+init.proj.oven1 <- function(chain) list(lambda=runif(1, 20, 75), alpha=runif(1, 0, 20), 
+  K=runif(1, 10, 50), iota=runif(1), r=runif(1, 0, 0.1), 
+  p0=rnorm(1), p1=rnorm(1), p2=rnorm(1), p3=rnorm(1), sigma.p=runif(1),
+  N=fakeN, .RNG.name="base::Wichmann-Hill", .RNG.seed=year1 + aou + chain)
+# Three options on what parameters to trace
+pars.proj.oven1 <- c("lambda", "alpha", "K", "r", "iota", "p0", "p1", "p2", 
+  "p3", "sigma.p")
+pars.proj.oven1.1 <- c("lambda", "alpha", "K", "r", "iota", "p0", "p1", "p2", 
+  "p3", "sigma.p", "meanN", "PQE", "N1", "N2", "N3")
+pars.proj.oven1.2 <- c("lambda", "alpha", "K", "r", "iota", "p0", "p1", "p2", 
+  "p3", "sigma.p", "N")
+
+set.seed(year1 + aou)
+jm.proj.oven1 <- jags.model("dm.proj.oven1.txt", dat.proj.oven1, 
+  init.proj.oven1, n.chains=nChain, n.adapt=nAdapt)
+update(jm.proj.oven1, n.iter=nBurn)
+jc.proj.oven1 <- coda.samples(jm.proj.oven1, pars.proj.oven1.1, n.iter=nRun,
+  thin=nThin)
+summary(jc.proj.oven1)
+save.image(file.name)
+png(paste(mod.name, 'converge.\%i.png', sep='.'))
+plot(jc.proj.oven1)
+dev.off()
+gelman.diag(jc.proj.oven1)
+save(jc.proj.oven1, file=file.name2)
+\end{verbatim}
+
 \subsection{BBS analysis code}
+\begin{verbatim}
+# Read in input files
+ydata <- as.matrix(read.csv('oven3a.csv', row.names=1))
+obs5 <- as.matrix(read.csv('observers.csv', row.names=1))
+first <- as.matrix(read.csv('first.run.csv', row.names=1))
+wind.factor <- as.matrix(read.csv('wind.factor.csv', row.names=1))
+# Fix formatting
+wind.factor[wind.factor==" 0"] <- "0"
+wind.factor[wind.factor==" 1"] <- "1"
+wind.factor[wind.factor==" 2"] <- "2"
+
+library(rjags)
+
+# Settings
+year1 <- 66
+species <- 'oven'
+nSites <- nrow(ydata)
+nYears <- ncol(ydata)
+
+fillN3 <- function(N, p, lambda, alpha) { # Need to replace NAs in initial N with values, adjust others by detection probability
+  for (i in 1:nrow(N)) {
+    notNA <- which(!is.na(N[i, ]))
+    size <- N[i, notNA] 
+    size[size==0] <- 0.5
+    addTo <- rnbinom(length(notNA), size=size, prob=p)
+    N[i, notNA] <- N[i, notNA] + addTo
+    for (t in 1:ncol(N))
+      if (is.na(N[i,t])) {
+        if (t == ncol(N) || is.na(N[i, t+1])) {
+          if (t==1) {
+            N[i, t] <- rnbinom(1, size=alpha, mu=lambda) #If no previous or subsequent value, use supplied default
+          }
+          else {
+            N[i, t] <- N[i, t-1] #If no subsequent value, use previous
+          }
+        }
+        else if (t==1)
+          N[i, t] <- N[i, t+1] #If no previous value, use subsequent
+        else
+          N[i, t] <- ceiling((N[i, t-1] + N[i, t+1])/2) #Otherwise, use mean
+      }
+  }
+  return(N)
+}
+
+N <- fillN3(ydata, 0.18, 33.4, 0.65)
+head(N)
+summary(N)
+
+# Other input variables
+nObs <- max(as.vector(obs5), na.rm=T) + 1
+obs5[is.na(obs5)] <- nObs
+wind1 <- matrix(as.integer(!is.na(wind.factor) & wind.factor=="1"), nSites, nYears)
+summary(wind1)
+wind2 <- matrix(as.integer(!is.na(wind.factor) & wind.factor=="2"), nSites, nYears)
+wind3 <- matrix(as.integer(!is.na(wind.factor) & wind.factor=="3+"), nSites, nYears)
+first[is.na(first)] <- 1
+
+# Repeat top model from frequentist analysis
+# Pick filename based on model
+mixture <- c('P','NB','ZIP')[2]
+dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[4]
+p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[4]] 
+r.model <- list(~1)[[1]] 
+K.model <- list(~1)[[1]] 
+lam.model <- list(~1)[[1]] 
+immigration <- TRUE
+iota.model <- list(~1) [[1]]
+p.rand <- c("None","Normal","Beta")[1]
+mod.name <- paste(paste(species, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'r', substring(deparse(r.model), 2), 
+  'K', substring(deparse(K.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'p', substring(deparse(p.model), 2), 'rand', p.rand, 'JAGS', sep='.') 
+file.name <- paste(mod.name, 'RData', sep='.')
+file.name
+
+sink(file="dm.ricki1.txt")
+cat("
+model {
+lambda ~ dunif(0, 200)
+alpha ~ dunif(0, 200)
+p.nb <- alpha/(alpha + lambda)
+r ~ dunif(0, 5)
+K ~ dunif(0, 200)
+p0 ~ dnorm(0, 0.01)
+p1 ~ dnorm(0, 0.01)
+p2 ~ dnorm(0, 0.01)
+p3 ~ dnorm(0, 0.01)
+p1st ~ dnorm(0, 0.01)
+iota ~ dunif(0, 15)
+for(i in 1:nSites) {
+  N[i,1] ~ dnegbin(p.nb, alpha)
+  for(t in 2:nYears) {
+      muN[i,t-1] <- N[i,t-1]*exp(r*(1-N[i,t-1]/K))+iota
+      N[i,t] ~ dpois(muN[i,t-1])
+    }
+  for(t in 1:nYears) {
+    logit(p[i,t]) <- p0 + p1*wind1[i,t] + p2*wind2[i,t] + p3*wind3[i,t] + p1st*first[i,t]
+    y[i,t] ~ dbin(p[i,t], N[i,t])
+    }
+  }
+}
+", fill=TRUE)
+sink()
+
+dat.ricki1 <- list(nSites=nSites, nYears=nYears, y=ydata, wind1=wind1, 
+  wind2=wind2, wind3=wind3, first=first)
+init.ricki1 <- function() list(lambda=runif(1, 10, 60), alpha=runif(1, 0, 60), 
+  r=runif(1), K=runif(1, 30, 120), iota=runif(1),
+  p0=rnorm(1, -1.5), p1=rnorm(1), p2=rnorm(1), p3=rnorm(1), p1st=rnorm(1), N=N)
+pars.ricki1 <- c("lambda", "alpha", "r", "K", "p0", "p1", "p2", "p3", "p1st", "iota")
+
+jm.ricki1 <- jags.model("dm.ricki1.txt", dat.ricki1, init.ricki1, n.chains=5, n.adapt=5000)
+jc.ricki1 <- coda.samples(jm.ricki1, pars.ricki1, n.iter=100000)
+
+summary(jc.ricki1)
+windows(record=T)
+plot(jc.ricki1)
+warnings()
+gelman.diag(jc.ricki1)
+
+jc.ricki1a <- window(jc.ricki1, 51001)
+summary(jc.ricki1a)
+gelman.diag(jc.ricki1a)
+
+save.image(file.name)
+
+# Add random observer effects
+rm(jc.ricki1)
+rm(jc.ricki1a)
+p.rand <- c("None","Normal","Beta")[2]
+mod.name <- paste(paste(species, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'r', substring(deparse(r.model), 2), 
+  'K', substring(deparse(K.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'p', substring(deparse(p.model), 2), 'rand', p.rand, 'JAGS', sep='.') 
+file.name = paste(mod.name, 'RData', sep='.')
+file.name
+
+sink(file="dm.ricki2.txt")
+cat("
+model {
+lambda ~ dunif(0, 200)
+alpha ~ dunif(0, 200)
+p.nb <- alpha/(alpha + lambda)
+r ~ dunif(0, 5)
+K ~ dunif(0, 200)
+p0 ~ dnorm(0, 0.01)
+p1 ~ dnorm(0, 0.01)
+p2 ~ dnorm(0, 0.01)
+p3 ~ dnorm(0, 0.01)
+p1st ~ dnorm(0, 0.01)
+iota ~ dunif(0, 15)
+sigma.p ~ dunif(0,15)
+tau.p <- 1 / (sigma.p * sigma.p)
+
+for (o in 1:nObs) {
+  eta[o] ~ dnorm(0,tau.p)
+}
+
+for(i in 1:nSites) {
+  N[i,1] ~ dnegbin(p.nb, alpha)
+  for(t in 2:nYears) {
+      muN[i,t-1] <- N[i,t-1]*exp(r*(1-N[i,t-1]/K))+iota
+      N[i,t] ~ dpois(muN[i,t-1])
+    }
+  for(t in 1:nYears) {
+    logit(p[i,t]) <- p0 + p1*wind1[i,t] + p2*wind2[i,t] + p3*wind3[i,t] + p1st*first[i,t] + eta[obsID[i,t]]
+    y[i,t] ~ dbin(p[i,t], N[i,t])
+    }
+  }
+}
+", fill=TRUE)
+sink()
+
+dat.ricki2 <- list(nSites=nSites, nYears=nYears, y=ydata, wind1=wind1, 
+  wind2=wind2, wind3=wind3, first=first, nObs=nObs, obsID=obs5)
+init.ricki2 <- function() list(lambda=runif(1, 10, 60), alpha=runif(1, 0, 60), 
+  r=runif(1, 0, 0.1), K=runif(1, 40, 70),
+  p0=rnorm(1, -1.5), p1=rnorm(1), p2=rnorm(1), p3=rnorm(1), p1st=rnorm(1), 
+  iota=runif(1), sigma.p=runif(1, 0, 2), N=N)
+pars.ricki2 <- c("lambda", "alpha", "r", "K", "iota", "p0", "p1", "p2", "p3", 
+  "p1st", "sigma.p")
+
+jm.ricki2 <- jags.model("dm.ricki2.txt", dat.ricki2, init.ricki2, n.chains=5, n.adapt=3000)
+jc.ricki2 <- coda.samples(jm.ricki2, pars.ricki2, n.iter=40000)
+
+summary(jc.ricki2)
+windows(record=T)
+plot(jc.ricki2)
+warnings()
+
+save.image(file.name)
+
+# Regional environmental stochasticity
+rm(jc.ricki2)
+p.rand <- c("None","Normal","Beta")[1]
+mod.name <- paste(paste(species, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'r', substring(deparse(r.model), 2), 
+  'K', substring(deparse(K.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'p', substring(deparse(p.model), 2), 'rand', p.rand, 'JAGS.ES', sep='.') 
+file.name = paste(mod.name, 'RData', sep='.')
+file.name
+
+sink(file="dm.ricki3.txt")
+cat("
+model {
+lambda ~ dunif(0, 400)
+alpha ~ dunif(0, 200)
+p.nb <- alpha/(alpha + lambda)
+r ~ dunif(0, 3)
+sigma.nu ~ dunif(0, 10)
+tau.nu <- 1 / (sigma.nu * sigma.nu)
+K ~ dunif(0, 400)
+p0 ~ dnorm(0, 0.01)
+p1 ~ dnorm(0, 0.01)
+p2 ~ dnorm(0, 0.01)
+p3 ~ dnorm(0, 0.01)
+p1st ~ dnorm(0, 0.01)
+iota ~ dunif(0, 15)
+
+for(t in 2:nYears) {
+  nu[t-1] ~ dnorm(0, tau.nu)
+}
+
+for(i in 1:nSites) {
+  N[i,1] ~ dnegbin(p.nb, alpha)
+  for(t in 2:nYears) {
+      muN[i,t-1] <- N[i,t-1]*exp(nu[t-1]+r*(1-N[i,t-1]/K))+iota
+      N[i,t] ~ dpois(muN[i,t-1])
+    }
+  for(t in 1:nYears) {
+    logit(p[i,t]) <- p0 + p1*wind1[i,t] + p2*wind2[i,t] + p3*wind3[i,t] + p1st*first[i,t]
+    y[i,t] ~ dbin(p[i,t], N[i,t])
+    }
+  }
+}
+", fill=TRUE)
+sink()
+
+dat.ricki3 <- list(nSites=nSites, nYears=nYears, y=ydata, wind1=wind1, 
+  wind2=wind2, wind3=wind3, first=first) 
+init.ricki3 <- function() list(lambda=runif(1, 10, 60), alpha=runif(1, 0, 60), 
+  r=runif(1, 0, 0.2), sigma.nu=runif(1), K=runif(1, 30, 120), iota=runif(1),
+  p0=rnorm(1, -1.5), p1=rnorm(1), p2=rnorm(1), p3=rnorm(1), p1st=rnorm(1), N=N)
+pars.ricki3 <- c("lambda", "alpha", "r", "sigma.nu", "K", "iota", 
+  "p0", "p1", "p2", "p3", "p1st")
+
+jm.ricki3 <- jags.model("dm.ricki3.txt", dat.ricki3, init.ricki3, n.chains=5, n.adapt=1000)
+update(jm.ricki3, 50000)
+jc.ricki3 <- coda.samples(jm.ricki3, pars.ricki3, n.iter=130000)
+
+summary(jc.ricki3)
+save.image(file.name)
+HPDinterval(jc.ricki3)
+gelman.diag(jc.ricki3)
+png(paste(mod.name, '%i.png', sep='.'))
+plot(jc.ricki3)
+dev.off()
+
+# Regional environmental stochasticity AND random observer effects
+summary(jc.ricki3)
+p.rand <- c("None","Normal","Beta")[2]
+mod.name <- paste(paste(species, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'r', substring(deparse(r.model), 2), 
+  'K', substring(deparse(K.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'p', substring(deparse(p.model), 2), 'rand', p.rand, 'JAGS.ES', sep='.') 
+file.name <- paste(mod.name, 'RData', sep='.')
+file.name
+
+sink(file="dm.ricki4.txt")
+cat("
+model {
+lambda ~ dunif(0, 400)
+alpha ~ dunif(0, 200)
+p.nb <- alpha/(alpha + lambda)
+r ~ dunif(0, 3)
+sigma.nu ~ dunif(0, 10)
+tau.nu <- 1 / (sigma.nu * sigma.nu)
+K ~ dunif(0, 1000)
+p0 ~ dnorm(0, 0.01)
+p1 ~ dnorm(0, 0.01)
+p2 ~ dnorm(0, 0.01)
+p3 ~ dnorm(0, 0.01)
+p1st ~ dnorm(0, 0.01)
+iota ~ dunif(0, 15)
+sigma.p ~ dunif(0,15)
+tau.p <- 1 / (sigma.p * sigma.p)
+
+for (o in 1:nObs) {
+  eta[o] ~ dnorm(0,tau.p)
+}
+
+for(t in 2:nYears) {
+  nu[t-1] ~ dnorm(0, tau.nu)
+}
+
+for(i in 1:nSites) {
+  N[i,1] ~ dnegbin(p.nb, alpha)
+  for(t in 2:nYears) {
+      muN[i,t-1] <- N[i,t-1]*exp(nu[t-1]+r*(1-N[i,t-1]/K))+iota
+      N[i,t] ~ dpois(muN[i,t-1])
+    }
+  for(t in 1:nYears) {
+    logit(p[i,t]) <- p0 + p1*wind1[i,t] + p2*wind2[i,t] + p3*wind3[i,t] + p1st*first[i,t] + eta[obsID[i,t]]
+    y[i,t] ~ dbin(p[i,t], N[i,t])
+    }
+  }
+}
+", fill=TRUE)
+sink()
+
+dat.ricki4 <- list(nSites=nSites, nYears=nYears, y=ydata, wind1=wind1, 
+  wind2=wind2, wind3=wind3, nObs=nObs, obsID=obs5, first=first)
+init.ricki4 <- function() list(lambda=runif(1, 10, 60), alpha=runif(1, 0, 60), 
+  r=runif(1, 0, 0.2), sigma.nu=runif(1), K=runif(1, 30, 120), iota=runif(1),
+  p0=rnorm(1, -1.5), p1=rnorm(1), p2=rnorm(1), p3=rnorm(1), p1st=rnorm(1), sigma.p=runif(1), N=N)
+pars.ricki4 <- c("lambda", "alpha", "r", "sigma.nu", "K", "iota", 
+  "p0", "p1", "p2", "p3", "p1st", "sigma.p")
+
+jm.ricki4 <- jags.model("dm.ricki4.txt", dat.ricki4, init.ricki4, n.chains=5, n.adapt=1000)
+update(jm.ricki4, 50000)
+jc.ricki4 <- coda.samples(jm.ricki4, pars.ricki4, n.iter=130000)
+
+summary(jc.ricki4)
+save.image(file.name)
+HPDinterval(jc.ricki4)
+gelman.diag(jc.ricki4)
+
+jpeg(paste(mod.name, '%i.jpg', sep='.'))
+plot(jc.ricki4)
+dev.off()
+
+# Track population sizes (includes random observer effects but not environmental stochasticity)
+rm(jc.ricki4)
+p.rand <- c("None","Normal","Beta")[2]
+mod.name <- paste(paste(species, year1, sep=''), mixture, dynamics, 'lam', 
+  substring(deparse(lam.model), 2), 'r', substring(deparse(r.model), 2), 
+  'K', substring(deparse(K.model), 2), 
+  ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+  'p', substring(deparse(p.model), 2), 'rand', p.rand, 'JAGS.N', sep='.') 
+file.name <- paste(mod.name, 'RData', sep='.')
+file.name
+
+sink(file="dm.ricki5.txt")
+cat("
+var Ndecade[nSites,5]
+model {
+lambda ~ dunif(0, 200)
+alpha ~ dunif(0, 200)
+p.nb <- alpha/(alpha + lambda)
+r ~ dunif(0, 5)
+K ~ dunif(0, 200)
+p0 ~ dnorm(0, 0.01)
+p1 ~ dnorm(0, 0.01)
+p2 ~ dnorm(0, 0.01)
+p3 ~ dnorm(0, 0.01)
+p1st ~ dnorm(0, 0.01)
+iota ~ dunif(0, 15)
+sigma.p ~ dunif(0,15)
+tau.p <- 1 / (sigma.p * sigma.p)
+
+for (o in 1:nObs) {
+  eta[o] ~ dnorm(0,tau.p)
+}
+
+for(i in 1:nSites) {
+  N[i,1] ~ dnegbin(p.nb, alpha)
+  for(t in 2:nYears) {
+      muN[i,t-1] <- N[i,t-1]*exp(r*(1-N[i,t-1]/K))+iota
+      N[i,t] ~ dpois(muN[i,t-1])
+    }
+  for(t in 1:nYears) {
+    logit(p[i,t]) <- p0 + p1*wind1[i,t] + p2*wind2[i,t] + p3*wind3[i,t] + p1st*first[i,t] + eta[obsID[i,t]]
+    y[i,t] ~ dbin(p[i,t], N[i,t])
+    }
+  }
+
+for(t in 1:nYears) {
+  Nmean[t] <- mean(N[,t])
+}
+
+for (t2 in 1:length(decades)) {
+  for(i in 1:nSites) {
+    Ndecade[i,t2] <- N[i,decades[t2]]
+  }
+}
+}
+", fill=TRUE)
+sink()
+
+dat.ricki5 <- list(nSites=nSites, nYears=nYears, y=ydata, wind1=wind1, 
+  wind2=wind2, wind3=wind3, nObs=nObs, obsID=obs5, first=first,
+  decades=as.integer(seq(5, 45, 10)))
+init.ricki5 <- function() list(lambda=runif(1, 10, 60), alpha=runif(1, 0, 60), 
+  r=runif(1), K=runif(1, 30, 120), iota=runif(1),
+  p0=rnorm(1, -1.5), p1=rnorm(1), p2=rnorm(1), p3=rnorm(1), p1st=rnorm(1), sigma.p=runif(1), N=N)
+pars.ricki5 <- c("lambda", "alpha", "r", "K", "p0", "p1", "p2", "p3", "p1st", 
+  "sigma.p", "iota", "Nmean", "Ndecade")
+
+jm.ricki5 <- jags.model("dm.ricki5.txt", dat.ricki5, init.ricki5, n.chains=1, n.adapt=1000)
+update(jm.ricki5, n.iter=100000)
+jc.ricki5 <- coda.samples(jm.ricki5, pars.ricki5, n.iter=40000)
+
+# Summarize results
+mat.ricki5 <- as.matrix(jc.ricki5)
+NmeanCols<- which(colnames(mat.ricki5)=="Nmean[1]") + 1:nYears - 1
+Nmean.frame<-data.frame(year=1966:2010, Mean=colMeans(mat.ricki5[,NmeanCols]), HPDinterval(jc.ricki5[,NmeanCols]))
+Nmean.frame
+mean.N1 <- colMeans(mat.ricki5[,2:(122*5+1)])
+mean.N1
+mean.N2 <- matrix(mean.N1, nrow=122, ncol=5)
+mean.N2
+colnames(mean.N2) <- paste("N", seq(1970,2010,10), sep="")
+max(mean.N2)
+max.plot <- round(max(mean.N2)/5, -1)*5
+mean.N2 <- with(rt2a, data.frame(statenum, Route, Lati, Longi, mean.N2))
+head(mean.N2)
+
+# Measures of variability
+sd.N1 <- apply(mat.ricki5[,2:(122*45+1)], 2, sd)
+sd.N1
+mean(sd.N1)
+mean(mean.N1)
+mean(mat.ricki5[,"lambda"])
+sd(mat.ricki5[,"lambda"])
+sd(mat.ricki5[,"lambda"])/mean(mat.ricki5[,"lambda"])
+sd(mat.ricki5[,"alpha"])/mean(mat.ricki5[,"alpha"])
+sd(mat.ricki5[,"iota"])/mean(mat.ricki5[,"iota"])
+sd(mat.ricki5[,"r"])/mean(mat.ricki5[,"r"])
+summary(sd.N1/mean.N1)
+sd.N2 <- apply(mean.N2, 2, sd) #Variation between sites in a year
+sd.N2
+
+# Make plot
+library(ggplot2)
+library(gridExtra)
+library(maps)
+
+all.states <- map_data("state")
+states <- subset(all.states, region %in% c('maryland', 'virginia'))
+
+max.radius <- 4
+leg.sizes <- c(1, 10, 100, 250)
+gN.sp <- list()
+gN.sp[[1]] <- ggplot(mean.N2, aes(x=Longi,y=Lati)) + theme_bw() + labs(x="", y="")+
+  geom_polygon(data=states, aes(x=long, y=lat, group=group), color="black", fill="white") +
+  geom_point(aes(size=N1970), alpha=0.5, color="black")+scale_area(name="Abundance", range=c(1, max.radius), breaks=leg.sizes)+
+  geom_text(x=-83.2, y=39.75, label="1970")
+gN.sp[[2]] <- ggplot(mean.N2, aes(x=Longi,y=Lati)) + theme_bw() + labs(x="", y="")+
+  geom_polygon(data=states, aes(x=long, y=lat, group=group), color="black", fill="white") +
+  geom_point(aes(size=N1980), alpha=0.5, color="black")+scale_area(name="Abundance", range=c(1, max.radius), breaks=leg.sizes)+
+  geom_text(x=-83.2, y=39.75, label="1980")
+gN.sp[[3]] <- ggplot(mean.N2, aes(x=Longi,y=Lati)) + theme_bw() + labs(x="", y="")+
+  geom_polygon(data=states, aes(x=long, y=lat, group=group), color="black", fill="white") +
+  geom_point(aes(size=N1990), alpha=0.5, color="black")+scale_area(name="Abundance", range=c(1, max.radius), breaks=leg.sizes)+
+  geom_text(x=-83.2, y=39.75, label="1990")
+gN.sp[[4]] <- ggplot(mean.N2, aes(x=Longi,y=Lati)) + theme_bw() + labs(x="", y="")+
+  geom_polygon(data=states, aes(x=long, y=lat, group=group), color="black", fill="white") +
+  geom_point(aes(size=N2000), alpha=0.5, color="black")+scale_area(name="Abundance", range=c(1, max.radius), breaks=leg.sizes)+
+  geom_text(x=-83.2, y=39.75, label="2000")
+gN.sp[[5]] <- ggplot(mean.N2, aes(x=Longi,y=Lati)) + theme_bw() + labs(x="", y="")+
+  geom_polygon(data=states, aes(x=long, y=lat, group=group), color="black", fill="white") +
+  geom_point(aes(size=N2010), alpha=0.5, color="black")+scale_area(name="Abundance", range=c(1, max.radius), breaks=leg.sizes)+
+  geom_text(x=-83.2, y=39.75, label="2010")
+
+gN.mean <- ggplot(Nmean.frame.f, aes(x=year,y=Mean)) + theme_bw() + labs(x="Year",
+  y="Mean Abundance")+
+  geom_ribbon(aes(ymin=lower,ymax=upper),fill="black",
+  alpha=0.5,colour=NA) + geom_line(colour="black", size=1)
+
+png(filename = "OVEN_N_by_route_year6.png", height = 6.5, width = 6.5, units = "in", res=600)
+print(arrangeGrob(gN.sp[[1]] + theme(legend.position=c(0.145, 0.56), 
+  legend.background=element_blank(), axis.text=element_blank(), 
+  axis.line=element_blank(), axis.ticks=element_blank(), panel.border=element_blank(),
+  panel.grid=element_blank(), plot.margin=unit(rep(0.1, 4), "lines")),
+  gN.sp[[2]] + theme(legend.position="none", axis.text=element_blank(), 
+  axis.line=element_blank(), axis.ticks=element_blank(), panel.border=element_blank(),
+  panel.grid=element_blank(), plot.margin=unit(rep(0.1, 4), "lines")),
+  gN.sp[[3]] + theme(legend.position="none", axis.text=element_blank(), 
+  axis.line=element_blank(), axis.ticks=element_blank(), panel.border=element_blank(),
+  panel.grid=element_blank(), plot.margin=unit(rep(0.1, 4), "lines")),
+  gN.sp[[4]] + theme(legend.position="none", axis.text=element_blank(), 
+  axis.line=element_blank(), axis.ticks=element_blank(), panel.border=element_blank(),
+  panel.grid=element_blank(), plot.margin=unit(rep(0.1, 4), "lines")),
+  gN.sp[[5]] + theme(legend.position="none", axis.text=element_blank(), 
+  axis.line=element_blank(), axis.ticks=element_blank(), panel.border=element_blank(),
+  panel.grid=element_blank(), plot.margin=unit(rep(0.1, 4), "lines")),
+  gN.mean, main=" ", nrow=3, ncol=2))
+dev.off()
+
+save.image(file.name)
+\end{verbatim}
 

### 2013-09-18 supp/bbs_estimates_appendix.tex
Edits from Scott plus appendices updated

\documentclass{article}
\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[pdftex]{hyperref}
\usepackage{soul}
\usepackage{graphicx}

\begin{document}
\title{Appendix 3: BBS Results}
\maketitle
\renewcommand*\thetable{S3.\arabic{table}}
\renewcommand*\thefigure{S3.\arabic{figure}}
\begin{table}[htb]
  \centering
  \small
\caption{Model selection table for ovenbirds in Maryland and Virginia,
    1966-2010.  We present model name and number, number of 
parameters (Par.), and difference in Akaike's
information criterion between each model and the top model of
that set ($\Delta$AIC).  The first section compares
models for initial abundance, the second for detection
probability, and the third for dynamics.}
  \begin{tabular}[h]{lcc}
\hline
Model	&Par.	&$\Delta$AIC	\\
\hline
A. Initial Abundance && \\
A.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0\\
A.2. P[$\Lambda$(.)]Exponential[$r$(.)]$p$(.)	&3	&1262.7\\
A.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4 &1264.7\\
\hline
B. Detection Probability && \\
B.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &0	\\
B.2. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind) &7 &0.9\\
B.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(1st)	&5	&5.0
\\B.4. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&6.4\\
\hline
C. Dynamics && \\
C.1. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&0	\\
C.2. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&8.4 \\
C.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(wind+1st) &9	&36.5\\
C.4. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st) &10	&38.6\\
C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&192.8\\
C.6. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&195.1\\
C.7. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &271.3\\
C.8. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&273.7\\
C.9. NB[$\Lambda$(.)$\alpha$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&1856.7\\
\hline
\end{tabular}
\end{table}
\clearpage

[Andy and Richard, which do you prefer for parameter estimates?  
Table or figure?  Either one may need a bit of work still.]
% latex table generated in R 3.0.0 by xtable 1.7-1 package
% Wed Sep 18 14:26:41 2013
\begin{table}[ht]
\centering
  \caption{Parameter estimates for ovenbirds in Maryland and Virginia from BBS data, 1966-2010.  
MLE estimates come from the NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) 
model run in the \textbf{R} package \texttt{unmarked}; base estimates from the same model run in \textbf{JAGS};
obs estimates from the base model with random observer effects added; and ES estimates from the base
model with regional environmental stochasticity added.  Detectability parameters (intercept: $p_{0}$, 
effect of wind speed 1: $p_{1}$, effect of wind speed 2: $p_{2}$, effect of wind speed 3 or higher: $p_{3}$, 
effect of first run of a route by an observer: $p_{1st}$, and random observer effect: $\sigma_{p}$) are on the 
logit scale.}
\begin{tabular}{rllllll}
  \hline
 & Parameter & MLE & base & obs & ES & ES and obs \\ 
  \hline
1 & $\Lambda$ & 33.4 $\pm$ 4.7 & 31.1 $\pm$ 5.2 & 42.5 $\pm$ 7.3 & 33.2 $\pm$ 5.8 & 43.8 $\pm$ 7.9 \\ 
  2 & $\alpha$ & 0.654 $\pm$ 0.087 & 0.624 $\pm$ 0.108 & 0.730 $\pm$ 0.119 & 0.649 $\pm$ 0.110 & 0.753 $\pm$ 0.124 \\ 
  3 & $r$ & 0.026 $\pm$ 0.006 & 0.011 $\pm$ 0.005 & 0.018 $\pm$ 0.006 & 0.020 $\pm$ 0.010 & 0.023 $\pm$ 0.010 \\ 
  4 & $K$ & 56.9 $\pm$ 9.4 & 55.8 $\pm$ 18.2 & 101.8 $\pm$ 21.1 & 112.0 $\pm$ 58.8 & 141.5 $\pm$ 69.5 \\ 
  5 & $\iota$ & 0.283 $\pm$ 0.043 & 0.333 $\pm$ 0.043 & 0.306 $\pm$ 0.051 & 0.329 $\pm$ 0.046 & 0.303 $\pm$ 0.052 \\ 
  6 & $p_{0}$ & -1.492 $\pm$ 0.100 & -1.476 $\pm$ 0.097 & -1.994 $\pm$ 0.116 & -1.571 $\pm$ 0.099 & -2.056 $\pm$ 0.124 \\ 
  7 & $p_{1}$ & -0.012 $\pm$ 0.021 & -0.011 $\pm$ 0.021 & -0.014 $\pm$ 0.021 & -0.013 $\pm$ 0.021 & -0.015 $\pm$ 0.021 \\ 
  8 & $p_{2}$ & -0.075 $\pm$ 0.027 & -0.071 $\pm$ 0.027 & -0.056 $\pm$ 0.026 & -0.069 $\pm$ 0.027 & -0.055 $\pm$ 0.026 \\ 
  9 & $p_{3}$ & -0.133 $\pm$ 0.049 & -0.135 $\pm$ 0.049 & -0.145 $\pm$ 0.048 & -0.141 $\pm$ 0.049 & -0.144 $\pm$ 0.048 \\ 
  10 & $p_{1st}$ & -0.058 $\pm$ 0.028 & -0.047 $\pm$ 0.027 & -0.075 $\pm$ 0.029 & -0.045 $\pm$ 0.028 & -0.075 $\pm$ 0.029 \\ 
  31 & $\sigma_{p}$ & - & - & 0.353 $\pm$ 0.034 & - & 0.349 $\pm$ 0.034 \\ 
  42 & $\sigma_{\nu}$ & - & - & - & 0.074 $\pm$ 0.014 & 0.064 $\pm$ 0.014 \\ 
   \hline
\end{tabular}
\end{table}

\clearpage
\begin{figure}
  \centering
  \includegraphics[width=6.5in]{../figs/oven_par_est}
\caption{Parameter estimates for ovenbirds in Maryland and Virginia from BBS data, 1966-2010.  
MLE estimates come from the NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) 
model run in the \textbf{R} package \texttt{unmarked}; base estimates from the same model run in \textbf{JAGS};
obs estimates from the base model with random observer effects added; and ES estimates from the base
model with regional environmental stochasticity added.  Detectability parameters (intercept: $p_{0}$, 
effect of wind speed 1: $p_{1}$, effect of wind speed 2: $p_{2}$, effect of wind speed 3 or higher: $p_{3}$, 
effect of first run of a route by an observer: $p_{1st}$, and random observer effect: $\sigma_{p}$) are on the 
logit scale.}
\label{fig:oven_par_est}
\end{figure}

\end{document}

### 2013-09-18 supp/sim_tables_appendix.tex
Edits from Scott plus appendices updated

\documentclass{article}
\usepackage[landscape,a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[pdftex]{hyperref}
\usepackage{soul}

\begin{document}
\title{Appendix 2: Simulation Results}
\maketitle
\renewcommand*\thetable{S2.\arabic{table}}
\begin{table}[htb]
  \centering
  \footnotesize
  \caption{Simulation results from 27 cases of exponential dynamics with 1000 Monte Carlo replicates for each.}
  \begin{tabular}{cccccccccccc}
    \hline
    \multicolumn{3}{c}{Data generating values}&
    \multicolumn{3}{c}{Mean} &
    \multicolumn{3}{c}{RMSE} &
    \multicolumn{3}{c}{Coverage} \\
    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
    \multicolumn{3}{c}{\rule{4cm}{1pt}} \\
%    \hline
    $\lambda$ & $r$ & $p$ & $\lambda$ & $r$ & $p$ & $\lambda$ & $r$ &
    $p$ & $\lambda$ & $r$ & $p$ \\
    \hline
10	&-0.005	&0.05	&10.03924	&-0.00601	&0.05007	&1.05051	&0.00605	&0.00432	&0.959	&0.945	&0.96\\
10	&-0.005	&0.25	&10.0333	&-0.00573	&0.25&0.76926	&0.00598	&0.01545 &0.951	&0.942	&0.945\\
10	&-0.005	&0.5	&10.00858	&-0.00598	&0.49984	&0.61057	&0.00532	&0.02352	&0.949	&0.957	&0.945\\
10	&0.005	&0.05	&10.06802	&0.00424	&0.04996	&1.04574	&0.00576	&0.00428	&0.955	&0.939	&0.949\\
10	&0.005	&0.25	&10.03648	&0.00399	&0.25087	&0.80109	&0.00525	&0.01665	&0.934	&0.954	&0.921\\
10	&0.005	&0.5	&9.96719	&0.00431	&0.50244	&0.61777	&0.00492	&0.02539	&0.934	&0.957	&0.92\\
10	&0	&0.05	&10.06548	&-9e-04	        &0.05001	&1.06134	&0.00553	&0.0046	&0.961	&0.96	&0.944\\
10	&0	&0.25	&10.03522	&-0.00093	&0.25012	&0.76321	&0.00555	&0.01585	&0.953	&0.943	&0.936\\
10	&0	&0.5	&10.0036	&-0.00067	&0.50052	&0.58102	&0.00544	&0.02274	&0.949	&0.945	&0.948\\
1	&-0.005	&0.05	&1.04714	&-0.01254	&0.05045	&0.24087	&0.02411	&0.00915	&0.949	&0.953	&0.97\\
1	&-0.005	&0.25	&1.01614	&-0.01173	&0.25038	&0.15418	&0.02267	&0.02749	&0.952	&0.949	&0.945\\
1	&-0.005	&0.5	&1.01056	&-0.01136	&0.50008	&0.13529	&0.02287	&0.03846	&0.957	&0.936	&0.95\\
1	&0.005	&0.05	&1.0281	        &-5e-05	        &0.05032	&0.23616	&0.02057	&0.00954	&0.939	&0.948	&0.947\\
1	&0.005	&0.25	&1.01389	&-0.00143	&0.2502	&0.1609	&0.02246	&0.0268	&0.957	&0.937	&0.946\\
1	&0.005	&0.5	&1.00554	&-0.00114	&0.50194	&0.13693	&0.02064	&0.03746	&0.947	&0.945	&0.95\\
1	&0	&0.05	&1.03641	&-0.00685	&0.05017	&0.23656	&0.02457	&0.00928	&0.958	&0.934	&0.961\\
1	&0	&0.25	&1.01556	&-0.00587	&0.24993	&0.16434	&0.02134	&0.02592	&0.944	&0.945	&0.953\\
1	&0	&0.5	&1.00539	&-0.00533	&0.49994	&0.13367	&0.01982	&0.03812	&0.951	&0.947	&0.952\\
5	&-0.005	&0.05	&5.07525	&-0.00638	&0.04978	&0.65646	&0.00881	&0.00496	&0.945	&0.959	&0.965\\
5	&-0.005	&0.25	&5.02641	&-0.00619	&0.25013	&0.4727	&0.00779	&0.01735	&0.943	&0.966	&0.937\\
5	&-0.005	&0.5	&5.01265	&-0.00656	&0.50067	&0.36154	&0.00814	&0.02465	&0.96	&0.949	&0.961\\
5	&0.005	&0.05	&5.05623	&0.00396	&0.05002	&0.62418	&0.00786	&0.00505	&0.944	&0.957	&0.952\\
5	&0.005	&0.25	&5.01567	&0.0041	&0.25014	&0.44585	&0.00746	&0.01723	&0.948	&0.947	&0.942\\
5	&0.005	&0.5	&5.02672	&0.00368	&0.49908	&0.36657	&0.00727	&0.02475	&0.947	&0.95	&0.943\\
5	&0	&0.05	&5.06304	&-0.00162	&0.04994	&0.61073	&0.00838	&0.00492	&0.96	&0.96	&0.956\\
5	&0	&0.25	&5.04372	&-0.00141	&0.24927	&0.46542	&0.00812	&0.01709	&0.942	&0.938	&0.945\\
5	&0	&0.5	&5.0239	&-0.00117	&0.4991	&0.37326	&0.00747	&0.02487	&0.952	&0.964	&0.946\\
  \hline
  \end{tabular}
  \label{tab:simexp}
\end{table}
\clearpage
\begin{table}
  \centering
  \footnotesize
  \caption{Simulation results from 9 cases of Ricker dynamics with 1000 Monte Carlo replicates for each.}
  \begin{tabular}{cccccccccccccccc}
    \hline
    \multicolumn{4}{c}{Data generating values}&
    \multicolumn{4}{c}{Mean} &
    \multicolumn{4}{c}{RMSE} &
    \multicolumn{4}{c}{Coverage} \\
    \multicolumn{4}{c}{\rule{4cm}{1pt}} &
    \multicolumn{4}{c}{\rule{4cm}{1pt}} &
    \multicolumn{4}{c}{\rule{4cm}{1pt}} &
    \multicolumn{4}{c}{\rule{4cm}{1pt}} \\
%    \hline
    $\lambda$ & $r$ & $K$ & $p$ & $\lambda$ & $r$ & $K$ & $p$ & $\lambda$ & $r$ & $K$ &
    $p$ & $\lambda$ & $r$ & $K$ & $p$ \\
    \hline
10	&0.005	&5	&0.25	&9.914 &0.00982  &6.368 &0.25211	&0.78540 &0.01147   &5.56126 &0.01614	&0.955 &0.807 &0.776 &0.963\\
10	&0.005	&10	&0.25	&9.954 &0.00866 &10.989 &0.25179 &0.77529 &0.00929   &8.73913 &0.01613  &0.960 &0.831 &0.828 &0.943\\
10	&0.005	&20	&0.25	&9.935 &0.00827 &59.778 &0.25145 &0.76079 &0.00830 &418.37837 &0.01559  &0.955 &0.841 &0.943 &0.955\\
10	&0.05	&5	&0.25	&10.038 &0.05290  &4.682 &0.25160 &0.89588 &0.02748   &1.58258 &0.01884  &0.945 &0.922 &0.919 &0.951\\
10	&0.05	&10	&0.25	&10.009 &0.05270  &9.718 &0.25172 &0.83411 &0.01876   &1.74552 &0.01742  &0.961 &0.930 &0.960 &0.948\\
10	&0.05	&20	&0.25	&10.000 &0.05136 &19.790 &0.25082 &0.83906 &0.01316   &2.51289 &0.01670  &0.951 &0.943 &0.962 &0.940\\
10	&0.1	&5	&0.25	&10.046 &0.10127  &4.833 &0.25048 &0.91333 &0.03529   &0.87551 &0.01953  &0.948 &0.954 &0.944 &0.954\\
10	&0.1	&10	&0.25	&9.990 &0.10254  &9.893 &0.25143 &0.88385 &0.02267   &0.89598 &0.01800  &0.949 &0.954 &0.964 &0.950\\
10	&0.1	&20	&0.25	&9.988 &0.10185 &19.916 &0.25117 &0.85780 &0.01543   &1.62111 &0.01680  &0.947 &0.941 &0.951 &0.953\\
  \hline
  \end{tabular}
  \label{tab:simricker}
\end{table}

\clearpage
\begin{table}
  \centering
  \footnotesize
  \caption{Simulation results from 3 cases of Ricker-logistic and immigration dynamics with 1000 Monte Carlo replicates for each.}
  \begin{tabular}{cccccccccccccccccccc}
    \hline
    \multicolumn{5}{c}{Data generating values}&
    \multicolumn{5}{c}{Mean} &
    \multicolumn{5}{c}{RMSE} &
    \multicolumn{5}{c}{Coverage} \\
    \multicolumn{5}{c}{\rule{4cm}{1pt}} &
    \multicolumn{5}{c}{\rule{7cm}{1pt}} &
    \multicolumn{5}{c}{\rule{7cm}{1pt}} &
    \multicolumn{5}{c}{\rule{4cm}{1pt}} \\
%    \hline
    $\lambda$ & $r$ & $K$ & $\iota$ & $p$ & $\lambda$ & $r$ & $K$ & $\iota$ & $p$ & $\lambda$ & $r$ & $K$ &
    $\iota$ & $p$ & $\lambda$ & $r$ & $K$ & $\iota$ & $p$ \\
    \hline
10 &0.05 &10 &0.005 &0.25  &9.997 &0.05148 &9.687 &0.00510 &0.25077 &0.87742 &0.01805 &1.77679 &0.00315 &0.01795  &0.951 &0.944 &0.964 &0.947 &0.942\\
10 &0.05 &10 &0.050 &0.25 &10.015 &0.05166 &9.758 &0.05039 &0.25070 &0.87490 &0.01731 &1.72376 &0.01055 &0.01839  &0.943 &0.932 &0.958 &0.940 &0.943\\
10 &0.05 &10 &0.500 &0.25 &10.000 &0.05057 &9.648 &0.50811 &0.25110 &0.96491 &0.01942 &2.68264 &0.07763 &0.02050  &0.947 &0.940 &0.952 &0.951 &0.941\\
  \hline
  \end{tabular}
  \label{tab:simricker}
\end{table}

\end{document}

### 2013-10-14 Dail_Mad_ext.tex
Edits from Andy

@@
 \usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
-\usepackage[mathlines,displaymath]{lineno}
+\usepackage[mathlines]{lineno} %,displaymath
 \usepackage{setspace}
@@
 
+%\doublespacing
 \author{Jeffrey A. Hostetler\\ %\footnote{Contact:
@@
 
-\vspace{0.5cm}
+%\vspace{0.5cm}
 
@@
 %\newpage
+\vspace{0.5cm}
 
@@
 modeled using a population growth model, $\mu(N_{t-1})$
-and random process variation denoted $\grave{o}_t$.
+and random process variation denoted $\eta_t$.
+% Richard, is switching this from $\grave{o}$ to $\eta$ okay?  Andy thought
+% the o with accent was weird and I think I agree.  $\nu$ seems standard,
+% but we use it for a slightly different meaning (ES only) later.  
 The growth model may be density-dependent, as in the case of the 
@@
     N_1 &= X_1 \label{eq:ss1a} \\
-N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \;
+N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
 t=2,\hdots,T \label{eq:ss1b} \\
@@
 \end{subequations}
-where $\grave{o}_t$ is the random effect allowing for process
+where $\eta_t$ is the random effect allowing for process
 variation unaccounted for by the deterministic model. In classical
@@
 are assumed to be i.i.d. Gaussian deviates:
-$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
+$\eta_t \sim \mathrm{N}(0, \sigma)$ and
 $\zeta_t \sim \mathrm{N}(0, \tau)$. Standard practice is to
@@
 Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
-which $N_1$ is a random variable. Thus, uncertainty exists
+although $N_1$ is a random variable. Thus, uncertainty exists
 that should be accounted for in the model.  
 %are needed to fully specify the model.
-%To be more specific, a fully-specified state-space model requires at least
-%three probability distributions, which we represent
-%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
-%as the probability distribution of the random variable $Y$ given the
-%parameter $\Omega$). The three probability distributions required for a
-%state-space model are:
+To be more specific, a fully-specified state-space model requires at least
+three probability distributions, which we represent
+using bracket notation (i.e., $[Y|\Omega]$ is interpreted
+as the probability distribution of the random variable $Y$ given the
+parameter $\Omega$). The three probability distributions required for a
+state-space model are:
 %%%% JEFF, does this extra sentence help? This notation comes from
@@
 %to save space by trimming a bunch of this discussion back down.]}
-%\begin{subequations}
-%  \label{eq:ss2}
-%  \begin{align}
-%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
-%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
-%\label{eq:ss2c}
-%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
-%  \end{align}
-%\end{subequations}
-%where $\bm{\theta}$ are the process variation parameters for the
-%initial state, $\bm{\Theta}$ are the process variation parameters
-%influencing how abundance changes over time, and $\bm{p}$ are the
-%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
-%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
-%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
+\begin{subequations}
+  \label{eq:ss2}
+  \begin{align}
+    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
+[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
+\label{eq:ss2c}
+    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
+  \end{align}
+\end{subequations}
+where $\bm{\theta}$ are the process variation parameters for the
+initial state, $\bm{\Theta}$ are the process variation parameters
+influencing how abundance changes over time, and $\bm{p}$ are the
+observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
+and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
+models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
 %The three lines of general model
@@
 %we allow $N_1$ to be a random variable. But what should its
-%istribution, $[N_1|\bm{\theta}]$ be?
-%And how could $\bm{\theta}$ be estimated given that there is only a single observation
-%available? Often, these problems are simply assumed away. For
+%distribution, $[N_1|\bm{\theta}]$ be?
+And how can $\bm{\theta}$ be estimated given that there is only a single observation
+available? 
+%Often, these problems are simply assumed away. For
 %example, sometimes researchers assume that there is no process variation in
@@
 %determining why a population varies over time.
+%% OR
+%Sometimes researchers assume that there is no process variation in 
+%the first year or that the population is at equilibrium such that the expected
+%value of $N_t$ is constant through time, but these approaches are unsatisfactory.
 
@@
 %In the case where no secondary sampling was conducted, i.e. $J=1$, 
-Let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
+Let $X_{i,t}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
 at site $i$ and primary period $t$. 
@@
 %j=1,\hdots,J$. 
-Typically, we know $X_{it}$ will be lower than the
+Typically, we know $X_{i,t}$ will be lower than the
 actual quantity of interest, abundance,
-denoted $N_{it}$. In cases where detection probability is perfect,
-abundance is observed directly such that $X_{it} \equiv N_{it}$.
+denoted $N_{i,t}$. In cases where detection probability is perfect,
+abundance is observed directly such that $X_{i,t} \equiv N_{i,t}$.
 
@@
 \begin{gather}
-N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
+N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
 \text{or} \nonumber \\
-N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
+N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
 \label{eq:N1}
@@
 year 1.
-The Poisson distribution assumes that the mean of $N_{i1}$ is
+The Poisson distribution assumes that the mean of $N_{i,1}$ is
 equal to its variance, whereas the negative binomial distribution allows the
@@
 however, in each case they modeled $N_t$ as the sum of two random variables:
-$S_{it}$, the number of indivduals surviving from $t-1$ and not
-emigrating; and $G_{it}$ the number of new individuals entering
+$S_{i,t}$, the number of indivduals surviving from $t-1$ and not
+emigrating; and $G_{i,t}$ the number of new individuals entering
 the population. Their most general model was
@@
 \left.\begin{aligned}
-S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
-G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
-N_{it} &= S_{it}+G_{it}
+S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
+G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
+N_{i,t} &= S_{i,t}+G_{i,t}
 \end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
@@
 where $\omega$ is the apparent survival probability and $\gamma$
-is the recruitment rate (which can depend on $N_{it-1}$).
+is the recruitment rate (which can depend on $N_{i,t-1}$).
 \citet{dail_madsen:2011} proposed three
 models for $\gamma$: the constant model,
-$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
-depend on $N_{it-1}$, and which simulates a ``propagule rain'' of new
-individuals; the autoregressive model, $G_{it} \sim
-\mathrm{Pois}(\gamma(N_{it-1}))$, which
+$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
+depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new
+individuals; the autoregressive model, $G_{i,t} \sim
+\mathrm{Pois}(\gamma(N_{i,t-1}))$, which
 simulates geometric or density independent growth; and the
@@
 \begin{equation}
-  X_{it} \sim \mathrm{Bin}(N_{it}, p)
+  X_{i,t} \sim \mathrm{Bin}(N_{i,t}, p)
   \label{eq:p1}
@@
 ecology. The exponential growth model is the simplest: 
-$f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
+$f(N_{i,t-1}) = N_{i,t-1}e^r$ where $r$ is the intrinsic
 rate of increase. %Although demographic and environmental stochasticity
@@
 %simpler way of 
-To allow for demographic stochasticity, we can regard $N_{it}$ as a Poisson 
+To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
 random variable, simplifying Eq.~\ref{eq:Nt} to:
 \begin{equation}
-  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
+  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
 \label{eq:exp}
@@
 \begin{equation}
-  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K)))
+  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
 \label{eq:rick}
@@
 \begin{equation}
-N_{it} \sim
-\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1))))
+N_{i,t} \sim
+\mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
 \label{eq:gomp}
@@
 Because a single Poisson distribution controls the
-distribution of $N_{it}$ in each of these models, the discrete
-convolution used by \citet{dail_madsen:2011} is not required,
+distribution of $N_{i,t}$ in each of these models, the discrete
+convolution used by \citet{dail_madsen:2011} to construct the likelihood is not required,
 speeding up processing time.
@@
 \begin{equation}
-  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
+  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1} + \iota)
   \label{eq:expimm2}
 \end{equation}
-where $\iota$ represents the immigration rate, and is
+where $\iota$ represents the average number of immigrants per year, and is
 constrained to be
@@
 \begin{equation}
-G_{it} \sim \left\{
+G_{i,t} \sim \left\{
 \begin{aligned}
 0 &\; \text{with probability} \; \psi \\
-\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
+\mathrm{Pois}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
 \label{eq:ZIPts}
@@
 environmental stochasticity is often modeled as $r_t \sim
-\text{Norm}(\mu_r, \sigma_r)$.
+\mathrm{Norm}(\mu_r, \sigma_r)$.
 \end{comment}
@@
 binomial draws for survival and Poisson draws for recruitment.  Our population
-growth extensions use a single Poisson draw for abundance: $N_{it} \sim 
-\text{Pois}(f(N_{it-1}))$.  This is also a commonly used (and mathematically equivalent) 
+growth extensions use a single Poisson draw for abundance: $N_{i,t} \sim 
+\mathrm{Pois}(f(N_{i,t-1}))$.  This is also a commonly used (and mathematically equivalent) 
 method of modeling demographic stochasticity 
@@
 \begin{gather}
-\log(r_{it}) = \beta^{r}_{0} + \beta^{r}_{1}T_{it} \nonumber \\
-\log(K_{it}) = \beta^{K}_{0} + \beta^{K}_{1}R_{it}  
+\log(r_{i,t}) = \beta^{r}_{0} + \beta^{r}_{1}T_{i,t} \nonumber \\
+\log(K_{i,t}) = \beta^{K}_{0} + \beta^{K}_{1}R_{i,t}  
 %N_{it} \sim
@@
 \begin{equation}
-N_{it} \sim
-\text{Pois}(N_{it-1}\exp(\nu_{it} + r(1-N_{it-1}/K)) + \iota)
+N_{i,t} \sim
+\mathrm{Pois}(N_{i,t-1}\exp(\nu_{i,t} + r(1-N_{i,t-1}/K)) + \iota)
 \label{eq:nuRand}
@@
 between sites, and might be appropriate where sites are widely dispersed.  In the second, 
-all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{it} = \nu_{t}$),
+all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{i,t} = \nu_{t}$),
 which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
 Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
-for $\nu_{it}$ with distance-dependent (geographic or otherwise) correlation coefficients
+for $\nu_{i,t}$ with distance-dependent (geographic or otherwise) correlation coefficients
 between sites in a year might be appropriate.
@@
 \begin{gather}
-X_{ijt} \sim \mathrm{Bin}(N_{it}, p_j) \nonumber \\
-\mathrm{logit}(p_j) \sim \mathrm{Normal}(\mu_p, \sigma_p)
+X_{i,j,t} \sim \mathrm{Bin}(N_{i,t}, p_j) \nonumber \\
+\mathrm{logit}(p_j) \sim \mathrm{N}(\mu_p, \sigma_p)
 \label{eq:pobs}
 \end{gather}
-where $X_{ijt}$ is the number of bird recorded at site $i$ by
+where $X_{i,j,t}$ is the number of bird recorded at site $i$ by
 observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
@@
 
-We ran a series of frequentist models, 
+We fitted a series of frequentist models, 
 and then a series of Bayesian models. We started by testing
@@
 \end{figure}
-
+%\arglebargle
 \end{document}

### 2013-11-11 Dail_Mad_ext.tex
abstract edits

@@
 
-\title{Improved state-space models for inference about spatial
-and temporal variation in abundance from count data}
+\title{Improved state-space models for inference about 
+spatial and temporal variation in abundance from count data}
 
@@
         \small Washington, DC 20013-7012\\
-   \and Richard B. Chandler \footnote{Contact:
-     \url{rbchan@uga.edu}, +001-706-542-5815} \\
+   \and Richard B. Chandler \footnote{Current address: Warnell School of
+     Forestry and Natural Resources, University of Georgia,
+     \url{rchandler@warnell.uga.edu}, +001-706-542-5815} \\
         \small USGS Patuxent Wildlife Research Center \\
-        \small Laurel, MD 20708}
+        \small Laurel, MD 20708
+}
 \maketitle
@@
 \begin{enumerate}
-\item Models of population dynamics play a central role in theoretical and
-applied ecology, where they are used for purposes such as testing 
+\item Models of population dynamics %play a central role in theoretical and
+%applied ecology, where they 
+are frequently used for purposes such as testing 
 hypotheses about density dependence and predicting species' responses 
-to future environmental change or conservation actions. Failure to
-account for observation error in such models can result in 
-bias. Therefore, ecologists have increasingly relied on state-space
-models to directly model both the observation error and the ecological
+to future environmental change or conservation actions. Fitting models
+of population dynamics to field data is challenging because most data
+sets are characterized by missing data and observation error, which
+can inflate estimates of process variation if ignored.  
+%Failure to account for observation error in such models can result in 
+%bias. Therefore, ecologists have increasingly relied on 
+Recently, state-space models have been developed to deal with these problems
+%these probdirectly model 
+by directly modeling both the observation error and the ecological
 process of interest.
@@
 in many common situations, (2) they do not admit spatial variation in
-population dynamics, and (3) we have no clear interpretation of the
+population dynamics, and (3) they assert an implausible model for
+%we have no clear interpretation of the
 observation error. We demonstrate how each of these problems can be
@@
 of these ideas is possible using either frequentist or Bayesian
-methods. We provide code to fit these models using the \textbf{R} package
-\texttt{unmarked} and \textbf{JAGS} is also provided.
-\item A simulation study indicated that bias was negligible and coverage
+methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code.%. We provide code to fit these models using the \textbf{R} package
+%\texttt{unmarked} and \textbf{JAGS} is also provided.
+\item Results of a simulation study suggest that bias is %was 
+negligible and coverage
 nominal for the proposed model extensions. An analysis of data from
-the North American Breeding Bird Survey highlighted how these methods
-can be readily applied to existing data, but it also suggested 
-that precision would be low when we lack direct information about detection
+the North American Breeding Bird Survey highlights how these methods
+can be readily applied to existing data, but it also suggests 
+that precision will be low when direct information about detection
 probability (such as is collected using distance sampling or
-replicated counts).
+replicated counts) is lacking.
 \end{enumerate}
@@
 detectability, Gompertz-logistic model, immigration, open population point 
-count models, random observer effects, Ricker-logistic model, zero-inflated
+count models, population dynamics, random observer effects, Ricker-logistic model, zero-inflated
 

### 2013-11-11 Dail_Mad_ext.tex
intro edits

@@
 
-Theoretical ecology requires models of population dynamics for testing
-hypotheses regarding spatial and temporal variation in
-abundance. For example, much theoretical work has focused on
-understanding the importance and existence of phenomenon such as
-density-dependent population regulation, population cycling, and spatial synchrony
-\citep{may:1975,turchin:1990,bjornstad_etal:1999}, %,royama:1977,dennis_taper:1994
-and population models are required to evaluate associated hypotheses.
+%Theoretical ecology requires models of population dynamics for testing
+%hypotheses regarding spatial and temporal variation in
+%abundance. For example, much theoretical work has focused on
+Models of population dynamics are vital in both theoretical and
+applied ecological research. For instance, %the have been used to assess
+the importance and existence of phenomenon such as
+density-dependent population regulation, population cycling, and
+spatial synchrony have been studied by comparing data from natural
+populations with theoretical models
+\citep{may:1975,turchin:1990,bjornstad_etal:1999,royama:1977}. %,dennis_taper:1994
+%and population models are required to evaluate associated hypotheses.
 In applied contexts, population models are used for estimating extinction probabilities
@@
 population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
-In order to address these questions, two complicating factors must
-be confronted when fitting population models to data. 
+
+Several challenges are routinely encountered when trying to fit models
+of population dynamics to data from field studies. % is that 
+%In order to address these questions, two complicating factors must
+%be confronted when fitting population models to data. 
 First, deterministic models of population dynamics are virtually always inadequate due to process
@@
 
-State-space models are a widely used approach for
-studying population dynamics while accounting for process variation
+Recently developed state-space models %are a widely used approach for
+have made it possible to %become the primary tool for 
+study population dynamics while accounting for both process variation
 and observation error \citep{devalpine_hastings:2002,
@@
 models are time-series models in which the true state of the
-system (e.g. population size during each year) is observed
-imperfectly. One reason for the widespread adoption of
-state-space models in ecology is that failure to account for process
-variation and observation error can bias estimators of abundance and
+system (e.g. population size during each year) is modeled as a hidden
+process, and the observed data are modeled conditional on this process
+and the observation error. One reason for the widespread adoption of
+state-space models in ecology is that failure to account for 
+%process variation and 
+observation error can bias estimators of abundance and
 population growth parameters. For instance, the strength of
@@
 
-A simple state-space models can be described as follows.
+A simple state-space model can be described as follows.
 Let $N_t$ be the abundance of a species during year $t$, for
@@
 %, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
-The full model can now be written as:
+A generic model can be written: %The full model can now be written as:
 \begin{subequations}
@@
 X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
-t=1,\hdots,T \label{eq:ss1c}
+t=1,\hdots,T. \label{eq:ss1c}
   \end{align}
 \end{subequations}
-where $\eta_t$ is the random effect allowing for process
-variation unaccounted for by the deterministic model. In classical
+%where $\eta_t$ is the random effect allowing for process
+%variation unaccounted for by the deterministic model. 
+In classical
 state-space models, the two sets of random effects
@@
 are typically integer-valued, as in the case of count data, raising
-concerns about the use of Gaussian distribution for the random
-effects. Second, the model for observation error has little biological basis. 
+concerns about the Gaussian assumption for the random
+effects. Second, the model for observation error is implausible. %has little biological basis. 
 Third, the model does not account for spatial variation in abundance. Lastly, some of the
@@
 
-Use of the Gaussian distribution for modeling random process variation
-and observation error is motivated by convenience rather than
-biology. Specifically, the Gaussian assumptions
-allow for parameter estimation using the Kalman filter
-\citep{dennis_etal:2006}, which is much more computationally efficient
-than estimation methods when random effects are not Gaussian distributed
+%Use of the Gaussian distribution for modeling random process variation
+%and observation error is often motivated by convenience rather than
+%biology. Specifically, t
+The assumption that random effects are Gaussian distributed makes it
+easier to estimate parameters because methods such as the Kalman
+filter can be applied %The Gaussian assumption allows for efficient methods of parameter
+%estimation, such as the Kalman filter
+\citep{dennis_etal:2006}, which are not available otherwise. % which is much more computationally efficient
+%than estimation methods when random effects are not Gaussian distributed
 \citep{devalpine_hastings:2002}. %The problem with this is that it
-This approach allows for negative and non-integer values of the state-variable, which is
-inconsistent with the observed data, and may result in implausible predictions.
-Transforming the count data and the underlying abundance variables do
-little to resolve these issues \citep{ohara_kotze:2010}.
-
-Another problem with standard state-space models is that the
-observation error has no clear interpretation. The use of a mean zero
-Gaussian distribution implies that $X_t$ will be
-higher than $N_t$ as often as it is lower than $N_t$. 
-Identifying a biological mechanism that would cause such symmetric errors
-is difficult.
+However, these models effectively assume that abundance can be take on any
+real value, including negative values, 
+%This approach allows for negative and non-integer values of the state-variable, which is
+%inconsistent with the observed data, and may 
+and these models can result in implausible predictions.
+Transforming the count data and the underlying abundance variables 
+do little to resolve these issues \citep{ohara_kotze:2010}.
+
+Another problem with standard state-space models is that the model for
+the mean zero Gaussian distribution for the observation error
+implies that false positive detections are
+equally likely as false negatives. To our knowledge, all studies to
+date have shown that false negative detections are much more common
+that false positives (CITE DAVE MILLER PAPERS). 
+%will be
+%higher than $N_t$ as often as it is lower than $N_t$. 
+%Identifying a biological mechanism that would cause such symmetric errors
+%is difficult.
 A more likely form of observation error, and one that has been recognized for well
@@
 A more serious problem %than the ones associated with the Gaussian assumptions 
-is that the parameters of the simple state-space models
+is that the parameters of simple state-space models
 such as Eq~\ref{eq:ss1} are not identifiable in many
-circumstances \citep{polansky_etal:2009}. Recall that the
-Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
-although $N_1$ is a random variable. Thus, uncertainty exists
-that should be accounted for in the model.  
-%are needed to fully specify the model.
-To be more specific, a fully-specified state-space model requires at least
-three probability distributions, which we represent
-using bracket notation (i.e., $[Y|\Omega]$ is interpreted
-as the probability distribution of the random variable $Y$ given the
-parameter $\Omega$). The three probability distributions required for a
-state-space model are:
+circumstances \citep{polansky_etal:2009}. Specifically, abundance in
+the first year cannot be estimated because there is only one data
+point (the count in the first year) that carries any information about
+it. 
+%Recall that the
+%Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
+%although $N_1$ is a random variable. Thus, uncertainty exists
+%that should be accounted for in the model.  
+%%are needed to fully specify the model.
+%To be more specific, a fully-specified state-space model requires at least
+%three probability distributions, which we represent
+%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
+%as the probability distribution of the random variable $Y$ given the
+%parameter $\Omega$). The three probability distributions required for a
+%state-space model are:
 %%%% JEFF, does this extra sentence help? This notation comes from
@@
 %distribution, $[N_1|\bm{\theta}]$ be?
-And how can $\bm{\theta}$ be estimated given that there is only a single observation
-available? 
+%And how can $\bm{\theta}$ be estimated given that there is only a single observation
+%vailable? 
 %Often, these problems are simply assumed away. For
@@
 overcome the limitations described above. \citet{devalpine_hastings:2002} and
-\citet{kery_etal:2009} described methods for fitting models with non-Gaussian
+\citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
 distributions for the process and observation errors. Observation models with
 more intuitive interpretations, such as those that explicitly model
-detection probability, have been proposed by \citet{kery_etal:2009}. \citet{lele_etal:1998} and
+detection probability, have been proposed by 
+\citet{kery_etal:2009}. \citet{lele_etal:1998} and 
 \citet{kery_etal:2009} developed models allowing for inference about
@@
 the work by \citet{kery_etal:2009} addressed several of these limitations
-simultaneously. However, their model did not include serial
-dependence, which is a hallmark of population models. This
-limits the utility of their model for making inferences about explicit
-population processes.
+simultaneously; however, their model did not include serial
+dependence, which is a hallmark of population models. %This
+%limits the utility of their model for making inferences about explicit
+%population processes.
 
 In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
-that simultaneous resolves each of the aforementioned problems with
-traditional state-space models, and is designed for simple count data.
-In the following section, we describe the DM model in its original
-form and explain how it resolves each of the deficiencies with
-standard state-space models. In Section~\ref{sec:ext}, we extend the
+that resolves many of the problems with classical state-space models.
+%simultaneous resolves each of the aforementioned problems with
+%traditional state-space models, and is designed for simple count data.
+%In the following section, we describe the DM model in its original
+%form and explain how it resolves each of the deficiencies with
+%standard state-space models. 
+In Section~\ref{sec:ext}, we extend the
 model to accommodate classical models of population growth and
 to handle several features common to ecological
-time-series. Specifically, we describe methods for accommodating
-excess zeros and nuisance variables such as random observer effects.
+time-series, such as excess zeros and unexplained random variation. 
+%Specifically, we describe methods for accommodating
+%excess zeros and nuisance variables such as random observer effects.
 Both frequentist and Bayesian methods of inference are discussed, and

### 2013-11-12 Dail_Mad_ext.tex
more edits to intro

@@
 \usepackage{verbatim}
-\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
+%\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
+\usepackage[pdftex,hidelinks,pdfstartview={XYZ 1 1 1.5},bookmarks=true]{hyperref}
 \usepackage{rotating}
@@
 \item Conventional state-space models, however, have
-three important limitations: (1) the parameters are not identifiable
-in many common situations, (2) they do not admit spatial variation in
-population dynamics, and (3) they assert an implausible model for
+several important limitations: (1) they assume that random effects are
+Gaussian distributed, which implies that abundance can be negative and
+that false negative observation errors are equally likely as false positive
+errors, %results in unrealistic predictions
+%(e.g. negative values of abundance) and an implausible mode
+%and observation error are Gaussian random effects
+%the parameters are not identifiable
+%in many common situations, 
+(2) they do not admit spatial variation in
+population dynamics, and (3) some of the parameters of the model are not
+estimable. % they assert an implausible model for
 %we have no clear interpretation of the
-observation error. We demonstrate how each of these problems can be
+%observation error. 
+We demonstrate how each of these problems can be
 resolved using a class of hierarchical models proposed by
-\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
-time-series data that attributes observation error to imperfect detection.
+\citet[Biometrics]{dail_madsen:2011} 
+%for spatially-replicated time series data 
+that attributes observation error to imperfect detection.
 \item We expand this class of models to accommodate classical growth
 models (e.g. exponential and Ricker-logistic), zero-inflation,
-and random effects such as observer-specific
-detection probabilities. We also present methods for forecasting
+and random effects. 
+% such as observer-specific detection probabilities. 
+We also present methods for forecasting
 population size under future environmental conditions. Implementation
@@
   buckland_etal:2004, dennis_etal:2006}. Classical state-space
-models are time-series models in which the true state of the
+models are time series models in which the true state of the
 system (e.g. population size during each year) is modeled as a hidden
@@
 $\eta_t \sim \mathrm{N}(0, \sigma)$ and
-$\zeta_t \sim \mathrm{N}(0, \tau)$. Standard practice is to
-ignore process variation associated with $N_1$, as
-indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
-assumed to be at equilibrium such that the $N_1$ can be regarded as an
-outcome of the equilibrium distribution.
+$\zeta_t \sim \mathrm{N}(0, \tau)$. %Standard practice is to
+%ignore 
+The process variation associated with $N_1$ is often ignored, or
+%, as indicated by Eq~\ref{eq:ss1a}. Alternatively, 
+the population is assumed to be at equilibrium such that the $N_1$ can
+be regarded as an outcome of the equilibrium distribution.
 
 Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
-are among the most widespread approaches for modeling population dynamics
-using time-series data, several problems exist. First,
-time-series data and the underlying abundance parameters of interest
-are typically integer-valued, as in the case of count data, raising
-concerns about the Gaussian assumption for the random
-effects. Second, the model for observation error is implausible. %has little biological basis. 
-Third, the model does not account for spatial variation in abundance. Lastly, some of the
-parameters of the model are not estimable. We briefly discuss each of
-these points before describing a general approach to resolving these
-shortcomings.
-
+overcome many of the limitations associated with the application of
+classical models of population dynamics, 
+%are among the most widespread approaches for modeling population dynamics
+%using time series data, 
+several problems exist. %, which we explain in the following paragraphs. 
+% First,
+% time series data and the underlying abundance parameters of interest
+% are typically integer-valued, as in the case of count data, raising
+% concerns about the Gaussian assumption for the random
+% effects. Second, the model for observation error is implausible. %has little biological basis. 
+% Third, the model does not account for spatial variation in abundance. Lastly, some of the
+% parameters of the model are not estimable. We briefly discuss each of
+% these points before describing a general approach to resolving these
+% shortcomings.
 %Use of the Gaussian distribution for modeling random process variation
@@
 %estimation, such as the Kalman filter
-\citep{dennis_etal:2006}, which are not available otherwise. % which is much more computationally efficient
+\citep{dennis_etal:2006}; however, % which are not available otherwise % which is much more computationally efficient
 %than estimation methods when random effects are not Gaussian distributed
-\citep{devalpine_hastings:2002}. %The problem with this is that it
-However, these models effectively assume that abundance can be take on any
-real value, including negative values, 
+%\citep{devalpine_hastings:2002}. %The problem with this is that it
+%However, 
+these models effectively assume that abundance can be take on any
+real value, and hence they can predict negative values of abundance. %including negative values, 
 %This approach allows for negative and non-integer values of the state-variable, which is
 %inconsistent with the observed data, and may 
-and these models can result in implausible predictions.
+%and these models can result in implausible predictions.
 Transforming the count data and the underlying abundance variables 
@@
 
-Another problem with standard state-space models is that the model for
-the mean zero Gaussian distribution for the observation error
+With respect to the observation error, the Gaussian assumption
+%Another problem with standard state-space models is that the model for
+%the mean zero Gaussian distribution for the observation error
 implies that false positive detections are
-equally likely as false negatives. To our knowledge, all studies to
-date have shown that false negative detections are much more common
-that false positives (CITE DAVE MILLER PAPERS). 
+equally likely as false negatives, a pattern inconsistent with most
+findings \citep{miller_etal:2011}. 
+%. To our knowledge, all studies to
+%date have shown that false negative detections are much more common
+%that false positives \citep{miller_etal:2011}.
 %will be
@@
 this form of observation error, rarely have these methods been
-integrated into state-space models \citep[but
+integrated into state-space models of population dynamics \citep[but
 see][]{buckland_etal:2004}.
 
-Common formulations of state space models do not admit
-spatial variation. This reduces the scope of the inferences that can
-be drawn from the models. Moreover, space is an important factor in the
-regulation of natural populations \citep{murdoch:1994,
-pulliam:1988, gill_etal:2001}.
-
-A more serious problem %than the ones associated with the Gaussian assumptions 
-is that the parameters of simple state-space models
-such as Eq~\ref{eq:ss1} are not identifiable in many
-circumstances \citep{polansky_etal:2009}. Specifically, abundance in
+Other limitations of state-space models, as commonly applied in ecology,
+include the failure to admit spatial variation in abundance and
+the fact that some parameters are not identifiable
+\citep{polansky_etal:2009}. 
+% without making 
+%Another limitation of many state-space models is that they do 
+%Common formulations of state space models do 
+%not admit spatial variation in abundance or demographic
+%parameters. 
+Ignorning spatial variation simply reduces the scope of the
+inferences, 
+% that can 
+%be drawn from the models, it disregards the fact that space is an
+%important factor in the regulation of natural populations \citep{murdoch:1994,
+%pulliam:1988, gill_etal:2001}.
+but identifiability problems require %lack of identifiability because
+model constraints or the assertion of additional assumptions. 
+%A more serious problem %than the ones associated with the Gaussian assumptions 
+%is that the parameters of simple state-space models
+%such as Eq~\ref{eq:ss1} are not identifiable in many
+%circumstances \citep{polansky_etal:2009}. 
+Specifically, abundance in
 the first year cannot be estimated because there is only one data
 point (the count in the first year) that carries any information about
-it. 
+it. To remedy this, pratitioners either assume that there is no
+observation error in the first year, or they assume the population is
+at equilibrium, which may defeat the purpose of many studies of
+population dynamics. 
 %Recall that the
@@
 In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
-that resolves many of the problems with classical state-space models.
+that resolves each of the problems outlined above. 
 %simultaneous resolves each of the aforementioned problems with
@@
 %standard state-space models. 
-In Section~\ref{sec:ext}, we extend the
+%In Section~\ref{sec:ext}, 
+Our aim is to extend the
 model to accommodate classical models of population growth and
 to handle several features common to ecological
-time-series, such as excess zeros and unexplained random variation. 
+time series, such as excess zeros and unexplained random variation. 
 %Specifically, we describe methods for accommodating
@@
 Both frequentist and Bayesian methods of inference are discussed, and
-code for fitting models is presented in the appendices.
+we evaluate the performance of the model using a simulation study by
+analyzing data from the North American Breeding Bird Survey (BBS), one of
+the most spatially and temporally extensive sets of count data on
+vertebrate populations \citep{robbins_etal:1986}.
+%code for fitting models is presented in the appendices.
 %The Bayesian method
@@
 %collected as part of many existing monitoring programs.
-In Section~\ref{sec:app}, we evaluate the performance of the model
-extensions using a simulation study and by analyzing data from the
-North American Breeding Bird Survey (BBS), one of
-the most spatially and temporally extensive sets of count data on
-vertebrate populations \citep{robbins_etal:1986}. The overarching aim
-of the paper is to provide ecologists with flexible and accessible means of
-addressing important questions related to the variation of abundance in
-space and time. 
+%In Section~\ref{sec:app}, we evaluate the performance of the model
+%extensions using a simulation study and by analyzing data from the
+%North American Breeding Bird Survey (BBS), one of
+%the most spatially and temporally extensive sets of count data on
+%vertebrate populations \citep{robbins_etal:1986}. %The overarching aim
+%of the paper is to provide ecologists with flexible and accessible means of
+%addressing important questions related to the variation of abundance in
+%space and time. 
 %believe these extensions will increase the utility of the model for
@@
   assumptions and does not acknowledge many of the sources of
-variation inherent to existing ecological time-series data.  We  propose variations on these models that are likely to be more
+variation inherent to existing ecological time series data.  We  propose variations on these models that are likely to be more
 realistic and useful in many cases. We demonstrate the
@@
 %the framework. Second, we have demonstrated methods of
-%accounting for zero-inflation in the time-series. Third, we
+%accounting for zero-inflation in the time series. Third, we
 %have illustrated how additional random effects such as

### 2013-11-12 Dail_Mad_ext.tex
model development sectino

@@
 %application to existing datasets such the Breeding Bird Survey.
-\begin{comment}
-  The DM model allows for inference about spatial and temporal
-dynamics in abundance while accounting for observation error
-using only spatially and temporally replicated count data. Their model
-is an extension of the closed-population $N$-mixture model
-\citep{royle:2004biom}, which was designed specifically to
-address the problem of modeling spatial variation in abundance in the
-  presence of observation error, specifically, when detection
-  probability is less than one. The DM extension relaxes the
-assumption of population closure and includes explicit
-parameters describing population change as a first-order Markovian
-  process. [more details]
-
-[applications / potential etc... ] The DM model is relevant to a
-vast number of problems confronting basic and applied ecologicalresearch. Answering questions about climate change etc...
-require
-spatially and temporally extensive datasets such as result from
-the
-  North American Breeding Bird Survey (BBS)
-\citep{robbins_etal:1986}. However, most of the datasets of theappropriate scale are collected by volunteers and using
-protocols
-  that are not amenable to traditional approaches of modeling
-  abundance and detection probability based on standard
-capture-recapture methods. Instead they have, at best, simple
-count
-data for which, when it can be interpreted as counts of unique  individuals, the DM model can be applied.. Yadi ya
-
-Although the DM model was designed explicitly for these
-purposes,
-the original formulation of the model makes strict
-distributional
-  assumptions and does not acknowledge many of the sources of
-variation inherent to existing ecological time series data.  We  propose variations on these models that are likely to be more
-realistic and useful in many cases. We demonstrate the
-usefulness
-  of these new models with simulated and real BBS data.
-
-In this paper, we demonstrate how the DM model resolves each ofthese four limitations of conventional state-space models, and
-we
-extend the DM model in several important ways. First, we
-demonstrate
-  how classical population models can be embedded within the DM
-model. Then we present methods for handling excess zeros as well
-as
-nuisance parameters such as random observer effects on detection probability. A simulation study
-\end{comment}
 
@@
 
-The DM model is an extension of the $N$-mixture model
-\citep{royle:2004biom}, which allows for inference about spatial
+The DM model is an extension of the 
+\citet{royle:2004biom} $N$-mixture model, which allows for inference about spatial
 variation in abundance when individuals cannot be detected with
@@
 
-\subsection{The Data}
+%\subsection{The Data}
 
@@
 %j=1,\hdots,J$. 
-Typically, we know $X_{i,t}$ will be lower than the
-actual quantity of interest, abundance,
-denoted $N_{i,t}$. In cases where detection probability is perfect,
+In general, $X_{i,t}$ will be lower than abundance, 
+%actual quantity of interest, abundance,
+%denoted 
+$N_{i,t}$, but in cases where detection is perfect,
 abundance is observed directly such that $X_{i,t} \equiv N_{i,t}$.
@@
 
-Like traditional state-space models, the DM model includes the
-three conditionally related processes shown in Eq.~\ref{eq:ss1},
-which correspond to (1) initial abundance, i.e. the
+Like traditional state-space models, the DM model includes 
+three conditionally related processes corresponding to: 
+% shown in Eq.~\ref{eq:ss1}, which correspond to 
+(1) initial abundance, i.e. the
 abundance at site $i$ during the first primary period,
@@
 
-Recall that conventional state-space models assume that the
-distribution for the initial time period,
-$[N_1 | \bm{\theta}]$, is either the equilibrium
+%Recall that 
+Conventional state-space models assume that the
+distribution for the initial time period %,
+%$[N_1 | \bm{\theta}]$, 
+is either the equilibrium
 distribution, or has zero variance.
@@
 The DM model assumes that abundance in time $t$ is a function of
-abundance in time $t-1$, i.e. abundance at each site evolves as a
-first order Markovian process. %, although higher order processes are also possible.
+abundance at time $t-1$, i.e. abundance at each site evolves as a
+first order Markov process. %, although higher order processes are also possible.
 \citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
@@
 simulates geometric or density independent growth; and the
-``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
-expected abundance constant over time. Covariates of
+``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which assumes the
+population is at equilibrium. Covariates of
 $\omega$ and $\gamma$ can be easily accommodated, for example
@@
 %(3) have parameters that are estimable.
-The one
-remaining task is to replace the observation model
-$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
-more easily interpreted. The observation
-model assumes that individuals are missed due to imperfect
-detection. The simplest model for imperfect detection is
+%The one
+%remaining task is to replace the observation model
+%$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
+%more easily interpreted. 
+The observation model adopted by \citet{dail_madsen:2011} is the same
+binomial model proposed by \citet{royle:2004biom}: % assumes that individuals are missed due to imperfect
+%detection. The simplest model for imperfect detection is
 \begin{equation}
@@
 \end{equation}
+where $p$ is the probability of detecting each individual. Variation
+in detection probability can be modeled as a function of site-specific
+or occasion-specific covariates using, for example, a logit-linear model.
 %or equivalently,
@@
 %\end{equation}
-Under this model, $\bm{p}$ is interpreted as the detection
-probability parameter.
+%Under this model, $\bm{p}$ is interpreted as the detection
+%probability parameter.
 

### 2013-11-12 Dail_Mad_ext.tex
up to sec2.4

@@
 
-\subsection{Bayesian inference}
-
-\citet{dail_madsen:2011} demonstrated how to fit these models using maximum 
-likelihood estimation. The likelihood can be maximized numerically using the
-\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
-are given in the Appendix.
-
-Bayesian inference, however, has several appealing features. First, it
-allows direct probability statements to be made about a hypothesis
-given data \citep{link_barker:2010}.    
-Second, Bayesian methods offer straight-forward approaches for
-combine data from multiple sources or use existing estimates of parameters as prior
-distributions. In practice, simulation methods such as Markov
-chain Monte Carlo (MCMC) are used to simulate the posterior
-distributions. %Although MCMC can be slow, our experience is that it
-%may actually converge faster than the amount of time required to
-%maximize the likelihood shown in Eq.\ref{eq:like}, especially
-%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
-%Furthermore, 
-MCMC may be the only possible option for estimating parameters
-using some of the extensions described below, such as allowing for
-additional random effects.
-
-Writing custom MCMC algorthims is often tedious and foreign to
-ecologists, but software packages such as %\textbf{WinBUGS} and
-\textbf{JAGS} \citep{plummer:2003} overcome the technical 
-difficulties by allowing users to specify the model using a simple symbolic 
-descriptions. Examples are given in the Appendix.
-
-%For example here is \textbf{BUGS} code for the ``geometric'' model described
-%in Eq.~\ref{eq:hm}:
-%\begin{verbatim}
-%model {
-%lambda ~ dunif(0, 5)
-%omega ~ dunif(0,1)
-%gamma ~ dunif(0, 10)
-%p ~ dunif(0,1)
-%for(i in 1:nSites) {
-%  N[i,1] ~ dpois(lambda)
-%  y[i,1] ~ dbin(p, N[i,1])
-%  for(t in 2:nYears) {
-%    S[i,t-1] ~ dbin(omega, N[i,t-1])
-%    G[i,t-1] ~ dpois(gamma*N[i,t-1])
-%    N[i,t] <- S[i,t-1] + G[i,t-1]
-%    y[i,t] ~ dbin(p, N[i,t])
-%    }
-%  }
-%}
-%\end{verbatim}
-
 \section{Model Extensions}
@@
 where $K$ is the stable equilibrium of the population and $r$ is
-the
-instantaneous population growth rate at low population
-densities, and
-both parameters are constrained to be positive. This is a
-stochastic version
-of the Ricker-logistic population growth model
-\citep{ricker:1954}. We also
-implemented a modified Gompertz-logistic density-dependent
-model \citep{hart_gotelli:2011}:
+the instantaneous population growth rate at low population
+densities, and both parameters are constrained to be positive. This is
+a stochastic version of the Ricker-logistic population growth model
+\citep{ricker:1954}. Another option is a modified Gompertz-logistic
+density-dependent model \citep{hart_gotelli:2011}:
 \begin{equation}
-N_{i,t} \sim
-\mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
+N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
 \label{eq:gomp}
@@
 Here the interpretations of $r$ and $K$ are similar to in the
-Ricker-logistic model.
-Because a single Poisson distribution controls the
-distribution of $N_{i,t}$ in each of these models, the discrete
-convolution used by \citet{dail_madsen:2011} to construct the likelihood is not required,
-speeding up processing time.
+Ricker-logistic model. Because a single Poisson distribution controls
+the distribution of $N_{i,t}$ in each of these models, the discrete 
+convolution used by \citet{dail_madsen:2011} to construct the
+likelihood is not required, speeding up processing time.
 
@@
 
-The geometric-recruitment, exponential, Ricker-logistic, and Gompertz-logistic
-versions of the DM models all share a common feature (or bug):
-once the population at a site reaches 0, it must remain at 0
-because all contributions to population growth are assumed to be
-local.  We therefore generalized these models to
-include both internal and external (immigration) contributions
-to population growth.  The exponential plus immigration
+One limitation of the geometric-recruitment, exponential,
+Ricker-logistic, and Gompertz-logistic versions of the DM models all
+%share a common feature (or bug): once the 
+is that they include no mechanism for a local population to recover
+after extinction. %at a site reaches 0, it must remain at 0
+%because all contributions to population growth are assumed to be
+%local.  
+However, these models can be generalized 
+%We therefore generalized these models 
+to include both internal and external (immigration) contributions
+to population growth. For example, an exponential plus immigration
 model is:
@@
 
-\citet{dail_madsen:2011} suggested two distributions for modeling
-initial abundance: Poisson and negative binomial. We have
-extended their model to include another distribution, the zero-inflated
-Poisson. This could be useful when, for example, one is modeling
-the abundance of several species with the same set of point
-count surveys for all species, but some survey sites are
-outside the range of some of the species. The distribution
-of initial abundances can be represented as:
+%\citet{dail_madsen:2011} suggested two distributions for modeling
+%initial abundance: Poisson and negative binomial. We have
+%extended their model to include another distribution, the zero-inflated
+%Poisson. This could be useful when, for example, one is modeling
+%the abundance of several species with the same set of point
+%count surveys for all species, but some survey sites are
+%outside the range of some of the species. 
+The negative binomial distribution used in $N$-mixture models can have
+a very long right tail, and its mean to variance ratio may not be
+ideal in all cases. An alternative is to consider zero-inflated
+distributions, such as the zero-inflated Poisson:
+%The distribution
+%of initial abundances can be represented as:
 \begin{equation}
@@
 \begin{aligned}
-0 &\; \text{with probability} \; \psi \\
-\Lambda &\; \text{with probability} \; (1-\psi)
+\mbox{Pois}(0) &\; \text{with probability} \; \psi \\
+\mbox{Pois}(\Lambda) &\; \text{with probability} \; (1-\psi)
 \end{aligned} \right.
@@
 
-These models allow three sources of zero counts by observers: a
-species was at a site but not detected; the site was within the
-species' range but no individuals occured at that site in that
-year; and the site was outside the species' range. Furthermore,
+%These models allow three sources of zero counts by observers: a
+%species was at a site but not detected; the site was within the
+%species' range but no individuals occured at that site in that
+%year; and the site was outside the species' range. Furthermore,
+Under this formulation, 
 detection, abundance, and zero-inflation can be modeled separately as
-functions of different (or the same) covariates. For example, detection of a
+functions of covariates. For example, detection of a
 species might depend on wind speed,
@@
 \begin{aligned}
-0 &\; \text{with probability} \; \psi \\
+\mathrm{Pois}(0) &\; \text{with probability} \; \psi \\
 \mathrm{Pois}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
@@
 ``missing'' in future years. An example of this is provided in the Appendix.
+
+\subsection{Bayesian inference}
+
+\citet{dail_madsen:2011} demonstrated how to fit these models using maximum 
+likelihood estimation. The likelihood can be maximized numerically using the
+\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
+are given in the Appendix.
+
+Bayesian inference, however, has several appealing features. First, it
+allows direct probability statements to be made about a hypothesis
+given data \citep{link_barker:2010}.    
+Second, Bayesian methods offer straight-forward approaches for
+combine data from multiple sources or use existing estimates of parameters as prior
+distributions. In practice, simulation methods such as Markov
+chain Monte Carlo (MCMC) are used to simulate the posterior
+distributions. %Although MCMC can be slow, our experience is that it
+%may actually converge faster than the amount of time required to
+%maximize the likelihood shown in Eq.\ref{eq:like}, especially
+%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
+%Furthermore, 
+MCMC may be the only possible option for estimating parameters
+using some of the extensions described below, such as allowing for
+additional random effects.
+
+Writing custom MCMC algorthims is often tedious and foreign to
+ecologists, but software packages such as %\textbf{WinBUGS} and
+\textbf{JAGS} \citep{plummer:2003} overcome the technical 
+difficulties by allowing users to specify the model using a simple symbolic 
+descriptions. Examples are given in the Appendix.
+
+%For example here is \textbf{BUGS} code for the ``geometric'' model described
+%in Eq.~\ref{eq:hm}:
+%\begin{verbatim}
+%model {
+%lambda ~ dunif(0, 5)
+%omega ~ dunif(0,1)
+%gamma ~ dunif(0, 10)
+%p ~ dunif(0,1)
+%for(i in 1:nSites) {
+%  N[i,1] ~ dpois(lambda)
+%  y[i,1] ~ dbin(p, N[i,1])
+%  for(t in 2:nYears) {
+%    S[i,t-1] ~ dbin(omega, N[i,t-1])
+%    G[i,t-1] ~ dpois(gamma*N[i,t-1])
+%    N[i,t] <- S[i,t-1] + G[i,t-1]
+%    y[i,t] ~ dbin(p, N[i,t])
+%    }
+%  }
+%}
+%\end{verbatim}
 

### 2013-11-12 Dail_Mad_ext.tex
to results

@@
 \end{equation}
-\begin{comment}
-When both initial abundance and dynamics are modeled as
-zero-inflated,
-$\psi$ can be modeled as time-varying or not (in either case,
-$\psi$ can also be
-modeled as varying over space). In the latter case, it makes
-sense
-for the model to "remember" which sites are inside the range of
-the
-species between the initial abundance stage and each time step
-of the
-dynamics stage (see code sample?).  Adding zero-inflation to the
-dynamics of the geometric-recruitment, exponential, Ricker-logistic, or Gompertz-logistic
-models
-(without immigration) should have no effect when initial
-abundance is
-also zero-inflated unless $\psi$ is time-varying (because these
-dynamic
-models cannot recover from zero abundance).
-\end{comment}
-We have implemented
-zero-inflated dynamics in the Bayesian framework using program
-\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
+%We have implemented
+%zero-inflated dynamics in the Bayesian framework using program
+%\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
 %(version 3.2.0; Plummer 2003)
@@
 \begin{comment}
-Even though the DM model is conceptually simple, even in
-its most stripped down form it includes many random effects (the $N$'s)
+Even though the DM model is conceptually simple, %even in
+%its most stripped down form 
+it includes many random effects (the $N$'s)
 making it challenging to compute the likelihood or implement MCMC
@@
 \end{comment}
-The original DM models include demographic stochasticity --- variation in population growth 
-rate due to the randomness of birth and death processes --- through random 
-binomial draws for survival and Poisson draws for recruitment.  Our population
-growth extensions use a single Poisson draw for abundance: $N_{i,t} \sim 
-\mathrm{Pois}(f(N_{i,t-1}))$.  This is also a commonly used (and mathematically equivalent) 
-method of modeling demographic stochasticity 
-\citep{bonsall_hastings:2004,melbourne_hastings:2008}.
+The original DM models allows for demographic stochasticity --- variation in population growth 
+rate due to the randomness of birth and death processes --- through %random 
+via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
+growth extensions use a single probability distribution for abundance: $N_{i,t} \sim 
+\mathrm{Pois}(f(N_{i,t-1}))$.  
+% JEFF: I'm not sure what you're driving at here. Do these papers
+% assume a Poisson distribution too?
+%This is also a commonly used (and mathematically equivalent) 
+%method of modeling demographic stochasticity 
+%\citep{bonsall_hastings:2004,melbourne_hastings:2008}.
 
@@
 
+% JEFF: It seems to me that this approach could be used to account for
+% either demographic or enviro stochasticity. How can you know what
+% the variation in attributable to?
 Another way to model environmental stochasticity is as lognormal variation in observed local population 
@@
 
-\subsection{Random effects of observers}
+\subsection{Random observer effects}
 
@@
 %
-To include random observer effects in DM models and these
-extensions, Eq.~\ref{eq:p1} can be modified to:
+To include random observer effects in DM models, 
+% and these extensions, 
+Eq.~\ref{eq:p1} can be modified to:
 \begin{gather}
@@
 \end{gather}
-where $X_{i,j,t}$ is the number of bird recorded at site $i$ by
+where $X_{i,j,t}$ is the number of individuals counted at site $i$ by
 observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
@@
 the standard deviation of the random observer effects (also on the logit scale). 
-We have implemented random observer effects in the Bayesian
-framework using program \textbf{JAGS} \citep[version 3.2.0, see Appendix]{plummer:2003}.
+%We have implemented random observer effects in the Bayesian
+%framework using program \textbf{JAGS} \citep[version 3.2.0, see Appendix]{plummer:2003}.
 % (version 3.2.0; Plummer 2003)
@@
 
-\subsection{Forecasting Future Population Size}
-
-State-space models allow for inference about future population size,
-a critical component of population viability analysis (PVA) as well
-as conservation planning. Let $N^+_{t}$ denote population size at some
-future point in time and $[N^+_t]$ denote the probability
-distribution for this random variable. If we can specify this
-distribution, we can make statements about the
-probability that $N^+_t = k$ where $k$ is any possible
-population size including zero.
+%\subsection{Forecasting Future Population Size}
+
+%State-space models allow for inference about future population size,
+%a critical component of population viability analysis (PVA) as well
+%as conservation planning. 
+%Let $N^+_{t}$ denote population size at some
+%future point in time and $[N^+_t]$ denote the probability
+%distribution for this random variable. If we can specify this
+%distribution, we can make statements about the
+%probability that $N^+_t = k$ where $k$ is any possible
+%population size including zero.
 
@@
 %empirical Bayes methods, 
-A fully Bayesian approach to PVA is straightforward to
-implement with MCMC. Because population size at any future point %Furthermore, b
-in time is regarded as a standard, unknown parameter, computing the
-posterior distribution of $N^+_t$ requires no additional steps
-in the MCMC algorithm, other than formally stating that the count data are
-``missing'' in future years. An example of this is provided in the Appendix.
-
-\subsection{Bayesian inference}
-
-\citet{dail_madsen:2011} demonstrated how to fit these models using maximum 
-likelihood estimation. The likelihood can be maximized numerically using the
+%A fully Bayesian approach to PVA is straightforward to
+%implement with MCMC. Because population size at any future point %Furthermore, b
+%in time is regarded as a standard, unknown parameter, computing the
+%posterior distribution of $N^+_t$ requires no additional steps
+%in the MCMC algorithm, other than formally stating that the count data are
+%``missing'' in future years. An example of this is provided in the Appendix.
+
+\subsection{Statistical inference}
+
+\citet{dail_madsen:2011} used likelihood based methods for estimating
+the parameters of their model. The same likelihood functions can be
+used to accomodate the alternative population growth models that we
+have proposed. Methods for implementing these models using maximum 
+likelihood estimation are included in the %. The likelihood can be maximized numerically using the
 \textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
@@
 
-Bayesian inference, however, has several appealing features. First, it
-allows direct probability statements to be made about a hypothesis
-given data \citep{link_barker:2010}.    
-Second, Bayesian methods offer straight-forward approaches for
-combine data from multiple sources or use existing estimates of parameters as prior
-distributions. In practice, simulation methods such as Markov
-chain Monte Carlo (MCMC) are used to simulate the posterior
-distributions. %Although MCMC can be slow, our experience is that it
+Bayesian inference is an alternative to classical inference with
+several appealing features. First, it allows direct probability
+statements to be made about a hypothesis given data
+\citep{link_barker:2010}.  Second, Bayesian methods offer
+straight-forward approaches for combining data from multiple sources
+or use existing estimates of parameters as prior distributions. Third,
+one common usage of state-space models is for predicting future
+population size, and this task easily accomplished using Bayesian
+methods.
+
+In
+practice, simulation methods such as Markov chain Monte Carlo (MCMC)
+are used to simulate the posterior distributions. While time
+consuming, MCMC may be the only viable approach for estimating
+parameters, especially for hierarchical models with many random
+effects.  
+% Although MCMC can be slow, our experience is that it
 %may actually converge faster than the amount of time required to
@@
 %Furthermore, 
-MCMC may be the only possible option for estimating parameters
-using some of the extensions described below, such as allowing for
-additional random effects.
-
-Writing custom MCMC algorthims is often tedious and foreign to
-ecologists, but software packages such as %\textbf{WinBUGS} and
-\textbf{JAGS} \citep{plummer:2003} overcome the technical 
-difficulties by allowing users to specify the model using a simple symbolic 
-descriptions. Examples are given in the Appendix.
+%MCMC may be the only possible option for estimating parameters
+%using some of the extensions described below, such as allowing for
+%additional random effects.
+%Writing custom MCMC algorthims is often tedious and foreign to
+%ecologists, but 
+Software packages such as %\textbf{WinBUGS} and
+\textbf{JAGS} \citep{plummer:2003} make these methods readily
+available to ecologists, and we have provided examples in the
+Appendix. 
+%overcome the technical 
+%difficulties by allowing users to specify the model using a simple symbolic 
+%descriptions. Examples are given in the Appendix.
 
@@
 
-We applied these models to North American Breeding Bird Survey
-(BBS) data from 1966-2010 in the bordering US states
+We applied our model extensions to North American Breeding Bird Survey
+(BBS) data collected from 1966-2010 in the bordering US states
 Maryland and Virginia. For our focal species, we selected the
 ovenbird (\textit{Seiurus aurocapilla}), an abundant and widespread,
-forest-breeding migratory songbird with a stable or increasing trend in the region \citep{porneluzi_etal:2011}.
+forest-breeding migratory songbird with a stable or increasing trend
+in the region \citep{porneluzi_etal:2011}. 
 %and golden-winged warblers (\textit{Vermivora chrysoptera}), which
@@
 
-We fitted a series of frequentist models, 
-and then a series of Bayesian models. We started by testing
+We fitted a series of models using maximum likelihood, 
+and then a series of models using MCMC. We started by testing
 three models of initial abundance (Poisson, negative binomial, and
@@
 %the minimum AIC dynamics model.
-As above, these models require a a finite upper bound ($N_{max}$) to integrate
+%As above, t
+These models require a a finite upper bound ($N_{max}$) to integrate
 over when run in a maximum likelihood
@@
 We ran the top ranked models from the maximum likelihood
-analyses in a Bayesian framework with non-informative priors.  We added random observer 
+analyses in a Bayesian framework with non-informative priors.  
+% JEFF: We need to be specific about the priors
+We added random observer 
 effects and environmental stochasticity, modeled as lognormal

### 2013-11-12 Dail_Mad_ext.tex
to results

@@
 on the Ricker-logistic + immigration dynamics model; here we fixed all
-parameters the same as the Ricker-logistic model (with r = 0.05 and K = 10) and
+parameters the same as the Ricker-logistic model (with $r$ = 0.05 and $K$ = 10) and
 simulated low, medium, and high values of immigration rate with 1000

### 2013-11-14 Dail_Mad_ext.tex
edits to results and discussion

@@
 
-We used the \texttt{unmarked} library in \textbf{R} to estimate the parameters 
+We used the \textbf{R} package \texttt{unmarked} to estimate the parameters 
 for each simulation with the same
@@
 We applied our model extensions to North American Breeding Bird Survey
-(BBS) data collected from 1966-2010 in the bordering US states
+(BBS) data collected from 1966-2010 in %the bordering US states
 Maryland and Virginia. For our focal species, we selected the
@@
 observers in the United States and Canada. An observer conducts 50
-3-minute point counts, with 400 m radii and 0.8 km apart from each other along a
+3-minute point counts, with 400 m radii and 0.8 km apart 
+%from each other 
+along a
 39.4 km route. We only used data marked as acceptable for use in the annual BBS
@@
 year specific abundance estimates.  Because of the memory requirements involved,
-we obtained these estimates from a single chain.  %We present these estimates for ovenbirds only.
+we used %obtained these estimates from 
+a single chain.  %We present these estimates for ovenbirds only.
 
@@
 482\% for non-zero true $r$) and biased low (between -0.008 and -0.005; relative bias between
--151\% and -101\% for non-zero true $r$) when the true value of $\Lambda$ was low (1). 
-Simulations of the Ricker-logistic model also generally performed well (Fig.~\ref{fig:rick_hists}; 
+-151\% and -101\% for non-zero true $r$) when the true value of
+$\Lambda$ was low. % (1). 
+Simulations of the Ricker-logistic model %also 
+generally performed well (Fig.~\ref{fig:rick_hists}; 
 Table S2.2 in Appendix 2), except when the true value of $r$ was low (0.005). In this case the 
@@
 run in the Bayesian framework, except that the estimate of $r$
-more than halved (Table S3.2 OR Figure S3.1, Appendix 3). %to 0.011 $\pm$ 0.005 [for
+more than halved 
+%% JEFF: How do we explain that????
+(Table S3.2 OR Figure S3.1, Appendix 3). %to 0.011 $\pm$ 0.005 [for
 %Bayesian model estimates we present mean and SD]). 
@@
 %for the estimate of mean initial abundance ($\Lambda$) was 0.15.
-As expected, uncertainty in estimates of route and year specific abundance was high.
+%As expected, uncertainty in estimates of route and year specific abundance was high.
 
@@
 
-Our work has highlighted the limitations of classical state-space
-models as applied to ecological time series data, and we have
-demonstrated how a class of open population $N$-mixture models
-proposed by \citet{dail_madsen:2011} overcomes these
-limitations. We extended their model in several important
+Our work %has highlighted the limitations of classical state-space
+%models as applied to ecological time series data, and we have
+%demonstrated 
+demonstrates how a class of open population $N$-mixture models
+proposed by \citet{dail_madsen:2011} overcomes 
+%these limitations. 
+many of the limitations of classical state-space models as applied to
+ecological time series data.
+We extended their model in several important
 ways to accomodate common features of ecological

### 2013-12-03 Dail_Mad_ext.tex
Incorporated comments from Richard plus formatted for Ecology

@@
-\documentclass[12pt]{article}
-\renewcommand*\abstractname{Summary}
-\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
+\documentclass[12pt,english]{article}
+%\renewcommand*\abstractname{Summary}
+\usepackage[verbose,letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
 \usepackage[mathlines]{lineno} %,displaymath
+\linenumbers
 \usepackage{setspace}
@@
 \usepackage{graphicx}
-\usepackage[round,colon,authoryear]{natbib}
+\usepackage[round,comma,authoryear]{natbib}
 \usepackage{color, soul}
@@
 \usepackage{verbatim}
+\usepackage{indentfirst}
+\usepackage[format=plain,justification=raggedright,labelsep=period,font={stretch=1.9}]{caption}
 %\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
-\usepackage[pdftex,hidelinks,pdfstartview={XYZ 1 1 1.5},bookmarks=true]{hyperref}
+\usepackage[pdftex,hidelinks,pdfstartview={XYZ 1 1 1.5}]{hyperref}
 \usepackage{rotating}
-
+\usepackage[compact]{titlesec}
+\titleformat{\section}[display]{\normalfont\bfseries}{}{0em}{\centering}
+\titleformat{\subsection}[display]{\normalfont\itshape}{}{0em}{\centering}
 \bibliographystyle{ecology}
+\setlength{\bibsep}{0.0pt}
+\bibpunct{(}{)}{,}{a}{}{,}
 %\hyphenpenalty=100000
+\setlength{\jot}{-1.2ex}
+\usepackage{etoolbox}
+\AtBeginEnvironment{tabular}{\doublespacing}
 
 \begin{document}
-
+\renewcommand{\thefootnote}{\arabic{footnote}}
 \title{Improved state-space models for inference about 
@@
 %\doublespacing
-\author{Jeffrey A. Hostetler\\ %\footnote{Contact:
-%     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
-        \small Migratory Bird Center \\
-        \small Smithsonian Conservation Biology Institute \\
-        \small National Zoological Park, MRC 5503 \\
-        \small Washington, DC 20013-7012\\
-   \and Richard B. Chandler \footnote{Current address: Warnell School of
+\author{Jeffrey A. Hostetler\thanks{Migratory Bird Center, Smithsonian 
+Conservation Biology Institute, National Zoological Park, MRC 5503, 
+Washington, DC 20013-7012}
+   \and Richard B. Chandler\thanks{USGS Patuxent Wildlife Research Center 
+        Laurel, MD 20708} \footnote{Current address: Warnell School of
      Forestry and Natural Resources, University of Georgia,
-     \url{rchandler@warnell.uga.edu}, +001-706-542-5815} \\
-        \small USGS Patuxent Wildlife Research Center \\
-        \small Laurel, MD 20708
+     \url{rchandler@warnell.uga.edu}, +001-706-542-5815}        
 }
+\date{} %remove date
 \maketitle
-
-\abstract{
-\noindent
-\begin{enumerate}
-\item Models of population dynamics %play a central role in theoretical and
+\begin{spacing}{1.9}
+\begin{flushleft}
+\abstract{%\noindent 
+Models of population dynamics %play a central role in theoretical and
 %applied ecology, where they 
@@
 of population dynamics to field data is challenging because most data
-sets are characterized by missing data and observation error, which
+sets are characterized by observation error, which
 can inflate estimates of process variation if ignored.  
@@
 %bias. Therefore, ecologists have increasingly relied on 
-Recently, state-space models have been developed to deal with these problems
+Recently, state-space models have been developed to deal with this problem
 %these probdirectly model 
@@
 process of interest.
-\item Conventional state-space models, however, have
+Conventional state-space models, however, have
 several important limitations: (1) they assume that random effects are
 Gaussian distributed, which implies that abundance can be negative and
-that false negative observation errors are equally likely as false positive
-errors, %results in unrealistic predictions
+that false positive observation errors are equally likely as false negative
+errors; %results in unrealistic predictions
 %(e.g. negative values of abundance) and an implausible mode
@@
 (2) they do not admit spatial variation in
-population dynamics, and (3) some of the parameters of the model are not
+population dynamics; and (3) some of the parameters of the model are not
 estimable. % they assert an implausible model for
@@
 that attributes observation error to imperfect detection.
-\item We expand this class of models to accommodate classical growth
+We expand this class of models to accommodate classical growth
 models (e.g. exponential and Ricker-logistic), zero-inflation,
@@
 of these ideas is possible using either frequentist or Bayesian
-methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code.%. We provide code to fit these models using the \textbf{R} package
+methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code. %We provide code to fit these models using the \textbf{R} package
 %\texttt{unmarked} and \textbf{JAGS} is also provided.
-\item Results of a simulation study suggest that bias is %was 
+Results of a simulation study suggest that bias is %was 
 negligible and coverage
@@
 replicated counts) is lacking.
-\end{enumerate}
 }
@@
 
-\textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
+\textit{Key words}: abundance, Dail and Madsen model, density-dependence,
 detectability, Gompertz-logistic model, immigration, open population point 
@@
 
+\section*{Introduction}
 %Theoretical ecology requires models of population dynamics for testing
@@
 population growth parameters. For instance, the strength of
-density dependence will be over-estimated if observation error is
+density dependence will be overestimated if observation error is
 ignored \citep{link_nichols:1994,shenk_etal:1998}.
@@
 A generic model can be written: %The full model can now be written as:
-\begin{subequations}
+\begin{linenomath*}
+\begin{gather}
   \label{eq:ss1}
   \begin{align}
-    N_1 &= X_1 \label{eq:ss1a} \\
+    N_1 &= X_1 \nonumber \\
 N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
-t=2,\hdots,T \label{eq:ss1b} \\
+t=2,\hdots,T  \\
 X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
-t=1,\hdots,T. \label{eq:ss1c}
+t=1,\hdots,T. \nonumber 
   \end{align}
-\end{subequations}
+\end{gather}
+\end{linenomath*}
 %where $\eta_t$ is the random effect allowing for process
@@
 Both frequentist and Bayesian methods of inference are discussed, and
-we evaluate the performance of the model using a simulation study by
+we evaluate the performance of the model using a simulation study and by
 analyzing data from the North American Breeding Bird Survey (BBS), one of
@@
 
-\section{The Dail-Madsen Model}
+\section*{The Dail-Madsen Model}
 \label{sec:dm}
@@
 
-\subsection{The Original Formulation of the Model}
+%\subsection{The Original Formulation of the Model}
 
@@
 
-\subsubsection{Initial abundance}
+\subsection*{Initial abundance}
 
@@
 as either a Poisson or negative binomial random variable:
+\begin{linenomath*}
 \begin{gather}
@@
 \end{gather}
+\end{linenomath*}
 where $\Lambda_i$ is the expected abundance at site $i$ during
@@
 
-\subsubsection{Abundance in subsequent time periods}
+\subsection*{Abundance in subsequent time periods}
 
@@
 the population. Their most general model was
+\begin{linenomath*}
 \begin{equation}
@@
 \end{equation}
+\end{linenomath*}
 where $\omega$ is the apparent survival probability and $\gamma$
@@
 
-\subsubsection{Observation process}
+\subsection*{Observation process}
 
@@
 %detection. The simplest model for imperfect detection is
+\begin{linenomath*}
 \begin{equation}
@@
 \end{equation}
+\end{linenomath*}
 where $p$ is the probability of detecting each individual. Variation
@@
 
-\section{Model Extensions}
+\section*{Model Extensions}
 \label{sec:ext}
 
-\subsection{Population growth models}
+\subsection*{Population growth models}
 
@@
 %total model combinations, and possibly more realistic estimates.
-If the mechanistic model must be sacrificed, we suggest replacing it
-with classical population growth models that have a long history in
-ecology. The exponential growth model is the simplest: 
-$f(N_{i,t-1}) = N_{i,t-1}e^r$ where $r$ is the intrinsic
+When the mechanistic model is unrealistic, we suggest replacing it
+with classical population growth models. %that have a long history in ecology.
+The exponential growth model is the simplest: 
+$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
 rate of increase. %Although demographic and environmental stochasticity
@@
 random variable, simplifying Eq.~\ref{eq:Nt} to:
+\begin{linenomath*}
 \begin{equation}
@@
 \end{equation}
-This,
-like the geometric-recruitment version of the model, is a variant on a simple
-density-independent, exponential model of population growth.
+\end{linenomath*}
+
+%This,
+%like the geometric-recruitment version of the model, is a variant on a simple
+%density-independent, exponential model of population growth.
 Density-dependent versions of the model are also possible.  For
 example:
+\begin{linenomath*}
 \begin{equation}
@@
 \end{equation}
+\end{linenomath*}
 where $K$ is the stable equilibrium of the population and $r$ is
@@
 density-dependent model \citep{hart_gotelli:2011}:
+\begin{linenomath*}
 \begin{equation}
@@
 \end{equation}
+\end{linenomath*}
 Here the interpretations of $r$ and $K$ are similar to in the
@@
 
-\subsection{Immigration models}
+\subsection*{Immigration models}
 
@@
 model is:
+\begin{linenomath*}
 \begin{equation}
@@
 \end{equation}
+\end{linenomath*}
 where $\iota$ represents the average number of immigrants per year, and is
@@
 
-We have implemented all preceding models in
-a maximum likelihood framework by extending the
-\texttt{unmarked} package
-\citep{fiske_chandler:2011} in \textbf{R} \citep{R-2012}.
+%We have implemented all preceding models in
+%a maximum likelihood framework by extending the
+%\texttt{unmarked} package
+%\citep{fiske_chandler:2011} in \textbf{R} \citep{R-2012}.
 
 %\subsection{Range model of initial abundance}
-\subsection{Excess zeros}
+\subsection*{Excess zeros}
 
@@
 %of initial abundances can be represented as:
+\begin{linenomath*}
 \begin{equation}
@@
 \end{equation}
+\end{linenomath*}
 where $\psi$ represents the proportion of extra zeros. 
+Mechanistically, this distribution might be appropriate when not all sites are 
+within the range of a species.
 
@@
 (Eq.~\ref{eq:Nt}) can be modified as follows:
+\begin{linenomath*}
 \begin{equation}
@@
 \end{equation}
+\end{linenomath*}
 %We have implemented
@@
 
-\subsection{Environmental and demographic stochasticity}
+\subsection*{Environmental and demographic stochasticity}
 \begin{comment}
@@
 The original DM models allows for demographic stochasticity --- variation in population growth 
-rate due to the randomness of birth and death processes --- through %random 
+rate due to the randomness of birth and death processes --- %through random 
 via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
-growth extensions use a single probability distribution for abundance: $N_{i,t} \sim 
-\mathrm{Pois}(f(N_{i,t-1}))$.  
+growth extensions use a single Poisson distribution for abundance. %: $N_{i,t} \sim 
+%\mathrm{Pois}(f(N_{i,t-1}))$.  
 % JEFF: I'm not sure what you're driving at here. Do these papers
 % assume a Poisson distribution too?
-%This is also a commonly used (and mathematically equivalent) 
-%method of modeling demographic stochasticity 
-%\citep{bonsall_hastings:2004,melbourne_hastings:2008}.
+This is also a commonly used and mathematically equivalent 
+method of modeling demographic stochasticity 
+\citep{bonsall_hastings:2004,melbourne_hastings:2008}.
 
@@
 as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
-function of temperature ($T$) and $K$ a function of rainfall ($R$):
+function of temperature ($C$) and $K$ a function of rainfall ($M$):
+\begin{linenomath*}
 \begin{gather}
-\log(r_{i,t}) = \beta^{r}_{0} + \beta^{r}_{1}T_{i,t} \nonumber \\
-\log(K_{i,t}) = \beta^{K}_{0} + \beta^{K}_{1}R_{i,t}  
+\log(r_{i,t}) = \beta^{r}_{0} + \beta^{r}_{1}C_{i,t} \nonumber \\
+\log(K_{i,t}) = \beta^{K}_{0} + \beta^{K}_{1}M_{i,t}  
 %N_{it} \sim
@@
 \end{gather}
+\end{linenomath*}
 
@@
 For example, for the Ricker-logistic + immigration model:
+\begin{linenomath*}
 \begin{equation}
@@
 \end{equation}
+\end{linenomath*}
 where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
@@
 Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
-for $\nu_{i,t}$ with distance-dependent (geographic or otherwise) correlation coefficients
+for $\nu_{i,t}$ with distance-dependent correlation coefficients %(geographic or otherwise) 
 between sites in a year might be appropriate.
 
-\subsection{Random observer effects}
+\subsection*{Random observer effects}
 
@@
 Eq.~\ref{eq:p1} can be modified to:
+\begin{linenomath*}
 \begin{gather}
@@
 \end{gather}
+\end{linenomath*}
 where $X_{i,j,t}$ is the number of individuals counted at site $i$ by
@@
 
-\subsection{Statistical inference}
+\subsection*{Statistical inference}
 
@@
 the parameters of their model. The same likelihood functions can be
-used to accomodate the alternative population growth models that we
+used to accomodate the alternative population growth, immigration, and 
+zero-inflation (initial abundance only) models that we
 have proposed. Methods for implementing these models using maximum 
@@
 \textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
-are given in the Appendix.
+are given in Appendix 1.
 
@@
 one common usage of state-space models is for predicting future
-population size, and this task easily accomplished using Bayesian
+population size, and this task is easily accomplished using Bayesian
 methods.
@@
 \textbf{JAGS} \citep{plummer:2003} make these methods readily
-available to ecologists, and we have provided examples in the
-Appendix. 
+available to ecologists, and we have provided examples in 
+Appendix 1. 
 %overcome the technical 
@@
 
-\section{Applications}
+\section*{Applications}
 \label{sec:app}
 
-\subsection{Simulation Study}
+\subsection*{Simulation Study}
 
@@
 
-\subsection{Analysis of Breeding Bird Survey Data}
+\subsection*{Analysis of Breeding Bird Survey Data}
 
@@
 observers in the United States and Canada. An observer conducts 50
-3-minute point counts, with 400 m radii and 0.8 km apart 
+3-minute point counts with 400 m radii, 0.8 km apart 
 %from each other 
@@
 We ran the top ranked models from the maximum likelihood
-analyses in a Bayesian framework with non-informative priors.  
+analyses in a Bayesian framework with non-informative priors ($\Lambda$,
+$\alpha$, and $K  \sim \mathrm{U}(0, 200)$, $r  \sim \mathrm{U}(0, 5)$,
+$\iota  \sim \mathrm{U}(0, 15)$, and logit-linear $p$ coefficients $\sim \mathrm{N}(0, 100)$).  
 % JEFF: We need to be specific about the priors
@@
 least 3,000 adaptation and tuning samples.  
-We also used these models to obtain route and 
+We also used the random observer effects model to obtain route and 
 year specific abundance estimates.  Because of the memory requirements involved,
@@
 
-\section{Results}
-
-\subsection{Simulation Study}
+\section*{Results}
+
+\subsection*{Simulation Study}
 
@@
 Simulations of the Ricker-logistic model %also 
-generally performed well (Fig.~\ref{fig:rick_hists}; 
-Table S2.2 in Appendix 2), except when the true value of $r$ was low (0.005). In this case the 
+generally performed well (Fig. S2.1 and Table S2.2 in Appendix 2), except when the true value of $r$ was low (0.005). In this case the 
 estimates of $r$ and $K$ were inaccurate (relative RMSE between 87.4\% and 2091.8\%),
@@
 91.9\% and 96.4\%.  All three cases simulated for the Ricker-logistic + immigration model 
-performed fairly well (Fig.~\ref{fig:ricki_hists}; Table S2.3 in Appendix 2; relative RMSE between 7.2\% 
+performed fairly well (Fig. S2.2 and Table S2.3 in Appendix 2; relative RMSE between 7.2\% 
 and 62.9\%, bias between -3.5\% and 3.3\%, and coverage between
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 Although incorporating these random effects can increase
-the difficulty of achieving model convergence and decrease estimate precision, they are certainly important.
+the difficulty of achieving model convergence and decrease estimate precision, they can be important.
 For example, the estimate of semi-equilibrium abundance for the ovenbird increased by more than 150\%
@@
 ecologists lacking skills in computer programming and advanced
-statistics. We have included an appendix with code to
+statistics. Appendix 1 includes code to
 simulate and fit many of the extensions discussed in this paper.
@@
 
-\section{Acknowledgements}
+\section*{Acknowledgements}
 
@@
 
-\vfill
-%\clearpage
-\newpage
+%\vfill
+%%\clearpage
+%\newpage
 
@@
 
-%\newpage
-
 %%% Figures
+\clearpage
+
+\section*{Figure Legends}
+\noindent Figure ~\ref{fig:exp_hists}. Histograms of 1000 parameter estimates for each of 27
+simulation cases with exponential dynamics. The vertical lines are the 
+data-generating values.
+
+%\noindent Figure ~\ref{fig:rick_hists}. 
+%
+%\noindent Figure ~\ref{fig:ricki_hists}. Histograms of 1000 parameter estimates for each of 3
+%simulation cases with Ricker-logistic + immigration dynamics. The vertical lines are the 
+%data-generating values.
+
+\noindent Figure ~\ref{fig:oven_N}. Maps of BBS route and year specific estimated 
+abundances for ovenbirds, for the years 1970, 1980, 1990, 2000, and 2010.  
+Lower right panel depicts estimated mean route abundance by year
+(mean and 95\% credible interval).
 
 \begin{figure}
+\caption{}
   \centering
   \includegraphics[height=8in]{figs/exp_hists}
-\caption{Histograms of 1000 parameter estimates for each of 27
-simulation cases with exponential dynamics. The vertical lines are the 
-data-generating values.}
 \label{fig:exp_hists}
@@
 
-\begin{sidewaysfigure}
-  \centering
-  \includegraphics{figs/rick_hists}
-\caption{Histograms of 1000 parameter estimates for each of 9
-simulation cases with Ricker-logistic dynamics. The vertical lines are the 
-data-generating values.}
-\label{fig:rick_hists}
-\end{sidewaysfigure}
-
-\begin{sidewaysfigure}
-  \centering
-  \includegraphics[width=8.5in]{figs/ricki_hists}
-\caption{Histograms of 1000 parameter estimates for each of 3
-simulation cases with Ricker-logistic + immigration dynamics. The vertical lines are the 
-data-generating values.}
-\label{fig:ricki_hists}
-\end{sidewaysfigure}
-
 \begin{figure}
+\caption{}
   \centering
   \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year6}
-\caption{Maps of BBS route and year specific estimated abundances for ovenbirds, for the years 1970,
-1980, 1990, 2000, and 2010.  Lower right panel depicts estimated mean route abundance by year
-(mean and 95\% credible interval).}
 \label{fig:oven_N}
 \end{figure}
-%\arglebargle
+\end{flushleft}
+\end{spacing}
 \end{document}

### 2014-06-25 Dail_Mad_ext.tex
Response to reviewer's comments

@@
 (Table S3.1C); the best supported of these was the Ricker-logistic + immigration 
-(see Appendix 3, Table S3.2 OR Figure S3.1 for estimates).
+(see Appendix 3, Fig. S3.1 for estimates).
 %This model estimated $\Lambda$ at 33.4 $\pm$ 4.7, $r$ at 0.026 $\pm$ 0.006, 
@@
 %% JEFF: How do we explain that????
-(Table S3.2 OR Figure S3.1, Appendix 3). %to 0.011 $\pm$ 0.005 [for
+(Appendix 3, Fig. S3.1). %to 0.011 $\pm$ 0.005 [for
 %Bayesian model estimates we present mean and SD]). 
@@
 
-We thank J. Andrew Royle and T. Scott Sillett for helpful
-comments and suggestions. We gratefully acknowledge the U. S. National Park 
-Service, the Smithsonian Institution, and [Richard's funding sources] for funding support.
+We thank J. Andrew Royle and T. Scott Sillett for helpful suggestions on our analysis
+and manuscript. We gratefully acknowledge the U. S. National Park 
+Service, the Smithsonian Institution, and USGS's Breeding Bird Survey program for funding support.
+We thank K. Pardieck and D. Zilowkowski for help with data acquisition. J. Sauer 
+provided helpful feedback about our analysis.
 

### 2014-06-25 improved_state_space.tex
Response to reviewer's comments

--- 2014-06-25 Dail_Mad_ext.tex
@@
 \documentclass[12pt,english]{article}
-%\renewcommand*\abstractname{Summary}
 \usepackage[verbose,letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
-\usepackage[mathlines]{lineno} %,displaymath
+\usepackage[mathlines]{lineno} 
 \linenumbers
@@
 \usepackage[format=plain,justification=raggedright,labelsep=period,font={stretch=1.9}]{caption}
-%\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
-\usepackage[pdftex,hidelinks,pdfstartview={XYZ 1 1 1.5}]{hyperref}
+\usepackage[pdftex,hidelinks]{hyperref} 
 \usepackage{rotating}
@@
 \bibpunct{(}{)}{,}{a}{}{,}
-%\hyphenpenalty=100000
 \setlength{\jot}{-1.2ex}
@@
 \AtBeginEnvironment{tabular}{\doublespacing}
-
-\begin{document}
-\renewcommand{\thefootnote}{\arabic{footnote}}
+\makeatletter
+\let\@fnsymbol\@arabic
+\makeatother
 \title{Improved state-space models for inference about 
@@
 
-%\doublespacing
 \author{Jeffrey A. Hostetler\thanks{Migratory Bird Center, Smithsonian 
 Conservation Biology Institute, National Zoological Park, MRC 5503, 
-Washington, DC 20013-7012}
-   \and Richard B. Chandler\thanks{USGS Patuxent Wildlife Research Center 
-        Laurel, MD 20708} \footnote{Current address: Warnell School of
-     Forestry and Natural Resources, University of Georgia,
-     \url{rchandler@warnell.uga.edu}, +001-706-542-5815}        
+Washington, DC 20013-7012} \textsuperscript{,}\footnote{Email: \href{mailto:hostetlerj@si.edu}{hostetlerj@si.edu}}
+   \and Richard B. Chandler\footnote{USGS Patuxent Wildlife Research Center 
+        Laurel, MD 20708} \textsuperscript{,}\thanks{Current address: Warnell School of
+     Forestry and Natural Resources, University of Georgia}        
 }
 \date{} %remove date
+\begin{document}
 \maketitle
+\vspace{-1cm}
 \begin{spacing}{1.9}
 \begin{flushleft}
-\abstract{%\noindent 
-Models of population dynamics %play a central role in theoretical and
-%applied ecology, where they 
+\abstract{
+Models of population dynamics 
 are frequently used for purposes such as testing 
@@
 can inflate estimates of process variation if ignored.  
-%Failure to account for observation error in such models can result in 
-%bias. Therefore, ecologists have increasingly relied on 
 Recently, state-space models have been developed to deal with this problem
-%these probdirectly model 
 by directly modeling both the observation error and the ecological
@@
 that false positive observation errors are equally likely as false negative
-errors; %results in unrealistic predictions
-%(e.g. negative values of abundance) and an implausible mode
-%and observation error are Gaussian random effects
-%the parameters are not identifiable
-%in many common situations, 
-(2) they do not admit spatial variation in
+errors; (2) they do not admit spatial variation in
 population dynamics; and (3) some of the parameters of the model are not
-estimable. % they assert an implausible model for
-%we have no clear interpretation of the
-%observation error. 
-We demonstrate how each of these problems can be
+estimable. We demonstrate how each of these problems can be
 resolved using a class of hierarchical models proposed by
 \citet[Biometrics]{dail_madsen:2011} 
-%for spatially-replicated time series data 
 that attributes observation error to imperfect detection.
@@
 and random effects. 
-% such as observer-specific detection probabilities. 
 We also present methods for forecasting
@@
 of these ideas is possible using either frequentist or Bayesian
-methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code. %We provide code to fit these models using the \textbf{R} package
-%\texttt{unmarked} and \textbf{JAGS} is also provided.
-Results of a simulation study suggest that bias is %was 
+methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code. 
+Results of a simulation study suggest that bias is 
 negligible and coverage
@@
 
-%\vspace{0.5cm}
-
 \textit{Key words}: abundance, Dail and Madsen model, density-dependence,
@@
 
-%\newpage
-\vspace{0.5cm}
+%\vspace{0.5cm}
 
 \section*{Introduction}
-%Theoretical ecology requires models of population dynamics for testing
-%hypotheses regarding spatial and temporal variation in
-%abundance. For example, much theoretical work has focused on
 Models of population dynamics are vital in both theoretical and
-applied ecological research. For instance, %the have been used to assess
+applied ecological research. For instance, 
 the importance and existence of phenomenon such as
@@
 populations with theoretical models
-\citep{may:1975,turchin:1990,bjornstad_etal:1999,royama:1977}. %,dennis_taper:1994
-%and population models are required to evaluate associated hypotheses.
+\citep{may:1975,turchin:1990,bjornstad_etal:1999}. 
 In applied contexts, population models are used for estimating extinction probabilities
@@
 Several challenges are routinely encountered when trying to fit models
-of population dynamics to data from field studies. % is that 
-%In order to address these questions, two complicating factors must
-%be confronted when fitting population models to data. 
+of population dynamics to data from field studies. 
 First, deterministic models of population dynamics are virtually always inadequate due to process
@@
 
-Recently developed state-space models %are a widely used approach for
-have made it possible to %become the primary tool for 
+Recently developed state-space models 
+have made it possible to
 study population dynamics while accounting for both process variation
@@
 state-space models in ecology is that failure to account for 
-%process variation and 
 observation error can bias estimators of abundance and
@@
 and random process variation denoted $\eta_t$.
-% Richard, is switching this from $\grave{o}$ to $\eta$ okay?  Andy thought
-% the o with accent was weird and I think I agree.  $\nu$ seems standard,
-% but we use it for a slightly different meaning (ES only) later.  
 The growth model may be density-dependent, as in the case of the 
@@
 is exponential.
-%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
-A generic model can be written: %The full model can now be written as:
+A generic model can be written: 
 \begin{linenomath*}
@@
 \end{linenomath*}
-%where $\eta_t$ is the random effect allowing for process
-%variation unaccounted for by the deterministic model. 
 In classical
@@
 $\eta_t \sim \mathrm{N}(0, \sigma)$ and
-$\zeta_t \sim \mathrm{N}(0, \tau)$. %Standard practice is to
-%ignore 
+$\zeta_t \sim \mathrm{N}(0, \tau)$. 
 The process variation associated with $N_1$ is often ignored, or
-%, as indicated by Eq~\ref{eq:ss1a}. Alternatively, 
 the population is assumed to be at equilibrium such that the $N_1$ can
@@
 classical models of population dynamics, 
-%are among the most widespread approaches for modeling population dynamics
-%using time series data, 
-several problems exist. %, which we explain in the following paragraphs. 
-% First,
-% time series data and the underlying abundance parameters of interest
-% are typically integer-valued, as in the case of count data, raising
-% concerns about the Gaussian assumption for the random
-% effects. Second, the model for observation error is implausible. %has little biological basis. 
-% Third, the model does not account for spatial variation in abundance. Lastly, some of the
-% parameters of the model are not estimable. We briefly discuss each of
-% these points before describing a general approach to resolving these
-% shortcomings.
-%Use of the Gaussian distribution for modeling random process variation
-%and observation error is often motivated by convenience rather than
-%biology. Specifically, t
+several problems exist. 
 The assumption that random effects are Gaussian distributed makes it
 easier to estimate parameters because methods such as the Kalman
-filter can be applied %The Gaussian assumption allows for efficient methods of parameter
-%estimation, such as the Kalman filter
-\citep{dennis_etal:2006}; however, % which are not available otherwise % which is much more computationally efficient
-%than estimation methods when random effects are not Gaussian distributed
-%\citep{devalpine_hastings:2002}. %The problem with this is that it
-%However, 
+filter can be applied 
+\citep{dennis_etal:2006}; however, 
 these models effectively assume that abundance can be take on any
-real value, and hence they can predict negative values of abundance. %including negative values, 
-%This approach allows for negative and non-integer values of the state-variable, which is
-%inconsistent with the observed data, and may 
-%and these models can result in implausible predictions.
+real value, and hence they can predict negative values of abundance. 
 Transforming the count data and the underlying abundance variables 
@@
 With respect to the observation error, the Gaussian assumption
-%Another problem with standard state-space models is that the model for
-%the mean zero Gaussian distribution for the observation error
 implies that false positive detections are
@@
 findings \citep{miller_etal:2011}. 
-%. To our knowledge, all studies to
-%date have shown that false negative detections are much more common
-%that false positives \citep{miller_etal:2011}.
-%will be
-%higher than $N_t$ as often as it is lower than $N_t$. 
-%Identifying a biological mechanism that would cause such symmetric errors
-%is difficult.
 A more likely form of observation error, and one that has been recognized for well
@@
 present. Imperfect detection may be attributable to
-characteristics of the species under study, such as its elusiveness,
-or to individuals being missed by the ecologist collecting the data in the field.
+characteristics of the species under study, such as it being present but unavailable for detection,
+or to individuals being available but missed by the ecologist collecting the data in the field.
 Although a vast number of methods have been devised for accounting for
@@
 \citep{polansky_etal:2009}. 
-% without making 
-%Another limitation of many state-space models is that they do 
-%Common formulations of state space models do 
-%not admit spatial variation in abundance or demographic
-%parameters. 
-Ignorning spatial variation simply reduces the scope of the
-inferences, 
-% that can 
-%be drawn from the models, it disregards the fact that space is an
-%important factor in the regulation of natural populations \citep{murdoch:1994,
-%pulliam:1988, gill_etal:2001}.
-but identifiability problems require %lack of identifiability because
+Ignoring spatial variation reduces the scope of the
+inferences and can lead to underestimation of variances. 
+Identifiability problems require 
 model constraints or the assertion of additional assumptions. 
-%A more serious problem %than the ones associated with the Gaussian assumptions 
-%is that the parameters of simple state-space models
-%such as Eq~\ref{eq:ss1} are not identifiable in many
-%circumstances \citep{polansky_etal:2009}. 
 Specifically, abundance in
@@
 point (the count in the first year) that carries any information about
-it. To remedy this, pratitioners either assume that there is no
+it. To remedy this, practitioners either assume that there is no
 observation error in the first year, or they assume the population is
@@
 population dynamics.
-%Recall that the
-%Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
-%although $N_1$ is a random variable. Thus, uncertainty exists
-%that should be accounted for in the model.  
-%%are needed to fully specify the model.
-%To be more specific, a fully-specified state-space model requires at least
-%three probability distributions, which we represent
-%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
-%as the probability distribution of the random variable $Y$ given the
-%parameter $\Omega$). The three probability distributions required for a
-%state-space model are:
-%%%% JEFF, does this extra sentence help? This notation comes from
-%%%% several of the original state-space papers and is widespread in
-%%%% books like Andy's and Bill's.
-%\hl{[I'm not familiar with bracket notation,
-% so these equations are confusing to me. How widely used is this
-% notation in ecology? Will others likely understand this? Also, we might want
-%to save space by trimming a bunch of this discussion back down.]}
-%\begin{subequations}
-%  \label{eq:ss2}
-%  \begin{align}
-%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
-%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
-%\label{eq:ss2c}
-%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
-%  \end{align}
-%\end{subequations}
-%where $\bm{\theta}$ are the process variation parameters for the
-%initial state, $\bm{\Theta}$ are the process variation parameters
-%influencing how abundance changes over time, and $\bm{p}$ are the
-%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
-%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
-%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
-%The three lines of general model
-%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
-%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
-%we allow $N_1$ to be a random variable. But what should its
-%distribution, $[N_1|\bm{\theta}]$ be?
-%And how can $\bm{\theta}$ be estimated given that there is only a single observation
-%vailable? 
-%Often, these problems are simply assumed away. For
-%example, sometimes researchers assume that there is no process variation in
-%the first year, although this seems unjustified. Alternatively, the
-%population may be assumed to be at equilibrium such that the expected
-%value of $N_t$ is constant through time. Although this 
-%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
-%defeats the objective of many studies of population dynamics, namely
-%determining why a population varies over time.
-%% OR
-%Sometimes researchers assume that there is no process variation in 
-%the first year or that the population is at equilibrium such that the expected
-%value of $N_t$ is constant through time, but these approaches are unsatisfactory.
 
@@
 simultaneously; however, their model did not include serial
-dependence, which is a hallmark of population models. %This
-%limits the utility of their model for making inferences about explicit
-%population processes.
+dependence, which is a hallmark of population models.
 
@@
 that resolves each of the problems outlined above. 
-%simultaneous resolves each of the aforementioned problems with
-%traditional state-space models, and is designed for simple count data.
-%In the following section, we describe the DM model in its original
-%form and explain how it resolves each of the deficiencies with
-%standard state-space models. 
-%In Section~\ref{sec:ext}, 
 Our aim is to extend the
@@
 time series, such as excess zeros and unexplained random variation. 
-%Specifically, we describe methods for accommodating
-%excess zeros and nuisance variables such as random observer effects.
 Both frequentist and Bayesian methods of inference are discussed, and
@@
 vertebrate populations \citep{robbins_etal:1986}.
-%code for fitting models is presented in the appendices.
-%The Bayesian method
-%is attractive in that it can accommodate prior information about
-%detection parameters, which is available for many species, yet not
-%collected as part of many existing monitoring programs.
-%In Section~\ref{sec:app}, we evaluate the performance of the model
-%extensions using a simulation study and by analyzing data from the
-%North American Breeding Bird Survey (BBS), one of
-%the most spatially and temporally extensive sets of count data on
-%vertebrate populations \citep{robbins_etal:1986}. %The overarching aim
-%of the paper is to provide ecologists with flexible and accessible means of
-%addressing important questions related to the variation of abundance in
-%space and time. 
-%believe these extensions will increase the utility of the model for
-%application to existing datasets such the Breeding Bird Survey.
 
@@
 
-%\subsection{The Data}
-
 The DM model requires count data collected at $R$ sites, each of
 which is surveyed on $T$ primary sampling periods. 
-%A site is ideally a
-%well-defined region of a study area, such as a wetland or a
-%patch of early-successional habitat, although it could be an arbitrarily
-%defined region such as a randomly located survey plots. The time
-%frame of the study is arbitrary, but in general
-%it will be sufficiently long to allow for temporal variation in
-%abundance. For instance, a primary sampling period could be a
-%one month long breeding season, which could be surveyed once per
-%year for $T$ years. If the a site is surveyed on $J$ occasions during a
-%primary period, these are called secondary sampling occasions.
-%In the case where no secondary sampling was conducted, i.e. $J=1$, 
 Let $X_{i,t}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
 at site $i$ and primary period $t$. 
-%If secondary sampling periods were
-%used, an extra dimension is added so that we have $X_{ijt}:
-%j=1,\hdots,J$. 
 In general, $X_{i,t}$ will be lower than abundance, 
-%actual quantity of interest, abundance,
-%denoted 
 $N_{i,t}$, but in cases where detection is perfect,
@@
 
-%\subsection{The Original Formulation of the Model}
-
 Like traditional state-space models, the DM model includes 
 three conditionally related processes corresponding to: 
-% shown in Eq.~\ref{eq:ss1}, which correspond to 
 (1) initial abundance, i.e. the
 abundance at site $i$ during the first primary period,
-denoted (2) abundance at time $t$ (for $t>1$) which depends upon
+denoted $N_{i,1}$, (2) abundance at time $t$ (for $t>1$) which depends upon
 abundance at $t-1$, and (3) the
@@
 The first two processes describe the state process---the
-variation in abundance in space and time. The third process %simply 
+variation in abundance in space and time. The third process 
 describes the the relationship between
@@
 
-%Recall that 
 Conventional state-space models assume that the
-distribution for the initial time period %,
-%$[N_1 | \bm{\theta}]$, 
+distribution for the initial time period 
 is either the equilibrium
 distribution, or has zero variance.
-In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
+In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i,1}$
 as either a Poisson or negative binomial random variable:
@@
 overdispersion determined by the parameter $\alpha$.
-%In the Poisson case, the equivalent expression of
-%$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
-%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
-%\begin{equation}
-%[N_{i1} | \Lambda] =
-%\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
-%  \label{eq:N1br}
-%\end{equation}
-% Maybe we don't need the Poisson distriubtion equation? - Jeff
 
@@
 \beta^{\Lambda}_1
-x_{i}$. The second important point is that the spatial
+x_{i}$. Second, the spatial
 replicates resolve the
 problem of parameter non-identifiability that are common to
-standard state-space models because, as demonstrated by
-\citet{royle:2004biom},
+standard state-space models because
+%, as demonstrated by \citet{royle:2004biom},
 process variation and observation error can be estimated from
-spatially-replicated count data. Hence, the first component of
+spatially-replicated count data \citep{royle:2004biom}. Hence, the first component of
 the model addresses both the issues of spatial inference and
@@
 abundance at time $t-1$, i.e. abundance at each site evolves as a
-first order Markov process. %, although higher order processes are also possible.
+first order Markov process. 
 \citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
 however, in each case they modeled $N_t$ as the sum of two random variables:
-$S_{i,t}$, the number of indivduals surviving from $t-1$ and not
+$S_{i,t}$, the number of individuals surviving from $t-1$ and not
 emigrating; and $G_{i,t}$ the number of new individuals entering
@@
 
-%To make the connection between conventional state-space models
-%and the DM model, we need to replace $[N_t|N_{t-1}, \bm{\Theta}]$,
-%Eq.~\ref{eq:ss2b}, with an expression derived from
-%Eq.~\ref{eq:Nt}. This requires summing over all possible values
-%of $S$ and $G$, which is accomplished using the discrete convolution:
-%\begin{equation}
-%[N_{it}|N_{it-1}, \omega, \gamma] =
-%  \sum_{S_{it-1}=0}^{\min(N_{it},N_{it-1})}
-%\binom{N_{it-1}}{S_{it-1}}\omega^{S_{it-1}}(1-\omega)^{N_{it-1}-S_{it-1}}
-%\times
-%\frac{\gamma^{N_{it}-S_{it-1}}e^{-\gamma}}{(N_{it}-S_{it-1})!}  \label{eq:P}
-%\end{equation}
-%where here we assumed no dependence of $\gamma$ on $N_{it-1}$.
-%
-%In the absence of movement, $\omega$ is exactly the probability of surviving
-%from year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate
-%(under the geometric growth model). This ability to directly estimate
-%demographic parameters from simple count data is one of the DM
-%model's most appearling features. However,
-%in the more common scenario when immigration and emigration
-%occur, $\omega$ and $\gamma$ no longer can be interpreted as vital
-%rates. Rather, $\omega$ is the probability of surviving and not
-%emigrating, and $\gamma$ is the sum of the birth rate and the immigration rate,
-%i.e. the recruitment rate. In such cases, we propose
-%replacing Eq.~\ref{eq:Nt} with standard population models.
-
-\begin{comment}
-Irrespective of the model for population dynamics, it is
-worthwhile
-to note that the discrete distributions used in the DM model
-avoid
-the issues associated with the Gaussian distribution used to
-model
-  process variation in standard state-space models.
-\end{comment}
-
 \subsection*{Observation process}
 
-%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
-%i.e. the model for abundance dynamics. This state
-%model has allowed us to replace the initial abundance
-%distribution and transition models of conventional state-space models, with
-%alternatives that (1) allow for modeling spatial variation in
-%abundance, (2) respect the discrete nature of the counts, and
-%(3) have parameters that are estimable.
-%The one
-%remaining task is to replace the observation model
-%$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
-%more easily interpreted. 
 The observation model adopted by \citet{dail_madsen:2011} is the same
-binomial model proposed by \citet{royle:2004biom}: % assumes that individuals are missed due to imperfect
-%detection. The simplest model for imperfect detection is
+binomial model proposed by \citet{royle:2004biom}: 
 \begin{linenomath*}
@@
 or occasion-specific covariates using, for example, a logit-linear model.
-%or equivalently,
-%\begin{equation}
-%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
-%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
-%  \label{eq:1}
-%\end{equation}
-%Under this model, $\bm{p}$ is interpreted as the detection
-%probability parameter.
-
-%%%% Save for discussion?
-%Surprisingly detection probability, $p$, can be estimated without
-%secondary sampling occasions. However, this is only possible
-%if the parametric assumptions of the population dynamics are met
-%exactly. In practice, it is preferable to obtain direct information
-%about $p$ using secondary sampling occasions or some other auxiliary
-%data, \hl{as is demonstrated in section....}
-
-%The entire model can now be written:
-%\begin{gather}
-%  N_{i1} \sim \text{Pois}(\Lambda) \nonumber \\
-%S_{it}|N_{it-1} \sim \mathrm{Bin}(N_{it-1}, \omega) \nonumber \\  
-%G_{it}|N_{it-1}\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
-%  N_{it} = S_{it}+G_{it} \nonumber \\
-%  X_{it} \sim \text{Bin}(N_{it}, p) \nonumber
-%  \label{eq:hm}
-%\end{gather}
-
-%\subsection{Statistical Inference}
-
-%\subsubsection{Maximum likelihood}
-
-%Maximum likelihood estimation of parameters in models with
-%discrete random effects requires removing them by
-%summation. For Markovian models, this can be done recursively,
-%making the likelihood tractable \citep{dail_madsen:2011}. 
-%A general expression of the likelihood can be written:
-%\begin{multline}
-%  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
-%  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
-%  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
-%%  \prod_{t=T}^2 \left\{
-%  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
-%  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
-%  \right\} \right\}
-%  \label{eq:like}
-%\end{multline}
-%where $\infty$ is replaced by a finite upper bound $N_{max}$, which
-%%% Can we rename the upper bound?  We're also using K
-%%% for equilibrium abundance.
-%should be high enough such that the MLEs are not affected by increasing
-%it further. 
-%The likelihood can be maximized numerically using the
-%\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
-%are given in the Appendix.
 
@@
 \label{sec:ext}
-
 \subsection*{Population growth models}
-
-% Preliminary DM model runs for several species tended to lead to
-% estimates of survival that were unrealistically high and recruitment
-% that were unrealistically low, or the reverse (compared to independent
-% demographic analyses). The DM models are able to partition changes in
-% abundance to survival and recruitment in part by making strong
-% distributional assumptions. % (Dail and Madsen 2011).
-% When those assumptions are heavily violated the models may proportion
-% population growth incorrectly into survival and recruitment, even if
-% they estimates population growth accurately.
-
+\begin{samepage}
 Partitioning population growth into survival and recruitment
@@
 especially when the sites are not closed with respect to immigration
-and emigration. %Furthermore, a
-%simpler model would have several merits: faster running time, fewer
-%total model combinations, and possibly more realistic estimates.
+and emigration. 
 When the mechanistic model is unrealistic, we suggest replacing it
-with classical population growth models. %that have a long history in ecology.
+with classical population growth models that have a long history in ecology.
 The exponential growth model is the simplest: 
 $N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
-rate of increase. %Although demographic and environmental stochasticity
-%in $r$ could be modeled as functions of covariates or random effects, a
-%simpler way of 
+rate of increase. 
 To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
@@
 
-%This,
-%like the geometric-recruitment version of the model, is a variant on a simple
-%density-independent, exponential model of population growth.
 Density-dependent versions of the model are also possible.  For
@@
 \end{linenomath*}
-Here the interpretations of $r$ and $K$ are similar to in the
+Here the interpretations of $r$ and $K$ are similar to those in the
 Ricker-logistic model. Because a single Poisson distribution controls
@@
 likelihood is not required, speeding up processing time.
+\end{samepage}
 
@@
 Ricker-logistic, and Gompertz-logistic versions of the DM models all
-%share a common feature (or bug): once the 
 is that they include no mechanism for a local population to recover
-after extinction. %at a site reaches 0, it must remain at 0
-%because all contributions to population growth are assumed to be
-%local.  
+after extinction. 
 However, these models can be generalized 
-%We therefore generalized these models 
 to include both internal and external (immigration) contributions
@@
 positive (this is equivalent to separate Poisson processes for
-growth and immigration).  %This model is close to
-%the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
-%of $\omega$ and $\iota$ instead of $\gamma$, except that the first
-%process is Poisson distributed instead of binomial. 
+growth and immigration).  
 The geometric-recruitment, Ricker-logistic,
@@
 refer to $K$ in these models as the semi-equilibrium abundance.
-\begin{comment}
-Similarly, the Ricker-logistic + immigration model can be represented as:
-the
-Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
-\end{comment}
-
-% Save for discussion
-\begin{comment}
-These immigration models are based on the assumption that number
-of
-immigrant is not dependent on local density; however, extendingthese models in this way is conceptually straight-forward, a
-Other
-  versions of these models with different assumptions are
-possible. The Ricker-logistic + immigration and Gompertz-logistic + immigration
-models
-can be justified by the further assumption that immigrants
-arrive as
-  adults immediately before the point-count survey (so that
-density-dependent processes do not have a chance to act on thenumbers of immigrants; Otto and Day 2007 p.
-161).
-Deterministicversions of these models do reach stable equilibriums, but theequilibriums are not at K and cannot be solved for analytically  (Otto and Day 2007).
-\end{comment}
-
-%We have implemented all preceding models in
-%a maximum likelihood framework by extending the
-%\texttt{unmarked} package
-%\citep{fiske_chandler:2011} in \textbf{R} \citep{R-2012}.
-
-%\subsection{Range model of initial abundance}
+
 \subsection*{Excess zeros}
 
-%\citet{dail_madsen:2011} suggested two distributions for modeling
-%initial abundance: Poisson and negative binomial. We have
-%extended their model to include another distribution, the zero-inflated
-%Poisson. This could be useful when, for example, one is modeling
-%the abundance of several species with the same set of point
-%count surveys for all species, but some survey sites are
-%outside the range of some of the species. 
 The negative binomial distribution used in $N$-mixture models can have
@@
 distributions, such as the zero-inflated Poisson:
-%The distribution
-%of initial abundances can be represented as:
 \begin{linenomath*}
@@
 
-%These models allow three sources of zero counts by observers: a
-%species was at a site but not detected; the site was within the
-%species' range but no individuals occured at that site in that
-%year; and the site was outside the species' range. Furthermore,
 Under this formulation, 
@@
 climate.
-%This approach combines elements of occupancy modeling
-%\citep{mackenzie_etal:2006} and abundance modeling.
-%, and could prove useful
-%for distinguishing species' fundamental and realized niches
-%(Hutchinson 1957).
-
-%The zero inflation factor can also affect
-%recruitment or population dynamics (see below).
-%We also considered a zero-inflated negative binomial model of initial
-%abundance, but this model did not perform well in preliminary tests.
-
-%\subsection{Dynamic range models}
 
@@
 \end{linenomath*}
-%We have implemented
-%zero-inflated dynamics in the Bayesian framework using program
-%\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
-%(version 3.2.0; Plummer 2003)
 
@@
 \begin{comment}
-Even though the DM model is conceptually simple, %even in
-%its most stripped down form 
+Even though the DM model is conceptually simple, 
 it includes many random effects (the $N$'s)
@@
 The original DM models allows for demographic stochasticity --- variation in population growth 
-rate due to the randomness of birth and death processes --- %through random 
+rate due to the randomness of birth and death processes --- 
 via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
-growth extensions use a single Poisson distribution for abundance. %: $N_{i,t} \sim 
-%\mathrm{Pois}(f(N_{i,t-1}))$.  
-% JEFF: I'm not sure what you're driving at here. Do these papers
-% assume a Poisson distribution too?
-This is also a commonly used and mathematically equivalent 
-method of modeling demographic stochasticity 
-\citep{bonsall_hastings:2004,melbourne_hastings:2008}.
+growth extensions use a single Poisson distribution for abundance. 
+Conceptually, this is an appropriate method of modeling demographic stochasticity 
+for semelparous organisms with one generation per sampling interval (year)
+\citep{bonsall_hastings:2004,melbourne_hastings:2008},
+but may actually perform better for iteroparous organisms as well (see Discussion).  
+Dynamics of semelparous organisms with more than one generation per year 
+(many insects) may be overdispersed compared to the Poisson.  If the number of generations per
+year is known, then this can be accounted for in our models by adding extra $N$'s between each
+count (see Supplement 1).
 
@@
 \log(K_{i,t}) = \beta^{K}_{0} + \beta^{K}_{1}M_{i,t}  
-%N_{it} \sim
-%\text{Pois}(N_{it-1}\exp(r_{it-1}(1-\log(N_{it-1}+1)/\log(K_{it-1}+1)))) \nonumber
 \label{eq:weather}
@@
 
-% JEFF: It seems to me that this approach could be used to account for
-% either demographic or enviro stochasticity. How can you know what
-% the variation in attributable to?
 Another way to model environmental stochasticity is as lognormal variation in observed local population 
@@
 where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
-Three variations suggest themselves.  The first has independent environmental stochasticity 
+We suggest three variations for spatial variation in environmental stochasticity.  The first has independent environmental stochasticity 
 between sites, and might be appropriate where sites are widely dispersed.  In the second, 
@@
 Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
-for $\nu_{i,t}$ with distance-dependent correlation coefficients %(geographic or otherwise) 
+for $\nu_{i,t}$ with distance-dependent correlation coefficients 
 between sites in a year might be appropriate.
@@
 in avian point count surveys such as the BBS 
-\citep{robbins_etal:1986,sauer_etal:1994auk,campbell_francis:2011}.%,alldredge_etal:2007auk %diefenbach_etal:2003,
-%Estimating a separate detection probability for each observer
-%can be difficult and reduces one's ability to estimate the quantities
-%of interest.  This problem is compounded by the fact that
-%observers differ greatly in the number of surveys they have
-%run (so that many observers' separate detection probabilities
-%could not be accurately estimated).
+\citep{robbins_etal:1986,sauer_etal:1994auk,campbell_francis:2011}.
 
 Current BBS trend estimators deal with this problem by
-treating observer identity as a random %(as opposed to a fixed)
+treating observer identity as a random 
 effect \citep{link_sauer:2002,sauer_link:2011}.
-%(Link and Sauer 2002, Sauer and Link 2011).
-%This allows observer-specific differences in detection probability,
-%but assumes that observers are selected at random from a pool
-%of potential observers.  Models that contain both random and
-%fixed effects are referred to as mixed models.  Often in mixed
-%models (as in this case), the random effect is modeled not
-%because it of interest in itself, but to avoid bias in the
-%estimates of the fixed effects.
-%
 To include random observer effects in DM models, 
-% and these extensions, 
 Eq.~\ref{eq:p1} can be modified to:
@@
 the standard deviation of the random observer effects (also on the logit scale).
-%We have implemented random observer effects in the Bayesian
-%framework using program \textbf{JAGS} \citep[version 3.2.0, see Appendix]{plummer:2003}.
-% (version 3.2.0; Plummer 2003)
-%with the \textbf{R} package \texttt{rjags} \citep{plummer:2011,R-2012} interface.
-
-%\subsection{Forecasting Future Population Size}
-
-%State-space models allow for inference about future population size,
-%a critical component of population viability analysis (PVA) as well
-%as conservation planning. 
-%Let $N^+_{t}$ denote population size at some
-%future point in time and $[N^+_t]$ denote the probability
-%distribution for this random variable. If we can specify this
-%distribution, we can make statements about the
-%probability that $N^+_t = k$ where $k$ is any possible
-%population size including zero.
-
-%The classical approach to computing $[N^+_t]$ is based upon
-%Empirical Bayes methods [CITE]. 
-%Note that neither Empiriis not actually a
-%Bayesian method, it is used to estim.
-%We take the parameter estimates and plug them back in.... For
-%example to obtain the posterior distribution for $[N_{i1}]$
-%\begin{equation}
-%[\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},
-%\hat{p}]}{\sum_{N_{i1}=0}^K
-%[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
-%  \label{eq:eb1}
-%\end{equation}
-%Once we have estimated the posterior distribution for the first
-%time period, we can then proceed to estimate the posterior
-%distribution for subsequent time periods using recursion.
-%\begin{equation}
-%  [\hat{N}_{it}] =
-%  \frac{[N_{it}|N_{it-1},\hat{\bm{\Theta}}][X_{it}|N_{it},
-%\hat{p}]}{\sum_{N_{it}=0}^K [N_{it}|N_{it-1},
-%\hat{\Theta}][X_{it}|N_{it},\hat{p}]}
-%  \label{eq:eb2}
-%\end{equation}
-%
-%Note that when $X_{it}$ is not observed, including for future
-%time periods, the distribution $[X_{it}|N_{it}]$ is taken to be
-%uniform such that it has no influence on the posterior. In other words,
-%the posterior for years with no observed data is determined entirely
-%by the estimated posterior from the previous year and the
-%population dynamics parameters.
-%
-%If covariates of the dynamics parameters are present, the problem
-%becomes much harder because, to fully account for uncertainty,
-%we need a model for the future values of the covariates as well. When
-%the covariates are continuous, this requires integrating over their
-%possible values, which becomes very challenging
-%computationally. Indeed, simulation methods such as MCMC are the
-%only viable options when numerous covariates are present, which will
-%often be the case in many ecological studies.
-
-%% COULD INCLUDE THIS PARAGRAPH, OR NOT.
-%A standard approach to PVA uses previously derived estimates of current 
-%population size, vital rate or population growth parameters, and environmental
-%stochasticity to stochastically project the population forward and calculate probabilities
-%of extinction or other measures of risk.  Estimates from DM models and the extensions
-%given here can be used in this context.  Uncertainty in extinction or quasi-extinction probabilities
-%can be estimated by either drawing values for parameters from their posterior distributions
-%(if estimated using MCMC) or randomly drawing values for parameters using their means and standard
-%errors (if estimated using maximum likelihood), and running a number of simulations with each
-%set of parameter values \citep{white:2000,hostetler_etal:2012}.
-
-%Relative to the computational challenges associated with the
-%empirical Bayes methods, 
-%A fully Bayesian approach to PVA is straightforward to
-%implement with MCMC. Because population size at any future point %Furthermore, b
-%in time is regarded as a standard, unknown parameter, computing the
-%posterior distribution of $N^+_t$ requires no additional steps
-%in the MCMC algorithm, other than formally stating that the count data are
-%``missing'' in future years. An example of this is provided in the Appendix.
 
@@
 the parameters of their model. The same likelihood functions can be
-used to accomodate the alternative population growth, immigration, and 
+used to accommodate the alternative population growth, immigration, and 
 zero-inflation (initial abundance only) models that we
 have proposed. Methods for implementing these models using maximum 
-likelihood estimation are included in the %. The likelihood can be maximized numerically using the
+likelihood estimation are included in the 
 \textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
-are given in Appendix 1.
+are given in Supplement 1.
 
@@
 effects.  
-% Although MCMC can be slow, our experience is that it
-%may actually converge faster than the amount of time required to
-%maximize the likelihood shown in Eq.\ref{eq:like}, especially
-%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
-%Furthermore, 
-%MCMC may be the only possible option for estimating parameters
-%using some of the extensions described below, such as allowing for
-%additional random effects.
-%Writing custom MCMC algorthims is often tedious and foreign to
-%ecologists, but 
-Software packages such as %\textbf{WinBUGS} and
+Software packages such as 
 \textbf{JAGS} \citep{plummer:2003} make these methods readily
 available to ecologists, and we have provided examples in 
-Appendix 1. 
-%overcome the technical 
-%difficulties by allowing users to specify the model using a simple symbolic 
-%descriptions. Examples are given in the Appendix.
-
-%For example here is \textbf{BUGS} code for the ``geometric'' model described
-%in Eq.~\ref{eq:hm}:
-%\begin{verbatim}
-%model {
-%lambda ~ dunif(0, 5)
-%omega ~ dunif(0,1)
-%gamma ~ dunif(0, 10)
-%p ~ dunif(0,1)
-%for(i in 1:nSites) {
-%  N[i,1] ~ dpois(lambda)
-%  y[i,1] ~ dbin(p, N[i,1])
-%  for(t in 2:nYears) {
-%    S[i,t-1] ~ dbin(omega, N[i,t-1])
-%    G[i,t-1] ~ dpois(gamma*N[i,t-1])
-%    N[i,t] <- S[i,t-1] + G[i,t-1]
-%    y[i,t] ~ dbin(p, N[i,t])
-%    }
-%  }
-%}
-%\end{verbatim}
+Supplement 1.
 
@@
 
-\subsection*{Simulation Study}
-
-We simulated data for 100 sites over 40 years.  All
+\subsection*{Proof of Concept Simulation Study}
+
+We simulated data for 100 sites over 40 years, with 1000 simulations of 
+each combination of parameter values.  All
 simulations assumed initial abundance was Poisson distributed
@@
 detection probability.  Our first series of simulations
-assumed dynamics were exponential.  We ran 1000 simulations for
+assumed dynamics were exponential.  We ran 
 each combination of low, medium, and high $\Lambda \in
 \{1,5,10\}$, $r \in \{-0.005, 0, 0.005\}$, and
-$p \in \{0.05, 0.25, 0.5\}$. % (Table 1; 27 total combinations).
+$p \in \{0.05, 0.25, 0.5\}$. 
 Our second series of simulations changed dynamics to the Ricker-logistic
@@
 of 0.25, and simulated low, medium, and high values of equilibrium
-abundance and maximum growth rate, each with 1000 simulations (Table 1;
+abundance and maximum growth rate, (Table A1;
 9 total combinations). Our third series of simulations was based
@@
 parameters the same as the Ricker-logistic model (with $r$ = 0.05 and $K$ = 10) and
-simulated low, medium, and high values of immigration rate with 1000
-simulation each (Table 1).
+simulated low, medium, and high values of immigration rate (Table A1).
 
@@
 initial abundance (Poisson) and dynamics models as were simulated.
-When run in a maximum likelihood framework, these models require a
-a finite upper bound $N_{max}$,
-\citep{royle:2004biom,dail_madsen:2011};
-%(Royle 2004, Dail and Madsen 2011);
-we used 200.  We report bias of estimates, root
+We report bias of estimates, root
 mean squared error, and coverage (percentage of 95\% confidence
@@
 
-\begin{comment}
-Ideas:
-I. Test importance of random effects modeling, and other
-differences between JAGS and unmarked
-II.	Model with low p values, or a range of p values
-III. Ability to distinguish correct initial abundance and
-dynamics models
-IV.	When can you detect DD if it exists (power)?
-V.	Building on the last two: what if your model set doesn't
-include the correct DD model? When will the wrong DD model be
-favored
-over the wrong DID model?
-\end{comment}
+\subsection*{Robustness Simulation Study}
+
+We tested our models against violations of the Poisson process assumption for
+annual dynamics.  We ran three alternative scenarios: 1) a geometric-recruitment model with
+$\omega = \gamma = 0.5$; 2) a Leslie matrix model with six age classes; and 3) 
+a Gompertz-logistic model with two non-overlapping generations per year (sampling occasion).   
+The first two scenarios are underdispersed compared to the Poisson and the third is overdispersed.
+In the first two cases, we estimated parameters with both the exponential and geometric-recruitment
+models using the \textbf{R} package \texttt{unmarked}.  We compared the bias,
+root mean squared error, and coverage of the model estimates as well as their Akaike's Information
+Criterion (AIC) values.
+For the third scenario, we estimated parameters with the Gompertz-logistic model with 
+one generation per year using both maximum likelihood and MCMC approaches, as well
+as allowing for two generations per year with MCMC.  We compared the bias, root mean
+squared error, and coverage of all three model estimates.  Details of the robustness simulation
+study can be found in Appendix A.
 
@@
 We applied our model extensions to North American Breeding Bird Survey
-(BBS) data collected from 1966-2010 in %the bordering US states
+(BBS) data collected from 1966-2010 in 
 Maryland and Virginia. For our focal species, we selected the
 ovenbird (\textit{Seiurus aurocapilla}), an abundant and widespread,
-forest-breeding migratory songbird with a stable or increasing trend
+forest- breeding migratory songbird with a stable or increasing trend
 in the region \citep{porneluzi_etal:2011}.
-%and golden-winged warblers (\textit{Vermivora chrysoptera}), which
-%only breeds in the western parts of these states and has been declining in the
-%region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?
-
-%GENERAL ABOUT BBS DATA.
+
 The BBS is an annual roadside survey implemented by trained
@@
 3-minute point counts with 400 m radii, 0.8 km apart 
-%from each other 
 along a
@@
 each route on the Beaufort scale as a categorical predictor. 
-%BBS volunteers record wind conditions
-%at the beginning and end of each route on the Beaufort Scale
-%\citep[start and end wind]{robbins_etal:1986}.
-%%(start and end wind; Robbins et al. 1986)(start and end wind;0-9; cite).
-%When start or end wind was not recorded we imputed those values with
-%the start or end mean. We took the mean of start and end wind for
-%each route and put this average wind scale value into four
-%categories: 0 $\leq$ wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind
-%$\geq$ 3 (maximum of 3.5).
 Following \citet{link_sauer:2002},
-%Link and Sauer (2002),
 we also included the first time an observer ran a route as a predictor variable for detection
@@
 zero-inflated Poisson) with exponential dynamics
-(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
-Criterion (AIC) model from that set to test three additional
+(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum AIC model from that set to test three additional
 models for p: wind, first, and wind + first. We selected the minimum
@@
 the previous three with immigration, and exponential + immigration.
-%geometric-recruitment +
-%immigration, exponential + immigration, Ricker-logistic +
-%immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
-%minimum temperature for June and July on all dynamics parameters from
-%the minimum AIC dynamics model.
-%As above, t
 These models require a a finite upper bound ($N_{max}$) to integrate
 over when run in a maximum likelihood
-framework; we used 600. % for ovenbirds and 350 for golden-winged warblers.
+framework; we used 600.
 
@@
 $\iota  \sim \mathrm{U}(0, 15)$, and logit-linear $p$ coefficients $\sim \mathrm{N}(0, 100)$).  
-% JEFF: We need to be specific about the priors
 We added random observer 
 effects and environmental stochasticity, modeled as lognormal
-variation in regional population growth rate, Eq.~\ref{eq:nuRand}. % and, where appropriate, zero-inflated dynamics.  
+variation in regional population growth rate, Eq.~\ref{eq:nuRand}.
 We tested for lack of convergence using
@@
 year specific abundance estimates.  Because of the memory requirements involved,
-we used %obtained these estimates from 
-a single chain.  %We present these estimates for ovenbirds only.
+we used a single chain.
 
@@
 
-\subsection*{Simulation Study}
+\subsection*{Proof of Concept Simulation Study}
 
 Estimator bias and error were generally low with the exponential model, and 
-coverage was nominal (93.4\% - 97.0\%; Fig.~\ref{fig:exp_hists}; Table S2.1 in Appendix 2). 
+coverage was nominal (93.4\% - 97.0\%; Fig.~\ref{fig:exp_hists}; Table A1 in Appendix A). 
 However, estimates of $r$ were 
@@
 -151\% and -101\% for non-zero true $r$) when the true value of
-$\Lambda$ was low. % (1). 
-Simulations of the Ricker-logistic model %also 
-generally performed well (Fig. S2.1 and Table S2.2 in Appendix 2), except when the true value of $r$ was low (0.005). In this case the 
+$\Lambda$ was low. 
+Simulations of the Ricker-logistic model 
+generally performed well (Fig. A1 and Table A2 in Appendix A), except when the true value of $r$ was low (0.005). In this case the 
 estimates of $r$ and $K$ were inaccurate (relative RMSE between 87.4\% and 2091.8\%),
@@
 91.9\% and 96.4\%.  All three cases simulated for the Ricker-logistic + immigration model 
-performed fairly well (Fig. S2.2 and Table S2.3 in Appendix 2; relative RMSE between 7.2\% 
+performed fairly well (Fig. A2 and Table A3 in Appendix A; relative RMSE between 7.2\% 
 and 62.9\%, bias between -3.5\% and 3.3\%, and coverage between
@@
 
+\subsection*{Robustness Simulation Study}
+In simulations of the geometric-recruitment model estimated with exponential model,
+estimates of $\Lambda$ were biased high %(2.71, relative bias 27.1\%) 
+and $p$ biased low (Table A5). % (-0.05, relative bias -21.1\%).  
+%As a result, RMSE for these parameters were high %(2.88 and 0.05, respectively) 
+%and coverages low. % (13.0\% and 3.9\%).
+However, estimates of $r$ had low bias %(-0.0005) 
+and RMSE, %(0.005)
+with nominal coverage. % (95.7\%).  
+The estimates from the geometric-recruitment model for $\Lambda$ and $p$ had low bias, but the estimates
+for $\gamma$ and $\omega$ were somewhat biased and imprecise (Table A5 and Fig. A3).
+The geometric- recruitment model had lower
+AIC 71.3\% of the time; when the geometric-recruitment model was favored, the
+mean $\Delta$AIC was 6.0 and when the exponential model was favored
+the mean was 1.4.  %Results from the Leslie matrix model simulations were similar, but more dramatic.
+For the Leslie matrix simulations, bias in estimates of $\Lambda$ and $p$ from both the exponential and geometric-recruitment models
+were high, but the exponential biases were higher (Table A6 and Fig. A3).  
+The geometric-recruitment model was favored in 100\% of the simulations, 
+with a mean $\Delta$AIC of 89.0.
+
+The Gompertz-logistic model with a single generation per year fitted with both maximum likelihood and
+MCMC using data simulated from a Gompertz-logistic model with two generations per year underestimated 
+$\Lambda$  and $K$ and %(bias -1.61 and -1.68, respectively) (bias -8.73 and -5.42)  (bias 0.10 and 0.11)
+overestimated $p$ (Table A7 and Figure A4).  As a result, RMSEs for these parameters 
+were all high and coverages all low, except the coverage of $K$ by the maximum 
+likelihood model was 100\%.  However, the model with two generations per year 
+estimated parameters well, with minimal bias and RMSE and nominal coverage.
+
 \subsection*{Analysis of Breeding Bird Survey Data}
@@
 initial abundance over the Poisson and zero-inflated Poisson
-(Table S3.1A in Appendix 3). %Even compared to the Poisson, there was
-%no evidence to support a zero-inflation factor for this species.
+(Table B1A in Appendix B). 
 The best supported model for $p$ included additive effects of wind speed
-and first run of a route by an observer (Table S3.1B, model B.1). First run 
+and first run of a route by an observer (Table B1B, model B.1). First run 
 and increasing wind speeds both decreased $p$. All dynamics models 
 with immigration were better supported than models without immigration 
-(Table S3.1C); the best supported of these was the Ricker-logistic + immigration 
-(see Appendix 3, Fig. S3.1 for estimates).
-%This model estimated $\Lambda$ at 33.4 $\pm$ 4.7, $r$ at 0.026 $\pm$ 0.006, 
-%$K$ at  56.9 $\pm$ 9.4, and $\iota$ at 0.283 $\pm$ 0.043.
-%The best supported model of those without immigration was the
-%Gompertz-logistic, and the geometric-recruitment and the constant model had the least
-%support.
-%There was little support for an effect of minimum June and July
-% temperature on dynamics parameters of the Ricker + Immigration model
-%(Table 2A, models A.17 - A.18).
-%For both species, 
-Constant-recruitment dynamics models estimated unrealisticly high
-survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05), whereas % for ovenbird and
-%0.935 $\pm$ 0.011 for
-%golden-winged warbler). 
+(Table B1C); the best supported of these was the Ricker-logistic + immigration 
+(see Appendix B, Fig. B1 for estimates).
+Constant- recruitment dynamics models estimated unrealistically high
+survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05), whereas 
 the geometric-recruitment and geometric-recruitment + immigration
-models estimated nonviably low survival probabilities %for ovenbird 
+models estimated nonviably low survival probabilities 
 ($\omega$ =
 0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively).
-%but not for golden-winged warbler ($\omega$ = 0.715 $\pm$ 0.288 for both models).
-
-Estimates for the top ranked model %for ovenbird
-%(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
+
+Estimates for the top ranked model 
 were similar when
 run in the Bayesian framework, except that the estimate of $r$
-more than halved 
-%% JEFF: How do we explain that????
-(Appendix 3, Fig. S3.1). %to 0.011 $\pm$ 0.005 [for
-%Bayesian model estimates we present mean and SD]). 
+more than halved (Appendix B, Fig. B1). 
 When random observer effects
 were added, estimates for $\Lambda$ and $K$ increased dramatically
-%(to 42.5 $\pm$ 7.3 and 101.8 $\pm$ 21.1,
-%respectively),
-and the estimate for r was intermediate. The %(0.018 $\pm$ 0.006)
-estimate of the intercept for $p$ dropped. % from -1.5 $\pm$
-%0.1 to -2.0 $\pm$ 0.1 (on the logit scale), and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
-When environmental stochasticity was added, the estimate for $K$ increased. %(to 112 $\pm$ 58.8) and $r$ was
-%again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
-%on estimate of $p$ or $\Lambda$. 
+and the estimate for r was intermediate. The 
+estimate of the intercept for $p$ dropped. 
+When environmental stochasticity was added, the estimate for $K$ increased. 
 The highest estimates of $\Lambda$ and $K$ came 
-from the model that included environmental stochasticity and random observer effects. %(43.8 $\pm$
-%7.9 and 141.5 $\pm$ 69.5, respectively).  
+from the model that included environmental stochasticity and random observer effects. 
 Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
@@
 more evenly distributed by 2010.
-%These aren't up to date, and it doesn't look like people use CV as a measure of 
-%uncertainty in MCMC estimates anyway.
-%Coefficient of variation (CV) for route and year specific
-%abundance estimates varied between 0.11 and 2.31 (mean = 0.46).  By comparison, CV
-%for the estimate of mean initial abundance ($\Lambda$) was 0.15.
-%As expected, uncertainty in estimates of route and year specific abundance was high.
-
-%There was slightly more support for the ZIP distribution for initial
-%abundance of golden-winged warbler than for the negative binomial; both were strongly
-%supported over the Poisson (Table 3A). The best
-%supported model for $p$ was an effect of first run (Table 3B, model
-%B.1). The best supported dynamics model was exponential
-%(Table 3C, model C.1). %; the estimate of r from this model was -0.058
-%$\pm$ 0.017 (SE). There was no support for immigration added to the
-%exponential or geometric-recruitment models (Table 3C, models C.4 and
-%C.6).
-%However, the Gompertz-logistic + immigration and Ricker-logistic + immigration
-%models were supported over the corresponding models without immigration
-%(Table 3C, models C.4, C.5, C.7, and C.8), despite low estimates
-%of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$
-%0.002, respectively).  
-%There was considerable support for an effect of
-%average minimum June and July temperature on r in the population
-%growth model (Table 1B, model B.17), with a negative slope.
-%Because of the closeness of rankings for the ZIP and negative binomial
-%distributions for golden-winged warbler, we also ran the subsequent models with the
-%negative binomial initial abundance; rankings of subsequent
-%models were similar. %, but with a $p$ covariate (first) the negative
-%binomial distribution outranked the ZIP.
-
-%Estimates for the top ranked model for golden-winged warbler
-%%(ZIP[Λ(.)α(.)]Exponential[r(.)]p(1st))
-%were similar when run in the Bayesian framework. When random
-%observer effects were added, the estimate for $\Lambda$ increased dramatically,
-%and the estimate of the intercept for $p$ dropped.
-%%(from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
-%%for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
-%%estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. 
-%Adding environmental stochasticity also increased the estimate of $\Lambda$ 
-%%(to 42.9 $\pm$ 17.9) 
-%and the model with both random effects had the highest estimate of $\Lambda$.
-%% (74.6 $\pm$ 32.9). 
-%Adding zero-inflation to the dynamics (as well as
-%the initial abundance) has no effect for exponential models, because populations
-%that started at 0 would stay at 0 in any case. Therefore, we tested
-%zero-inflated dynamics on the Exponential + Immigration model. Adding zero-inflation to the dynamics
-%increased the estimate of $\iota$, but had only minor % from 0 to 0.091 $\pm$ 0.049
-%effects on the other parameters. Gelman and Rubin diagnostics and visual
-%examination of the chain trajectory and density plots provided evidence for a lack of convergence in
-%estimates of $\Lambda$ for all Bayesian models ran, and for most parameters when random observer
-%effects were included.
-
-%Estimated detection probabilities were generally low. The estimated probability
-%of detecting an ovenbird in the top model (Table 2, model C.1) varied between 0.157
-%$\pm$ 0.014 and 0.184 $\pm$ 0.015, depending on wind speed and whether the
-%observer had run that route before. The estimated probability of detecting a
-%golden-winged warbler went from 0.013 $\pm$ 0.006 for first time observers to 0.028 $\pm$ 0.008 for
-%repeat observers for the top model (Table 3, model C.1). Accounting for random
-%observer effects decreased average detection probability for both species.
 
@@
 
-\begin{comment}
-  The model (1) recognizes
-the discrete nature of count data and the underlying true
-abundance
-state, (2) includes a realistic observation model with explicitdetection probability parameters (3) has estimable parameters,including those describing the initial state at time $t=1$, and
-(4)
-allows for inference about both temporal and spatial variation
-in
-  abundance.
-\end{comment}
-
-Our work %has highlighted the limitations of classical state-space
-%models as applied to ecological time series data, and we have
-%demonstrated 
-demonstrates how a class of open population $N$-mixture models
+Our work demonstrates how a class of open population $N$-mixture models
 proposed by \citet{dail_madsen:2011} overcomes 
-%these limitations. 
 many of the limitations of classical state-space models as applied to
@@
 We extended their model in several important
-ways to accomodate common features of ecological
+ways to accommodate common features of ecological
 datasets, including sparse counts, zero-inflation, environmental
 stochasticity, and random variation in observation error.
-%We made four important developments of the open population
-%N-mixture model proposed by \citet{dail_madsen:2011} that make the
-%model more applicable to long-term data such as is collected
-%by the North American Breeding Bird Survey.
-%First, we have
-%reconciled the objectives of traditional state-space models
-%with open population $N$-mixture models by illustrating how
-%classical models of population dynamics can be embedded within
-%the framework. Second, we have demonstrated methods of
-%accounting for zero-inflation in the time series. Third, we
-%have illustrated how additional random effects such as
-%observer-specific detection probabilities can be
-%accommodated. Fourth, we presented a Bayesian analysis of the
-%model, which, makes it possible to incorporate prior information when
-%available, and in some cases, facilitates parameter estimation.
 We also demonstrated how many of the objectives of conventional
@@
 population growth models can be embedded in the DM model.
+
 Although we view this as an important connection to make between the two
 different classes of state-space models, we believe that
-%In some ways,
 such population growth models are more phenomenological
@@
 lack of support for the mechanistic models relative to the population
-growth models.
-
-One strategy for adhereing to the original formulation of the
+growth models.  A separate analysis of BBS data for eleven other forest passerines
+found similar poor support and unrealistic estimates for the mechanistic 
+models (J. Hostetler, unpublished analyses), suggesting that some process, possibly
+individual heterogeneity \citep{vindenes_etal:2008}, is increasing variances
+above the levels expected from demographic stochasticity alone.
+
+One strategy for adhering to the original formulation of the
 model that included survival and recruitment parameters would be to
@@
 
-%In contrast, recent work has sought to use state-space
-%models for inference about population processes such as mortality and
-%recruitment. To do so, methods have been developed to combine multiple
-%sources of information, such as counts and mark-recapture data
-%\citep{besbeas_etal:2002, buckland_etal:2004,
-%  schaub_etal:2007}.
-
 Integrated population models may be viewed as the gold standard in
@@
 are produced by many of the largest monitoring programs in the
-world, such as the BBS. %North American Breeding Bird Survey
-%\citep[BBS;][]{robbins_etal:1986}.
+world, such as the BBS. 
 In the absence of direct information about demographic parameters, and
@@
 may be of interest.  We have shown a few approaches for incorporating some of these
-(environmental stochasticity and random observer effects).  %Other approaches to modeling
-%each of these may also be worth trying.  
+(environmental stochasticity and random observer effects).  
 Although incorporating these random effects can increase
@@
 ecologists lacking skills in computer programming and advanced
-statistics. Appendix 1 includes code to
+statistics. Supplement 1 includes code to
 simulate and fit many of the extensions discussed in this paper.
@@
 extensions that we believe would be computationally prohibitive in many
-cases. Nonetheless, we will discuss one---spatially-explict models of
+cases. Nonetheless, we will discuss one---spatially-explicit models of
 immigration. 
@@
 Bayesian framework.
-% Could use a better reference than Hastings 1991 for the total abundance 
-% immigration models, but haven't been able to find one.  And haven't found
-% any references for nearby site immigration (maybe that's because it's not a great idea).
-\begin{comment}
-
-For data sources such as the BBS, where stops are nested within
-routes, spatially hierarchical versions of these models could be
-the best way to model data from individual stops over many
-routes. These models would allow the analyst to account for the
-potential spatial dependence of stops within a route, by having
-the route as another random effect.
-
-Other models of variability in growth or recruitment (such as
-negative binomial or generalized Poisson) could be informative.
-Stochasticity (represented as variation in the numbers of
-animals occurring in a site and year around what the
-deterministic dynamics predict) could be over or underdispersed,
-compared to the Poisson distribution. Alternatively,
-stochasticity could possibly be applied directly to population
-model parameters (such as r and K), instead of numbers of
-animals.
-\end{comment}
-
-%[paragraph on spatially-explicit dispersal]
-%[Other possible extensions we could mention:
-%Spatially hierarchical model (so can use individual stops)
-%Other models of variability in growth or recruitment (such as negative binomial)
-%Add density dependence to survival and recruitment models]
-%[Discuss causes of low p values (and other BBS results) here or in BBS
-%results?]
-
-Our models generally estimated parameters well in analysis of simulated data.
+
+Our models generally estimated parameters well in analysis of data simulated with the same model.
 Two exceptions were the estimates of instantaneous growth rate in exponential model when the 
@@
 The failure of the first case was likely due to a large number of sites starting at abundance 0 or reaching it early in the simulation.  
-%[Richard, maybe come up with a summary statistic for this based on the simulation N values?]
 Because a site that hits abundance 0 in this simulation stayed at 0, this provided little information with
@@
 
-%We have presented a test case for these models using data from
-%the BBS for ovenbirds. We demonstrate that these models can be
-%used to model and estimate initial abundance, detection
-%probability, and density-independent and dependent functions for
-%population dynamics. We tested these models in both the
-%frequentist and Bayesian frameworks and found strengths and
-%weaknesses of each approach. The Bayesian analyses permit more
-%flexibility and allow the incorporation of random observer
-%effects, environmental stochasticity, and zero-inflated dynamics. 
-%Our results suggest ignoring
-%variation between observers can lead to underestimation of
-%initial abundance and equilibrium population sizes. On the other
-%hand, our Bayesian analyses did have some convergence problems.
-%%particularly for the species (golden-winged warbler) with sparser data. 
-%It's likely that poor model fits are related to the low estimates of
-%detection probability, which in turn are probably caused by the
-%large 400 m radius used for BBS point counts. Again, we believe
-%that direct information on detectability (such as repeated
-%visits within a season) would greatly improve the fit of these
-%models in both frameworks.  We also were able to obtain year
-%and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
-%although with generally lower precision than more global estimates.  These sort of estimates
-%may be more reliable when abundance and dynamics are modeled 
-%as functions of habitat covariates, which we did not do here.
-\begin{comment}
-[Alternatively, could put discussions of our results where
-they fit in by subject area, rather than separately. For
-instance, could discuss BBS population dynamics model selection
-in population models paragraph, BBS Bayesian vs.
-frequentist strengths and weakness in the Bayesian paragraph, 
-and golden-winged warbler ZIP results in ZIP
-paragraph.]
-\end{comment}
+Our models generally did not estimate parameters well in simulations when data were underdispersed
+or overdispersed compared to the model assumptions.  However, we present models for
+both Poisson distributed and overdispersed data, and the original DM model and the 
+stage-structured model of \citet{zipkin_etal:2014} are available for underdispersed data.  AIC model
+selection did a good job of picking the most appropriate distribution.  Density-dependent models
+for underdispersed data still remain to be developed.
 
@@
 example,
-%it is possible to test hypotheses about temporal and spatial
-%population regulation. Furthermore, the Bayesian approach is very
-%useful in that it can be used to combine multiple sources of data to
-%develop mechanistic models of population dynamics. Thus one can test
-it allows for inference %hypotheses
+it allows for inference 
 about the effects of climate change on either explicit
 demographic parameters or in derived parameters such as
-population growth rate. Furthermore, %under the Bayesian approach, population
+population growth rate. Furthermore, 
 in addition to assessing past climate change on population
@@
 possible. Thus, we expect this modeling methodology can be used
-to make quantiative assessments of species vulnerability to climate
+to make quantitative assessments of species vulnerability to climate
 change and other important human-induced drivers of population
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 
@@
 
-\bibliography{Dail_Mad_ext}
+\bibliography{improved_state_space}
+
+\begin{center}
+\noindent{}\textbf{Appendix A}: Simulation Details and Results.
+
+\noindent{}\textbf{Appendix B}: BBS Results.
+
+\noindent{}\textbf{Supplement 1}: Code Samples.
+\end{center}
 
 \newpage
-
-% Tables
-
-\begin{table}[t]
-  \centering
-\caption{Changed parameter values, by series of simulations.  We
-simulated 1000 sets of data each combination of parameter values
-for 100 sites over 40 years.  We assumed that initial abundance
-($\Lambda$) was Poisson distributed.  For the exponential 
-simulations we included all combinations of low, medium, and
-high $\Lambda$, growth rate ($r$), and detection probability ($p$).
-For the Ricker-logistic model simulations we used $\Lambda$ = 10 and $p$ = 0.25, and
-simulated low, medium, and high values of $r$ and equilibrium density ($K$).
-For the Ricker-logistic + immigration dynamics model we fixed all parameters the
-same as the Ricker-logistic model (with $r$ = 0.05 and $K$ = 10) and
-simulated low, medium, and high values of immigration rate ($\iota$).}  
-\begin{tabular}{lcccccccc}
-    \hline
-    & \multicolumn{3}{c}{Exponential Growth} && \multicolumn{2}{c}{Ricker-logistic} &&
-    Ricker-logistic + Immigration \\
-    \cline{2-4}     \cline{6-7}    \cline{9-9}
-% & \multicolumn{3}{c}{\rule{25mm}{.1pt}} & \rule{5cm}{.5pt} & \multicolumn{2}{c}{\rule{20mm}{.1pt}} \\
-& $\Lambda$ & $r$ & $p$ && $r$  & $K$ && $\iota$  \\    
-\hline
-    Low	        &1	&-0.01	&0.05	&&0.005	 &5	&&0.005  \\
-    Med	        &5	&0	&0.25	&&0.05	&10	&&0.05   \\
-    High		 &10 &0.005	&0.5	&&0.1	&20	&&0.5    \\
-    \hline
-  \end{tabular}
-\end{table}
-
-%\vfill
-%\clearpage
-\newpage
-
-% Perhaps stack the species
-\begin{comment}
-\begin{sidewaystable}
-  \centering
-  \small
-\caption{Model selection table for ovenbirds and golden-winged
-warblers in Maryland and Virginia, 1966-2010. We present model
-name
-and number, number of parameters (Par.), and difference in
-Akaike's
-information criterion between each model and the top model of
-that
-set ($\Delta$AIC).  For each species, the first section comparesmodels for initial abundance, the second for detection
-probability,
-    and the third for dynamics.}
-  \begin{tabular}[h]{lcclcc}
-\hline
-A. Ovenbirds	&&&		B. Golden-Winged Warblers \\
-\hline
-Model	&Par.	&$\Delta$AIC	&Model	&Par.	&$\Delta$AIC \\
-A.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0
-&B.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&0 \\
-A.2. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&1262.7
-&B.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0.4 \\A.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4
-&1264.7	&B.3. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&37.8 \\
-A.4. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
-&0	&B.4. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)	&5
-&0 \\
-A.5. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind)	&7	&0.9
-&B.5. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&3.5 \\
-A.6. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(1st)	&5	&5.0
-&B.6. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
-&5.9 \\
-A.7. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&6.4
-&B.7. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind) &7 &9.3
-\\
-A.8.
-NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
-&10	&0	&B.8. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)
-&5	&0 \\
-A.9.
-NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
-&10	&8.4
-&B.9.
-ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st)
-&6	&1.7 \\
-A.10.
-NB[$\Lambda$(.)$\alpha$(.)]Growth+Immigration[$r$(.)$\iota$(.)]$p$(wind+1st)
-&9	&36.5
-&B.10.
-ZIP[$\Lambda$(.)$\psi$(.)]Growth+Immigration[$r$(.)$\iota$(.)]$p$(1st)
-&6	&2.0 \\
-A.11.
-NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
-&10	&38.6
-&B.11.
-ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
-&7	&2.5 \\
-A.12.
-NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(wind+1st)
-&9	&192.8
-&B.12. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(1st)
-&6
-&3.0 \\
-A.13.
-NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(wind+1st)
-&9	&195.1
-&B.13.
-ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
-&7	&3.7 \\
-A.14. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
-&271.3
-&B.14.
-ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
-&7	&4.2 \\
-A.15.
-NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
-&9	&273.7
-&B.15. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(1st)	&7&5.2 \\
-A.16.
-NB[$\Lambda$(.)$\alpha$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
-&9	&1856.7
-&B.16.
-ZIP[$\Lambda$(.)$\psi$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st)
-&6	&12.5 \\
-\hline
-\end{tabular}
-\end{sidewaystable}
-\end{comment}
-
-% How about separate tables?  Also updated the Gompertz-logistic results.
-%\begin{table}
-%  \centering
-%  \small
-%\caption{Model selection table for ovenbirds in Maryland and Virginia,
-%    1966-2010.  We present model name and number, number of 
-%parameters (Par.), and difference in Akaike's
-%information criterion between each model and the top model of
-%that set ($\Delta$AIC).  The first section compares
-%models for initial abundance, the second for detection
-%probability, and the third for dynamics.}
-%  \begin{tabular}[h]{lcc}
-%\hline
-%Model	&Par.	&$\Delta$AIC	\\
-%\hline
-%A. Initial Abundance && \\
-%A.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0\\
-%A.2. P[$\Lambda$(.)]Exponential[$r$(.)]$p$(.)	&3	&1262.7\\
-%A.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4 &1264.7\\
-%\hline
-%B. Detection Probability && \\
-%B.1. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &0	\\
-%B.2. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind) &7 &0.9\\
-%B.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(1st)	&5	&5.0
-%\\B.4. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&6.4\\
-%\hline
-%C. Dynamics && \\
-%C.1. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&0	\\
-%C.2. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&8.4 \\
-%C.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(wind+1st) &9	&36.5\\
-%C.4. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st) &10	&38.6\\
-%C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&192.8\\
-%C.6. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&195.1\\
-%C.7. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &271.3\\
-%C.8. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&273.7\\
-%C.9. NB[$\Lambda$(.)$\alpha$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(wind+1st) &9	&1856.7\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-%\vfill
-%%\clearpage
-%\newpage
-
-%\begin{table}
-%  \centering
-%  \small
-%  \caption{Model selection table for golden-winged
-%warblers in Maryland and Virginia, 1966-2010. We present model
-%name and number, number of parameters (Par.), and difference in
-%Akaike's information criterion between each model and the top model of
-%that set ($\Delta$AIC).  The first section compares
-%models for initial abundance, the second for detection
-%probability, and the third for dynamics.}
-%  \begin{tabular}[h]{lcc}
-%\hline
-%Model	&Par.	&$\Delta$AIC\\
-%\hline
-%A. Initial Abundance && \\
-%A.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4	&0 \\
-%A.2. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(.)	&4	&0.4 \\
-%A.3. P[$\Lambda$(.)]Exponential[$r$(.)]$p$(.)	&3	&37.8 \\
-%\hline
-%B. Detection Probability && \\
-%B.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(1st)	&5 &0 \\
-%B.2. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(.)	&4	&3.5 \\
-%B.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &5.9 \\
-%B.4. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(wind) &7 &9.3\\
-%\hline
-%C. Dynamics && \\
-%C.1. ZIP[$\Lambda$(.)$\psi$(.)]Exponential[$r$(.)]$p$(1st) &5	&0 \\
-%C.2. ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&1.7 \\
-%C.3. ZIP[$\Lambda$(.)$\psi$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(1st) &6	&2.0 \\
-%C.4. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st) &7	&2.5 \\
-%C.5. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(1st)	&6&3.0 \\
-%C.6. ZIP[$\Lambda$(.)$\psi$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st) &7	&3.7 \\
-%C.7. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(1st) &7	&4.2 \\
-%C.8. ZIP[$\Lambda$(.)$\psi$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(1st)	&7 &5.2 \\
-%C.9. ZIP[$\Lambda$(.)$\psi$(.)]Constant-recruitment[$\gamma$(.)$\omega$(.)]$p$(1st) &6	&12.5 \\
-%\hline
-%\end{tabular}
-%\end{table}
 
@@
 data-generating values.
-
-%\noindent Figure ~\ref{fig:rick_hists}. 
-%
-%\noindent Figure ~\ref{fig:ricki_hists}. Histograms of 1000 parameter estimates for each of 3
-%simulation cases with Ricker-logistic + immigration dynamics. The vertical lines are the 
-%data-generating values.
 

### 2014-06-25 supp/bbs_estimates_appendix.tex
Response to reviewer's comments

@@
 \begin{document}
-\title{Appendix 3: BBS Results}
+\title{Appendix B: BBS Results}
 \maketitle
-\renewcommand*\thetable{S3.\arabic{table}}
-\renewcommand*\thefigure{S3.\arabic{figure}}
+\renewcommand*\thetable{B\arabic{table}}
+\renewcommand*\thefigure{B\arabic{figure}}
 \begin{table}[htb]
@@
 
-[Andy and Richard, which do you prefer for parameter estimates?  
-Table or figure?  Either one may need a bit of work still.]
-% latex table generated in R 3.0.0 by xtable 1.7-1 package
-% Wed Sep 18 14:26:41 2013
-\begin{table}[ht]
-\centering
-  \caption{Parameter estimates for ovenbirds in Maryland and Virginia from BBS data, 1966-2010.  
-MLE estimates come from the NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) 
-model run in the \textbf{R} package \texttt{unmarked}; base estimates from the same model run in \textbf{JAGS};
-obs estimates from the base model with random observer effects added; and ES estimates from the base
-model with regional environmental stochasticity added.  Detectability parameters (intercept: $p_{0}$, 
-effect of wind speed 1: $p_{1}$, effect of wind speed 2: $p_{2}$, effect of wind speed 3 or higher: $p_{3}$, 
-effect of first run of a route by an observer: $p_{1st}$, and random observer effect: $\sigma_{p}$) are on the 
-logit scale.}
-\begin{tabular}{rllllll}
-  \hline
- & Parameter & MLE & base & obs & ES & ES and obs \\ 
-  \hline
-1 & $\Lambda$ & 33.4 $\pm$ 4.7 & 31.1 $\pm$ 5.2 & 42.5 $\pm$ 7.3 & 33.2 $\pm$ 5.8 & 43.8 $\pm$ 7.9 \\ 
-  2 & $\alpha$ & 0.654 $\pm$ 0.087 & 0.624 $\pm$ 0.108 & 0.730 $\pm$ 0.119 & 0.649 $\pm$ 0.110 & 0.753 $\pm$ 0.124 \\ 
-  3 & $r$ & 0.026 $\pm$ 0.006 & 0.011 $\pm$ 0.005 & 0.018 $\pm$ 0.006 & 0.020 $\pm$ 0.010 & 0.023 $\pm$ 0.010 \\ 
-  4 & $K$ & 56.9 $\pm$ 9.4 & 55.8 $\pm$ 18.2 & 101.8 $\pm$ 21.1 & 112.0 $\pm$ 58.8 & 141.5 $\pm$ 69.5 \\ 
-  5 & $\iota$ & 0.283 $\pm$ 0.043 & 0.333 $\pm$ 0.043 & 0.306 $\pm$ 0.051 & 0.329 $\pm$ 0.046 & 0.303 $\pm$ 0.052 \\ 
-  6 & $p_{0}$ & -1.492 $\pm$ 0.100 & -1.476 $\pm$ 0.097 & -1.994 $\pm$ 0.116 & -1.571 $\pm$ 0.099 & -2.056 $\pm$ 0.124 \\ 
-  7 & $p_{1}$ & -0.012 $\pm$ 0.021 & -0.011 $\pm$ 0.021 & -0.014 $\pm$ 0.021 & -0.013 $\pm$ 0.021 & -0.015 $\pm$ 0.021 \\ 
-  8 & $p_{2}$ & -0.075 $\pm$ 0.027 & -0.071 $\pm$ 0.027 & -0.056 $\pm$ 0.026 & -0.069 $\pm$ 0.027 & -0.055 $\pm$ 0.026 \\ 
-  9 & $p_{3}$ & -0.133 $\pm$ 0.049 & -0.135 $\pm$ 0.049 & -0.145 $\pm$ 0.048 & -0.141 $\pm$ 0.049 & -0.144 $\pm$ 0.048 \\ 
-  10 & $p_{1st}$ & -0.058 $\pm$ 0.028 & -0.047 $\pm$ 0.027 & -0.075 $\pm$ 0.029 & -0.045 $\pm$ 0.028 & -0.075 $\pm$ 0.029 \\ 
-  31 & $\sigma_{p}$ & - & - & 0.353 $\pm$ 0.034 & - & 0.349 $\pm$ 0.034 \\ 
-  42 & $\sigma_{\nu}$ & - & - & - & 0.074 $\pm$ 0.014 & 0.064 $\pm$ 0.014 \\ 
-   \hline
-\end{tabular}
-\end{table}
+%[Andy and Richard, which do you prefer for parameter estimates?  
+%Table or figure?  Either one may need a bit of work still.]
+%% latex table generated in R 3.0.0 by xtable 1.7-1 package
+%% Wed Sep 18 14:26:41 2013
+%\begin{table}[ht]
+%\centering
+%  \caption{Parameter estimates for ovenbirds in Maryland and Virginia from BBS data, 1966-2010.  
+%MLE estimates come from the NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) 
+%model run in the \textbf{R} package \texttt{unmarked}; base estimates from the same model run in \textbf{JAGS};
+%obs estimates from the base model with random observer effects added; and ES estimates from the base
+%model with regional environmental stochasticity added.  Detectability parameters (intercept: $p_{0}$, 
+%effect of wind speed 1: $p_{1}$, effect of wind speed 2: $p_{2}$, effect of wind speed 3 or higher: $p_{3}$, 
+%effect of first run of a route by an observer: $p_{1st}$, and random observer effect: $\sigma_{p}$) are on the 
+%logit scale.}
+%\begin{tabular}{rllllll}
+%  \hline
+% & Parameter & MLE & base & obs & ES & ES and obs \\ 
+%  \hline
+%1 & $\Lambda$ & 33.4 $\pm$ 4.7 & 31.1 $\pm$ 5.2 & 42.5 $\pm$ 7.3 & 33.2 $\pm$ 5.8 & 43.8 $\pm$ 7.9 \\ 
+%  2 & $\alpha$ & 0.654 $\pm$ 0.087 & 0.624 $\pm$ 0.108 & 0.730 $\pm$ 0.119 & 0.649 $\pm$ 0.110 & 0.753 $\pm$ 0.124 \\ 
+%  3 & $r$ & 0.026 $\pm$ 0.006 & 0.011 $\pm$ 0.005 & 0.018 $\pm$ 0.006 & 0.020 $\pm$ 0.010 & 0.023 $\pm$ 0.010 \\ 
+%  4 & $K$ & 56.9 $\pm$ 9.4 & 55.8 $\pm$ 18.2 & 101.8 $\pm$ 21.1 & 112.0 $\pm$ 58.8 & 141.5 $\pm$ 69.5 \\ 
+%  5 & $\iota$ & 0.283 $\pm$ 0.043 & 0.333 $\pm$ 0.043 & 0.306 $\pm$ 0.051 & 0.329 $\pm$ 0.046 & 0.303 $\pm$ 0.052 \\ 
+%  6 & $p_{0}$ & -1.492 $\pm$ 0.100 & -1.476 $\pm$ 0.097 & -1.994 $\pm$ 0.116 & -1.571 $\pm$ 0.099 & -2.056 $\pm$ 0.124 \\ 
+%  7 & $p_{1}$ & -0.012 $\pm$ 0.021 & -0.011 $\pm$ 0.021 & -0.014 $\pm$ 0.021 & -0.013 $\pm$ 0.021 & -0.015 $\pm$ 0.021 \\ 
+%  8 & $p_{2}$ & -0.075 $\pm$ 0.027 & -0.071 $\pm$ 0.027 & -0.056 $\pm$ 0.026 & -0.069 $\pm$ 0.027 & -0.055 $\pm$ 0.026 \\ 
+%  9 & $p_{3}$ & -0.133 $\pm$ 0.049 & -0.135 $\pm$ 0.049 & -0.145 $\pm$ 0.048 & -0.141 $\pm$ 0.049 & -0.144 $\pm$ 0.048 \\ 
+%  10 & $p_{1st}$ & -0.058 $\pm$ 0.028 & -0.047 $\pm$ 0.027 & -0.075 $\pm$ 0.029 & -0.045 $\pm$ 0.028 & -0.075 $\pm$ 0.029 \\ 
+%  31 & $\sigma_{p}$ & - & - & 0.353 $\pm$ 0.034 & - & 0.349 $\pm$ 0.034 \\ 
+%  42 & $\sigma_{\nu}$ & - & - & - & 0.074 $\pm$ 0.014 & 0.064 $\pm$ 0.014 \\ 
+%   \hline
+%\end{tabular}
+%\end{table}
 
@@
 effect of wind speed 1: $p_{1}$, effect of wind speed 2: $p_{2}$, effect of wind speed 3 or higher: $p_{3}$, 
-effect of first run of a route by an observer: $p_{1st}$, and random observer effect: $\sigma_{p}$) are on the 
-logit scale.}
+effect of first run of a route by an observer: $p_{1st}$, and random observer SD: $\sigma_{p}$) are on the 
+logit scale. Error bars show SE for MLE estimates and SD for all others.}
 \label{fig:oven_par_est}

### 2014-06-25 supp/sim_tables_appendix.tex
Incorporated comments from Richard plus formatted for Ecology
Response to reviewer's comments

@@
 \documentclass{article}
-\usepackage[landscape,a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
+\usepackage[letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
 \usepackage[pdftex]{hyperref}
 \usepackage{soul}
+\usepackage{graphicx}
+\usepackage{amsmath}
+%\usepackage{rotating}
+\usepackage{pdflscape}
 
 \begin{document}
-\title{Appendix 2: Simulation Results}
+\title{Appendix A: Simulation Details and Results}
+\date{} %remove date
 \maketitle
-\renewcommand*\thetable{S2.\arabic{table}}
+\renewcommand*\thetable{A\arabic{table}}
+\renewcommand*\thefigure{A\arabic{figure}}
+\renewcommand*\theequation{A\arabic{equation}}
+
+\section*{Details}
+\subsection*{Proof of Concept Simulation Study}
+
+We used the \textbf{R} package \texttt{unmarked} to estimate the parameters 
+for each simulation with the same
+initial abundance (Poisson) and dynamics models as were simulated.
+When run in a maximum likelihood framework, these models require a
+a finite upper bound $N_{max}$; we used 200.
+
+\subsection*{Robustness Simulation Study}
+For the geometric-recruitment simulation, we ran 1000 simulations with the following input 
+parameters: $\Lambda = 10$, $\omega = \gamma = 0.5$, and $p = 0.25$.  
+For the maximum likelihood exponential and geometric-recruitment models fit to these data, 
+we used finite upper bound $N_{max} = 200$. 
+For the Leslie matrix simulation, we also ran 1000 simulations with $\Lambda = 10$ and $p = 0.25$,
+and we increased the number of sites to 200 and the number of years to 50.  We used
+six age classes with apparent annual survival probabilities ($S_{i}$) of 0.32, 0.64, 0.68, 0.68, 0.64, and 0.56,
+and fecundities ($m_{i}$) of 0.71, 1, 1.35, 1.7, 2.05, and 2.4.  This can be expressed as the post-breeding
+census Leslie matrix model:
+\begin{equation}
+A = \begin{bmatrix} 0.2272 & 0.64 & 0.918 & 1.156 & 1.312 & 1.344 \\ 
+0.32 & 0 & 0 & 0 & 0 & 0 \\
+0 & 0.64 & 0 & 0 & 0 & 0 \\
+0 & 0 & 0.68 & 0 & 0 & 0 \\
+0 & 0 & 0 & 0.68 & 0 & 0 \\
+0 & 0 & 0 & 0 & 0.64 & 0 \end{bmatrix}
+\label{eq:leslie}
+\end{equation}
+We incorporated demographic stochasticity into this model with binomial distribution for survivals
+(lower sub-diagonal of Eq.~\ref{eq:leslie}) and Poisson draws for fertilities (top row of 
+Eq.~\ref{eq:leslie}).  Counts and initial abundances were based on adults (i.e. all age classes except
+the first).  The starting age distribution was random Poisson with mean at the stable age distribution.
+For the maximum likelihood exponential and geometric-recruitment models fit to these data, 
+we used finite upper bound $N_{max} = 150$.
+
+The third simulation was a Gompertz-logistic model with two non-overlapping generations per year (sampling occasion).   
+We used 200 sites for 25 years, and data-generating values of $\Lambda = 10$, $r = 0.05$, $K = 10$, and $p = 0.5$.
+We estimated parameters with the Gompertz-logistic model with 
+one generation per year using both maximum likelihood and MCMC approaches, as well
+as allowing for two generations per year with MCMC.  
+For the maximum likelihood model, we used finite upper bound $N_{max} = 150$. 
+For the Bayesian framework we used non-informative priors ($\Lambda$ and $K  \sim \mathrm{U}(0, 30)$, $r  \sim \mathrm{U}(0, 5)$,
+and $p  \sim \mathrm{U}(0, 1)$.  
+For model and simulation we ran a single chain. We sampled the MCMC for 20,000 iterations, after 1,000 adaptation and tuning samples.
+
+We compared the bias, root mean
+squared error, and coverage of all three model estimates.  For the models fit with MCMC,
+bias and root mean squared error are based on the relationship between mean estimates
+and true values, and coverage
+based on percentage of 95\% credible intervals for parameters that overlap the true values.
+\newpage
+
+%\section*{Simulation Setup and Results Tables}
+\begin{table}[t]
+  \centering
+\caption{Changed parameter values, by series of simulations.  We
+simulated 1000 sets of data each combination of parameter values
+for 100 sites over 40 years.  We assumed that initial abundance
+($\Lambda$) was Poisson distributed.  For the exponential 
+simulations we included all combinations of low, medium, and
+high $\Lambda$, growth rate ($r$), and detection probability ($p$).
+For the Ricker-logistic model simulations we used $\Lambda$ = 10 and $p$ = 0.25, and
+simulated low, medium, and high values of $r$ and equilibrium density ($K$).
+For the Ricker-logistic + immigration dynamics model we fixed all parameters the
+same as the Ricker-logistic model (with $r$ = 0.05 and $K$ = 10) and
+simulated low, medium, and high values of immigration rate ($\iota$).}  
+\begin{tabular}{lcccccccc}
+    \hline
+    & \multicolumn{3}{c}{Exponential Growth} && \multicolumn{2}{c}{Ricker-logistic} &&
+    Ricker-logistic + Immigration \\
+    \cline{2-4}     \cline{6-7}    \cline{9-9}
+& $\Lambda$ & $r$ & $p$ && $r$  & $K$ && $\iota$  \\    
+\hline
+    Low	        &1	&-0.01	&0.05	&&0.005	 &5	&&0.005  \\
+    Med	        &5	&0	&0.25	&&0.05	&10	&&0.05   \\
+    High		 &10 &0.005	&0.5	&&0.1	&20	&&0.5    \\
+    \hline
+  \end{tabular}
+\end{table}
+
+%\clearpage
+
+\begin{landscape}
+
 \begin{table}[htb]
@@
 %    \hline
-    $\lambda$ & $r$ & $p$ & $\lambda$ & $r$ & $p$ & $\lambda$ & $r$ &
-    $p$ & $\lambda$ & $r$ & $p$ \\
+    $\Lambda$ & $r$ & $p$ & $\Lambda$ & $r$ & $p$ & $\Lambda$ & $r$ &
+    $p$ & $\Lambda$ & $r$ & $p$ \\
     \hline
@@
 \clearpage
+
 \begin{table}
@@
 %    \hline
-    $\lambda$ & $r$ & $K$ & $p$ & $\lambda$ & $r$ & $K$ & $p$ & $\lambda$ & $r$ & $K$ &
-    $p$ & $\lambda$ & $r$ & $K$ & $p$ \\
+    $\Lambda$ & $r$ & $K$ & $p$ & $\Lambda$ & $r$ & $K$ & $p$ & $\Lambda$ & $r$ & $K$ &
+    $p$ & $\Lambda$ & $r$ & $K$ & $p$ \\
     \hline
@@
 %    \hline
-    $\lambda$ & $r$ & $K$ & $\iota$ & $p$ & $\lambda$ & $r$ & $K$ & $\iota$ & $p$ & $\lambda$ & $r$ & $K$ &
-    $\iota$ & $p$ & $\lambda$ & $r$ & $K$ & $\iota$ & $p$ \\
+    $\Lambda$ & $r$ & $K$ & $\iota$ & $p$ & $\Lambda$ & $r$ & $K$ & $\iota$ & $p$ & $\Lambda$ & $r$ & $K$ &
+    $\iota$ & $p$ & $\Lambda$ & $r$ & $K$ & $\iota$ & $p$ \\
     \hline
@@
   \end{tabular}
-  \label{tab:simricker}
+  \label{tab:simricki}
+\end{table}
+\clearpage
+
+\begin{table}[htb]
+  \centering
+  \footnotesize
+  \caption{Parameters of geometric-recruitment simulation (1000 Monte Carlo replicates)
+  and results of exponential and geometric-recruitment models fit to these data.}
+  \begin{tabular}{lccccccccccccccccc}
+    \hline
+    & \multicolumn{5}{c}{Mean} &&
+    \multicolumn{5}{c}{RMSE} &&
+    \multicolumn{5}{c}{Coverage} \\
+    \cline{2-6}     \cline{8-12}    \cline{14-18}
+    Model & $\Lambda$ & $\gamma$ & $\omega$ & $r$\footnotemark[1] & $p$ &&
+    $\Lambda$ & $\gamma$ & $\omega$ & $r$ & $p$ && 
+    $\Lambda$ & $\gamma$ & $\omega$ & $r$ & $p$ \\
+    \hline
+Simulation & 
+10	&0.5	&0.5	&0	&0.25	&&-	&-	&-	&-	&-	&&-	&-	&-	&-	&-\\
+Exponential &
+12.710	&-	&- 	&-0.00052 	&0.19728	&&
+2.8808	&-	&- 	&0.00458	&0.05424	&&
+0.130	&-	&- 	&0.957 	&0.039 \\
+Geometric-recruitment &
+10.457	&0.57321	&0.42621	&- 	&0.24269	&&
+1.39877 &0.17905	&0.17944	&- 	&0.02925	&&
+0.851	&0.935	&0.932	&- 	&0.814 \\
+  \hline
+  \multicolumn{18}{c}{$^{1}$ Derived parameter: $r = \log(\gamma + \omega)$.}
+  \end{tabular}
+  \label{tab:simgeom}
 \end{table}
 
+\clearpage
+
+\begin{table}[htb]
+  \centering
+  \footnotesize
+  \caption{Results of exponential and geometric-recruitment models fit to Leslie matrix simulation (1000 Monte Carlo replicates).}
+  \begin{tabular}{lccccccccccccc}
+    \hline
+    & \multicolumn{5}{c}{Mean} &&
+    \multicolumn{3}{c}{RMSE} &&
+    \multicolumn{3}{c}{Coverage} \\
+    \cline{2-6}     \cline{8-10}    \cline{12-14}
+    Model & $\Lambda$ & $\gamma$ & $\omega$ & $r$\footnotemark[1] & $p$ &&
+    $\Lambda$ & $r$ & $p$ && 
+    $\Lambda$ & $r$ & $p$ \\
+    \hline
+Exponential &
+27.841	&-	&- 	&-0.00470	&0.09037	&&
+17.97942	&0.00186 	&0.15979	&&
+0.000 		&0.977 	&0.000 \\
+Geometric-recruitment &
+12.782	&0.21211	&0.78326	&-	&0.19720	&&
+2.92896 	&- 	&0.05438	&&
+0.057		&- 	&0.022 \\
+  \hline
+  \multicolumn{14}{c}{$^{1}$ Derived parameter: $r = \log(\gamma + \omega)$.}
+  \end{tabular}
+  \label{tab:simmat}
+\end{table}
+\clearpage
+
+\begin{table}[htb]
+  \centering
+  \footnotesize
+  \caption{Parameters of Gompertz-logistic simulation with two generation per year 
+(1000 Monte Carlo replicates)
+  and results of models fit to these data.}
+  \begin{tabular}{lcccccccccccccc}
+    \hline
+    & \multicolumn{4}{c}{Mean} &&
+    \multicolumn{4}{c}{RMSE} &&
+    \multicolumn{4}{c}{Coverage} \\
+    \cline{2-5}     \cline{7-10}    \cline{12-15}
+    Model & $\Lambda$ & $r$ & $K$ & $p$ &&
+    $\Lambda$ & $r$ & $K$ & $p$ && 
+    $\Lambda$ & $r$ & $K$ & $p$ \\
+    \hline
+Simulation & 
+10	&0.05	&10	&0.5	&&-	&-	&-	&-	&&-	&-	&-	&-\\
+Gompertz-logistic (MLE; 1 generation/year) &
+8.392	&0.00919	&1.269	&0.60241	&&
+1.64842	&-	&8.94192	&0.10395	&&
+0.013	&-	&1.000 	&0.000 \\
+Gompertz-logistic (MCMC; 1 generation/year) &
+8.315	&0.03044	&4.575	&0.60723	&&
+1.72224	&-	&5.70813	&0.10870	&&
+0.007	&-	&0.368		&0.000 \\
+Gompertz-logistic (MCMC; 2 generations/year) &
+9.969	&0.05405	&10.407	&0.50337	&&
+0.55084	&0.01697	&2.26087	&0.02318	&&
+0.921	&0.964	&0.967	&0.935 \\
+  \hline
+  \end{tabular}
+  \label{tab:simgomp}
+\end{table}
+
+\clearpage
+
+\begin{figure}
+\caption{Histograms of 1000 parameter estimates for each of nine
+simulation cases with Ricker-logistic dynamics. The blue vertical lines are the 
+data-generating values.}
+  \centering
+  \includegraphics{../figs/rick_hists}
+\label{fig:rick_hists}
+\end{figure}
+
+\begin{figure}
+\caption{Histograms of 1000 parameter estimates for each of three
+simulation cases with Ricker-logistic + immigration dynamics. The blue vertical lines are the 
+data-generating values.}
+  \centering
+  \includegraphics[width=8.5in]{../figs/ricki_hists}
+\label{fig:ricki_hists}
+\end{figure}
+
+\begin{figure}
+\caption{Histograms of 1000 parameter estimates for each of two
+simulation cases with geometric-recruitment or Leslie matrix dynamics, each fit with 
+exponential and geometric-recruitment models. The blue vertical lines are the 
+data-generating values; the data-generating values for $r$ are derived from
+the actual dynamic parameters of these models.  No data-generating values
+are provided for $\gamma$ or $\omega$ by the Leslie matrix model because
+each age class had its own survival and fecundity parameters.}
+  \centering
+  \includegraphics{../figs/geom_mat_hists}
+\label{fig:geom_mat_hists}
+\end{figure}
+
+\begin{figure}
+\caption{Histograms of 1000 parameter estimates simulated with Gompertz-logistic dynamics, 
+two generations per year, fit with three different Gompertz-logistic models.
+The vertical lines are the data-generating values; changing the number of generations
+changes the meaning of $r$, so no data-generating values of it are provided for the 
+single generation per year cases.}
+  \centering
+  \includegraphics{../figs/gomp_hists}
+\label{fig:gomp_hists}
+\end{figure}
+\end{landscape}
+
 \end{document}

### 2014-06-25 supp/code_samples/oven.jags.five.versions.r

@@ -1,0 +1,499 @@
+# BBS analysis code using JAGS
+# Read in input files
+ydata <- as.matrix(read.csv('oven3a.csv', row.names=1))
+obs5 <- as.matrix(read.csv('observers.csv', row.names=1))
+first <- as.matrix(read.csv('first.run.csv', row.names=1))
+wind.factor <- as.matrix(read.csv('wind.factor.csv', row.names=1))
+# Fix formatting
+wind.factor[wind.factor==" 0"] <- "0"
+wind.factor[wind.factor==" 1"] <- "1"
+wind.factor[wind.factor==" 2"] <- "2"
+
+library(rjags)
+
+# Settings
+year1 <- 66
+species <- 'oven'
+nSites <- nrow(ydata)
+nYears <- ncol(ydata)
+
+
+fillN3 <- function(N, p, lambda, alpha) { # Need to replace NAs in initial N with values, adjust others by detection probability
+  for (i in 1:nrow(N)) {
+    notNA <- which(!is.na(N[i, ]))
+    size <- N[i, notNA] 
+    size[size==0] <- 0.5
+    addTo <- rnbinom(length(notNA), size=size, prob=p)
+    N[i, notNA] <- N[i, notNA] + addTo
+    for (t in 1:ncol(N))
+      if (is.na(N[i,t])) {
+        if (t == ncol(N) || is.na(N[i, t+1])) {
+          if (t==1) {
+            N[i, t] <- rnbinom(1, size=alpha, mu=lambda) #If no previous or subsequent value, use supplied default
+          }
+          else {
+            N[i, t] <- N[i, t-1] #If no subsequent value, use previous
+          }
+        }
+        else if (t==1)
+          N[i, t] <- N[i, t+1] #If no previous value, use subsequent
+        else
+          N[i, t] <- ceiling((N[i, t-1] + N[i, t+1])/2) #Otherwise, use mean
+      }
+  }
+  return(N)
+}
+
+N <- fillN3(ydata, 0.18, 33.4, 0.65)
[…]

### 2014-06-25 supp/code_samples/oven.unmarked.model.selection.r

@@ -1,0 +1,305 @@
+library(unmarked)
+
+# Read in input files
+ydata <- as.matrix(read.csv('oven3a.csv', row.names=1))
+observers <- as.matrix(read.csv('observers.csv', row.names=1))
+first.run <- as.matrix(read.csv('first.run.csv', row.names=1))
+wind.factor <- as.matrix(read.csv('wind.factor.csv', row.names=1))
+# Fix formatting
+wind.factor[wind.factor==" 0"] <- "0"
+wind.factor[wind.factor==" 1"] <- "1"
+wind.factor[wind.factor==" 2"] <- "2"
+# Combine obsCovs together into list
+obs4 <- list(observers = observers, first.run = first.run, wind.factor = 
+  wind.factor)
+  
+# Basic settings and labels
+year1 <- 66
+aou <- 'oven'
+gam.model <- list(~1)[[1]] 
+om.model <- list(~1)[[1]] 
+lam.model <- list(~1)[[1]] 
+K <- 600
+iota.model <- list(~1) [[1]]
+
+# Input data for pcountOpen
+bird.frame <- unmarkedFramePCO(ydata, obsCovs=obs4, numPrimary=45)
+
+# Run mixture models for initial abundance  
+mods1 <- rep('', 3)
+outputs1 = list()
+for (i in 1:3) {
+  mixture <- c('P','NB','ZIP')[i]
+  dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[3]
+  p.model <- list(~1, ~first.run, ~wind.factor, ~wind.factor + first.run)[[1]]
+  immigration <- FALSE
+  mod.name <- paste(gsub("\\*", ".", gsub(".stand", "", gsub("[ 1]", "", 
+    paste(paste(aou, year1, sep=''), mixture, dynamics, 'lam', 
+    substring(deparse(lam.model), 2), 'gam', substring(deparse(gam.model), 2), 
+    'om', substring(deparse(om.model), 2), 'p', substring(deparse(p.model), 2), 
+    ifelse(immigration, "iotaT", "iotaF"), substring(deparse(iota.model), 2),
+    'K', sep='.')))), K, sep="")
+  mods1[i] <- mod.name
+  file.name <- paste(mod.name, 'gzip', sep='.')
+  print(file.name)
+  fm <- pcountOpen(data=bird.frame, mixture=mixture, gammaformula=gam.model, 
[…]

### 2014-06-25 supp/code_samples/ricker_r0.1K5.r

@@ -1,0 +1,60 @@
+# Simulate data under Ricker model
+
+# TO MODIFY
+# (1) Change simulator... source(SOMETHING ELSE)
+# (2) Change stub
+# (3) Change simout
+# (4) Change simulation parameters
+# (5) Change pcountOpen arguments: "dynamics" and "K"
+# (6) Results
+
+ls()
+stub <- "ricker_r0.1K5"  # (2)
+setwd(paste('simulated', stub, sep='/'))
+source("../sim_ricker.r")  # (1)
+ls()
+
+library(unmarked)
+
+nsim <- 1000
+simout.mle <- simout.cover <- matrix(NA, nsim, 4)  # (3)
+colnames(simout.mle) <- colnames(simout.cover) <- c("Lambda", "r", "K", "p") # (3)
+
+# Simulate
+set.seed(345489)
+lambda <- 10 # (4)
+r <- 0.1     # (4)
+K <- 5       # (4)
+p <- 0.25    # (4)
+for(i in 1:nsim) {
+    cat("\ndoing", i, "at", format(Sys.time()), "\n")
+    data.name <- paste("data_", stub, "_", i, sep="")
+    data.file <- paste("data/", data.name, ".gzip", sep="")
+    sim.i <- sim.ricker(lambda=lambda, r=r, K=K, p=p, nSites=100, nYears=40) # (1)
+    cat("   max(N)=", max(sim.i$N), "\n", sep="")
+    assign(data.name, sim.i)
+    save(list=data.name, file=data.file) # save simulated data
+    umf.i <- unmarkedFramePCO(y=sim.i$y, numPrimary=ncol(sim.i$y))
+    fm.i <- pcountOpen(~1, ~1, ~1, ~1, umf.i, K=200, dynamics="ricker",  # (5)
+                       starts=c(log(lambda), log(r), log(K), qlogis(p))) # (5)
+    model.name <- paste("fm_", stub, "_", i, sep="")
+    model.file <- paste("fm/", model.name, ".gzip", sep="")
+    assign(model.name, fm.i)
+    save(list=model.name, file=model.file) # save fitted models
+    mle.i <- coef(fm.i)
+    simout.mle[i,] <- c(exp(mle.i[1]), exp(mle.i[2]), exp(mle.i[3]), plogis(mle.i[4])) # (6)
+    cat("   mle=", round(simout.mle[i,], 4), "\n")
[…]

### 2014-06-25 supp/code_samples/sample.proj.r

@@ -1,0 +1,173 @@
+# Forecast model
+# Since the forecast model can be implemented identically to a purely retrospective model
+# as far as JAGS code goes, we are including R code as well to show the differences in implementation.
+library(rjags)
+
+# Settings
+nRun <- 100000 # Number of times to run the MCMC, after burnin
+nThin <- 25 # Thinning rate (only store 1 iteration out of every nThin)
+nChain <- 3 # Number of separate chains to run MCMC (for testing convergence)
+nBurn <- 10000 # Number of times to run the MCMC before getting output
+nAdapt <- 1000 # Number of times to run the MCMC for adaptation at beginning
+year1 <- 1966 # First year of estimation data
+year2 <- 2010 # Last year of estimation data/first year of projection
+year3 <- 2100 # Last year of projection
+thresholds <- c(1,10,50) # Quasi-extirpation levels to monitor
+
+# Read in input files
+ydata <- as.matrix(read.csv('oven3a.csv', row.names=1))
+observers <- as.matrix(read.csv('observers.csv', row.names=1))
+wind.factor <- as.matrix(read.csv('wind.factor.csv', row.names=1))
+# Fix formatting
+wind.factor[wind.factor==" 0"] <- "0"
+wind.factor[wind.factor==" 1"] <- "1"
+wind.factor[wind.factor==" 2"] <- "2"
+
+# More settings
+species = 'oven'
+aou <- 6740
+nSites <- nrow(ydata)
+nYears <- year3 - year1 + 1
+nPast <- year2 - year1 + 1
+nFuture <- year3 - year2
+
+# File names to save
+mod.name <- paste(species, 'sample.proj.JAGS', sep='.')
+file.name <- paste(mod.name, 'RData', sep='.')
+file.name2 <- paste(mod.name, 'gzip', sep='.')
+file.name2
+
+# For projection, fill in the future counts with all NA
+ydata <- cbind(ydata, matrix(NA, nrow=nSites, ncol=nFuture))
+
+fakeN <- ydata + 10 # Try to set up plausible initial values for N
+fillNAs <- function(N, default) { # Need to replace NAs in initial N with values
+  tMax <- ncol(N)
+  defaults <- which(is.na(N[,1]) & is.na(N[,2]))
[…]

### 2014-06-25 supp/code_samples/sim_exp.r

@@ -1,0 +1,15 @@
+
+# Simulate data under exponential growth model
+sim.exp <- function(lambda, r, p, nSites=100, nYears=45) {
+    y <- N <- matrix(NA, nSites, nYears)
+    N[,1] <- rpois(nSites, lambda)
+    for(t in 2:nYears) {
+        mu <- N[,t-1]*exp(r)
+        N[,t] <- rpois(nSites, mu)
+    }
+    y[] <- rbinom(nSites*nYears, N, p)
+    return(list(N=N, y=y, lambda=lambda, r=r, p=p,
+                seed=.Random.seed, seed.kind=RNGkind()))
+}
+
+

### 2014-07-31 improved_state_space.tex
edits to intro and new bibitem

@@
 \textit{Key words}: abundance, Dail and Madsen model, density-dependence,
-detectability, Gompertz-logistic model, immigration, open population point 
-count models, population dynamics, random observer effects, Ricker-logistic model, zero-inflated
+detectability, Gompertz-logistic model, immigration, 
+%% open population point count models, 
+population dynamics, 
+%% random observer effects, 
+Ricker-logistic model, zero-inflated
 
@@
 \citep{may:1975,turchin:1990,bjornstad_etal:1999}. 
-In applied contexts, population models are used for estimating extinction probabilities
+In applied contexts, population models are used for estimating extinction risk % probabilities
 \citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
@@
 models are time series models in which the true state of the
-system (e.g. population size during each year) is modeled as a hidden
+system (e.g. population size during each year) is modeled as a latent %hidden
 process, and the observed data are modeled conditional on this process
@@
 Ricker-logistic model, or it might be density-independent, such as when growth
-is exponential.
+is exponential. %% Exponential or geometric?
 A generic model can be written: 
@@
     N_1 &= X_1 \nonumber \\
-N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
+N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; %% Should this be \eta_t?
 t=2,\hdots,T  \\
@@
 state-space models, the two sets of random effects
-are assumed to be i.i.d. Gaussian deviates:
+are assumed to be i.i.d. Gaussian variables: %deviates:
 $\eta_t \sim \mathrm{N}(0, \sigma)$ and
@@
 or to individuals being available but missed by the ecologist collecting the data in the field.
-Although a vast number of methods have been devised for accounting for
+Although a vast number of methods have been devised to account for %for accounting for
 this form of observation error, rarely have these methods been
-integrated into state-space models of population dynamics \citep[but
-see][]{buckland_etal:2004}.
+integrated into state-space models of population dynamics 
+%\citep[but see][]{buckland_etal:2004}. 
+Of the exceptions that exist, none provide software to facilitate model fitting 
+\citep{buckland_etal:2004}.
 
@@
 Ignoring spatial variation reduces the scope of the
-inferences and can lead to underestimation of variances. 
-Identifiability problems require 
-model constraints or the assertion of additional assumptions. 
-Specifically, abundance in
-the first year cannot be estimated because there is only one data
-point (the count in the first year) that carries any information about
-it. To remedy this, practitioners either assume that there is no
+inferences and reduces the information about the parameters of interest \citep{dennis_etal:2010}.
+%% Identifiability problems require 
+%% model constraints or the assertion of additional assumptions. 
+%% Specifically, 
+Even more problematic is the issue that without some form of replication in the first year, 
+abundance cannot be estimated without making strong assumptions about 
+the initial distribution. 
+%% because there is only one data
+%% point (the count in the first year) that carries any information about
+%%it. To remedy this, 
+For example, practitioners either assume that there is no
 observation error in the first year, or they assume the population is

### 2014-07-31 improved_state_space.tex
on to the results

@@
 variation in abundance when individuals cannot be detected with
-certainty. To estimate parameters associated with both abundance 
-and the detection process, the original $N$-mixture model uses replicate
+certainty. To estimate both abundance and detection parameters, 
+%% associated with both abundance and the detection process, 
+the original $N$-mixture model uses replicate
 observations at each site. The observations are collected during sufficiently
@@
 
-The DM model assumes that abundance in time $t$ is a function of
+The DM model assumes that abundance at time $t>1$ is a function of
 abundance at time $t-1$, i.e. abundance at each site evolves as a
@@
 where $\omega$ is the apparent survival probability and $\gamma$
-is the recruitment rate (which can depend on $N_{i,t-1}$).
+is the recruitment rate, which can depend on $N_{i,t-1}$.
 \citet{dail_madsen:2011} proposed three
@@
 individuals; the autoregressive model, $G_{i,t} \sim
-\mathrm{Pois}(\gamma(N_{i,t-1}))$, which
-simulates geometric or density independent growth; and the
+\mathrm{Pois}(\gamma(N_{i,t-1}))$
+%% , which simulates 
+of geometric or density independent growth; and the
 ``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which assumes the
@@
 Partitioning population growth into survival and recruitment
-is ideal in terms of providing a mechanistic description of population
+is useful %%ideal 
+in terms of providing a mechanistic description of population
 dynamics, but this is not always possible using simple count data,
@@
 with classical population growth models that have a long history in ecology.
-The exponential growth model is the simplest: 
+The exponential growth model is the simplest: %% Should this geometric instead of exponential?
 $N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
@@
 densities, and both parameters are constrained to be positive. This is
-a stochastic version of the Ricker-logistic population growth model
+a stochastic version of the 
+Ricker-logistic population growth model %% Could we just call this the "Ricker model"?
 \citep{ricker:1954}. Another option is a modified Gompertz-logistic
@@
 One limitation of the geometric-recruitment, exponential,
-Ricker-logistic, and Gompertz-logistic versions of the DM models all
+Ricker-logistic, and Gompertz-logistic versions of the DM models %%all
 is that they include no mechanism for a local population to recover
@@
 where $\psi$ represents the proportion of extra zeros. 
-Mechanistically, this distribution might be appropriate when not all sites are 
-within the range of a species.
-
+%% Mechanistically, this distribution might be appropriate when not all sites are 
+%% within the range of a species.
+%%
 Under this formulation, 
@@
 growth extensions use a single Poisson distribution for abundance. 
-Conceptually, this is an appropriate method of modeling demographic stochasticity 
-for semelparous organisms with one generation per sampling interval (year)
-\citep{bonsall_hastings:2004,melbourne_hastings:2008},
-but may actually perform better for iteroparous organisms as well (see Discussion).  
-Dynamics of semelparous organisms with more than one generation per year 
-(many insects) may be overdispersed compared to the Poisson.  If the number of generations per
-year is known, then this can be accounted for in our models by adding extra $N$'s between each
-count (see Supplement 1).
+% NOTE: I would save this for the discussion
+% Conceptually, this is an appropriate method of modeling demographic stochasticity 
+% for semelparous organisms with one generation per sampling interval (year)
+% \citep{bonsall_hastings:2004,melbourne_hastings:2008},
+% but may actually perform better for iteroparous organisms as well (see Discussion).  
+% Dynamics of semelparous organisms with more than one generation per year 
+% (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
+% year is known, then this can be accounted for in our models by adding extra $N$'s between each
+% count (see Supplement 1).
 
@@
 as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
-function of temperature ($C$) and $K$ a function of rainfall ($M$):
-\begin{linenomath*}
-\begin{gather}
-\log(r_{i,t}) = \beta^{r}_{0} + \beta^{r}_{1}C_{i,t} \nonumber \\
-\log(K_{i,t}) = \beta^{K}_{0} + \beta^{K}_{1}M_{i,t}  
-\label{eq:weather}
-\end{gather}
-\end{linenomath*}
-
-Another way to model environmental stochasticity is as lognormal variation in observed local population 
-growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  This could be applied to both density-independent and density-dependent models.
+function of temperature %($C$) 
+and $K$ a function of rainfall. %($M$):
+%% NOTE: I don't think these equations are necessary
+% \begin{linenomath*}
+% \begin{gather}
+% \log(r_{i,t}) = \beta^{r}_{0} + \beta^{r}_{1}C_{i,t} \nonumber \\
+% \log(K_{i,t}) = \beta^{K}_{0} + \beta^{K}_{1}M_{i,t}  
+% \label{eq:weather}
+% \end{gather}
+% \end{linenomath*}
+% 
+Another way to model environmental stochasticity is as lognormal variation in 
+the %observed local 
+population growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  
+This could be applied to both density-independent and density-dependent models.
 For example, for the Ricker-logistic + immigration model:
@@
 where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
-We suggest three variations for spatial variation in environmental stochasticity.  The first has independent environmental stochasticity 
+We suggest three variations for spatial variation in environmental stochasticity.  
+%% These are interesting ideas, but do we actually evaluate them? If not, it might be 
+%% better to save them for the discussion.
+The first has independent environmental stochasticity 
 between sites, and might be appropriate where sites are widely dispersed.  In the second, 
@@
 
-\subsection*{Proof of Concept Simulation Study}
+%\subsection*{Proof of Concept Simulation Study}
+\subsection*{Simulation Study to Assess Estimator Accuracy}
 
@@
 a Gompertz-logistic model with two non-overlapping generations per year (sampling occasion).   
+%% Couldn't we expand our model to deal with these "mis-specifications"? Like Elise and I did in a couple of recent papers?
 The first two scenarios are underdispersed compared to the Poisson and the third is overdispersed.
@@
 
-We fitted a series of models using maximum likelihood, 
+We fit a series of models using maximum likelihood, 
 and then a series of models using MCMC. We started by testing
@@
 We ran the top ranked models from the maximum likelihood
-analyses in a Bayesian framework with non-informative priors ($\Lambda$,
+analyses in a Bayesian framework with vague %non-informative 
+priors ($\Lambda$,
 $\alpha$, and $K  \sim \mathrm{U}(0, 200)$, $r  \sim \mathrm{U}(0, 5)$,
@@
 five Markov chains for each model \citep{gelman_rubin:1992}.
-For each chain we sampled the MCMC for at least 40,000 iterations, after at
-least 3,000 adaptation and tuning samples.  
+For each chain, we drew at least 40,000 samples, 
+%the MCMC for at least 40,000 iterations, 
+after at
+least 3,000 adaptation iterations. %and tuning samples.  
 We also used the random observer effects model to obtain route and 

### 2014-07-31 improved_state_space.tex
finished reviewing ms. Now to the response letter

@@
 \subsection*{Robustness Simulation Study}
-In simulations of the geometric-recruitment model estimated with exponential model,
+In simulations of the geometric-recruitment model estimated with the exponential model,
 estimates of $\Lambda$ were biased high %(2.71, relative bias 27.1\%) 
@@
 run in the Bayesian framework, except that the estimate of $r$
-more than halved (Appendix B, Fig. B1). 
+more than halved (Appendix B, Fig. B1). %% What's up with that??
 When random observer effects
@@
 model that included survival and recruitment parameters would be to
-use either prior information or additional data. For example, in a
+use either prior information or additional data \citep{zipkin_etal:2014}. For example, in a 
 Bayesian analysis existing estimates of demographic parameters
@@
 demographic data are often referred to as integrated population models
-\citep[IPM;][]{besbeas_etal:2002,
-buckland_etal:2004,schaub_etal:2007}.
+\citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
 We believe that a DM modeling framework could work well in this context,
@@
 that multiple other options are available as well, such as removal,
-multiple observer, and distance modeling \citep{williams_etal:2002}. These could be
+multiple observer, and distance sampling \citep{williams_etal:2002}. These could be
 accomplished by extending the open population $N$-mixture model
@@
 from the receiving site \citep{hastings:1991,hanski:1998}. These models would
-be based on the assumptions that either the study sites
-encompass the whole range of the metapopulation or that the
-annual changes in the populations sampled are representative of
-a larger metapopulation. Initial tests suggest that it would be
+require that all sites in the metapopulation are sampled, or that abundance at 
+unsampled locations can be estimated \citep{lele_etal:1998}. 
+%be based on the assumptions that either the study sites
+%encompass the whole range of the metapopulation or that the
+%annual changes in the populations sampled are representative of
+%a larger metapopulation. 
+Initial tests suggest that it would be
 possible to fit some or all of these models, at least in the

### 2014-07-31 improved_state_space.tex
response edits

@@
 simultaneously; however, their model did not include serial
-dependence, which is a hallmark of population models.
+dependence, which is a hallmark of classical population growth models as well as state-space models.
 

### 2014-08-03 improved_state_space.tex
Integrated Richard's revisions and shortened

@@
 \linenumbers
-\usepackage{setspace}
+\usepackage[nodisplayskipstretch]{setspace}
 \usepackage{amsmath}
@@
 \setlength{\jot}{-1.2ex}
+\setstretch{1}
+\widowpenalty 2
+\clubpenalty 2
 \usepackage{etoolbox}
@@
 }
-\date{} %remove date
+\date{} 
 \begin{document}
@@
 \textit{Key words}: abundance, Dail and Madsen model, density-dependence,
-detectability, Gompertz-logistic model, immigration, 
-%% open population point count models, 
-population dynamics, 
-%% random observer effects, 
-Ricker-logistic model, zero-inflated
+detectability, Gompertz-logistic, immigration, % model
+population dynamics, Ricker-logistic, zero-inflated % model
 
@@
 \citep{may:1975,turchin:1990,bjornstad_etal:1999}. 
-In applied contexts, population models are used for estimating extinction risk % probabilities
-\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
+In applied contexts, population models are used for estimating extinction risk 
+\citep{nadeem_lele:2011,hostetler_etal:2012} and for predicting the %schoener_spiller:1992,
 effects of future environmental conditions or conservation actions on
@@
 
-Several challenges are routinely encountered when trying to fit models
+Two challenges are routinely encountered when trying to fit models %Several
 of population dynamics to data from field studies. 
-First, deterministic models of population dynamics are virtually always inadequate due to process
-variation, the inherent stochasticity in demographic parameters and environmental
+First, deterministic models of population dynamics are generally inadequate due to process %virtually always
+variation, the stochasticity in demographic parameters and environmental % inherent
 conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
@@
 models are time series models in which the true state of the
-system (e.g. population size during each year) is modeled as a latent %hidden
+system (e.g. population size during each year) is modeled as a latent 
 process, and the observed data are modeled conditional on this process
@@
 The growth model may be density-dependent, as in the case of the 
-Ricker-logistic model, or it might be density-independent, such as when growth
+Ricker-logistic model (henceforth Ricker model), or it might be 
+density-independent, such as when growth
 is exponential. %% Exponential or geometric?
@@
   \begin{align}
-    N_1 &= X_1 \nonumber \\
-N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; %% Should this be \eta_t?
+    N_1 &= x_0 \nonumber \\ %X_1
+N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
 t=2,\hdots,T  \\
@@
 \end{linenomath*}
-In classical
+where $x_0$ is the model for initial abundance.  In classical
 state-space models, the two sets of random effects
-are assumed to be i.i.d. Gaussian variables: %deviates:
+are assumed to be i.i.d. Gaussian variables: 
 $\eta_t \sim \mathrm{N}(0, \sigma)$ and
@@
 or to individuals being available but missed by the ecologist collecting the data in the field.
-Although a vast number of methods have been devised to account for %for accounting for
+Although a vast number of methods have been devised to account for 
 this form of observation error, rarely have these methods been
-integrated into state-space models of population dynamics 
-%\citep[but see][]{buckland_etal:2004}. 
+integrated into state-space models of population dynamics. 
 Of the exceptions that exist, none provide software to facilitate model fitting 
@@
 Ignoring spatial variation reduces the scope of the
-inferences and reduces the information about the parameters of interest \citep{dennis_etal:2010}.
-%% Identifiability problems require 
-%% model constraints or the assertion of additional assumptions. 
-%% Specifically, 
+inferences, reduces the information about the parameters of interest,
+and can lead to underestimation of variances \citep{dennis_etal:2010}.
 Even more problematic is the issue that without some form of replication in the first year, 
@@
 the initial distribution. 
-%% because there is only one data
-%% point (the count in the first year) that carries any information about
-%%it. To remedy this, 
 For example, practitioners either assume that there is no
-observation error in the first year, or they assume the population is
+observation error in the first year ($x_0 = X_1$), or they assume the population is
 at equilibrium, which may defeat the purpose of many studies of
@@
 \citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
-distributions for the process and observation errors. Observation models with
-more intuitive interpretations, such as those that explicitly model
-detection probability, have been proposed by 
+distributions for the process and observation errors. An observation model %with
+%more intuitive interpretations, such as those that 
+explicitly modeling
+detection probability has been proposed by 
 \citet{kery_etal:2009}. \citet{lele_etal:1998} and 
@@
 certainty. To estimate both abundance and detection parameters, 
-%% associated with both abundance and the detection process, 
 the original $N$-mixture model uses replicate
@@
 standard state-space models because
-%, as demonstrated by \citet{royle:2004biom},
 process variation and observation error can be estimated from
@@
 $G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
-depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new
+depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new %
 individuals; the autoregressive model, $G_{i,t} \sim
-\mathrm{Pois}(\gamma(N_{i,t-1}))$
-%% , which simulates 
+\mathrm{Pois}(\gamma N_{i,t-1})$
 of geometric or density independent growth; and the
-``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which assumes the
+no-trend model, $\gamma = (1-\omega)\Lambda$, which assumes the
 population is at equilibrium. Covariates of
-$\omega$ and $\gamma$ can be easily accommodated, for example
+$\omega$ and $\gamma$ can be accommodated, e.g., % easily %for example
 using logit- and log-linear models respectively.
@@
 \subsection*{Population growth models}
-\begin{samepage}
+%\begin{samepage}
 Partitioning population growth into survival and recruitment
-is useful %%ideal 
+is useful  
 in terms of providing a mechanistic description of population
@@
 a stochastic version of the 
-Ricker-logistic population growth model %% Could we just call this the "Ricker model"?
-\citep{ricker:1954}. Another option is a modified Gompertz-logistic
-density-dependent model \citep{hart_gotelli:2011}:
+Ricker model \citep{ricker:1954}. Another option is a 
+modified Gompertz-logistic
+density-dependent model \citep[henceforth Gompertz model,][]{hart_gotelli:2011}:
 \begin{linenomath*}
@@
 Here the interpretations of $r$ and $K$ are similar to those in the
-Ricker-logistic model. Because a single Poisson distribution controls
+Ricker model. Because a single Poisson distribution controls
 the distribution of $N_{i,t}$ in each of these models, the discrete 
@@
 likelihood is not required, speeding up processing time.
-\end{samepage}
+%\end{samepage}
 
@@
 One limitation of the geometric-recruitment, exponential,
-Ricker-logistic, and Gompertz-logistic versions of the DM models %%all
+Ricker, and Gompertz versions of the DM models 
 is that they include no mechanism for a local population to recover
@@
 growth and immigration).  
-The geometric-recruitment, Ricker-logistic,
-and Gompertz-logistic models can be extended to allow for immigration in the
-same way.   Deterministic versions of the Ricker-logistic + immigration and the 
-Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
+The geometric-recruitment, Ricker,
+and Gompertz models can be extended to allow for immigration in the
+same way.   Deterministic versions of the Ricker + immigration and the 
+Gompertz + immigration do reach stable equilibriums, but the equilibriums 
 are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
@@
 where $\psi$ represents the proportion of extra zeros. 
-%% Mechanistically, this distribution might be appropriate when not all sites are 
-%% within the range of a species.
-%%
 Under this formulation, 
@@
 % NOTE: I would save this for the discussion
-% Conceptually, this is an appropriate method of modeling demographic stochasticity 
-% for semelparous organisms with one generation per sampling interval (year)
-% \citep{bonsall_hastings:2004,melbourne_hastings:2008},
+Conceptually, this is an appropriate method of modeling demographic stochasticity 
+ for semelparous organisms with one generation per sampling interval (year)
+ \citep{bonsall_hastings:2004},
+and the DM models appropriate for modeling the demographic stochasticity of iteroparous organisms,
+if they are underdispersed compared to the Poisson (but see Discussion).
 % but may actually perform better for iteroparous organisms as well (see Discussion).  
-% Dynamics of semelparous organisms with more than one generation per year 
-% (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
-% year is known, then this can be accounted for in our models by adding extra $N$'s between each
-% count (see Supplement 1).
+ Dynamics of semelparous organisms with more than one generation per year 
+ (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
+ year is known, then this can be accounted for in our models by adding extra $N$'s between each
+ count (see Supplement 1).
 
@@
 rate due to stochastic environmental conditions --- is by explicitly
-modeling dynamic parameters as functions of environmental covariates, such
-as weather.  For example, in the Gompertz-logistic model (Eq.~\ref{eq:gomp}), $r$ could be a 
-function of temperature %($C$) 
-and $K$ a function of rainfall. %($M$):
-%% NOTE: I don't think these equations are necessary
-% \begin{linenomath*}
-% \begin{gather}
-% \log(r_{i,t}) = \beta^{r}_{0} + \beta^{r}_{1}C_{i,t} \nonumber \\
-% \log(K_{i,t}) = \beta^{K}_{0} + \beta^{K}_{1}M_{i,t}  
-% \label{eq:weather}
-% \end{gather}
-% \end{linenomath*}
-% 
+modeling dynamic parameters as functions of environmental covariates. %, such as weather.  
+For example, in the Gompertz model (Eq.~\ref{eq:gomp}), $r$ could be a 
+function of temperature  
+and $K$ a function of rainfall.
+
 Another way to model environmental stochasticity is as lognormal variation in 
-the %observed local 
-population growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  
+the population growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  
 This could be applied to both density-independent and density-dependent models.
-For example, for the Ricker-logistic + immigration model:
+For example, for the Ricker + immigration model:
 \begin{linenomath*}
@@
 We suggest three variations for spatial variation in environmental stochasticity.  
-%% These are interesting ideas, but do we actually evaluate them? If not, it might be 
-%% better to save them for the discussion.
 The first has independent environmental stochasticity 
@@
 $\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
-the standard deviation of the random observer effects (also on the logit scale).
+the logit-scale standard deviation of the random observer effect. % (also on the logit scale).
 
@@
 
-In
-practice, simulation methods such as Markov chain Monte Carlo (MCMC)
-are used to simulate the posterior distributions. While time
+%In practice, 
+Simulation methods such as Markov chain Monte Carlo (MCMC)
+are used to estimate the posterior distributions. While time %simulate
 consuming, MCMC may be the only viable approach for estimating
@@
 
-%\subsection*{Proof of Concept Simulation Study}
 \subsection*{Simulation Study to Assess Estimator Accuracy}
@@
 $p \in \{0.05, 0.25, 0.5\}$. 
-Our second series of simulations changed dynamics to the Ricker-logistic
+Our second series of simulations changed dynamics to the Ricker
 model. We used an initial abundance of 10 and a detection probability
@@
 9 total combinations). Our third series of simulations was based
-on the Ricker-logistic + immigration dynamics model; here we fixed all
-parameters the same as the Ricker-logistic model (with $r$ = 0.05 and $K$ = 10) and
+on the Ricker + immigration dynamics model; here we fixed all
+parameters the same as the Ricker model (with $r$ = 0.05 and $K$ = 10) and
 simulated low, medium, and high values of immigration rate (Table A1).
@@
 $\omega = \gamma = 0.5$; 2) a Leslie matrix model with six age classes; and 3) 
-a Gompertz-logistic model with two non-overlapping generations per year (sampling occasion).   
+a Gompertz model with two non-overlapping generations per year (sampling occasion).   
 %% Couldn't we expand our model to deal with these "mis-specifications"? Like Elise and I did in a couple of recent papers?
@@
 Criterion (AIC) values.
-For the third scenario, we estimated parameters with the Gompertz-logistic model with 
+For the third scenario, we estimated parameters with the Gompertz model with 
 one generation per year using both maximum likelihood and MCMC approaches, as well
@@
 Maryland and Virginia. For our focal species, we selected the
-ovenbird (\textit{Seiurus aurocapilla}), an abundant and widespread,
-forest- breeding migratory songbird with a stable or increasing trend
+ovenbird (\textit{Seiurus aurocapilla}), an abundant, %and widespread,
+forest-breeding migratory songbird with a stable or increasing trend
 in the region \citep{porneluzi_etal:2011}.
@@
 AIC model from that set to test eight additional models of dynamics:
-constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, 
+constant-recruitment, geometric-recruitment, Ricker, Gompertz, 
 the previous three with immigration, and exponential + immigration.
-These models require a a finite upper bound ($N_{max}$) to integrate
+These models require a a finite upper bound to integrate
 over when run in a maximum likelihood
@@
 We ran the top ranked models from the maximum likelihood
-analyses in a Bayesian framework with vague %non-informative 
+analyses in a Bayesian framework with vague  
 priors ($\Lambda$,
@@
 For each chain, we drew at least 40,000 samples, 
-%the MCMC for at least 40,000 iterations, 
-after at
-least 3,000 adaptation iterations. %and tuning samples.  
+after at least 3,000 adaptation iterations.   
 We also used the random observer effects model to obtain route and 
@@
 
-\subsection*{Proof of Concept Simulation Study}
+\subsection*{Simulation Study to Assess Estimator Accuracy}
 
@@
 $\Lambda$ was low. 
-Simulations of the Ricker-logistic model 
+Simulations of the Ricker model 
 generally performed well (Fig. A1 and Table A2 in Appendix A), except when the true value of $r$ was low (0.005). In this case the 
@@
 biased high (between 65.4\% and 198.9\%), and had low coverage (between 77.6\% and 94.3\%).
-In other cases for the Ricker-logistic model, relative RMSE was between 6.7\%
+In other cases for the Ricker model, relative RMSE was between 6.7\%
 and 55\%, bias was between -6.4\% and 5.8\%, and coverage was between
-91.9\% and 96.4\%.  All three cases simulated for the Ricker-logistic + immigration model 
+91.9\% and 96.4\%.  All three cases simulated for the Ricker + immigration model 
 performed fairly well (Fig. A2 and Table A3 in Appendix A; relative RMSE between 7.2\% 
@@
 In simulations of the geometric-recruitment model estimated with the exponential model,
-estimates of $\Lambda$ were biased high %(2.71, relative bias 27.1\%) 
-and $p$ biased low (Table A5). % (-0.05, relative bias -21.1\%).  
-%As a result, RMSE for these parameters were high %(2.88 and 0.05, respectively) 
-%and coverages low. % (13.0\% and 3.9\%).
-However, estimates of $r$ had low bias %(-0.0005) 
-and RMSE, %(0.005)
-with nominal coverage. % (95.7\%).  
+estimates of $\Lambda$ were biased high 
+and $p$ biased low (Table A5). 
+However, estimates of $r$ had low bias and RMSE, with nominal coverage. 
 The estimates from the geometric-recruitment model for $\Lambda$ and $p$ had low bias, but the estimates
@@
 mean $\Delta$AIC was 6.0 and when the exponential model was favored
-the mean was 1.4.  %Results from the Leslie matrix model simulations were similar, but more dramatic.
+the mean was 1.4.  
 For the Leslie matrix simulations, bias in estimates of $\Lambda$ and $p$ from both the exponential and geometric-recruitment models
@@
 
-The Gompertz-logistic model with a single generation per year fitted with both maximum likelihood and
-MCMC using data simulated from a Gompertz-logistic model with two generations per year underestimated 
-$\Lambda$  and $K$ and %(bias -1.61 and -1.68, respectively) (bias -8.73 and -5.42)  (bias 0.10 and 0.11)
+The Gompertz model with a single generation per year fitted with both maximum likelihood and
+MCMC using data simulated from a Gompertz model with two generations per year underestimated 
+$\Lambda$  and $K$ and 
 overestimated $p$ (Table A7 and Figure A4).  As a result, RMSEs for these parameters 
@@
 (Table B1A in Appendix B). 
-The best supported model for $p$ included additive effects of wind speed
-and first run of a route by an observer (Table B1B, model B.1). First run 
+The best supported model for $p$ included effects of wind speed %additive 
+and an observer's first run of a route (Table B1B, model B.1). First run 
 and increasing wind speeds both decreased $p$. All dynamics models 
 with immigration were better supported than models without immigration 
-(Table B1C); the best supported of these was the Ricker-logistic + immigration 
+(Table B1C); the best supported was Ricker + immigration % of these the
 (see Appendix B, Fig. B1 for estimates).
-Constant- recruitment dynamics models estimated unrealistically high
+Constant-recruitment dynamics models estimated unrealistically high
 survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05), whereas 
-the geometric-recruitment and geometric-recruitment + immigration
+the geometric- recruitment and geometric-recruitment + immigration
 models estimated nonviably low survival probabilities 
@@
 run in the Bayesian framework, except that the estimate of $r$
-more than halved (Appendix B, Fig. B1). %% What's up with that??
+more than halved (Appendix B, Fig. B1). 
 When random observer effects
@@
 extensions that we believe would be computationally prohibitive in many
-cases. Nonetheless, we will discuss one---spatially-explicit models of
+cases. Nonetheless, we will discuss one --- spatially-explicit models of
 immigration. 
@@
 unsampled locations can be estimated \citep{lele_etal:1998}. 
-%be based on the assumptions that either the study sites
-%encompass the whole range of the metapopulation or that the
-%annual changes in the populations sampled are representative of
-%a larger metapopulation. 
 Initial tests suggest that it would be
@@
 true value of initial abundance was low, and the estimates of equilibrium abundance and maximum
-growth rate in the Ricker-logistic model when the true value of maximum growth rate was low.  
+growth rate in the Ricker model when the true value of maximum growth rate was low.  
 The failure of the first case was likely due to a large number of sites starting at abundance 0 or reaching it early in the simulation.  
@@
 stage-structured model of \citet{zipkin_etal:2014} are available for underdispersed data.  AIC model
-selection did a good job of picking the most appropriate distribution.  Density-dependent models
-for underdispersed data still remain to be developed.
+selection did a good job of picking the most appropriate distribution.  Justifiable 
+density-dependent models
+for underdispersed data remain to be developed.
 
@@
 
-We thank J. Andrew Royle and T. Scott Sillett for helpful suggestions on our analysis
-and manuscript. We gratefully acknowledge the U. S. National Park 
+We thank J. Andrew Royle, T. Scott Sillett, and John Sauer for helpful suggestions on our analysis. 
+We gratefully acknowledge the U. S. National Park 
 Service, the Smithsonian Institution, and USGS's Breeding Bird Survey program for funding support.
-We thank K. Pardieck and D. Zilowkowski for help with data acquisition. J. Sauer 
-provided helpful feedback about our analysis.
+We thank Keith Pardieck and David Ziolkowski for help with data acquisition. We are
+grateful for the numerous helpful suggestions for manuscript
+revision provided by J. Andrew Royle, T. Scott Sillett, and an anonymous
+referee.
 

### 2014-08-04 improved_state_space.tex
Small revisions; added line numbers to response

@@
 \textit{Key words}: abundance, Dail and Madsen model, density-dependence,
-detectability, Gompertz-logistic, immigration, % model
-population dynamics, Ricker-logistic, zero-inflated % model
+detectability, Gompertz model, immigration, %-logistic
+population dynamics, Ricker model, zero-inflated %-logistic
 
@@
 \end{linenomath*}
-where $x_0$ is the model for initial abundance.  In classical
+where $x_0$ is the distribution for initial abundance.  In classical
 state-space models, the two sets of random effects
@@
 $\zeta_t \sim \mathrm{N}(0, \tau)$. 
-The process variation associated with $N_1$ is often ignored, or
-the population is assumed to be at equilibrium such that the $N_1$ can
-be regarded as an outcome of the equilibrium distribution.
+The process variation associated with $N_1$ is often ignored ($x_0 = X_1$), or
+the population is assumed to be at stochastic equilibrium such that the $x_0$ can
+be regarded as an outcome of the equilibrium distribution \citep{devalpine_hastings:2002}.
 
@@
 For example, practitioners either assume that there is no
-observation error in the first year ($x_0 = X_1$), or they assume the population is
+observation error in the first year, or they assume the population is
 at equilibrium, which may defeat the purpose of many studies of
@@
 \subsection*{Environmental and demographic stochasticity}
-\begin{comment}
-Even though the DM model is conceptually simple, 
-it includes many random effects (the $N$'s)
-making it challenging to compute the likelihood or implement MCMC
-algorithms. Nonetheless, in some cases additional random effects
-may be of interest. For example, in conventional state-space models,
-environmental stochasticity is often modeled as $r_t \sim
-\mathrm{Norm}(\mu_r, \sigma_r)$.
-\end{comment}
+\label{subsec:stoch}
 The original DM models allows for demographic stochasticity --- variation in population growth 
@@
 straight-forward approaches for combining data from multiple sources
-or use existing estimates of parameters as prior distributions. Third,
-one common usage of state-space models is for predicting future
-population size, and this task is easily accomplished using Bayesian
+or using existing estimates of parameters as prior distributions. Third,
+a common usage of state-space models is to predict future
+population size, and this is easily accomplished using Bayesian
 methods.
@@
 
-\subsection*{Simulation Study to Assess Estimator Accuracy}
+\subsection*{Simulation study to assess estimator accuracy}
 
@@
 
-\subsection*{Robustness Simulation Study}
+\subsection*{Robustness simulation study}
 
-We tested our models against violations of the Poisson process assumption for
-annual dynamics.  We ran three alternative scenarios: 1) a geometric-recruitment model with
+Most of our DM model extensions assume Poisson process error for annual dynamics.
+Although we provide a variation for data overdispersed compared to the Poisson (see \textit{Environmental and demographic stochasticity}), 
+and the original DM models assume underdispersed data,  
+we tested our models against violations of the Poisson assumption.  
+We ran three alternative scenarios: 1) a geometric-recruitment model with
 $\omega = \gamma = 0.5$; 2) a Leslie matrix model with six age classes; and 3) 
 a Gompertz model with two non-overlapping generations per year (sampling occasion).   
-%% Couldn't we expand our model to deal with these "mis-specifications"? Like Elise and I did in a couple of recent papers?
 The first two scenarios are underdispersed compared to the Poisson and the third is overdispersed.
@@
 
-\subsection*{Analysis of Breeding Bird Survey Data}
+\subsection*{Analysis of Breeding Bird Survey data}
 
@@
 
-\subsection*{Robustness Simulation Study}
+\subsection*{Robustness simulation study}
 In simulations of the geometric-recruitment model estimated with the exponential model,
@@
 seperate the effects of movement versus births and deaths.
-Alternatively, ecologists could combine count data with data
-from, for example,
-capture-recapture data. Models that are fitted to count data and
-demographic data are often referred to as integrated population models
+Alternatively, ecologists could combine count data with demographic data,
+%from, for example,
+%capture-recapture data. Models that are fitted to count data and
+%demographic data are 
+often referred to as an integrated population model
 \citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
@@
 stage-structured model of \citet{zipkin_etal:2014} are available for underdispersed data.  AIC model
-selection did a good job of picking the most appropriate distribution.  Justifiable 
-density-dependent models
-for underdispersed data remain to be developed.
+selection did a good job of picking the most appropriate distribution.  Appropriate 
+density-dependent models for underdispersed data remain to be developed.
 

### 2014-08-04 supp/bbs_estimates_appendix.tex
Small revisions; added line numbers to response

@@
 \title{Appendix B: BBS Results}
+\date{} %remove date
 \maketitle
@@
 C. Dynamics && \\
-C.1. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&0	\\
-C.2. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&8.4 \\
+C.1. NB[$\Lambda$(.)$\alpha$(.)]Ricker+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&0	\\
+C.2. NB[$\Lambda$(.)$\alpha$(.)]Gompertz+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) &10	&8.4 \\
 C.3. NB[$\Lambda$(.)$\alpha$(.)]Exponential+Immigration[$r$(.)$\iota$(.)]$p$(wind+1st) &9	&36.5\\
 C.4. NB[$\Lambda$(.)$\alpha$(.)]Geometric-recruitment+Immigration[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st) &10	&38.6\\
-C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&192.8\\
-C.6. NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic[$r$(.)$K$(.)]$p$(wind+1st) &9	&195.1\\
+C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz[$r$(.)$K$(.)]$p$(wind+1st) &9	&192.8\\
+C.6. NB[$\Lambda$(.)$\alpha$(.)]Ricker[$r$(.)$K$(.)]$p$(wind+1st) &9	&195.1\\
 C.7. NB[$\Lambda$(.)$\alpha$(.)]Exponential[$r$(.)]$p$(wind+1st)	&8 &271.3\\
@@
 %  \caption{Parameter estimates for ovenbirds in Maryland and Virginia from BBS data, 1966-2010.  
-%MLE estimates come from the NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) 
+%MLE estimates come from the NB[$\Lambda$(.)$\alpha$(.)]Ricker+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) 
 %model run in the \textbf{R} package \texttt{unmarked}; base estimates from the same model run in \textbf{JAGS};
@@
 \caption{Parameter estimates for ovenbirds in Maryland and Virginia from BBS data, 1966-2010.  
-MLE estimates come from the NB[$\Lambda$(.)$\alpha$(.)]Ricker-logistic+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) 
+MLE estimates come from the NB[$\Lambda$(.)$\alpha$(.)]Ricker+Immigration[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st) 
 model run in the \textbf{R} package \texttt{unmarked}; base estimates from the same model run in \textbf{JAGS};

### 2014-08-04 supp/sim_tables_appendix.tex
Small revisions; added line numbers to response

@@
 \usepackage[letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
-\usepackage[pdftex]{hyperref}
+\usepackage[pdftex,hidelinks]{hyperref}
 \usepackage{soul}
@@
 
-The third simulation was a Gompertz-logistic model with two non-overlapping generations per year (sampling occasion).   
+The third simulation was a Gompertz model with two non-overlapping generations per year (sampling occasion).   
 We used 200 sites for 25 years, and data-generating values of $\Lambda = 10$, $r = 0.05$, $K = 10$, and $p = 0.5$.
-We estimated parameters with the Gompertz-logistic model with 
+We estimated parameters with the Gompertz model with 
 one generation per year using both maximum likelihood and MCMC approaches, as well
@@
 high $\Lambda$, growth rate ($r$), and detection probability ($p$).
-For the Ricker-logistic model simulations we used $\Lambda$ = 10 and $p$ = 0.25, and
+For the Ricker model simulations we used $\Lambda$ = 10 and $p$ = 0.25, and
 simulated low, medium, and high values of $r$ and equilibrium density ($K$).
-For the Ricker-logistic + immigration dynamics model we fixed all parameters the
-same as the Ricker-logistic model (with $r$ = 0.05 and $K$ = 10) and
+For the Ricker + immigration dynamics model we fixed all parameters the
+same as the Ricker model (with $r$ = 0.05 and $K$ = 10) and
 simulated low, medium, and high values of immigration rate ($\iota$).}  
@@
     \hline
-    & \multicolumn{3}{c}{Exponential Growth} && \multicolumn{2}{c}{Ricker-logistic} &&
-    Ricker-logistic + Immigration \\
+    & \multicolumn{3}{c}{Exponential Growth} && \multicolumn{2}{c}{Ricker} &&
+    Ricker + Immigration \\
     \cline{2-4}     \cline{6-7}    \cline{9-9}
@@
   \footnotesize
-  \caption{Simulation results from 3 cases of Ricker-logistic and immigration dynamics with 1000 Monte Carlo replicates for each.}
+  \caption{Simulation results from 3 cases of Ricker and immigration dynamics with 1000 Monte Carlo replicates for each.}
   \begin{tabular}{cccccccccccccccccccc}
@@
   \footnotesize
-  \caption{Parameters of Gompertz-logistic simulation with two generation per year 
+  \caption{Parameters of Gompertz simulation with two generation per year 
 (1000 Monte Carlo replicates)
@@
 10	&0.05	&10	&0.5	&&-	&-	&-	&-	&&-	&-	&-	&-\\
-Gompertz-logistic (MLE; 1 generation/year) &
+Gompertz (MLE; 1 generation/year) &
 8.392	&0.00919	&1.269	&0.60241	&&
@@
 0.013	&-	&1.000 	&0.000 \\
-Gompertz-logistic (MCMC; 1 generation/year) &
+Gompertz (MCMC; 1 generation/year) &
 8.315	&0.03044	&4.575	&0.60723	&&
@@
 0.007	&-	&0.368		&0.000 \\
-Gompertz-logistic (MCMC; 2 generations/year) &
+Gompertz (MCMC; 2 generations/year) &
 9.969	&0.05405	&10.407	&0.50337	&&
@@
 \caption{Histograms of 1000 parameter estimates for each of nine
-simulation cases with Ricker-logistic dynamics. The blue vertical lines are the 
+simulation cases with Ricker dynamics. The blue vertical lines are the 
 data-generating values.}
@@
 \caption{Histograms of 1000 parameter estimates for each of three
-simulation cases with Ricker-logistic + immigration dynamics. The blue vertical lines are the 
+simulation cases with Ricker + immigration dynamics. The blue vertical lines are the 
 data-generating values.}
@@
 \begin{figure}
-\caption{Histograms of 1000 parameter estimates simulated with Gompertz-logistic dynamics, 
-two generations per year, fit with three different Gompertz-logistic models.
+\caption{Histograms of 1000 parameter estimates simulated with Gompertz dynamics, 
+two generations per year, fit with three different Gompertz models.
 The vertical lines are the data-generating values; changing the number of generations

### 2014-08-04 improved_state_space.tex
Removed comments

@@
 \textit{Key words}: abundance, Dail and Madsen model, density-dependence,
-detectability, Gompertz model, immigration, %-logistic
-population dynamics, Ricker model, zero-inflated %-logistic
-
-%\vspace{0.5cm}
+detectability, Gompertz model, immigration, 
+population dynamics, Ricker model, zero-inflated
 
@@
 In applied contexts, population models are used for estimating extinction risk 
-\citep{nadeem_lele:2011,hostetler_etal:2012} and for predicting the %schoener_spiller:1992,
+\citep{nadeem_lele:2011,hostetler_etal:2012} and for predicting the 
 effects of future environmental conditions or conservation actions on
@@
 
-Two challenges are routinely encountered when trying to fit models %Several
+Two challenges are routinely encountered when trying to fit models 
 of population dynamics to data from field studies. 
-First, deterministic models of population dynamics are generally inadequate due to process %virtually always
-variation, the stochasticity in demographic parameters and environmental % inherent
+First, deterministic models of population dynamics are generally inadequate due to process 
+variation, the stochasticity in demographic parameters and environmental 
 conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
@@
 density-independent, such as when growth
-is exponential. %% Exponential or geometric?
+is exponential. 
 A generic model can be written: 
@@
   \begin{align}
-    N_1 &= x_0 \nonumber \\ %X_1
+    N_1 &= x_0 \nonumber \\ 
 N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
@@
 \citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
-distributions for the process and observation errors. An observation model %with
-%more intuitive interpretations, such as those that 
+distributions for the process and observation errors. An observation model 
 explicitly modeling
@@
 $G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
-depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new %
+depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new 
 individuals; the autoregressive model, $G_{i,t} \sim
@@
 population is at equilibrium. Covariates of
-$\omega$ and $\gamma$ can be accommodated, e.g., % easily %for example
+$\omega$ and $\gamma$ can be accommodated, e.g., 
 using logit- and log-linear models respectively.
@@
 \subsection*{Population growth models}
-%\begin{samepage}
 Partitioning population growth into survival and recruitment
@@
 with classical population growth models that have a long history in ecology.
-The exponential growth model is the simplest: %% Should this geometric instead of exponential?
+The exponential growth model is the simplest: 
 $N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
@@
 likelihood is not required, speeding up processing time.
-%\end{samepage}
 
@@
 growth extensions use a single Poisson distribution for abundance. 
-% NOTE: I would save this for the discussion
 Conceptually, this is an appropriate method of modeling demographic stochasticity 
@@
 if they are underdispersed compared to the Poisson (but see Discussion).
-% but may actually perform better for iteroparous organisms as well (see Discussion).  
  Dynamics of semelparous organisms with more than one generation per year 
@@
 rate due to stochastic environmental conditions --- is by explicitly
-modeling dynamic parameters as functions of environmental covariates. %, such as weather.  
+modeling dynamic parameters as functions of environmental covariates. 
 For example, in the Gompertz model (Eq.~\ref{eq:gomp}), $r$ could be a 
@@
 $\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
-the logit-scale standard deviation of the random observer effect. % (also on the logit scale).
+the logit-scale standard deviation of the random observer effect.
 
@@
 
-%In practice, 
 Simulation methods such as Markov chain Monte Carlo (MCMC)
-are used to estimate the posterior distributions. While time %simulate
+are used to estimate the posterior distributions. While time 
 consuming, MCMC may be the only viable approach for estimating
@@
 Maryland and Virginia. For our focal species, we selected the
-ovenbird (\textit{Seiurus aurocapilla}), an abundant, %and widespread,
+ovenbird (\textit{Seiurus aurocapilla}), an abundant, 
 forest-breeding migratory songbird with a stable or increasing trend
@@
 (Table B1A in Appendix B). 
-The best supported model for $p$ included effects of wind speed %additive 
+The best supported model for $p$ included effects of wind speed  
 and an observer's first run of a route (Table B1B, model B.1). First run 
@@
 with immigration were better supported than models without immigration 
-(Table B1C); the best supported was Ricker + immigration % of these the
+(Table B1C); the best supported was Ricker + immigration 
 (see Appendix B, Fig. B1 for estimates).
@@
 Alternatively, ecologists could combine count data with demographic data,
-%from, for example,
-%capture-recapture data. Models that are fitted to count data and
-%demographic data are 
 often referred to as an integrated population model

### 2014-10-14 improved_state_space.tex
starting to address new reviewer comments

@@
 variation in abundance in space and time. The third process 
-describes the the relationship between
+describes the relationship between
 abundance and the observed count data.
@@
 depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new 
-individuals; the autoregressive model, $G_{i,t} \sim
+individuals; the ``autoregressive'' model, $G_{i,t} \sim
 \mathrm{Pois}(\gamma N_{i,t-1})$
@@
 Lower right panel depicts estimated mean route abundance by year
-(mean and 95\% credible interval).
+(+/- 95\% credible interval).
 

### 2014-10-14 supp/code_samples/oven.unmarked.model.selection.r

@@ -1,3 +1,35 @@
+################################################################################
+# Code supplement for Hostetler, J. A. and R. B. Chandler. In revision. 
+# Improved state-space models for inference about spatial and temporal variation
+# in abundance from count data. Ecology.
+# 
+# Our manuscript extends the open population N-mixture models developed by Dail 
+# and Madsen (2011; Biometrics) to include classical population growth and
+# density dependence models, immigration, zero-inflation, and random observer
+# effects.  It is implemented in extensions to the R package unmarked and 
+# in program JAGS.  
+#
+# This program reads in data and fits models in unmarked to ovenbird Maryland 
+# and Virginia BBS data.  Requires comma-delimited files oven3a.csv,
+# observers.csv, first.run.csv, and wind.factor.csv.  The procedure first
+# selects the lowest AIC mixture model for initial abundance, then the lowest
+# AIC model for detection probability, and then the lowest AIC model for
+# population dynamics.
+#
+# Code by Jeffrey A. Hostetler, last updated September 9, 2014
+################################################################################
+
+################################################################################
+# IMPORTANT NOTE:
+# The currently released version of unmarked does not yet include our extensions
+# for density dependence or immigration.  Therefore, this program will not work
+# with that version of the package.  If you would like to run this code, let
+# me know what version of R you are using and I will send you a zip file with
+# a version of unmarked that this code will work with, as well as the input 
+# files.  Also note that it will take a while to run.
+################################################################################
+
+
 library(unmarked)
 
 # Read in input files
@@ -27,7 +59,7 @@ bird.frame <- unmarkedFramePCO(ydata, obsCovs=obs4, numPrimary=45)
 
 # Run mixture models for initial abundance  
 mods1 <- rep('', 3)
-outputs1 = list()
+outputs1 <- vector("list", 3)
 for (i in 1:3) {
   mixture <- c('P','NB','ZIP')[i]
   dynamics <- c("constant", "autoreg", "trend", "ricker", "gompertz")[3]
@@ -73,7 +105,8 @@ mix.num
 
 # Run models for detection probability  
[…]

### 2014-10-14 response to revision comments.tex
starting to address new reviewer comments

\documentclass[12pt,english]{article}
\usepackage[letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage{setspace}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage[round,comma,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[format=plain,justification=raggedright,labelsep=period,font={stretch=1.9}]{caption}
\usepackage[pdftex,hidelinks]{hyperref} 
\bibliographystyle{ecology}
\bibpunct{(}{)}{,}{a}{}{,}
\usepackage{etoolbox}
\AtBeginEnvironment{tabular}{\doublespacing}
\makeatletter
\let\@fnsymbol\@arabic
\makeatother
\title{Improved state-space models for inference about 
spatial and temporal variation in abundance from count data: Response to revision comments}

\author{Jeffrey A. Hostetler
   \and Richard B. Chandler    
}
\date{} 
\begin{document}
\maketitle
\vspace{-1cm}
\begin{spacing}{1.9}
\begin{flushleft}
\renewcommand*\thetable{R\arabic{table}}
\renewcommand*\thefigure{R\arabic{figure}}
\renewcommand*\theequation{R\arabic{equation}}

\section*{Reviewer 1 (Comments to Author):} 
\label{sec:rev1}

\subsection*{General Comments}
In this paper, the authors propose to extend the \citet{dail_madsen:2011} model to 
include classical models of population growth. They also propose extensions to 
deal with zero-inflated data, immigration and variation in observation errors.

This work is novel and important because it proposes to estimate population 
growth characteristics by accounting for imperfect detection of individuals. Otherwise, 
actually, population growth models are embedded in state-based models that assume 
normally distributed errors, as a consequence probability of detection cannot be estimated. 
Besides, in some models of population growth, the errors of observations are not taken 
into account.

The authors check for the reliability of their model to recover their parameters by using a 
simulation study. The simulation study covers a large range of different population 
dynamics as well as additional cases where the simulated data deviates from the assumption 
of the model used for the analysis. In addition, to the simulation study, the models are 
applied to real data.

\vspace{0.5cm}
\textit{Thank you for your positive comments!}
\vspace{0.5cm}

In a general way, effort should be put on tightening the text and better focus
on the main goal of the paper, that is about including models of population
growth in a framework that account for imperfect detection of individuals. The
choice of the model of population growth might be subjective and/or motivated by
the knowledge of the life history of the species under study. Even thought, some
studies, proposed some procedures based on model selection to discriminate among
the different models of population growth. Hence, the Ms would benefit of a
better description of the underlying ecological assumptions of the different
models of population growth. For example, the exponential model describes
population growth in an unlimited environment, where $r$ is the maximum per capita
of increase, etc… . In Ricker model, the density-dependence becomes stronger at
higher densities, due to the exponential function relating to carrying capacity.
The Gompertz model is similar to the Ricker model except that there is a
constant linear decrease in the instantaneous growth rate ($r$) as the natural
logarithm of population increase, etc…..

\vspace{0.5cm}
\textit{[I can do this.]}
\vspace{0.5cm}

\subsection*{Specific Comments}
I am unclear about the title of the manuscript because it does not reflect the real aim of the
paper, which is well stated by the authors, lines 417-418: ``one of the primary aims of our
paper was to demonstrate how classical population growth models can be embedded in the
DM model''.

I find that basing the introduction upon the notion of state-based model is rather confusing
regarding the real goal of the paper (see above). The Dail and Madsen model and statespace
models are defined in two different ``statistical worlds''. On the one hand, the constant
Dail and Madsen model (2011) has a Markov property so the variables can only take on a
finite set of discrete values for a finite period of observation. The distinguish feature of
Markov chains is that the conditional probability of $x_{t}$ given the entire past depends only on
the immediately preceding variable $x_{t-1}$. In addition, the errors of observation are well
identified because they are related to imperfect detection of individuals and expressed with
a discrete probability distribution. On the other hand, the state-space model framework
provides a structure for extending time-series models to handle both observation and
process errors. The term ``state-space'' model refers to models where the process model has
an autoregressive property and where the errors have a Gaussian distribution.
Autoregressive model are defined for continuous variable, and so the errors are normally
distributed. A space-state model, being given its properties, can be estimated by using a
Kalman filter. The Kalman filter is an updating procedure whereby the prior estimate of the
state is corrected by how well the next observation can be predicted.

Even if the Dail and Madsen model also provides a structure to account for time dependence
as well as for ``process errors'' through the birth and death process and for imperfect
detection, the Dail and Madsen model and state-space models are defined in two different
``statistical worlds''. Hence, the analytical results of the Dail and Madsen model and a state-space
model would be different. That is why, in my point view, it is unclear and misleading to
relate the Dail and Madsen model to a state-space model.

\vspace{0.5cm}
\textit{[Richard, can you address these three paragraphs?]}
\vspace{0.5cm}

Lines 118 - 174: the description of the model could be shortened (or moved in part in the
appendices) because it has been described now in further publications.

\vspace{0.5cm}
\textit{[I can work on this.]}
\vspace{0.5cm}

Line 165: \textit{``propagule rain''} is not an appropriate term here; this term is used in
Mainland/Island model to describe something like a continuous source of migrants which
can colonise an empty site. A large stable mainland population provides migrants, which is
not the case in Dail and Madsen model. Consistency with the original terminology defined in
Dail and Madsen (2011) should be kept as much as possible.

\vspace{0.5cm}
\textit{[I can see the reviewer's point (there's no actual mainland), although I think it really
behaves the same way.  Should I just cut/change this?]}
\vspace{0.5cm}

Line 165: \textit{``autoregressive''} model should be kept in quote as it is p. 580 of Dail and Madsen
(2011).

\vspace{0.5cm}
\textit{We have made this change.}
\vspace{0.5cm}

Line 135: typography error in \textit{``The third process describes the relationships between''} should
be \textit{``The third process describes the relationships between.''}

\vspace{0.5cm}
\textit{We have fixed the typographic error (surplus ``the'') here.}
\vspace{0.5cm}

Line 190: More explanations on the Gompertz model should be added here, especially on
the link between the Ricker model and the Gompertz model. The original reference of the
Gompertz logistic growth equation should be added here.

\vspace{0.5cm}
\textit{[I can do this.]}
\vspace{0.5cm}

Line 199: I am unclear about the meaning of \textit{``internal and external (immigration)
contributions to population growth.''} Then, the authors proposed to account for immigration
but they do not mention anything about emigration. In the Dail and Madsen model, the
survival rate is a reduced demographic parameter that accounts for both death and
emigration. Some words, in the discussion, about how emigration is (or not) handled in this
novel extension would be welcome.

\vspace{0.5cm}
\textit{[I can work on this.]}
\vspace{0.5cm}

Line 225: I am unclear about \textit{``semelparous organisms with one generation per sampling
interval (year).''} Semelparous refers to organisms that died after reproduction, (i.e. they can
have solely one single reproduction event in their life because they die after this
reproduction event). So, semelparous organisms cannot have ``one generation per sampling
interval'' but only one generation.

Lines 225 - 229 – This paragraph needs to be clarified – semelparous organisms cannot have
more than one generation per year because they can have only one single reproduction
event in their life (see above).

\vspace{0.5cm}
\textit{[I can work on these two.]}
\vspace{0.5cm}

Lines 281 - 296: The paragraph about the simulation of the data needs to be clarified to
understand better how the data are simulated. Some references to the different equations
of the model should help to clarify this section. For the simulation study, the time series
which are simulated are rather long (40 years) and also the length of the time series of the
application are about 40 years. If the time series were shorter, how the estimation of the
parameters would be affected.

\vspace{0.5cm}
\textit{[I can work on this.  Should I just speculate in the discussion about fewer years
of data?  I would think less precision would be the result.  Fewer than 10 years of data
could especially be a problem for density-dependent models, but again that's speculation.]}
\vspace{0.5cm}

Discussion: The text of the discussion needs to be reduced and tightened. The authors
should better focus on discussing the results of this study and relate them to other studies
on models of population growth.

\vspace{0.5cm}
\textit{[I can work on this.  Not sure yet what we can/should cut though.]}
\vspace{0.5cm}

The figure A1 and A2 describe results of the simulation based on the Ricker and Gompertz
models, and both are much more ``realistic'' than the model of exponential growth when
applied to population ecology. So, the Fig. A1 and A2 represent core results of the
manuscript and should be rather presented in the main text than in appendices.

\vspace{0.5cm}
\textit{[Between this comment and the one by the other reviewer, sounds like we should 
put all three of these figures in the main text.  If we still have to keep it to 30 pages,
that requires at least two pages of cutting, which is a lot.  Probably more, since we'll also 
be adding a few things.  This reviewer suggests cutting
text from DM model description and discussion.  I will work on those, but maybe we
should also consider emailing Evan Cooch about the length limit.]}
\vspace{0.5cm}

\section*{Reviewer 2 (Comments to Author):} 
\label{sec:rev2}
\subsection*{General Comments}
The authors build on hierarchical state-space models that describe population
dynamics proposed by Dail and Madsen (2011). The paper is clear, well-written,
and provides important advances and tests of the original DM formulation and
extensions. The authors cover a lot of ground, but to me, the principal novel
contribution of the paper is the embedding of classical population-growth and
density-dependence functions into the population dynamics model, rather than
relying on the original DM formulation that models population change as a
function of demographic (survival and recruitment) parameters. The authors also
propose a sort of hybrid set of models that combine a functional form for
population growth with a parameter representing immigration from outside of the
indexed spatial location. The paper also introduces negative binomial and
zero-inflated versions of the state model as well as models that include
demographic or environmental stochasticity. An important extension that
the authors propose that would greatly enhance the practical utility of these
models in ecological applications is the modeling of dynamics parameters as
functions of site-time specific environmental covariates or random effects. The
authors state that code will be provided for implementing models with maximum
likelihood and Bayesian McMC methods using freely available software packages,
which should facilitate the use of these models in the future. Overall, the
paper advances the state of the art of this class of models in many directions,
and the authors do a good job in highlighting both the potential and limitations
of the various model flavors.

\vspace{0.5cm}
\textit{Thank you for your positive comments!}
\vspace{0.5cm}

\subsection*{Specific comments} 
Model Extensions: Immigration models 
The immigration models are an interesting idea, combining classical growth and
density-dependence functions or the mechanistic geometric growth functions with
an additional parameter that represents immigration. As the authors note, an
important limitation of several proposed models without the immigration
parameter (including Dail and Madsen's geometric growth model) is that they do
not accommodate recovery following extinction, so the introduction of the
immigration parameter would seem a critical component in those models in cases
where local extinction is likely. I must admit, however, that I don't completely
understand how the parameters of these models are informed. For example, in the
exponential, Ricker, and Gompertz models, the r parameter includes a mixture of
local apparent survival (i.e., site fidelity + true survival) + local
recruitment (through birth processes) dynamics, while the iota parameter
supposedly reflects immigration from outside the site. How is there
information to separate recruitment components of r from immigration components
of iota? Similarly, how are both gamma and iota identifiable in the geometric
growth model? I suppose that somehow because the iota parameter does not depend
on the any of the other population parameters, it can be estimated separately.
But this is not entirely clear to me. Maybe a sentence or 2 could clarify this a
bit?

\vspace{0.5cm}
\textit{[I can work on this.  Not sure yet what the answer is though.]}
\vspace{0.5cm}

Line 347-348: As re: memory requirements, why not do multiple single-chain runs
(to yield multiple chains)?

\vspace{0.5cm}
\textit{[Will this work?  I should try it.]}
\vspace{0.5cm}

Results: Why present the graphical display of simulation results for the
exponential models in the main text (Fig 1), but relegate all of the tables and
graphical summaries for other models to Appendix A? I would prefer to see all of
the summary tables (first choice) or all of the figures (but not both) in the
main text.

\vspace{0.5cm}
\textit{[Between this comment and the one by the other reviewer, sounds like we should 
put all three of these figures in the main text.]}
\vspace{0.5cm}

Fig. 1: for initial abundance on x-axis, replace little lambda with big lambda
to match text.

\vspace{0.5cm}
\textit{[Richard, can you do this, and also increase the resolution of this figure?
I don't have the simulation results from the exponential model.  Alternatively,
you could send me the results and I can replot it (I think I have your plotting code).
It looks like you used a file called process.RData to store the results.]}
\vspace{0.5cm}

Fig. 2 legend, in parentheses could replace ``mean and'' with +/-, given that it
is already stated that the panel shows the mean.

\vspace{0.5cm}
\textit{We have made the suggested change.}
\vspace{0.5cm}

Line 358: replace 'inaccurate' with 'imprecise'.

\vspace{0.5cm}
\textit{These terms have multiple meanings and are defined differently in
different places.  We like the definition of \citet{williams_etal:2002}:
``Accuracy combines both bias and precision in an assessment of estimator
performance.''  On the same page (45) those authors describe mean squared error 
(MSE) as a measure of estimator accuracy.  Therefore root mean squared error (RMSE) and
relative RMSE are best described as measures of accuracy as well.  
In another case (line ?) we used the
word ``imprecise'' when the RMSE was high, but only after we established that
bias was low.  In this case, since bias is high, we feel it is more correct to describe
the estimates as ``inaccurate.''}
\vspace{0.5cm}

Lines 424-429: Could difficulty in estimating vital rates derives from having
survival and recruitment parameters taking similar values? For example many
small songbirds have apparent survival rates of $\sim .5$; for a stable population,
recruitment would be similar, which in lieu of direct information on vital
rates, might make the 2 parameters difficult to separate. In the American Robin
example in Dail and Madsen (2011), survival and recruitment were estimated at
reasonable values of .7 and .5 (although precision was still fairly low). Could
low detection probability have something to do with this problem? From the p
estimates in Fig. B1, detection probabilities were $< .2$ (which seems low to me
given that ovenbirds are loud and sing a lot, although I suppose it has to do
with large sample radius). In the simulations in the Dail-Madsen paper, lowest
coverage for gamma and omega was found at lowest p values (p = .25). I realize
I'm stretching a bit here, but it was a little
unsatisfying to me that the vital rates parameters could not be estimated, and
the reason for why this was so wasn't too clear. The authors suggest individual
heterogeneity might be an issue and suggest incorporating prior information or
data via integrated population models to improve vital rate estimation. However,
I wonder if some additional conjecture might be added here?

\vspace{0.5cm}
\textit{[Unsatisfying to us too!  I can work on this.  Richard, I seem to remember that 
you had done some simulations on why $p$ estimates are so low from BBS data 
(and it was exactly because of large radius).  Did you publish those results somewhere
we can cite?  If not, maybe we should not address that point here.
The additional speculations
the reviewer provides seem possible, although in our robustness testing we tried
DM simulations with $\omega = \gamma = 0.5$ and $p = 0.25$, and though the DM
estimator didn't do perfect, they weren't nearly as bad as the real data.  Maybe 
they would have been with a lower simulated $p$, but it seems more likely to me that
the real data somehow came from a different process than the DM model assumes.]}
\vspace{0.5cm}

\bibliography{improved_state_space}

\end{flushleft}
\end{spacing}
\end{document}

### 2014-11-04 improved_state_space.tex
continuing to address new reviewer comments

@@
 certainty. To estimate both abundance and detection parameters, 
-the original $N$-mixture model uses replicate
-observations at each site. The observations are collected during sufficiently
-short time intervals such that the population can safely be
+the $N$-mixture model uses replicate % original
+observations at each site. The observations are made during sufficiently
+short time intervals such that the population can be %safely
 assumed to be closed with respect to births, deaths, and movement. The DM
@@
 abundance at $t-1$, and (3) the
-detection process \citep{dail_madsen:2011}.
-The first two processes describe the state process---the
+detection process. % \citep{dail_madsen:2011}.
+The first two processes describe the %the state process---
 variation in abundance in space and time. The third process 
@@
 $G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
-depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new 
-individuals; the ``autoregressive'' model, $G_{i,t} \sim
+depend on $N_{i,t-1}$; %and which simulates a ``propagule rain'' of new individuals; 
+the ``autoregressive'' model, $G_{i,t} \sim
 \mathrm{Pois}(\gamma N_{i,t-1})$
@@
 dynamics, but this is not always possible using simple count data,
-especially when the sites are not closed with respect to immigration
-and emigration. 
+especially when the sites are not closed with respect to movement.
+%immigration and emigration. 
 When the mechanistic model is unrealistic, we suggest replacing it
 with classical population growth models that have a long history in ecology.
-The exponential growth model is the simplest: 
-$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
+The exponential growth model, which describes population growth in an 
+unlimited environment, is the simplest: 
+$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic or maximum per capita
 rate of increase. 
@@
 a stochastic version of the 
-Ricker model \citep{ricker:1954}. Another option is a 
+Ricker model \citep{ricker:1954}. In the Ricker model, 
+density dependence becomes stronger as abundance increases well above $K$,
+leading to overcompensation    
+%due to the exponential function relating to carrying capacity 
+\citep[abundance goes from above $K$ to below it;][]{sabo_etal:2004}.
+Another option is a 
 modified Gompertz-logistic
-density-dependent model \citep[henceforth Gompertz model,][]{hart_gotelli:2011}:
+density-
+dependent model \citep[henceforth Gompertz model;][]{gompertz:1825,
+hart_gotelli:2011}:
 \begin{linenomath*}
@@
 Here the interpretations of $r$ and $K$ are similar to those in the
-Ricker model. Because a single Poisson distribution controls
+Ricker model, but dynamics can only be overcompensatory at high values of $r$
+(approximately $r > log(K)$).
+%population growth declines linearly with an increase in the natural logarithm of abundance,
+%rather than linearly with an increase in abundance. 
+Because a single Poisson distribution controls
 the distribution of $N_{i,t}$ in each of these models, the discrete 
@@
 However, these models can be generalized 
-to include both internal and external (immigration) contributions
+to allow immigration to contribute
 to population growth. For example, an exponential plus immigration
@@
 and Gompertz models can be extended to allow for immigration in the
-same way.   Deterministic versions of the Ricker + immigration and the 
+same way.  Because the population contribution from immigration does
+not depend on current abundance, $\iota$ and other growth parameters
+are separately estimable.
+ Deterministic versions of the Ricker + immigration and the 
 Gompertz + immigration do reach stable equilibriums, but the equilibriums 
@@
 Conceptually, this is an appropriate method of modeling demographic stochasticity 
- for semelparous organisms with one generation per sampling interval (year)
+ for semelparous populations with one generation per sampling interval (year)
  \citep{bonsall_hastings:2004},
-and the DM models appropriate for modeling the demographic stochasticity of iteroparous organisms,
+and the DM models appropriate for modeling the demographic stochasticity of iteroparous populations,
 if they are underdispersed compared to the Poisson (but see Discussion).
- Dynamics of semelparous organisms with more than one generation per year 
+ Dynamics of semelparous populations with more than one generation per year 
  (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
@@
 detection probability.  Our first series of simulations
-assumed dynamics were exponential.  We ran 
-each combination of low, medium, and high $\Lambda \in
+assumed dynamics were exponential (Eq.~\ref{eq:exp}).  We ran 
+each combination of low, medium, and high initial abundance, 
+growth rate, and detection probability ($\Lambda \in
 \{1,5,10\}$, $r \in \{-0.005, 0, 0.005\}$, and
-$p \in \{0.05, 0.25, 0.5\}$. 
+ $p \in \{0.05, 0.25, 0.5\}$). 
 Our second series of simulations changed dynamics to the Ricker
-model. We used an initial abundance of 10 and a detection probability
+model (Eq.~\ref{eq:rick}). We used $\Lambda$ of 10 and $p$
 of 0.25, and simulated low, medium, and high values of equilibrium
-abundance and maximum growth rate, (Table A1;
+abundance $(K)$ and $r$ (Table A1;
 9 total combinations). Our third series of simulations was based
@@
 parameters the same as the Ricker model (with $r$ = 0.05 and $K$ = 10) and
-simulated low, medium, and high values of immigration rate (Table A1).
+simulated low, medium, and high values of immigration rate ($\iota$) (Table A1).
 
@@
 variation in regional population growth rate, Eq.~\ref{eq:nuRand}.
-We tested for lack of convergence using
+We also used the random observer effects model to obtain route and 
+year specific abundance estimates.    
+We tested for lack of convergence using four or
 five Markov chains for each model \citep{gelman_rubin:1992}.
@@
 after at least 3,000 adaptation iterations.
-We also used the random observer effects model to obtain route and 
-year specific abundance estimates.  Because of the memory requirements involved,
-we used a single chain.
 
@@
 from the model that included environmental stochasticity and random observer effects.
-Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
-provided little evidence of lack of convergence for any of the Bayesian models ran.
 
@@
 abundance was highest in western and eastern Maryland in 1970 and 1980, but became
-more evenly distributed by 2010.
+more evenly distributed by 2010.  Gelman and Rubin diagnostics and visual 
+examination of the chain trajectory and density plots
+provided little evidence of lack of convergence for any of the Bayesian models ran.
 
@@
 
-One strategy for adhering to the original formulation of the
-model that included survival and recruitment parameters would be to
-use either prior information or additional data \citep{zipkin_etal:2014}. For example, in a 
-Bayesian analysis existing estimates of demographic parameters
-could be used as informative priors, which might make it possible to
-seperate the effects of movement versus births and deaths.
-Alternatively, ecologists could combine count data with demographic data,
-often referred to as an integrated population model
-\citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
-We believe that a DM modeling framework could work well in this context,
-and that the DM model could be viewed as a specific case of the
-IPM proposed by \citet{buckland_etal:2004}.
-However, we caution that when count data are affected by
-movement and vital rates, it would be unwise to ignore the movement
-processes.
-
-Integrated population models may be viewed as the gold standard in
-state-space modeling, but ecologists do not always have direct
-information about vital rates, especially at large spatial scales.
-Count data are much more common, relatively inexpensive to obtain, and
-are produced by many of the largest monitoring programs in the
-world, such as the BBS. 
-In the absence of direct information about demographic parameters, and
-when the original assumptions of the model do not hold, our extensions to the
-DM model can still be used for important purposes such as predicting future
-population size and changes in species distributions.
+%One strategy for adhering to the original formulation of the
+%model that included survival and recruitment parameters would be to
+%use either prior information or additional data \citep{zipkin_etal:2014}. For example, in a 
+%Bayesian analysis existing estimates of demographic parameters
+%could be used as informative priors, which might make it possible to
+%seperate the effects of movement versus births and deaths.
+%Alternatively, ecologists could combine count data with demographic data,
+%often referred to as an integrated population model
+%\citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
+%We believe that a DM modeling framework could work well in this context,
+%and that the DM model could be viewed as a specific case of the
+%IPM proposed by \citet{buckland_etal:2004}.
+%However, we caution that when count data are affected by
+%movement and vital rates, it would be unwise to ignore the movement
+%processes.
+%
+%Integrated population models may be viewed as the gold standard in
+%state-space modeling, but ecologists do not always have direct
+%information about vital rates, especially at large spatial scales.
+%Count data are much more common, relatively inexpensive to obtain, and
+%are produced by many of the largest monitoring programs in the
+%world, such as the BBS. 
+%In the absence of direct information about demographic parameters, and
+%when the original assumptions of the model do not hold, our extensions to the
+%DM model can still be used for important purposes such as predicting future
+%population size and changes in species distributions.
 
@@
 In our models, immigration is currently modeled as independent of
-population sizes in other sites. Several alternatives suggest
-themselves if we think about movement in a metapopulation
-context \citep{hanski:1998}: immigration conditional on total or mean
+population sizes in other sites, and emigration is included only
+as an implicit piece of apparent survival or population growth. 
+We suggest alternatives that put movement in a metapopulation
+context: immigration conditional on total or mean % \citep{hanski:1998}
 abundance across sites the previous time step; immigration
@@
 possible to fit some or all of these models, at least in the
-Bayesian framework.
+Bayesian framework. It may also be possible to model an explicit emigration
+process or rate and tie immigration at nearby sites to this rate.
 
@@
 that abundance or the rate of growth.  In other cases, bias was low and coverage nominal.
+However, we simulated 40 years of data; with fewer years of data these models might
+experience more problems.
 
@@
 grateful for the numerous helpful suggestions for manuscript
-revision provided by J. Andrew Royle, T. Scott Sillett, and an anonymous
-referee.
+revision provided by J. Andrew Royle, T. Scott Sillett, and anonymous referees.
 
@@
   \centering
-  \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year6}
+  \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year8}
 \label{fig:oven_N}

### 2014-11-04 response to revision comments.tex
continuing to address new reviewer comments

@@
 \vspace{0.5cm}
-\textit{[I can do this.]}
+\textit{[We have described the classical population growth models in more 
+detail (lines ).]}
 \vspace{0.5cm}
@@
 I find that basing the introduction upon the notion of state-based model is rather confusing
-regarding the real goal of the paper (see above). The Dail and Madsen model and statespace
+regarding the real goal of the paper (see above). The Dail and Madsen model and state-space
 models are defined in two different ``statistical worlds''. On the one hand, the constant
@@
 \vspace{0.5cm}
-\textit{[I can work on this.]}
+\textit{[Richard, any thoughts on this?  I can see the reviewer's point - there are a lot of DM
+papers out now.  And this is almost 3 pages.  But I like what we've got here, and if we cut or
+move it we'd have to be careful, because we refer to this section a good deal in our extensions.
+How we shorten it may depend on how we respond to the DM model as state-space model critique.
+Please take a stab at this if you have time and ideas on how to shorten or move.]}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[I can see the reviewer's point (there's no actual mainland), although I think it really
-behaves the same way.  Should I just cut/change this?]}
+\textit{We have made the suggested change.}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[I can do this.]}
+\textit{[We have described the Gompertz model in more 
+detail (lines ).]}
+
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[I can work on this.]}
+\textit{We have edited the text for clarity (lines ).  We have added discussion about emigration
+(lines ).}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[I can work on these two.]}
+\textit{We were referring to generations of the population or species, not the individual.  We have edited 
+this paragraph for clarity.}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[I can work on this.  Should I just speculate in the discussion about fewer years
-of data?  I would think less precision would be the result.  Fewer than 10 years of data
-could especially be a problem for density-dependent models, but again that's speculation.]}
+\textit{We have added clarification and equation references to this paragraph. We expect that
+the models would not perform as well with far fewer years of data, and have commented on
+this in the discussion (lines).}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[I can work on this.  Not sure yet what we can/should cut though.]}
+\textit{[Richard, I commented out the two paragraphs on IPMs.  What do you think?]}
 \vspace{0.5cm}
@@
 put all three of these figures in the main text.  If we still have to keep it to 30 pages,
-that requires at least two pages of cutting, which is a lot.  Probably more, since we'll also 
-be adding a few things.  This reviewer suggests cutting
-text from DM model description and discussion.  I will work on those, but maybe we
+that requires a page and a half of additional cutting, which is a lot.  This reviewer suggests cutting
+text from DM model description and discussion.  I have worked on those (see above), but maybe we
 should also consider emailing Evan Cooch about the length limit.]}
@@
 \vspace{0.5cm}
-\textit{[I can work on this.  Not sure yet what the answer is though.]}
+\textit{We believe the reviewer has hit upon it - the population contribution from immigration
+does not depend on current abundance, whereas the contributions from $r$ in the density-dependent
+models and from $\gamma$ in the geometric-recruitment model do (in different ways).  With sufficient
+temporal and spatial replication, these parameters are separately estimable.  We have added text to that
+effect (lines ).}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[Will this work?  I should try it.]}
+\textit{We now have multiple chains for this model as well, and the convergence is good
+with very similar results.}
 \vspace{0.5cm}

### 2014-11-07 response-to-revision-comments.tex
Response letter

\documentclass[12pt,english]{article}
\usepackage[letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage{setspace}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage[round,comma,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[format=plain,justification=raggedright,labelsep=period,font={stretch=1.9}]{caption}
\usepackage[pdftex,hidelinks]{hyperref} 
\bibliographystyle{ecology}
\bibpunct{(}{)}{,}{a}{}{,}
\usepackage{etoolbox}
\AtBeginEnvironment{tabular}{\doublespacing}
\makeatletter
\let\@fnsymbol\@arabic
\makeatother
\title{Improved state-space models for inference about 
spatial and temporal variation in abundance from count data: Response to revision comments}

\author{Jeffrey A. Hostetler
   \and Richard B. Chandler    
}
\date{} 
\begin{document}
\maketitle
\vspace{-1cm}
\begin{spacing}{1.9}
\begin{flushleft}
\renewcommand*\thetable{R\arabic{table}}
\renewcommand*\thefigure{R\arabic{figure}}
\renewcommand*\theequation{R\arabic{equation}}

\section*{Reviewer 1 (Comments to Author):} 
\label{sec:rev1}

\subsection*{General Comments}
In this paper, the authors propose to extend the \citet{dail_madsen:2011} model to 
include classical models of population growth. They also propose extensions to 
deal with zero-inflated data, immigration and variation in observation errors.

This work is novel and important because it proposes to estimate population 
growth characteristics by accounting for imperfect detection of individuals. Otherwise, 
actually, population growth models are embedded in state-based models that assume 
normally distributed errors, as a consequence probability of detection cannot be estimated. 
Besides, in some models of population growth, the errors of observations are not taken 
into account.

The authors check for the reliability of their model to recover their parameters by using a 
simulation study. The simulation study covers a large range of different population 
dynamics as well as additional cases where the simulated data deviates from the assumption 
of the model used for the analysis. In addition, to the simulation study, the models are 
applied to real data.

\vspace{0.5cm}
\textit{Thank you for your positive comments!}
\vspace{0.5cm}

In a general way, effort should be put on tightening the text and better focus
on the main goal of the paper, that is about including models of population
growth in a framework that account for imperfect detection of individuals. The
choice of the model of population growth might be subjective and/or motivated by
the knowledge of the life history of the species under study. Even thought, some
studies, proposed some procedures based on model selection to discriminate among
the different models of population growth. Hence, the Ms would benefit of a
better description of the underlying ecological assumptions of the different
models of population growth. For example, the exponential model describes
population growth in an unlimited environment, where $r$ is the maximum per capita
of increase, etc… . In Ricker model, the density-dependence becomes stronger at
higher densities, due to the exponential function relating to carrying capacity.
The Gompertz model is similar to the Ricker model except that there is a
constant linear decrease in the instantaneous growth rate ($r$) as the natural
logarithm of population increase, etc…..

\vspace{0.5cm}
\textit{[We have described the classical population growth models in more 
detail (lines ).]}
\vspace{0.5cm}

\subsection*{Specific Comments}
I am unclear about the title of the manuscript because it does not reflect the real aim of the
paper, which is well stated by the authors, lines 417-418: ``one of the primary aims of our
paper was to demonstrate how classical population growth models can be embedded in the
DM model''.

I find that basing the introduction upon the notion of state-based model is rather confusing
regarding the real goal of the paper (see above). The Dail and Madsen model and state-space
models are defined in two different ``statistical worlds''. On the one hand, the constant
Dail and Madsen model (2011) has a Markov property so the variables can only take on a
finite set of discrete values for a finite period of observation. The distinguish feature of
Markov chains is that the conditional probability of $x_{t}$ given the entire past depends only on
the immediately preceding variable $x_{t-1}$. In addition, the errors of observation are well
identified because they are related to imperfect detection of individuals and expressed with
a discrete probability distribution. On the other hand, the state-space model framework
provides a structure for extending time-series models to handle both observation and
process errors. The term ``state-space'' model refers to models where the process model has
an autoregressive property and where the errors have a Gaussian distribution.
Autoregressive model are defined for continuous variable, and so the errors are normally
distributed. A space-state model, being given its properties, can be estimated by using a
Kalman filter. The Kalman filter is an updating procedure whereby the prior estimate of the
state is corrected by how well the next observation can be predicted.

Even if the Dail and Madsen model also provides a structure to account for time dependence
as well as for ``process errors'' through the birth and death process and for imperfect
detection, the Dail and Madsen model and state-space models are defined in two different
``statistical worlds''. Hence, the analytical results of the Dail and Madsen model and a state-space
model would be different. That is why, in my point view, it is unclear and misleading to
relate the Dail and Madsen model to a state-space model.

\vspace{0.5cm}
Our definition of a state-space model is in agreement with the brand new
book on the subject ``Modeling Population Dynamics: Model Formulation,
Fittin, and Assessment Using State-Space Methods'' by Newman et
al (2014, Springer). They cover many state-space models that have
a first-order Markov state process and a non-Gaussian
observation process. For this reason, we think that our title connects
our work to the bigger picture, which is state-space
models of count data.  
\vspace{0.5cm}

Lines 118 - 174: the description of the model could be shortened (or moved in part in the
appendices) because it has been described now in further publications.

\vspace{0.5cm}
\textit{[Richard, any thoughts on this?  I can see the reviewer's point - there are a lot of DM
papers out now.  And this is almost 3 pages.  But I like what we've got here, and if we cut or
move it we'd have to be careful, because we refer to this section a good deal in our extensions.
How we shorten it may depend on how we respond to the DM model as state-space model critique.
Please take a stab at this if you have time and ideas on how to
shorten or move.]}
\textit{[Jeff, I don't think we should remove this, but I think we
  could put equations 2 and 4 in the text, so as to save space. We
  might also be to shorten some of the text related to covariate
  effects.]}
\vspace{0.5cm}

Line 165: \textit{``propagule rain''} is not an appropriate term here; this term is used in
Mainland/Island model to describe something like a continuous source of migrants which
can colonise an empty site. A large stable mainland population provides migrants, which is
not the case in Dail and Madsen model. Consistency with the original terminology defined in
Dail and Madsen (2011) should be kept as much as possible.

\vspace{0.5cm}
\textit{We have made the suggested change.}
\vspace{0.5cm}

Line 165: \textit{``autoregressive''} model should be kept in quote as it is p. 580 of Dail and Madsen
(2011).

\vspace{0.5cm}
\textit{We have made this change.}
\vspace{0.5cm}

Line 135: typography error in \textit{``The third process describes the relationships between''} should
be \textit{``The third process describes the relationships between.''}

\vspace{0.5cm}
\textit{We have fixed the typographic error (surplus ``the'') here.}
\vspace{0.5cm}

Line 190: More explanations on the Gompertz model should be added here, especially on
the link between the Ricker model and the Gompertz model. The original reference of the
Gompertz logistic growth equation should be added here.

\vspace{0.5cm}
\textit{[We have described the Gompertz model in more 
detail (lines ).]}

\vspace{0.5cm}

Line 199: I am unclear about the meaning of \textit{``internal and external (immigration)
contributions to population growth.''} Then, the authors proposed to account for immigration
but they do not mention anything about emigration. In the Dail and Madsen model, the
survival rate is a reduced demographic parameter that accounts for both death and
emigration. Some words, in the discussion, about how emigration is (or not) handled in this
novel extension would be welcome.

\vspace{0.5cm}
\textit{We have edited the text for clarity (lines ).  We have added discussion about emigration
(lines ).}
\vspace{0.5cm}

Line 225: I am unclear about \textit{``semelparous organisms with one generation per sampling
interval (year).''} Semelparous refers to organisms that died after reproduction, (i.e. they can
have solely one single reproduction event in their life because they die after this
reproduction event). So, semelparous organisms cannot have ``one generation per sampling
interval'' but only one generation.

Lines 225 - 229 – This paragraph needs to be clarified – semelparous organisms cannot have
more than one generation per year because they can have only one single reproduction
event in their life (see above).

\vspace{0.5cm}
\textit{We were referring to generations of the population or species, not the individual.  We have edited 
this paragraph for clarity.}
\vspace{0.5cm}

Lines 281 - 296: The paragraph about the simulation of the data needs to be clarified to
understand better how the data are simulated. Some references to the different equations
of the model should help to clarify this section. For the simulation study, the time series
which are simulated are rather long (40 years) and also the length of the time series of the
application are about 40 years. If the time series were shorter, how the estimation of the
parameters would be affected.

\vspace{0.5cm}
\textit{We have added clarification and equation references to this paragraph. We expect that
the models would not perform as well with far fewer years of data, and have commented on
this in the discussion (lines).}
\vspace{0.5cm}

Discussion: The text of the discussion needs to be reduced and tightened. The authors
should better focus on discussing the results of this study and relate them to other studies
on models of population growth.

\vspace{0.5cm}
\textit{[Richard, I commented out the two paragraphs on IPMs.  What do
  you think?

Jeff, fine by me]}

\vspace{0.5cm}

The figure A1 and A2 describe results of the simulation based on the Ricker and Gompertz
models, and both are much more ``realistic'' than the model of exponential growth when
applied to population ecology. So, the Fig. A1 and A2 represent core results of the
manuscript and should be rather presented in the main text than in appendices.

\vspace{0.5cm}
\textit{[Between this comment and the one by the other reviewer, sounds like we should 
put all three of these figures in the main text.  If we still have to keep it to 30 pages,
that requires a page and a half of additional cutting, which is a lot.  This reviewer suggests cutting
text from DM model description and discussion.  I have worked on those (see above), but maybe we
should also consider emailing Evan Cooch about the length limit.

Jeff, I agree that we should ask Evan. If we have to modify something,
I'm in favor of moving Fig. 1 to the appendix, and inserting a table that summarizes
all the results.

]}
\vspace{0.5cm}

\section*{Reviewer 2 (Comments to Author):} 
\label{sec:rev2}
\subsection*{General Comments}
The authors build on hierarchical state-space models that describe population
dynamics proposed by Dail and Madsen (2011). The paper is clear, well-written,
and provides important advances and tests of the original DM formulation and
extensions. The authors cover a lot of ground, but to me, the principal novel
contribution of the paper is the embedding of classical population-growth and
density-dependence functions into the population dynamics model, rather than
relying on the original DM formulation that models population change as a
function of demographic (survival and recruitment) parameters. The authors also
propose a sort of hybrid set of models that combine a functional form for
population growth with a parameter representing immigration from outside of the
indexed spatial location. The paper also introduces negative binomial and
zero-inflated versions of the state model as well as models that include
demographic or environmental stochasticity. An important extension that
the authors propose that would greatly enhance the practical utility of these
models in ecological applications is the modeling of dynamics parameters as
functions of site-time specific environmental covariates or random effects. The
authors state that code will be provided for implementing models with maximum
likelihood and Bayesian McMC methods using freely available software packages,
which should facilitate the use of these models in the future. Overall, the
paper advances the state of the art of this class of models in many directions,
and the authors do a good job in highlighting both the potential and limitations
of the various model flavors.

\vspace{0.5cm}
\textit{Thank you for your positive comments!}
\vspace{0.5cm}

\subsection*{Specific comments} 
Model Extensions: Immigration models 
The immigration models are an interesting idea, combining classical growth and
density-dependence functions or the mechanistic geometric growth functions with
an additional parameter that represents immigration. As the authors note, an
important limitation of several proposed models without the immigration
parameter (including Dail and Madsen's geometric growth model) is that they do
not accommodate recovery following extinction, so the introduction of the
immigration parameter would seem a critical component in those models in cases
where local extinction is likely. I must admit, however, that I don't completely
understand how the parameters of these models are informed. For example, in the
exponential, Ricker, and Gompertz models, the r parameter includes a mixture of
local apparent survival (i.e., site fidelity + true survival) + local
recruitment (through birth processes) dynamics, while the iota parameter
supposedly reflects immigration from outside the site. How is there
information to separate recruitment components of r from immigration components
of iota? Similarly, how are both gamma and iota identifiable in the geometric
growth model? I suppose that somehow because the iota parameter does not depend
on the any of the other population parameters, it can be estimated separately.
But this is not entirely clear to me. Maybe a sentence or 2 could clarify this a
bit?

\vspace{0.5cm}
\textit{We believe the reviewer has hit upon it - the population contribution from immigration
does not depend on current abundance, whereas the contributions from $r$ in the density-dependent
models and from $\gamma$ in the geometric-recruitment model do (in different ways).  With sufficient
temporal and spatial replication, these parameters are separately estimable.  We have added text to that
effect (lines ).}
\vspace{0.5cm}

Line 347-348: As re: memory requirements, why not do multiple single-chain runs
(to yield multiple chains)?

\vspace{0.5cm}
\textit{We now have multiple chains for this model as well, and the convergence is good
with very similar results.}
\vspace{0.5cm}

Results: Why present the graphical display of simulation results for the
exponential models in the main text (Fig 1), but relegate all of the tables and
graphical summaries for other models to Appendix A? I would prefer to see all of
the summary tables (first choice) or all of the figures (but not both) in the
main text.

\vspace{0.5cm}
\textit{[Between this comment and the one by the other reviewer, sounds like we should 
put all three of these figures in the main text.]}
\vspace{0.5cm}

Fig. 1: for initial abundance on x-axis, replace little lambda with big lambda
to match text.

\vspace{0.5cm}
\textit{[Richard, can you do this, and also increase the resolution of this figure?
I don't have the simulation results from the exponential model.  Alternatively,
you could send me the results and I can replot it (I think I have your plotting code).
It looks like you used a file called process.RData to store the
results.

Jeff, let's wait to hear back from Evan. If we move this to the
appendix, I don't think these changes will be necessary.

]}
\vspace{0.5cm}

Fig. 2 legend, in parentheses could replace ``mean and'' with +/-, given that it
is already stated that the panel shows the mean.

\vspace{0.5cm}
\textit{We have made the suggested change.}
\vspace{0.5cm}

Line 358: replace 'inaccurate' with 'imprecise'.

\vspace{0.5cm}
\textit{These terms have multiple meanings and are defined differently in
different places.  We like the definition of \citet{williams_etal:2002}:
``Accuracy combines both bias and precision in an assessment of estimator
performance.''  On the same page (45) those authors describe mean squared error 
(MSE) as a measure of estimator accuracy.  Therefore root mean squared error (RMSE) and
relative RMSE are best described as measures of accuracy as well.  
In another case (line ?) we used the
word ``imprecise'' when the RMSE was high, but only after we established that
bias was low.  In this case, since bias is high, we feel it is more correct to describe
the estimates as ``inaccurate.''}
\vspace{0.5cm}

Lines 424-429: Could difficulty in estimating vital rates derives from having
survival and recruitment parameters taking similar values? For example many
small songbirds have apparent survival rates of $\sim .5$; for a stable population,
recruitment would be similar, which in lieu of direct information on vital
rates, might make the 2 parameters difficult to separate. In the American Robin
example in Dail and Madsen (2011), survival and recruitment were estimated at
reasonable values of .7 and .5 (although precision was still fairly low). Could
low detection probability have something to do with this problem? From the p
estimates in Fig. B1, detection probabilities were $< .2$ (which seems low to me
given that ovenbirds are loud and sing a lot, although I suppose it has to do
with large sample radius). In the simulations in the Dail-Madsen paper, lowest
coverage for gamma and omega was found at lowest p values (p = .25). I realize
I'm stretching a bit here, but it was a little
unsatisfying to me that the vital rates parameters could not be estimated, and
the reason for why this was so wasn't too clear. The authors suggest individual
heterogeneity might be an issue and suggest incorporating prior information or
data via integrated population models to improve vital rate estimation. However,
I wonder if some additional conjecture might be added here?

\vspace{0.5cm}
\textit{[Unsatisfying to us too!  I can work on this.  Richard, I seem to remember that 
you had done some simulations on why $p$ estimates are so low from BBS data 
(and it was exactly because of large radius).  Did you publish those results somewhere
we can cite?  If not, maybe we should not address that point here.
The additional speculations
the reviewer provides seem possible, although in our robustness testing we tried
DM simulations with $\omega = \gamma = 0.5$ and $p = 0.25$, and though the DM
estimator didn't do perfect, they weren't nearly as bad as the real data.  Maybe 
they would have been with a lower simulated $p$, but it seems more likely to me that
the real data somehow came from a different process than the DM model
assumes.

Jeff, I don't have any paper about this that we could cite. However,
it seem intuitive and even necessary that p will decrease as sample
area increases. A distance sampling observation model would be better
here. Didn't we mention that in the paper?

]}
\vspace{0.5cm}

\bibliography{improved_state_space}

\end{flushleft}
\end{spacing}
\end{document}

### 2014-11-14 response-to-revision-comments.tex
Finishing new reviewer comments, moved tables and figures around

@@
 \vspace{0.5cm}
-\textit{[We have described the classical population growth models in more 
-detail (lines ).]}
+\textit{We have described the classical population growth models in more 
+detail (lines ).}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-Our definition of a state-space model is in agreement with the brand new
-book on the subject ``Modeling Population Dynamics: Model Formulation,
-Fittin, and Assessment Using State-Space Methods'' by Newman et
-al (2014, Springer). They cover many state-space models that have
+\textit{Our definition of a state-space model is in agreement with the brand new
+book on the subject \citep{newman_etal:2014}. They cover many state-space 
+models that have
 a first-order Markov state process and a non-Gaussian
@@
 our work to the bigger picture, which is state-space
-models of count data.  
+models of count data.} 
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[Richard, any thoughts on this?  I can see the reviewer's point - there are a lot of DM
-papers out now.  And this is almost 3 pages.  But I like what we've got here, and if we cut or
-move it we'd have to be careful, because we refer to this section a good deal in our extensions.
-How we shorten it may depend on how we respond to the DM model as state-space model critique.
-Please take a stab at this if you have time and ideas on how to
-shorten or move.]}
-\textit{[Jeff, I don't think we should remove this, but I think we
-  could put equations 2 and 4 in the text, so as to save space. We
-  might also be to shorten some of the text related to covariate
-  effects.]}
+\textit{We have shortened the DM model description section somewhat.}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[We have described the Gompertz model in more 
-detail (lines ).]}
+\textit{We have described the Gompertz model in more 
+detail (lines ).}
 
@@
 \vspace{0.5cm}
-\textit{[Richard, I commented out the two paragraphs on IPMs.  What do
-  you think?
-
-Jeff, fine by me]}
+\textit{We have reduced the text of the discussion to focus more on the results of this
+study and other studies of population growth.}
 
@@
 \vspace{0.5cm}
-\textit{[Between this comment and the one by the other reviewer, sounds like we should 
-put all three of these figures in the main text.  If we still have to keep it to 30 pages,
-that requires a page and a half of additional cutting, which is a lot.  This reviewer suggests cutting
-text from DM model description and discussion.  I have worked on those (see above), but maybe we
-should also consider emailing Evan Cooch about the length limit.
-
-Jeff, I agree that we should ask Evan. If we have to modify something,
-I'm in favor of moving Fig. 1 to the appendix, and inserting a table that summarizes
-all the results.
-
-]}
+\textit{We have put the exponential, Ricker, and Gompertz simulation results tables in the 
+main text, and moved the exponential simulation results figure to the appendix.}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[Between this comment and the one by the other reviewer, sounds like we should 
-put all three of these figures in the main text.]}
+\textit{We have put the exponential, Ricker, and Gompertz simulation results tables in the 
+main text, and moved the exponential simulation results figure to the appendix.}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[Richard, can you do this, and also increase the resolution of this figure?
-I don't have the simulation results from the exponential model.  Alternatively,
-you could send me the results and I can replot it (I think I have your plotting code).
-It looks like you used a file called process.RData to store the
-results.
-
-Jeff, let's wait to hear back from Evan. If we move this to the
-appendix, I don't think these changes will be necessary.
-
-]}
+\textit{This figure is no longer in the text.}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{[Unsatisfying to us too!  I can work on this.  Richard, I seem to remember that 
-you had done some simulations on why $p$ estimates are so low from BBS data 
-(and it was exactly because of large radius).  Did you publish those results somewhere
-we can cite?  If not, maybe we should not address that point here.
-The additional speculations
-the reviewer provides seem possible, although in our robustness testing we tried
-DM simulations with $\omega = \gamma = 0.5$ and $p = 0.25$, and though the DM
-estimator didn't do perfect, they weren't nearly as bad as the real data.  Maybe 
-they would have been with a lower simulated $p$, but it seems more likely to me that
-the real data somehow came from a different process than the DM model
-assumes.
-
-Jeff, I don't have any paper about this that we could cite. However,
-it seem intuitive and even necessary that p will decrease as sample
-area increases. A distance sampling observation model would be better
-here. Didn't we mention that in the paper?
-
-]}
+\textit{We have added some speculation in this direction (lines ).}
 \vspace{0.5cm}

### 2014-11-14 improved_state_space.tex
Finishing new reviewer comments, moved tables and figures around

@@
 In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i,1}$
-as either a Poisson or negative binomial random variable:
-\begin{linenomath*}
-\begin{gather}
-N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
-\text{or} \nonumber \\
-N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
-\label{eq:N1}
-\end{gather}
-\end{linenomath*}
+%as either a Poisson or negative binomial random variable:
+%\begin{linenomath*}
+%\begin{gather}
+%N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
+%\text{or} \nonumber \\
+%N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
+%\label{eq:N1}
+%\end{gather}
+%\end{linenomath*}
+%where $\Lambda_i$ is the expected abundance at site $i$ during
+%year 1.
+as either a Poisson ($N_{i,1} \sim \mathrm{Pois}(\Lambda_i)$) or 
+negative binomial random variable ($N_{i,1} \sim \mathrm{NB}(\Lambda_i, \alpha)$),
 where $\Lambda_i$ is the expected abundance at site $i$ during
@@
 The Poisson distribution assumes that the mean of $N_{i,1}$ is
-equal to its variance, whereas the negative binomial distribution allows the
+equal to its variance, and the negative binomial distribution allows the
 variance to be greater than the mean with the amount of
@@
 mechanism for characterizing spatial variation in abundance. For
-instance, one might consider the influence of some environmental
+instance, one could model the effect of an environmental
 covariate ($x_i$) on abundance using a log-linear
@@
 process variation and observation error can be estimated from
-spatially-replicated count data \citep{royle:2004biom}. Hence, the first component of
-the model addresses both the issues of spatial inference and
-parameter identifiability discussed previously.
+spatially-replicated count data \citep{royle:2004biom}. %Hence, the first component of
+%the model addresses both the issues of spatial inference and
+%parameter identifiability discussed previously.
 
@@
 Estimator bias and error were generally low with the exponential model, and 
-coverage was nominal (93.4\% - 97.0\%; Fig.~\ref{fig:exp_hists}; Table A1 in Appendix A). 
+coverage was nominal (93.4\% - 97.0\%; Table~\ref{tab:simexp}; Fig. A1 in Appendix A). 
 However, estimates of $r$ were 
@@
 Simulations of the Ricker model 
-generally performed well (Fig. A1 and Table A2 in Appendix A), except when the true value of $r$ was low (0.005). In this case the 
+generally performed well (Table~\ref{tab:simricker}; Fig. A2 in Appendix A), except when the true value of $r$ was low (0.005). In this case the 
 estimates of $r$ and $K$ were inaccurate (relative RMSE between 87.4\% and 2091.8\%),
@@
 91.9\% and 96.4\%.  All three cases simulated for the Ricker + immigration model 
-performed fairly well (Fig. A2 and Table A3 in Appendix A; relative RMSE between 7.2\% 
+performed fairly well (Table~\ref{tab:simricki}; Fig. A3 in Appendix A; relative RMSE between 7.2\% 
 and 62.9\%, bias between -3.5\% and 3.3\%, and coverage between
@@
 estimates of $\Lambda$ were biased high 
-and $p$ biased low (Table A5). 
+and $p$ biased low (Table A2). 
 However, estimates of $r$ had low bias and RMSE, with nominal coverage. 
 The estimates from the geometric-recruitment model for $\Lambda$ and $p$ had low bias, but the estimates
-for $\gamma$ and $\omega$ were somewhat biased and imprecise (Table A5 and Fig. A3).
+for $\gamma$ and $\omega$ were somewhat biased and imprecise (Table A2 and Fig. A4).
 The geometric- recruitment model had lower
@@
 For the Leslie matrix simulations, bias in estimates of $\Lambda$ and $p$ from both the exponential and geometric-recruitment models
-were high, but the exponential biases were higher (Table A6 and Fig. A3).  
+were high, but the exponential biases were higher (Table A3 and Fig. A4).  
 The geometric-recruitment model was favored in 100\% of the simulations, 
@@
 $\Lambda$  and $K$ and 
-overestimated $p$ (Table A7 and Figure A4).  As a result, RMSEs for these parameters 
+overestimated $p$ (Table A4 and Figure A5).  As a result, RMSEs for these parameters 
 were all high and coverages all low, except the coverage of $K$ by the maximum 
@@
 individual heterogeneity \citep{vindenes_etal:2008}, is increasing variances
-above the levels expected from demographic stochasticity alone.
+above the levels expected from demographic stochasticity alone.  It
+is also possible that similar values for recruitment and survival rates in songbirds and low 
+detection probabilities for the BBS have contributed %from the large sample radii 
+to the DM model issues.
 
@@
 
+\clearpage
+
+%%% Tables
+\begin{table}[htb]
+  \centering
+  \footnotesize
+  \caption{Simulation results from 27 cases of exponential dynamics with 1000 Monte Carlo replicates for each.
+We present the data generating values, the mean estimate, the root mean squared error (RMSE), and the coverage
+for the initial abundance ($\Lambda$), growth rate ($r$), and detection probability ($p$) parameters.}
+  \begin{tabular}{cccccccccccc}
+    \hline
+    \multicolumn{3}{c}{Data generating values}&
+    \multicolumn{3}{c}{Mean} &
+    \multicolumn{3}{c}{RMSE} &
+    \multicolumn{3}{c}{Coverage} \\
+    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
+    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
+    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
+    \multicolumn{3}{c}{\rule{4cm}{1pt}} \\
+    $\Lambda$ & $r$ & $p$ & $\Lambda$ & $r$ & $p$ & $\Lambda$ & $r$ &
+    $p$ & $\Lambda$ & $r$ & $p$ \\
+    \hline
+1	&-0.005	&0.05	&1.047	&-0.0125	&0.050	&0.241	&0.0241	&0.009	&0.949	&0.953	&0.97\\
+1	&-0.005	&0.25	&1.016	&-0.0117	&0.250	&0.154	&0.0227	&0.027	&0.952	&0.949	&0.945\\
+1	&-0.005	&0.5	&1.011	&-0.0114	&0.500	&0.135	&0.0229	&0.038	&0.957	&0.936	&0.95\\
+1	&0.005	&0.05	&1.028  &-5e-05	      &0.050	&0.236	&0.0206	&0.010	&0.939	&0.948	&0.947\\
+1	&0.005	&0.25	&1.014	&-0.0014	&0.250	&0.161	&0.0225	&0.027	&0.957	&0.937	&0.946\\
+1	&0.005	&0.5	&1.006	&-0.0011	&0.502	&0.137	&0.0206	&0.037	&0.947	&0.945	&0.95\\
+1	&0	&0.05	&1.036		&-0.0068	&0.050	&0.237	&0.0246	&0.009	&0.958	&0.934	&0.961\\
+1	&0	&0.25	&1.016		&-0.0059	&0.250	&0.164	&0.0213	&0.026	&0.944	&0.945	&0.953\\
+1	&0	&0.5	&1.005		&-0.0053	&0.500	&0.134	&0.0198	&0.038	&0.951	&0.947	&0.952\\
+5	&-0.005	&0.05	&5.075	&-0.0064	&0.050	&0.656	&0.0088	&0.005	&0.945	&0.959	&0.965\\
+5	&-0.005	&0.25	&5.026	&-0.0062	&0.250	&0.473	&0.0078	&0.017	&0.943	&0.966	&0.937\\
+5	&-0.005	&0.5	&5.013	&-0.0066	&0.501	&0.362	&0.0081	&0.025	&0.96	&0.949	&0.961\\
+5	&0.005	&0.05	&5.056	&0.0040		&0.050	&0.624	&0.0079	&0.005	&0.944	&0.957	&0.952\\
+5	&0.005	&0.25	&5.016	&0.0041		&0.250	&0.446	&0.0075	&0.017	&0.948	&0.947	&0.942\\
+5	&0.005	&0.5	&5.027	&0.0037		&0.499	&0.367	&0.0073	&0.025	&0.947	&0.95	&0.943\\
+5	&0	&0.05		&5.063	&-0.0016	&0.050	&0.611	&0.0084	&0.005	&0.96	&0.96	&0.956\\
+5	&0	&0.25		&5.044	&-0.0014	&0.249	&0.465	&0.0081	&0.017	&0.942	&0.938	&0.945\\
+5	&0	&0.5		&5.024	&-0.0012	&0.499	&0.373	&0.0075	&0.025	&0.952	&0.964	&0.946\\
+10	&-0.005	&0.05	&10.039	&-0.0060	&0.050	&1.050	&0.0061	&0.004	&0.959	&0.945	&0.96\\
+10	&-0.005	&0.25	&10.033	&-0.0057	&0.250	&0.769	&0.0060	&0.015 &0.951	&0.942	&0.945\\
+10	&-0.005	&0.5	&10.009	&-0.0060	&0.500	&0.611	&0.0053	&0.024	&0.949	&0.957	&0.945\\
+10	&0.005	&0.05	&10.068	&0.0042		&0.050	&1.046	&0.0058	&0.004	&0.955	&0.939	&0.949\\
+10	&0.005	&0.25	&10.036	&0.0040		&0.251	&0.801	&0.0053	&0.017	&0.934	&0.954	&0.921\\
+10	&0.005	&0.5	&9.967	&0.0043		&0.502	&0.618	&0.0049	&0.025	&0.934	&0.957	&0.92\\
+10	&0	&0.05	&10.065		&-9e-04	      &0.050	&1.061	&0.0055	&0.005	&0.961	&0.96	&0.944\\
+10	&0	&0.25	&10.035		&-0.0009	&0.250	&0.763	&0.0056	&0.016	&0.953	&0.943	&0.936\\
+10	&0	&0.5	&10.003		&-0.0007	&0.501	&0.581	&0.0054	&0.023	&0.949	&0.945	&0.948\\
+  \hline
+  \end{tabular}
+  \label{tab:simexp}
+\end{table}
+\clearpage
+
+\begin{sidewaystable}
+  \centering
+  \footnotesize
+  \caption{Simulation results from 9 cases of Ricker dynamics with 1000 Monte Carlo replicates for each.
+We present the data generating values, the mean estimate, the root mean squared error (RMSE), and the coverage
+for the initial abundance ($\Lambda$), maximum growth rate ($r$), equilibrium abundance ($K$), 
+and detection probability ($p$) parameters.}
+  \begin{tabular}{cccccccccccccccc}
+    \hline
+    \multicolumn{4}{c}{Data generating values}&
+    \multicolumn{4}{c}{Mean} &
+    \multicolumn{4}{c}{RMSE} &
+    \multicolumn{4}{c}{Coverage} \\
+    \multicolumn{4}{c}{\rule{4cm}{1pt}} &
+    \multicolumn{4}{c}{\rule{4cm}{1pt}} &
+    \multicolumn{4}{c}{\rule{4cm}{1pt}} &
+    \multicolumn{4}{c}{\rule{4cm}{1pt}} \\
+    $\Lambda$ & $r$ & $K$ & $p$ & $\Lambda$ & $r$ & $K$ & $p$ & $\Lambda$ & $r$ & $K$ &
+    $p$ & $\Lambda$ & $r$ & $K$ & $p$ \\
+    \hline
+10	&0.005	&5	&0.25	&9.914 &0.0098  &6.368 &0.252	&0.785 &0.0115  &5.561 &0.016	&0.955 &0.807 &0.776 &0.963\\
+10	&0.005	&10	&0.25	&9.954 &0.0087 &10.989 &0.252 	&0.775 &0.0093  &8.739 &0.016  	&0.960 &0.831 &0.828 &0.943\\
+10	&0.005	&20	&0.25	&9.935 &0.0083 &59.778 &0.251 	&0.761 &0.0083  &418.378 &0.016	&0.955 &0.841 &0.943 &0.955\\
+10	&0.05	&5	&0.25	&10.038 &0.0529  &4.682 &0.252 	&0.896 &0.0275  &1.583 &0.019  	&0.945 &0.922 &0.919 &0.951\\
+10	&0.05	&10	&0.25	&10.009 &0.0527  &9.718 &0.252 	&0.834 &0.0188  &1.746 &0.017  	&0.961 &0.930 &0.960 &0.948\\
+10	&0.05	&20	&0.25	&10.000 &0.0514 &19.790 &0.251 &0.839 &0.0132  &2.513 &0.017  	&0.951 &0.943 &0.962 &0.940\\
+10	&0.1	&5	&0.25	&10.046 &0.1013  &4.833 &0.250 	&0.913 &0.0353  &0.876 &0.020  	&0.948 &0.954 &0.944 &0.954\\
+10	&0.1	&10	&0.25	&9.990 &0.1025  &9.893 &0.251 	&0.884 &0.0227  &0.896 &0.018  	&0.949 &0.954 &0.964 &0.950\\
+10	&0.1	&20	&0.25	&9.988 &0.1018 &19.916 &0.251 	&0.858 &0.0154  &1.621 &0.017	&0.947 &0.941 &0.951 &0.953\\
+  \hline
+  \end{tabular}
+  \label{tab:simricker}
+\end{sidewaystable}
+
+\clearpage
+\begin{sidewaystable}
+  \centering
+  \footnotesize
+  \caption{Simulation results from 3 cases of Ricker and immigration dynamics with 1000 Monte Carlo replicates for each.
+We present the data generating values, the mean estimate, the root mean squared error (RMSE), and the coverage
+for the initial abundance ($\Lambda$), maximum growth rate ($r$), semi-equilibrium abundance ($K$), immigration ($\iota$),
+and detection probability ($p$) parameters.}
+  \begin{tabular}{cccccccccccccccccccc}
+    \hline
+    \multicolumn{5}{c}{Data generating values}&
+    \multicolumn{5}{c}{Mean} &
+    \multicolumn{5}{c}{RMSE} &
+    \multicolumn{5}{c}{Coverage} \\
+    \multicolumn{5}{c}{\rule{4.1cm}{1pt}} &
+    \multicolumn{5}{c}{\rule{5.5cm}{1pt}} &
+    \multicolumn{5}{c}{\rule{5.5cm}{1pt}} &
+    \multicolumn{5}{c}{\rule{5cm}{1pt}} \\
+    $\Lambda$ & $r$ & $K$ & $\iota$ & $p$ & $\Lambda$ & $r$ & $K$ & $\iota$ & $p$ & $\Lambda$ & $r$ & $K$ &
+    $\iota$ & $p$ & $\Lambda$ & $r$ & $K$ & $\iota$ & $p$ \\
+    \hline
+10 &0.05 &10 &0.005 &0.25	&9.997 	&0.0515 &9.687 &0.0051 &0.251 	&0.877 &0.0181 &1.777 &0.0031 &0.018  &0.951 &0.944 &0.964 &0.947 &0.942\\
+10 &0.05 &10 &0.050 &0.25 	&10.015	&0.0517 &9.758 &0.0504 &0.251 	&0.875 &0.0173 &1.724 &0.0105 &0.018  &0.943 &0.932 &0.958 &0.940 &0.943\\
+10 &0.05 &10 &0.500 &0.25 	&10.000	&0.0506 &9.648 &0.5081 &0.251 	&0.965 &0.0194 &2.683 &0.0776 &0.021  &0.947 &0.940 &0.952 &0.951 &0.941\\
+  \hline
+  \end{tabular}
+  \label{tab:simricki}
+\end{sidewaystable}
+\clearpage
+
 %%% Figures
-\clearpage
-
 \section*{Figure Legends}
-\noindent Figure ~\ref{fig:exp_hists}. Histograms of 1000 parameter estimates for each of 27
-simulation cases with exponential dynamics. The vertical lines are the 
-data-generating values.
-
 \noindent Figure ~\ref{fig:oven_N}. Maps of BBS route and year specific estimated 
@@
 (+/- 95\% credible interval).
-
-\begin{figure}
-\caption{}
-  \centering
-  \includegraphics[height=8in]{figs/exp_hists}
-\label{fig:exp_hists}
-\end{figure}
 

### 2014-11-14 supp/sim_tables_appendix.tex
Removed comments
Switched landscape to sideways in appendix
Finishing new reviewer comments, moved tables and figures around

@@
 \usepackage{amsmath}
-%\usepackage{rotating}
+\usepackage{rotating}
 \usepackage{pdflscape}
@@
 \title{Appendix A: Simulation Details and Results}
-\date{} %remove date
+\date{} 
 \maketitle
@@
 
-%\section*{Simulation Setup and Results Tables}
 \begin{table}[t]
@@
 
-%\clearpage
+%\begin{landscape}
 
-\begin{landscape}
-
-\begin{table}[htb]
-  \centering
-  \footnotesize
-  \caption{Simulation results from 27 cases of exponential dynamics with 1000 Monte Carlo replicates for each.}
-  \begin{tabular}{cccccccccccc}
-    \hline
-    \multicolumn{3}{c}{Data generating values}&
-    \multicolumn{3}{c}{Mean} &
-    \multicolumn{3}{c}{RMSE} &
-    \multicolumn{3}{c}{Coverage} \\
-    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
-    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
-    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
-    \multicolumn{3}{c}{\rule{4cm}{1pt}} \\
-%    \hline
-    $\Lambda$ & $r$ & $p$ & $\Lambda$ & $r$ & $p$ & $\Lambda$ & $r$ &
-    $p$ & $\Lambda$ & $r$ & $p$ \\
-    \hline
-10	&-0.005	&0.05	&10.03924	&-0.00601	&0.05007	&1.05051	&0.00605	&0.00432	&0.959	&0.945	&0.96\\
-10	&-0.005	&0.25	&10.0333	&-0.00573	&0.25&0.76926	&0.00598	&0.01545 &0.951	&0.942	&0.945\\
-10	&-0.005	&0.5	&10.00858	&-0.00598	&0.49984	&0.61057	&0.00532	&0.02352	&0.949	&0.957	&0.945\\
-10	&0.005	&0.05	&10.06802	&0.00424	&0.04996	&1.04574	&0.00576	&0.00428	&0.955	&0.939	&0.949\\
-10	&0.005	&0.25	&10.03648	&0.00399	&0.25087	&0.80109	&0.00525	&0.01665	&0.934	&0.954	&0.921\\
-10	&0.005	&0.5	&9.96719	&0.00431	&0.50244	&0.61777	&0.00492	&0.02539	&0.934	&0.957	&0.92\\
-10	&0	&0.05	&10.06548	&-9e-04	        &0.05001	&1.06134	&0.00553	&0.0046	&0.961	&0.96	&0.944\\
-10	&0	&0.25	&10.03522	&-0.00093	&0.25012	&0.76321	&0.00555	&0.01585	&0.953	&0.943	&0.936\\
-10	&0	&0.5	&10.0036	&-0.00067	&0.50052	&0.58102	&0.00544	&0.02274	&0.949	&0.945	&0.948\\
-1	&-0.005	&0.05	&1.04714	&-0.01254	&0.05045	&0.24087	&0.02411	&0.00915	&0.949	&0.953	&0.97\\
-1	&-0.005	&0.25	&1.01614	&-0.01173	&0.25038	&0.15418	&0.02267	&0.02749	&0.952	&0.949	&0.945\\
-1	&-0.005	&0.5	&1.01056	&-0.01136	&0.50008	&0.13529	&0.02287	&0.03846	&0.957	&0.936	&0.95\\
-1	&0.005	&0.05	&1.0281	        &-5e-05	        &0.05032	&0.23616	&0.02057	&0.00954	&0.939	&0.948	&0.947\\
-1	&0.005	&0.25	&1.01389	&-0.00143	&0.2502	&0.1609	&0.02246	&0.0268	&0.957	&0.937	&0.946\\
-1	&0.005	&0.5	&1.00554	&-0.00114	&0.50194	&0.13693	&0.02064	&0.03746	&0.947	&0.945	&0.95\\
-1	&0	&0.05	&1.03641	&-0.00685	&0.05017	&0.23656	&0.02457	&0.00928	&0.958	&0.934	&0.961\\
-1	&0	&0.25	&1.01556	&-0.00587	&0.24993	&0.16434	&0.02134	&0.02592	&0.944	&0.945	&0.953\\
-1	&0	&0.5	&1.00539	&-0.00533	&0.49994	&0.13367	&0.01982	&0.03812	&0.951	&0.947	&0.952\\
-5	&-0.005	&0.05	&5.07525	&-0.00638	&0.04978	&0.65646	&0.00881	&0.00496	&0.945	&0.959	&0.965\\
-5	&-0.005	&0.25	&5.02641	&-0.00619	&0.25013	&0.4727	&0.00779	&0.01735	&0.943	&0.966	&0.937\\
-5	&-0.005	&0.5	&5.01265	&-0.00656	&0.50067	&0.36154	&0.00814	&0.02465	&0.96	&0.949	&0.961\\
-5	&0.005	&0.05	&5.05623	&0.00396	&0.05002	&0.62418	&0.00786	&0.00505	&0.944	&0.957	&0.952\\
-5	&0.005	&0.25	&5.01567	&0.0041	&0.25014	&0.44585	&0.00746	&0.01723	&0.948	&0.947	&0.942\\
-5	&0.005	&0.5	&5.02672	&0.00368	&0.49908	&0.36657	&0.00727	&0.02475	&0.947	&0.95	&0.943\\
-5	&0	&0.05	&5.06304	&-0.00162	&0.04994	&0.61073	&0.00838	&0.00492	&0.96	&0.96	&0.956\\
-5	&0	&0.25	&5.04372	&-0.00141	&0.24927	&0.46542	&0.00812	&0.01709	&0.942	&0.938	&0.945\\
-5	&0	&0.5	&5.0239	&-0.00117	&0.4991	&0.37326	&0.00747	&0.02487	&0.952	&0.964	&0.946\\
-  \hline
-  \end{tabular}
-  \label{tab:simexp}
-\end{table}
-\clearpage
-
-\begin{table}
-  \centering
-  \footnotesize
-  \caption{Simulation results from 9 cases of Ricker dynamics with 1000 Monte Carlo replicates for each.}
-  \begin{tabular}{cccccccccccccccc}
-    \hline
-    \multicolumn{4}{c}{Data generating values}&
-    \multicolumn{4}{c}{Mean} &
-    \multicolumn{4}{c}{RMSE} &
-    \multicolumn{4}{c}{Coverage} \\
-    \multicolumn{4}{c}{\rule{4cm}{1pt}} &
-    \multicolumn{4}{c}{\rule{4cm}{1pt}} &
-    \multicolumn{4}{c}{\rule{4cm}{1pt}} &
-    \multicolumn{4}{c}{\rule{4cm}{1pt}} \\
-%    \hline
-    $\Lambda$ & $r$ & $K$ & $p$ & $\Lambda$ & $r$ & $K$ & $p$ & $\Lambda$ & $r$ & $K$ &
-    $p$ & $\Lambda$ & $r$ & $K$ & $p$ \\
-    \hline
-10	&0.005	&5	&0.25	&9.914 &0.00982  &6.368 &0.25211	&0.78540 &0.01147   &5.56126 &0.01614	&0.955 &0.807 &0.776 &0.963\\
-10	&0.005	&10	&0.25	&9.954 &0.00866 &10.989 &0.25179 &0.77529 &0.00929   &8.73913 &0.01613  &0.960 &0.831 &0.828 &0.943\\
-10	&0.005	&20	&0.25	&9.935 &0.00827 &59.778 &0.25145 &0.76079 &0.00830 &418.37837 &0.01559  &0.955 &0.841 &0.943 &0.955\\
-10	&0.05	&5	&0.25	&10.038 &0.05290  &4.682 &0.25160 &0.89588 &0.02748   &1.58258 &0.01884  &0.945 &0.922 &0.919 &0.951\\
-10	&0.05	&10	&0.25	&10.009 &0.05270  &9.718 &0.25172 &0.83411 &0.01876   &1.74552 &0.01742  &0.961 &0.930 &0.960 &0.948\\
-10	&0.05	&20	&0.25	&10.000 &0.05136 &19.790 &0.25082 &0.83906 &0.01316   &2.51289 &0.01670  &0.951 &0.943 &0.962 &0.940\\
-10	&0.1	&5	&0.25	&10.046 &0.10127  &4.833 &0.25048 &0.91333 &0.03529   &0.87551 &0.01953  &0.948 &0.954 &0.944 &0.954\\
-10	&0.1	&10	&0.25	&9.990 &0.10254  &9.893 &0.25143 &0.88385 &0.02267   &0.89598 &0.01800  &0.949 &0.954 &0.964 &0.950\\
-10	&0.1	&20	&0.25	&9.988 &0.10185 &19.916 &0.25117 &0.85780 &0.01543   &1.62111 &0.01680  &0.947 &0.941 &0.951 &0.953\\
-  \hline
-  \end{tabular}
-  \label{tab:simricker}
-\end{table}
-
-\clearpage
-\begin{table}
-  \centering
-  \footnotesize
-  \caption{Simulation results from 3 cases of Ricker and immigration dynamics with 1000 Monte Carlo replicates for each.}
-  \begin{tabular}{cccccccccccccccccccc}
-    \hline
-    \multicolumn{5}{c}{Data generating values}&
-    \multicolumn{5}{c}{Mean} &
-    \multicolumn{5}{c}{RMSE} &
-    \multicolumn{5}{c}{Coverage} \\
-    \multicolumn{5}{c}{\rule{4cm}{1pt}} &
-    \multicolumn{5}{c}{\rule{7cm}{1pt}} &
-    \multicolumn{5}{c}{\rule{7cm}{1pt}} &
-    \multicolumn{5}{c}{\rule{4cm}{1pt}} \\
-%    \hline
-    $\Lambda$ & $r$ & $K$ & $\iota$ & $p$ & $\Lambda$ & $r$ & $K$ & $\iota$ & $p$ & $\Lambda$ & $r$ & $K$ &
-    $\iota$ & $p$ & $\Lambda$ & $r$ & $K$ & $\iota$ & $p$ \\
-    \hline
-10 &0.05 &10 &0.005 &0.25  &9.997 &0.05148 &9.687 &0.00510 &0.25077 &0.87742 &0.01805 &1.77679 &0.00315 &0.01795  &0.951 &0.944 &0.964 &0.947 &0.942\\
-10 &0.05 &10 &0.050 &0.25 &10.015 &0.05166 &9.758 &0.05039 &0.25070 &0.87490 &0.01731 &1.72376 &0.01055 &0.01839  &0.943 &0.932 &0.958 &0.940 &0.943\\
-10 &0.05 &10 &0.500 &0.25 &10.000 &0.05057 &9.648 &0.50811 &0.25110 &0.96491 &0.01942 &2.68264 &0.07763 &0.02050  &0.947 &0.940 &0.952 &0.951 &0.941\\
-  \hline
-  \end{tabular}
-  \label{tab:simricki}
-\end{table}
-\clearpage
-
-\begin{table}[htb]
+\begin{sidewaystable}[htb]
   \centering
@@
   \label{tab:simgeom}
-\end{table}
+\end{sidewaystable}
 
@@
 
-\begin{table}[htb]
+\begin{sidewaystable}[htb]
   \centering
@@
   \label{tab:simmat}
-\end{table}
+\end{sidewaystable}
 \clearpage
 
-\begin{table}[htb]
+\begin{sidewaystable}[htb]
   \centering
@@
   \label{tab:simgomp}
-\end{table}
+\end{sidewaystable}
 
@@
 \begin{figure}
+\caption{Histograms of 1000 parameter estimates for each of 27
+simulation cases with exponential dynamics. The vertical lines are the 
+data-generating values.}
+  \centering
+  \includegraphics[height=8in]{../figs/exp_hists}
+\label{fig:exp_hists}
+\end{figure}
+
+\begin{sidewaysfigure}
 \caption{Histograms of 1000 parameter estimates for each of nine
@@
 \label{fig:rick_hists}
-\end{figure}
+\end{sidewaysfigure}
 
-\begin{figure}
+\begin{sidewaysfigure}
 \caption{Histograms of 1000 parameter estimates for each of three
@@
 \label{fig:ricki_hists}
-\end{figure}
+\end{sidewaysfigure}
 
-\begin{figure}
+\begin{sidewaysfigure}
 \caption{Histograms of 1000 parameter estimates for each of two
@@
 \label{fig:geom_mat_hists}
-\end{figure}
+\end{sidewaysfigure}
 
-\begin{figure}
+\begin{sidewaysfigure}
 \caption{Histograms of 1000 parameter estimates simulated with Gompertz dynamics, 
@@
 \label{fig:gomp_hists}
-\end{figure}
-\end{landscape}
+\end{sidewaysfigure}
+%\end{landscape}
 

### 2014-11-14 improved_state_space.tex
edits to model description in intro

@@
 Let $N_t$ be the abundance of a species during year $t$, for
-$t=1,\hdots,T$, and let $X_t$ be
-the observed data, which differs from $N_t$ due to observation error,
-a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
-modeled using a population growth model, $\mu(N_{t-1})$
-and random process variation denoted $\eta_t$.
-The growth model may be density-dependent, as in the case of the 
-Ricker-logistic model (henceforth Ricker model), or it might be 
-density-independent, such as when growth
-is exponential. 
-A generic model can be written: 
+$t=1,\hdots,T$, and let $X_t$ be the observed data. 
+%, which differs from $N_t$ due to observation error,
+%a random effect denoted $\zeta_t$. 
+Temporal variation in $N_t$ is
+modeled using a probability distribution whose expected value is
+determined by a population growth model, $\mu(N_{t-1}, {\bm \theta})$, such as 
+%and random process variation denoted $\eta_t$.
+%The growth model may be 
+a density-dependent Ricker-logistic model (henceforth Ricker model), or  
+a density-independent exponential model with parameter vector $\bm
+\theta$. Such a model can be written:  
 \begin{linenomath*}
@@
   \begin{align}
-    N_1 &= x_0 \nonumber \\ 
-N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
-t=2,\hdots,T  \\
-X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
-t=1,\hdots,T. \nonumber 
+%     N_1 &= x_0 \nonumber \\ 
+% N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
+% t=2,\hdots,T  \\
+% X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
+% t=1,\hdots,T. \nonumber 
+  N_1 &\sim \mathrm{Normal}(\mu_1, \sigma_0) \nonumber \\
+  N_t &\sim \mathrm{Normal}(\mu(N_{t-1}, {\bm \theta}), \sigma) \qquad
+  \; \text{for} \;  t=2,\hdots,T  \\
+  X_t &\sim \mathrm{Normal}(N_t, \tau) \qquad \qquad \quad \;\;\, \text{for} \;  t=1,\hdots,T  \nonumber
   \end{align}
@@
 \end{linenomath*}
-where $x_0$ is the distribution for initial abundance.  In classical
-state-space models, the two sets of random effects
-are assumed to be i.i.d. Gaussian variables: 
-$\eta_t \sim \mathrm{N}(0, \sigma)$ and
-$\zeta_t \sim \mathrm{N}(0, \tau)$. 
-The process variation associated with $N_1$ is often ignored ($x_0 = X_1$), or
-the population is assumed to be at stochastic equilibrium such that the $x_0$ can
-be regarded as an outcome of the equilibrium distribution \citep{devalpine_hastings:2002}.
+%where $x_0$ is the distribution for initial abundance.  In classical
+%state-space models, the two sets of random effects
+%are assumed to be i.i.d. Gaussian variables: 
+%$\eta_t \sim \mathrm{N}(0, \sigma)$ and
+%$\zeta_t \sim \mathrm{N}(0, \tau)$. 
+where $\sigma_0$ and $\sigma$ are variance parameters of the
+underlying abundance process, and $\tau$ describes variation in 
+observation error. 
+%Often, parameter estimation requires fixing $\sigma_0$ at zero or assuming
+%that initial abundance is an outcome of the theoretical stationary
+%distribution of the first-order Markov chain \citep{devalpine_hastings:2002}.  
+%The process variation associated with $N_1$ is often ignored ($x_0 = X_1$), or
+%the population is assumed to be at stochastic equilibrium such that the $x_0$ can
+%be regarded as an outcome of the equilibrium distribution \citep{devalpine_hastings:2002}.
 
@@
 overcome many of the limitations associated with the application of
-classical models of population dynamics, 
-several problems exist. 
-The assumption that random effects are Gaussian distributed makes it
-easier to estimate parameters because methods such as the Kalman
-filter can be applied 
-\citep{dennis_etal:2006}; however, 
-these models effectively assume that abundance can be take on any
-real value, and hence they can predict negative values of abundance. 
+classical models of population dynamics, several problems exist. 
+The Gaussian assumptions 
+% that random effects are Gaussian distributed 
+make it easier to estimate parameters because methods such as the Kalman
+filter can be applied \citep{dennis_etal:2006}; however, 
+these models allow for implausible preditions such as 
+% assume that abundance can be take on any real value, and hence they
+% can predict 
+negative values of abundance. 
 Transforming the count data and the underlying abundance variables 
@@
 Several extensions of state-space models have been proposed to
-overcome the limitations described above. \citet{devalpine_hastings:2002} and
+overcome the limitations described above \citep{newman_etal:2014}. \citet{devalpine_hastings:2002} and
 \citet{brooks_etal:2004} described methods for fitting models with non-Gaussian

### 2014-11-14 response-to-revision-comments.tex
Final revisions before resubmit

@@
 \vspace{0.5cm}
-\textit{Thank you for your positive comments!}
+\textit{We thank the reviewer for the positive comments!}
 \vspace{0.5cm}
@@
 \textit{We have described the classical population growth models in more 
-detail (lines ).}
+detail (lines 172--174, 180--182, and 184--186).}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{We have shortened the DM model description section somewhat.}
+\textit{Good suggestion!  We have shortened the DM model description section in a few places
+where it would not interfere with the descriptions of our extensions.}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{We have made the suggested change.}
+\textit{We have removed this term.}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{We have made this change.}
+\textit{We have made this change (line 155).}
 \vspace{0.5cm}
@@
 \textit{We have described the Gompertz model in more 
-detail (lines ).}
+detail and added the original reference (lines 182--186).}
 
@@
 \vspace{0.5cm}
-\textit{We have edited the text for clarity (lines ).  We have added discussion about emigration
-(lines ).}
+\textit{We have edited the text for clarity (lines 190--194).  We have added discussion about emigration
+(lines 464--465 and 473--474).}
 \vspace{0.5cm}
@@
 \textit{We were referring to generations of the population or species, not the individual.  We have edited 
-this paragraph for clarity.}
+this paragraph for clarity (lines 216--227).}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{We have added clarification and equation references to this paragraph. We expect that
+\textit{We have added clarification and equation references to this paragraph (lines 277--292). 
+We expect that
 the models would not perform as well with far fewer years of data, and have commented on
-this in the discussion (lines).}
+this in the discussion (lines 485--486).}
 \vspace{0.5cm}
@@
 temporal and spatial replication, these parameters are separately estimable.  We have added text to that
-effect (lines ).}
+effect (lines 198--200).}
 \vspace{0.5cm}
@@
 \vspace{0.5cm}
-\textit{We have made the suggested change.}
+\textit{We have made the suggested change (now Fig. 1 legend).}
 \vspace{0.5cm}
@@
 \textit{These terms have multiple meanings and are defined differently in
-different places.  We like the definition of \citet{williams_etal:2002}:
+different places.  We like the definitions of \citet{williams_etal:2002}:
+``The tendency of replicated estimates to be dispersed is an expression
+of estimator precision,'' and
 ``Accuracy combines both bias and precision in an assessment of estimator
-performance.''  On the same page (45) those authors describe mean squared error 
+performance.''  On the same pages (44--45) those authors describe mean squared error 
 (MSE) as a measure of estimator accuracy.  Therefore root mean squared error (RMSE) and
 relative RMSE are best described as measures of accuracy as well.  
-In another case (line ?) we used the
-word ``imprecise'' when the RMSE was high, but only after we established that
-bias was low.  In this case, since bias is high, we feel it is more correct to describe
+In a few other cases (lines 348 and 365) we used the
+word ``imprecise'' when the RMSE was high, but in those cases the estimates were highly dispersed
+(Figs. A1 and A4).  In this case, since the estimates for $r$ are not very dispersed (Fig. A2), 
+we feel it is more correct to describe
 the estimates as ``inaccurate.''}
@@
 \vspace{0.5cm}
-\textit{We have added some speculation in this direction (lines ).}
+\textit{We have added some conjectures in this direction (lines 425--427).}
 \vspace{0.5cm}

### 2014-11-14 supp/code_samples/oven.jags.five.versions.r

@@ -10,6 +10,7 @@ wind.factor[wind.factor==" 1"] <- "1"
 wind.factor[wind.factor==" 2"] <- "2"
 
 library(rjags)
+library(runjags)
 
 # Settings
 year1 <- 66
@@ -398,34 +399,40 @@ sink()
 dat.ricki5 <- list(nSites=nSites, nYears=nYears, y=ydata, wind1=wind1, 
   wind2=wind2, wind3=wind3, nObs=nObs, obsID=obs5, first=first,
   decades=as.integer(seq(5, 45, 10)))
-init.ricki5 <- function() list(lambda=runif(1, 10, 60), alpha=runif(1, 0, 60), 
+init.ricki5 <- function(chain) list(lambda=runif(1, 10, 60), alpha=runif(1, 0, 60), 
   r=runif(1), K=runif(1, 30, 120), iota=runif(1),
-  p0=rnorm(1, -1.5), p1=rnorm(1), p2=rnorm(1), p3=rnorm(1), p1st=rnorm(1), sigma.p=runif(1), N=N)
+  p0=rnorm(1, -1.5), p1=rnorm(1), p2=rnorm(1), p3=rnorm(1), p1st=rnorm(1), sigma.p=runif(1), 
+  N=N, .RNG.seed=chain, .RNG.name = c("base::Wichmann-Hill","base::Marsaglia-Multicarry",
+  "base::Super-Duper","base::Mersenne-Twister")[chain])
 pars.ricki5 <- c("lambda", "alpha", "r", "K", "p0", "p1", "p2", "p3", "p1st", 
   "sigma.p", "iota", "Nmean", "Ndecade")
 
-jm.ricki5 <- jags.model("dm.ricki5.txt", dat.ricki5, init.ricki5, n.chains=1, n.adapt=1000)
-update(jm.ricki5, n.iter=100000)
-jc.ricki5 <- coda.samples(jm.ricki5, pars.ricki5, n.iter=40000)
+jr.ricki5 <- run.jags("dm.ricki5.txt", data=dat.ricki5, inits=init.ricki5, 
+  n.chains=4, adapt=1000, burnin=99000, sample=40000, monitor=pars.ricki5, 
+  thin=2, method="parallel")
 
-# Summarize results
+print(jr.ricki5)
+save.image(file.name)
+
+jc.ricki5 <- as.mcmc(jr.ricki5)
[…]

### 2014-11-29 improved_state_space.tex
Final revisions before resubmit
Making normal distribution equations consistent

--- 2014-11-14 improved_state_space.tex
@@
 Let $N_t$ be the abundance of a species during year $t$, for
-$t=1,\hdots,T$, and let $X_t$ be
-the observed data, which differs from $N_t$ due to observation error,
-a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
-modeled using a population growth model, $\mu(N_{t-1})$
-and random process variation denoted $\eta_t$.
-The growth model may be density-dependent, as in the case of the 
-Ricker-logistic model (henceforth Ricker model), or it might be 
-density-independent, such as when growth
-is exponential. 
-A generic model can be written: 
+$t=1,\hdots,T$, and let $X_t$ be the observed data. 
+Temporal variation in $N_t$ is
+modeled using a probability distribution whose expected value is
+determined by a population growth model, $\mu(N_{t-1}, {\bm \theta})$, such as 
+a density-dependent Ricker-logistic model (henceforth Ricker model) or  
+a density-independent exponential model, with parameter vector $\bm
+\theta$. Such a model can be written:  
 \begin{linenomath*}
@@
   \begin{align}
-    N_1 &= x_0 \nonumber \\ 
-N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
-t=2,\hdots,T  \\
-X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
-t=1,\hdots,T. \nonumber 
+  N_1 &\sim \mathrm{Normal}(\mu_1, \sigma^{2}_0) \nonumber \\
+  N_t &\sim \mathrm{Normal}(\mu(N_{t-1}, {\bm \theta}), \sigma^{2}) \qquad
+  \; \text{for} \;  t=2,\hdots,T  \\
+  X_t &\sim \mathrm{Normal}(N_t, \tau^{2}) \qquad \qquad \quad \;\;\, \text{for} \;  t=1,\hdots,T  \nonumber
   \end{align}
@@
 \end{linenomath*}
-where $x_0$ is the distribution for initial abundance.  In classical
-state-space models, the two sets of random effects
-are assumed to be i.i.d. Gaussian variables: 
-$\eta_t \sim \mathrm{N}(0, \sigma)$ and
-$\zeta_t \sim \mathrm{N}(0, \tau)$. 
-The process variation associated with $N_1$ is often ignored ($x_0 = X_1$), or
-the population is assumed to be at stochastic equilibrium such that the $x_0$ can
-be regarded as an outcome of the equilibrium distribution \citep{devalpine_hastings:2002}.
+where $\sigma^{2}_0$ and $\sigma^{2}$ are variance parameters of the
+underlying abundance process, and $\tau^{2}$ describes variation in 
+observation error.
 
@@
 overcome many of the limitations associated with the application of
-classical models of population dynamics, 
-several problems exist. 
-The assumption that random effects are Gaussian distributed makes it
-easier to estimate parameters because methods such as the Kalman
-filter can be applied 
-\citep{dennis_etal:2006}; however, 
-these models effectively assume that abundance can be take on any
-real value, and hence they can predict negative values of abundance. 
+classical models of population dynamics, several problems exist. 
+The Gaussian assumptions 
+make it easier to estimate parameters because methods such as the Kalman
+filter can be applied \citep{dennis_etal:2006}; however, 
+these models allow for implausible preditions such as 
+negative values of abundance. 
 Transforming the count data and the underlying abundance variables 
@@
 certainty. To estimate both abundance and detection parameters, 
-the $N$-mixture model uses replicate % original
+the $N$-mixture model uses replicate 
 observations at each site. The observations are made during sufficiently
-short time intervals such that the population can be %safely
+short time intervals such that the population can be 
 assumed to be closed with respect to births, deaths, and movement. The DM
@@
 abundance at $t-1$, and (3) the
-detection process. % \citep{dail_madsen:2011}.
-The first two processes describe the %the state process---
+detection process. 
+The first two processes describe the 
 variation in abundance in space and time. The third process 
@@
 In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i,1}$
-%as either a Poisson or negative binomial random variable:
-%\begin{linenomath*}
-%\begin{gather}
-%N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
-%\text{or} \nonumber \\
-%N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
-%\label{eq:N1}
-%\end{gather}
-%\end{linenomath*}
-%where $\Lambda_i$ is the expected abundance at site $i$ during
-%year 1.
 as either a Poisson ($N_{i,1} \sim \mathrm{Pois}(\Lambda_i)$) or 
@@
 process variation and observation error can be estimated from
-spatially-replicated count data \citep{royle:2004biom}. %Hence, the first component of
-%the model addresses both the issues of spatial inference and
-%parameter identifiability discussed previously.
+spatially-replicated count data \citep{royle:2004biom}.
 
@@
 $G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
-depend on $N_{i,t-1}$; %and which simulates a ``propagule rain'' of new individuals; 
+depend on $N_{i,t-1}$; 
 the ``autoregressive'' model, $G_{i,t} \sim
@@
 especially when the sites are not closed with respect to movement.
-%immigration and emigration. 
 When the mechanistic model is unrealistic, we suggest replacing it
@@
 leading to overcompensation    
-%due to the exponential function relating to carrying capacity 
 \citep[abundance goes from above $K$ to below it;][]{sabo_etal:2004}.
@@
 (approximately $r > log(K)$).
-%population growth declines linearly with an increase in the natural logarithm of abundance,
-%rather than linearly with an increase in abundance. 
 Because a single Poisson distribution controls
@@
 \end{linenomath*}
-where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
+where $\nu$ is a variable with a normal distribution, mean value of 0, and variance $\sigma^{2}_\nu$.  
 We suggest three variations for spatial variation in environmental stochasticity.  
@@
 X_{i,j,t} \sim \mathrm{Bin}(N_{i,t}, p_j) \nonumber \\
-\mathrm{logit}(p_j) \sim \mathrm{N}(\mu_p, \sigma_p)
+\mathrm{logit}(p_j) \sim \mathrm{Normal}(\mu_p, \sigma^{2}_p)
 \label{eq:pobs}
@@
 observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
-$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
-the logit-scale standard deviation of the random observer effect.
+$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma^{2}_p$ is
+the logit-scale variance of the random observer effect.
 
@@
 is also possible that similar values for recruitment and survival rates in songbirds and low 
-detection probabilities for the BBS have contributed %from the large sample radii 
+detection probabilities for the BBS have contributed 
 to the DM model issues.
-
-%One strategy for adhering to the original formulation of the
-%model that included survival and recruitment parameters would be to
-%use either prior information or additional data \citep{zipkin_etal:2014}. For example, in a 
-%Bayesian analysis existing estimates of demographic parameters
-%could be used as informative priors, which might make it possible to
-%seperate the effects of movement versus births and deaths.
-%Alternatively, ecologists could combine count data with demographic data,
-%often referred to as an integrated population model
-%\citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
-%We believe that a DM modeling framework could work well in this context,
-%and that the DM model could be viewed as a specific case of the
-%IPM proposed by \citet{buckland_etal:2004}.
-%However, we caution that when count data are affected by
-%movement and vital rates, it would be unwise to ignore the movement
-%processes.
-%
-%Integrated population models may be viewed as the gold standard in
-%state-space modeling, but ecologists do not always have direct
-%information about vital rates, especially at large spatial scales.
-%Count data are much more common, relatively inexpensive to obtain, and
-%are produced by many of the largest monitoring programs in the
-%world, such as the BBS. 
-%In the absence of direct information about demographic parameters, and
-%when the original assumptions of the model do not hold, our extensions to the
-%DM model can still be used for important purposes such as predicting future
-%population size and changes in species distributions.
 
@@
 We suggest alternatives that put movement in a metapopulation
-context: immigration conditional on total or mean % \citep{hanski:1998}
+context: immigration conditional on total or mean 
 abundance across sites the previous time step; immigration
@@
     \multicolumn{3}{c}{Coverage} \\
-    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
-    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
-    \multicolumn{3}{c}{\rule{4cm}{1pt}} &
-    \multicolumn{3}{c}{\rule{4cm}{1pt}} \\
+    \multicolumn{3}{c}{\rule{3.5cm}{1pt}} &
+    \multicolumn{3}{c}{\rule{3.75cm}{1pt}} &
+    \multicolumn{3}{c}{\rule{3.75cm}{1pt}} &
+    \multicolumn{3}{c}{\rule{3.7cm}{1pt}} \\
     $\Lambda$ & $r$ & $p$ & $\Lambda$ & $r$ & $p$ & $\Lambda$ & $r$ &
@@
     \hline
+10	&-0.005	&0.05	&10.039	&-0.0060	&0.050	&1.050	&0.0061	&0.004	&0.959	&0.945	&0.96\\
+10	&-0.005	&0.25	&10.033	&-0.0057	&0.250	&0.769	&0.0060	&0.015 &0.951	&0.942	&0.945\\
+10	&-0.005	&0.5	&10.009	&-0.0060	&0.500	&0.611	&0.0053	&0.024	&0.949	&0.957	&0.945\\
+10	&0.005	&0.05	&10.068	&0.0042		&0.050	&1.046	&0.0058	&0.004	&0.955	&0.939	&0.949\\
+10	&0.005	&0.25	&10.036	&0.0040		&0.251	&0.801	&0.0053	&0.017	&0.934	&0.954	&0.921\\
+10	&0.005	&0.5	&9.967	&0.0043		&0.502	&0.618	&0.0049	&0.025	&0.934	&0.957	&0.92\\
+10	&0	&0.05	&10.065		&-9e-04	      &0.050	&1.061	&0.0055	&0.005	&0.961	&0.96	&0.944\\
+10	&0	&0.25	&10.035		&-0.0009	&0.250	&0.763	&0.0056	&0.016	&0.953	&0.943	&0.936\\
+10	&0	&0.5	&10.003		&-0.0007	&0.501	&0.581	&0.0054	&0.023	&0.949	&0.945	&0.948\\
 1	&-0.005	&0.05	&1.047	&-0.0125	&0.050	&0.241	&0.0241	&0.009	&0.949	&0.953	&0.97\\
@@
 5	&0	&0.5		&5.024	&-0.0012	&0.499	&0.373	&0.0075	&0.025	&0.952	&0.964	&0.946\\
-10	&-0.005	&0.05	&10.039	&-0.0060	&0.050	&1.050	&0.0061	&0.004	&0.959	&0.945	&0.96\\
-10	&-0.005	&0.25	&10.033	&-0.0057	&0.250	&0.769	&0.0060	&0.015 &0.951	&0.942	&0.945\\
-10	&-0.005	&0.5	&10.009	&-0.0060	&0.500	&0.611	&0.0053	&0.024	&0.949	&0.957	&0.945\\
-10	&0.005	&0.05	&10.068	&0.0042		&0.050	&1.046	&0.0058	&0.004	&0.955	&0.939	&0.949\\
-10	&0.005	&0.25	&10.036	&0.0040		&0.251	&0.801	&0.0053	&0.017	&0.934	&0.954	&0.921\\
-10	&0.005	&0.5	&9.967	&0.0043		&0.502	&0.618	&0.0049	&0.025	&0.934	&0.957	&0.92\\
-10	&0	&0.05	&10.065		&-9e-04	      &0.050	&1.061	&0.0055	&0.005	&0.961	&0.96	&0.944\\
-10	&0	&0.25	&10.035		&-0.0009	&0.250	&0.763	&0.0056	&0.016	&0.953	&0.943	&0.936\\
-10	&0	&0.5	&10.003		&-0.0007	&0.501	&0.581	&0.0054	&0.023	&0.949	&0.945	&0.948\\
   \hline

### 2014-11-29 improved_state_space.tex
Making normal distribution equations consistent

@@
 $\alpha$, and $K  \sim \mathrm{U}(0, 200)$, $r  \sim \mathrm{U}(0, 5)$,
-$\iota  \sim \mathrm{U}(0, 15)$, and logit-linear $p$ coefficients $\sim \mathrm{N}(0, 100)$).  
+$\iota  \sim \mathrm{U}(0, 15)$, and logit-linear $p$ coefficients $\sim \mathrm{Normal}(0, 100)$).  
 We added random observer 
